---
title: "Diagnostic Test Evaluation"
subtitle: "The Clinician's Toolkit for Measuring How Well Tests Work"
description: |
  Learn to evaluate diagnostic tests using 2×2 tables, sensitivity, specificity,
  predictive values, likelihood ratios, and ROC curves. Apply these tools to
  real Indian clinical scenarios including cervical cancer screening, TB diagnosis,
  and malaria testing.
categories:
  - Diagnostic Statistics
  - Clinical Epidemiology
  - ROC Analysis
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(kableExtra)
source("../R/mcq.R")
library(patchwork)
library(scales)

set.seed(42)

theme_clean <- function() {
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
}
```

::: {.callout-note appearance="minimal"}
**Lecture slides for this module:** [Open Slides](../slides/05-diagnostic-slides.html){target="_blank"}
:::

## Learning Objectives

By the end of this module, you will be able to:

1. Construct a **2×2 contingency table** with the index test in rows and gold standard in columns
2. Calculate **sensitivity, specificity, PPV, and NPV** from the 2×2 table
3. Explain how **prevalence transforms predictive values** — the same test behaves differently in different populations
4. Compute and interpret **likelihood ratios** (LR+ and LR−) as prevalence-independent measures of test performance
5. Use the **Fagan nomogram** to move from pre-test to post-test probability
6. Construct and interpret **ROC curves** and understand AUC
7. Choose the **optimal cutoff** for a test based on clinical context
8. Decide when to prioritise **sensitivity vs specificity** in test selection
9. Apply these concepts to Indian clinical scenarios: **cervical cancer screening, TB, and malaria**

::: {.callout-tip title="Builds on Module 4"}
This module assumes you understand **conditional probability**, **Bayes' theorem**, and **natural frequencies** from [Module 4: Probability and Clinical Reasoning](04-probability.qmd). If concepts like "pre-test probability" or "likelihood ratio" feel unfamiliar, review Module 4 first.
:::

---

:::{.clinical-hook}
## Real-World Dilemma: Cervical Cancer Screening in Rural India

India bears nearly a quarter of the global cervical cancer burden. In many rural districts, Pap smear infrastructure is unavailable — cytology labs are hours away and results take weeks. So the government promotes **VIA (Visual Inspection with Acetic Acid)** as an alternative screening test.

A community health worker in a tribal district of Madhya Pradesh screens 1000 women with VIA. The results are compared against colposcopy-directed biopsy (the gold standard).

**VIA test characteristics:**

- Sensitivity: ~80%
- Specificity: ~85%

**Pap smear test characteristics:**

- Sensitivity: ~60%
- Specificity: ~95%

The district medical officer asks: *"VIA is more sensitive, so isn't it the better test?"*

The answer is: **it depends on what you need the test to do** — and understanding that requires the full diagnostic test evaluation toolkit.

This module gives you that toolkit.
:::

---

## Part 1: The 2×2 Contingency Table

Every diagnostic test evaluation begins with comparing the **index test** (the test being evaluated) against a **gold standard** (the reference standard that defines "true" disease status).

### The Standard Layout

The conventional arrangement places the **index test result in rows** and the **gold standard in columns**:

```{r}
#| label: tbl-2x2-layout
#| echo: false
#| tbl-cap: "Standard 2×2 Table Layout: Index Test in Rows, Gold Standard in Columns"

layout_df <- data.frame(
  ` ` = c("**Test Positive**", "**Test Negative**", "**Total**"),
  `Disease Present` = c("True Positive (TP)", "False Negative (FN)", "TP + FN"),
  `Disease Absent` = c("False Positive (FP)", "True Negative (TN)", "FP + TN"),
  `Row Total` = c("TP + FP", "FN + TN", "N"),
  check.names = FALSE
)

kable(layout_df, format = "html", escape = FALSE, align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  add_header_above(c(" " = 1, "Gold Standard" = 2, " " = 1)) %>%
  column_spec(1, bold = TRUE, width = "20%") %>%
  row_spec(3, bold = TRUE, background = "#f0f0f0")
```

**Key definitions:**

- **True Positive (TP):** Test says positive, patient truly has the disease
- **False Positive (FP):** Test says positive, but patient does NOT have the disease *(right upper corner — "false alarm")*
- **False Negative (FN):** Test says negative, but patient truly HAS the disease *(left lower corner — "missed case")*
- **True Negative (TN):** Test says negative, patient truly does not have the disease

### Worked Example: VIA Screening in Rural MP

Let's construct the 2×2 table for the VIA cervical cancer screening scenario. Suppose in a population of 1000 women screened, cervical precancer/cancer prevalence is 5%.

```{r}
#| label: tbl-2x2-via
#| echo: false
#| tbl-cap: "2×2 Table: VIA Screening for Cervical Precancer (n = 1000, Prevalence = 5%)"

total_n <- 1000
prev <- 0.05
sens_via <- 0.80
spec_via <- 0.85

diseased <- total_n * prev
healthy <- total_n * (1 - prev)

tp_via <- round(diseased * sens_via)
fn_via <- diseased - tp_via
fp_via <- round(healthy * (1 - spec_via))
tn_via <- healthy - fp_via

via_2x2 <- data.frame(
  ` ` = c("**VIA Positive**", "**VIA Negative**", "**Total**"),
  `CIN2+ Present (Biopsy)` = c(tp_via, fn_via, diseased),
  `CIN2+ Absent (Biopsy)` = c(fp_via, tn_via, healthy),
  `Row Total` = c(tp_via + fp_via, fn_via + tn_via, total_n),
  check.names = FALSE
)

kable(via_2x2, format = "html", escape = FALSE, align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  add_header_above(c(" " = 1, "Colposcopy-Directed Biopsy (Gold Standard)" = 2, " " = 1)) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(3, bold = TRUE, background = "#f0f0f0")
```

**Reading the table:**

- Of `r diseased` women with CIN2+ (precancer), VIA correctly identified `r tp_via` (**true positives**) and missed `r fn_via` (**false negatives**)
- Of `r healthy` women without CIN2+, VIA correctly cleared `r tn_via` (**true negatives**) but falsely flagged `r fp_via` (**false positives**)

### Visual: The 2×2 Table as an Icon Array

```{r}
#| label: fig-via-icon-array
#| fig-cap: "1000 Women Screened with VIA: Each square is one woman. Green = correct result, Red = incorrect result. Note the large number of false positives (red, bottom-right cluster) even though specificity is 85%."
#| echo: false
#| fig-width: 12
#| fig-height: 6

# Create icon array data
icon_data <- tibble(
  id = 1:total_n,
  category = c(
    rep("True Positive", tp_via),
    rep("False Negative", fn_via),
    rep("False Positive", fp_via),
    rep("True Negative", tn_via)
  )
) %>%
  mutate(
    row = ((id - 1) %% 20) + 1,
    col = ((id - 1) %/% 20) + 1,
    category = factor(category,
                      levels = c("True Positive", "False Negative",
                                 "False Positive", "True Negative"))
  )

icon_colours <- c("True Positive" = "#27ae60",
                   "False Negative" = "#e74c3c",
                   "False Positive" = "#f39c12",
                   "True Negative" = "#bdc3c7")

ggplot(icon_data, aes(x = col, y = row, fill = category)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = icon_colours, name = "Result Category") +
  coord_fixed() +
  labs(
    title = "VIA Screening: 1000 Women in Rural MP (Prevalence = 5%)",
    subtitle = paste0("TP = ", tp_via, " | FN = ", fn_via,
                      " | FP = ", fp_via, " | TN = ", tn_via)
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40"),
    legend.position = "bottom"
  )
```

:::{.common-mistake}
### Common Mistake: Getting the 2×2 Layout Wrong

Many students place disease status in rows and test result in columns. While mathematically equivalent, the **standard convention** (as used in most textbooks, NEET PG, and USMLE) places the **index test in rows** and the **gold standard (reference) in columns**. This means:

- **FP is in the top-right cell** (test positive, disease absent)
- **FN is in the bottom-left cell** (test negative, disease present)

Getting this layout right is essential for exam questions.
:::

---

## Part 2: Sensitivity and Specificity — Intrinsic Properties of the Test

These two measures describe how well the test performs *by itself* — they don't change when you take the test to a different population (unlike PPV and NPV, which do).

:::{.formula-box}
### Sensitivity and Specificity

**Sensitivity (True Positive Rate):**

$$\text{Sensitivity} = \frac{TP}{TP + FN} = \frac{\text{True Positives}}{\text{All Diseased}}$$

*"Of everyone who truly HAS the disease, what fraction does the test correctly identify?"*

**Specificity (True Negative Rate):**

$$\text{Specificity} = \frac{TN}{FP + TN} = \frac{\text{True Negatives}}{\text{All Non-Diseased}}$$

*"Of everyone who truly does NOT have the disease, what fraction does the test correctly clear?"*
:::

### VIA vs Pap Smear: Head-to-Head

```{r}
#| label: fig-via-vs-pap
#| fig-cap: "VIA vs Pap Smear: A Trade-Off. VIA catches more cases (higher sensitivity) but produces more false alarms (lower specificity). Pap misses more cases but has fewer false alarms."
#| echo: false
#| fig-width: 12
#| fig-height: 6

# Pap smear data
sens_pap <- 0.60
spec_pap <- 0.95

tp_pap <- round(diseased * sens_pap)
fn_pap <- diseased - tp_pap
fp_pap <- round(healthy * (1 - spec_pap))
tn_pap <- healthy - fp_pap

comparison <- tibble(
  test = rep(c("VIA\n(Visual Inspection\nwith Acetic Acid)", "Pap Smear\n(Cytology)"), each = 4),
  metric = rep(c("True Positive\n(Correctly detected)", "False Negative\n(Missed case)",
                 "False Positive\n(False alarm)", "True Negative\n(Correctly cleared)"), 2),
  count = c(tp_via, fn_via, fp_via, tn_via,
            tp_pap, fn_pap, fp_pap, tn_pap),
  fill = rep(c("#27ae60", "#e74c3c", "#f39c12", "#3498db"), 2)
)

# Focus on the key clinical cells: TP, FN, FP
comparison_key <- comparison %>%
  filter(metric != "True Negative\n(Correctly cleared)")

ggplot(comparison_key, aes(x = metric, y = count, fill = fill)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = count), vjust = -0.5, fontface = "bold", size = 4.5) +
  scale_fill_identity() +
  facet_wrap(~ test) +
  scale_y_continuous(limits = c(0, 180)) +
  labs(
    title = "VIA vs Pap Smear: The Sensitivity-Specificity Trade-Off",
    subtitle = "VIA: fewer missed cases (FN=10) but more false alarms (FP=142). Pap: more missed (FN=20) but fewer false alarms (FP=48).",
    x = NULL, y = "Number of Women (out of 1000 screened)"
  ) +
  theme_clean() +
  theme(strip.text = element_text(face = "bold", size = 11),
        axis.text.x = element_text(size = 8, lineheight = 1.0))
```

### Worked Calculations

**VIA:**

$$\text{Sensitivity} = \frac{`r tp_via`}{`r tp_via` + `r fn_via`} = \frac{`r tp_via`}{`r diseased`} = `r round(sens_via * 100)`\%$$

$$\text{Specificity} = \frac{`r tn_via`}{`r fp_via` + `r tn_via`} = \frac{`r tn_via`}{`r healthy`} = `r round(spec_via * 100)`\%$$

**Pap smear:**

$$\text{Sensitivity} = \frac{`r tp_pap`}{`r tp_pap` + `r fn_pap`} = \frac{`r tp_pap`}{`r diseased`} = `r round(sens_pap * 100)`\%$$

$$\text{Specificity} = \frac{`r tn_pap`}{`r fp_pap` + `r tn_pap`} = \frac{`r tn_pap`}{`r healthy`} = `r round(spec_pap * 100)`\%$$

:::{.key-concept}
### The Clinical Mnemonics

- **SnNOut** — **S**ensitive test, **N**egative result, rules **Out** disease. If a highly sensitive test is negative, you can confidently say the patient doesn't have the disease.
- **SpPIn** — **Sp**ecific test, **P**ositive result, rules **In** disease. If a highly specific test is positive, you can confidently say the patient has the disease.

VIA is more useful for **SnNOut** (screening: don't miss cases). Pap is more useful for **SpPIn** (confirmation: don't overtreat).
:::

---

## Part 3: Predictive Values — How Prevalence Changes Everything

As you learned in Module 4, the same test behaves very differently depending on **how common the disease is** in the population being tested. Sensitivity and specificity are properties of the test. Predictive values are properties of the **test in a specific population**.

:::{.formula-box}
### Positive and Negative Predictive Values

**Positive Predictive Value (PPV):**

$$\text{PPV} = \frac{TP}{TP + FP} = \frac{\text{True Positives}}{\text{All Positive Tests}}$$

*"If the test is positive, what's the probability the patient actually has the disease?"*

**Negative Predictive Value (NPV):**

$$\text{NPV} = \frac{TN}{FN + TN} = \frac{\text{True Negatives}}{\text{All Negative Tests}}$$

*"If the test is negative, what's the probability the patient truly doesn't have the disease?"*
:::

### VIA Predictive Values at 5% Prevalence

```{r}
#| label: via-ppv-npv
#| echo: false

ppv_via <- tp_via / (tp_via + fp_via)
npv_via <- tn_via / (fn_via + tn_via)
ppv_pap <- tp_pap / (tp_pap + fp_pap)
npv_pap <- tn_pap / (fn_pap + tn_pap)
```

$$\text{PPV}_{\text{VIA}} = \frac{`r tp_via`}{`r tp_via` + `r fp_via`} = \frac{`r tp_via`}{`r tp_via + fp_via`} = `r round(ppv_via * 100, 1)`\%$$

$$\text{NPV}_{\text{VIA}} = \frac{`r tn_via`}{`r fn_via` + `r tn_via`} = \frac{`r tn_via`}{`r fn_via + tn_via`} = `r round(npv_via * 100, 1)`\%$$

**Interpretation:** Of every 100 women who screen VIA-positive, only about `r round(ppv_via * 100)` actually have CIN2+. The remaining ~`r 100 - round(ppv_via * 100)` will undergo unnecessary colposcopy — anxiety, cost, and discomfort for nothing.

### VIA vs Pap: Predictive Value Comparison

```{r}
#| label: tbl-via-pap-comparison
#| echo: false
#| tbl-cap: "Head-to-Head: VIA vs Pap Smear at 5% Prevalence"

comp_df <- data.frame(
  Metric = c("Sensitivity", "Specificity", "PPV", "NPV",
             "False Positives (per 1000)", "False Negatives (per 1000)"),
  VIA = c(paste0(round(sens_via*100), "%"),
          paste0(round(spec_via*100), "%"),
          paste0(round(ppv_via*100, 1), "%"),
          paste0(round(npv_via*100, 1), "%"),
          fp_via,
          fn_via),
  `Pap Smear` = c(paste0(round(sens_pap*100), "%"),
                   paste0(round(spec_pap*100), "%"),
                   paste0(round(ppv_pap*100, 1), "%"),
                   paste0(round(npv_pap*100, 1), "%"),
                   fp_pap,
                   fn_pap),
  `Clinical Implication` = c(
    "VIA catches more cases (better for screening)",
    "Pap has fewer false alarms (better for confirmation)",
    "Pap: fewer unnecessary colposcopies",
    "Both excellent for ruling out",
    "VIA: 142 unnecessary referrals vs Pap: 48",
    "VIA: 10 missed vs Pap: 20 missed"
  ),
  check.names = FALSE
)

kable(comp_df, format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(3, background = "#fef9e7") %>%
  row_spec(5:6, background = "#fdedec")
```

### The Prevalence Effect: Same Tests, Different Populations

```{r}
#| label: fig-ppv-prevalence-via-pap
#| fig-cap: "PPV vs Prevalence for VIA and Pap Smear. At low prevalence (screening camps), both tests have low PPV — but Pap's higher specificity gives it an advantage. At higher prevalence (referral clinics), both perform well."
#| echo: false
#| fig-width: 11
#| fig-height: 6

prev_range <- seq(0.005, 0.40, by = 0.005)

ppv_via_curve <- sapply(prev_range, function(p) {
  tp <- p * sens_via; fp <- (1-p) * (1-spec_via); tp / (tp + fp)
})
ppv_pap_curve <- sapply(prev_range, function(p) {
  tp <- p * sens_pap; fp <- (1-p) * (1-spec_pap); tp / (tp + fp)
})

ppv_curves <- tibble(
  prevalence = rep(prev_range * 100, 2),
  ppv = c(ppv_via_curve * 100, ppv_pap_curve * 100),
  test = rep(c("VIA (Sens=80%, Spec=85%)", "Pap (Sens=60%, Spec=95%)"),
             each = length(prev_range))
)

key_pts <- tibble(
  prevalence = rep(c(2, 5, 15, 30), 2),
  ppv = c(
    sapply(c(0.02, 0.05, 0.15, 0.30), function(p) {
      tp <- p*sens_via; fp <- (1-p)*(1-spec_via); tp/(tp+fp)*100 }),
    sapply(c(0.02, 0.05, 0.15, 0.30), function(p) {
      tp <- p*sens_pap; fp <- (1-p)*(1-spec_pap); tp/(tp+fp)*100 })
  ),
  test = rep(c("VIA (Sens=80%, Spec=85%)", "Pap (Sens=60%, Spec=95%)"), each = 4),
  setting = rep(c("Population\nscreening (2%)", "District\nhospital (5%)",
                  "Colposcopy\nclinic (15%)", "Symptomatic\nreferral (30%)"), 2)
)

ggplot(ppv_curves, aes(x = prevalence, y = ppv, colour = test)) +
  geom_line(linewidth = 1.3) +
  geom_point(data = key_pts, aes(x = prevalence, y = ppv), size = 3) +
  geom_text(data = key_pts %>% filter(test == "VIA (Sens=80%, Spec=85%)"),
            aes(x = prevalence, y = ppv + 5, label = setting),
            size = 2.8, fontface = "bold", colour = "#c0392b", lineheight = 0.9) +
  scale_colour_manual(values = c("VIA (Sens=80%, Spec=85%)" = "#e74c3c",
                                  "Pap (Sens=60%, Spec=95%)" = "#2980b9")) +
  scale_y_continuous(limits = c(0, 100), labels = function(y) paste0(y, "%")) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "PPV vs Prevalence: VIA vs Pap Smear for Cervical Cancer Screening",
    subtitle = "At low prevalence, Pap's higher specificity gives it a PPV advantage. At higher prevalence, VIA catches up.",
    x = "CIN2+ Prevalence in Screened Population",
    y = "Positive Predictive Value (PPV)",
    colour = NULL
  ) +
  theme_clean() +
  theme(legend.position = "bottom")
```

:::{.key-concept}
### Why India Chose VIA for Community Screening

Despite its lower specificity (more false positives), VIA was chosen for the national cervical cancer screening programme because:

1. **Higher sensitivity** catches more precancers — critical when follow-up is uncertain in rural settings
2. **"See and treat" approach** — VIA-positive women can receive cryotherapy the same day, avoiding loss to follow-up
3. **No lab infrastructure needed** — works with acetic acid, a speculum, and a trained health worker
4. **Lower specificity is acceptable** when the cost of a false positive (an unnecessary colposcopy) is much less than the cost of a false negative (missed cancer progressing to invasive disease)

The test selection depends on the **clinical context and healthcare system**, not just the numbers.
:::

---

## Part 4: Likelihood Ratios — The Prevalence-Independent Power Tool

Sensitivity and specificity describe the test. PPV and NPV depend on prevalence. Is there a single measure that captures test performance *and* works across all populations?

**Yes: Likelihood Ratios.** As introduced in Module 4, they connect directly to Bayesian updating.

:::{.formula-box}
### Likelihood Ratios

**Positive Likelihood Ratio (LR+):**

$$LR^+ = \frac{\text{Sensitivity}}{1 - \text{Specificity}} = \frac{TPR}{FPR}$$

*"How many times more likely is a positive result in someone with disease vs. without?"*

**Negative Likelihood Ratio (LR−):**

$$LR^- = \frac{1 - \text{Sensitivity}}{\text{Specificity}} = \frac{FNR}{TNR}$$

*"How much less likely is a negative result in someone with disease vs. without?"*
:::

### Interpretation Scale

```{r}
#| label: tbl-lr-interpretation
#| echo: false
#| tbl-cap: "Likelihood Ratio Interpretation Scale"

lr_df <- data.frame(
  `LR+ Value` = c("> 10", "5 – 10", "2 – 5", "1 – 2", "1"),
  `Shift in Probability` = c("Large increase (~+45%)",
                              "Moderate increase (~+30%)",
                              "Small increase (~+15%)",
                              "Minimal change",
                              "No change (useless test)"),
  `Clinical Action` = c("Near-diagnostic — strong rule-in",
                         "Often sufficient for treatment decisions",
                         "Helpful in combination with other evidence",
                         "Rarely useful alone",
                         "Discard the test"),
  check.names = FALSE
)

lr_neg_df <- data.frame(
  `LR- Value` = c("< 0.1", "0.1 – 0.2", "0.2 – 0.5", "0.5 – 1.0", "1"),
  `Shift in Probability` = c("Large decrease (~−45%)",
                              "Moderate decrease (~−30%)",
                              "Small decrease (~−15%)",
                              "Minimal change",
                              "No change (useless test)"),
  `Clinical Action` = c("Near-exclusion — strong rule-out",
                         "Often sufficient to exclude",
                         "Helpful in combination with other evidence",
                         "Rarely useful alone",
                         "Discard the test"),
  check.names = FALSE
)

kable(lr_df, format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(1, background = "#d5f5e3") %>%
  row_spec(5, background = "#fdedec")
```

### VIA and Pap Smear Likelihood Ratios

```{r}
#| label: lr-calculations
#| echo: false

lr_pos_via <- sens_via / (1 - spec_via)
lr_neg_via <- (1 - sens_via) / spec_via
lr_pos_pap <- sens_pap / (1 - spec_pap)
lr_neg_pap <- (1 - sens_pap) / spec_pap
```

| Test | LR+ | LR− | Interpretation |
|:-----|:----|:----|:---------------|
| **VIA** | `r round(lr_pos_via, 1)` | `r round(lr_neg_via, 2)` | Moderate rule-in (LR+ ~5), modest rule-out (LR− = 0.24) |
| **Pap Smear** | `r round(lr_pos_pap, 1)` | `r round(lr_neg_pap, 2)` | Strong rule-in (LR+ = 12), weak rule-out (LR− = 0.42) |

**Key insight:** Pap has a much higher LR+ (12 vs 5.3) — so a positive Pap is more convincing than a positive VIA. But VIA has a better LR− (0.24 vs 0.42) — so a negative VIA is more reassuring than a negative Pap.

This quantifies the SnNOut/SpPIn trade-off precisely.

---

## Part 5: The Fagan Nomogram — Visual Bayesian Updating

The Fagan nomogram is a graphical tool that lets you move from pre-test probability to post-test probability using likelihood ratios — without doing any calculation.

```{r}
#| label: fig-fagan-via-pap
#| fig-cap: "Pre-Test to Post-Test Probability for VIA and Pap Smear. Red lines = after positive test; Blue lines = after negative test. Pap's positive result (LR+ = 12) shifts probability more dramatically than VIA's (LR+ = 5.3)."
#| echo: false
#| fig-width: 12
#| fig-height: 7

pre_test <- seq(0.01, 0.99, by = 0.01)

calc_post <- function(pre, lr) {
  odds_pre <- pre / (1 - pre)
  odds_post <- odds_pre * lr
  odds_post / (1 + odds_post)
}

fagan_df <- tibble(
  pre = rep(pre_test * 100, 4),
  post = c(
    calc_post(pre_test, lr_pos_via) * 100,
    calc_post(pre_test, lr_neg_via) * 100,
    calc_post(pre_test, lr_pos_pap) * 100,
    calc_post(pre_test, lr_neg_pap) * 100
  ),
  result = rep(c("VIA Positive", "VIA Negative",
                  "Pap Positive", "Pap Negative"), each = length(pre_test)),
  line_type = rep(c("positive", "negative", "positive", "negative"),
                  each = length(pre_test))
)

# Key scenario: 5% pre-test probability
scenarios <- tibble(
  test = c("VIA+", "VIA−", "Pap+", "Pap−"),
  pre = 5,
  post = c(
    calc_post(0.05, lr_pos_via) * 100,
    calc_post(0.05, lr_neg_via) * 100,
    calc_post(0.05, lr_pos_pap) * 100,
    calc_post(0.05, lr_neg_pap) * 100
  ),
  colour = c("#e74c3c", "#3498db", "#c0392b", "#2471a3")
)

ggplot(fagan_df, aes(x = pre, y = post, colour = result)) +
  geom_line(linewidth = 1.1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              colour = "grey60", linewidth = 0.7) +
  # Mark 5% pre-test scenarios
  geom_point(data = scenarios, aes(x = pre, y = post),
             size = 4, inherit.aes = FALSE, colour = "#2c3e50") +
  geom_text(data = scenarios,
            aes(x = pre + 3, y = post, label = paste0(test, "\n", round(post, 1), "%")),
            size = 3.2, fontface = "bold", inherit.aes = FALSE, colour = "#2c3e50") +
  scale_colour_manual(values = c(
    "VIA Positive" = "#e74c3c",
    "VIA Negative" = "#3498db",
    "Pap Positive" = "#c0392b",
    "Pap Negative" = "#2471a3"
  )) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  scale_y_continuous(labels = function(y) paste0(y, "%")) +
  labs(
    title = "Fagan Nomogram: VIA vs Pap Smear",
    subtitle = "At 5% pre-test probability, Pap+ shifts to 39% while VIA+ shifts to only 22%. Pap's higher specificity = more convincing positive.",
    x = "Pre-test Probability",
    y = "Post-test Probability",
    colour = NULL
  ) +
  theme_clean() +
  theme(legend.position = "bottom")
```

:::{.callout-note title="Reading the Fagan Nomogram"}

1. Start on the **x-axis** at your pre-test probability (clinical estimate before any test)
2. Draw a vertical line up to the **relevant curve** (positive or negative result)
3. Read across to the **y-axis** — that's your post-test probability

At 5% pre-test probability:

- **VIA positive** → post-test ~22% (needs confirmation)
- **Pap positive** → post-test ~39% (stronger evidence, may warrant treatment)
- **VIA negative** → post-test ~1.2% (good rule-out)
- **Pap negative** → post-test ~2.2% (adequate rule-out, but worse than VIA)
:::

---

## Part 6: ROC Curves — Choosing the Best Cutoff

Many diagnostic tests don't give a simple "positive/negative" — they produce a **continuous measurement** (blood glucose, tumour marker, antibody titre). We choose a **cutoff** to classify results as positive or negative. Different cutoffs give different sensitivity-specificity trade-offs.

The **ROC (Receiver Operating Characteristic) curve** plots sensitivity vs (1 − specificity) for every possible cutoff, showing you the complete trade-off landscape.

### Building an ROC Curve: Fasting Glucose for Diabetes

```{r}
#| label: fig-roc-glucose
#| fig-cap: "ROC Curve for Fasting Blood Glucose as a Diabetes Screening Test. Each point is a different cutoff value. The curve shows the trade-off: lowering the cutoff catches more diabetics (higher sensitivity) but also more false positives (lower specificity)."
#| echo: false
#| fig-width: 11
#| fig-height: 7

set.seed(42)
n_diabetic <- 300
n_healthy <- 2700

# Fasting glucose distributions
glucose_diabetic <- rnorm(n_diabetic, mean = 145, sd = 35)
glucose_healthy <- rnorm(n_healthy, mean = 90, sd = 15)

# Distribution plot
dist_df <- tibble(
  glucose = c(glucose_diabetic, glucose_healthy),
  status = c(rep("Diabetic", n_diabetic), rep("Non-Diabetic", n_healthy))
)

p_dist <- ggplot(dist_df, aes(x = glucose, fill = status)) +
  geom_density(alpha = 0.5, colour = NA) +
  geom_vline(xintercept = c(100, 126), linetype = c("dashed", "solid"),
             colour = c("#f39c12", "#e74c3c"), linewidth = 0.8) +
  annotate("text", x = 100, y = 0.025, label = "IFG\n(100)",
           size = 3, fontface = "bold", colour = "#f39c12") +
  annotate("text", x = 126, y = 0.025, label = "DM\n(126)",
           size = 3, fontface = "bold", colour = "#e74c3c") +
  scale_fill_manual(values = c("Diabetic" = "#e74c3c", "Non-Diabetic" = "#3498db")) +
  labs(title = "Fasting Glucose Distributions",
       subtitle = "Overlap zone is where diagnostic uncertainty lives",
       x = "Fasting Glucose (mg/dL)", y = "Density", fill = NULL) +
  theme_clean() +
  theme(legend.position = "bottom")

# Calculate ROC
cutoffs <- seq(60, 220, by = 2)

roc_data <- lapply(cutoffs, function(c) {
  tpr <- sum(glucose_diabetic >= c) / n_diabetic
  fpr <- sum(glucose_healthy >= c) / n_healthy
  tibble(cutoff = c, tpr = tpr, fpr = fpr)
}) %>% bind_rows() %>% arrange(fpr)

# AUC
auc_val <- sum(diff(roc_data$fpr) *
               (roc_data$tpr[-1] + roc_data$tpr[-nrow(roc_data)]) / 2)

# Key cutoffs
cut_100 <- roc_data[which.min(abs(roc_data$cutoff - 100)), ]
cut_126 <- roc_data[which.min(abs(roc_data$cutoff - 126)), ]

p_roc <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
  geom_path(colour = "#2c3e50", linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              colour = "grey60", linewidth = 0.7) +
  # Key cutoffs
  geom_point(data = cut_100, colour = "#f39c12", size = 5) +
  geom_point(data = cut_126, colour = "#e74c3c", size = 5) +
  annotate("text", x = cut_100$fpr + 0.06, y = cut_100$tpr,
           label = paste0("Cutoff 100\nSens=", round(cut_100$tpr*100),
                          "%, Spec=", round((1-cut_100$fpr)*100), "%"),
           size = 3.2, fontface = "bold", colour = "#f39c12") +
  annotate("text", x = cut_126$fpr + 0.06, y = cut_126$tpr - 0.05,
           label = paste0("Cutoff 126\nSens=", round(cut_126$tpr*100),
                          "%, Spec=", round((1-cut_126$fpr)*100), "%"),
           size = 3.2, fontface = "bold", colour = "#e74c3c") +
  annotate("text", x = 0.55, y = 0.25,
           label = paste0("AUC = ", round(auc_val, 3)),
           size = 5, fontface = "bold", colour = "#2c3e50") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "ROC Curve: Fasting Glucose for Diabetes",
       x = "False Positive Rate (1 − Specificity)",
       y = "True Positive Rate (Sensitivity)") +
  theme_clean() +
  theme(aspect.ratio = 1)

p_dist + p_roc + plot_layout(widths = c(1, 1.2))
```

### Understanding the ROC Curve

**Moving along the curve:**

- **Top-left corner** = perfect test (sensitivity = 100%, FPR = 0%)
- **Bottom-right** = useless test (sensitivity = 0%, FPR = 100%)
- **Diagonal** = no discrimination (random coin flip)
- **Closer to top-left** = better discrimination

**The two cutoff points:**

- **Cutoff 100 mg/dL (IFG threshold):** High sensitivity (`r round(cut_100$tpr*100)`%), moderate specificity (`r round((1-cut_100$fpr)*100)`%) — good for screening (don't miss pre-diabetics)
- **Cutoff 126 mg/dL (DM threshold):** Lower sensitivity (`r round(cut_126$tpr*100)`%), high specificity (`r round((1-cut_126$fpr)*100)`%) — good for diagnosis (don't overdiagnose)

### AUC: Summarising Overall Test Performance

The **Area Under the ROC Curve (AUC)** captures the test's overall ability to discriminate between diseased and non-diseased individuals across *all* possible cutoffs.

```{r}
#| label: tbl-auc-interpretation
#| echo: false
#| tbl-cap: "AUC Interpretation Scale"

auc_df <- data.frame(
  `AUC Range` = c("0.90 – 1.00", "0.80 – 0.90", "0.70 – 0.80",
                   "0.60 – 0.70", "0.50"),
  Discrimination = c("Excellent", "Good", "Fair", "Poor", "None (coin flip)"),
  `Clinical Analogy` = c(
    "Troponin for MI — near-perfect separation",
    "GeneXpert for TB — reliably distinguishes",
    "CRP for infection — helpful but imperfect",
    "ESR for inflammation — barely better than chance",
    "Random guess — test is worthless"
  ),
  check.names = FALSE
)

kable(auc_df, format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(1, background = "#d5f5e3") %>%
  row_spec(5, background = "#fdedec")
```

**Practical meaning of AUC:** AUC = `r round(auc_val, 2)` means if you randomly pick one diabetic and one non-diabetic person, there's a `r round(auc_val * 100)`% chance the diabetic has a higher fasting glucose.

### Comparing Two Tests: ROC Curves Side by Side

```{r}
#| label: fig-roc-comparison
#| fig-cap: "Comparing Two Diagnostic Tests by ROC Curves. The curve closer to the top-left corner is the better test. The difference in AUC quantifies how much better."
#| echo: false
#| fig-width: 10
#| fig-height: 6

set.seed(42)

# Better test: HbA1c (better separation)
hba1c_diabetic <- rnorm(300, mean = 8.5, sd = 1.5)
hba1c_healthy <- rnorm(2700, mean = 5.5, sd = 0.6)

# Poorer test: random blood glucose (more overlap)
rbg_diabetic <- rnorm(300, mean = 200, sd = 60)
rbg_healthy <- rnorm(2700, mean = 120, sd = 40)

calc_roc <- function(diseased_vals, healthy_vals) {
  all_vals <- c(diseased_vals, healthy_vals)
  cutoffs_seq <- seq(min(all_vals) - 5, max(all_vals) + 5, by = (max(all_vals) - min(all_vals)) / 100)

  roc <- lapply(cutoffs_seq, function(c) {
    tpr <- sum(diseased_vals >= c) / length(diseased_vals)
    fpr <- sum(healthy_vals >= c) / length(healthy_vals)
    tibble(tpr = tpr, fpr = fpr)
  }) %>% bind_rows() %>% arrange(fpr)

  roc_sorted <- roc %>% distinct()
  auc <- sum(diff(roc_sorted$fpr) * (roc_sorted$tpr[-1] + roc_sorted$tpr[-nrow(roc_sorted)]) / 2)
  roc$auc <- auc
  return(roc)
}

roc_hba1c <- calc_roc(hba1c_diabetic, hba1c_healthy) %>%
  mutate(test = paste0("HbA1c (AUC = ", round(first(auc), 2), ")"))
roc_rbg <- calc_roc(rbg_diabetic, rbg_healthy) %>%
  mutate(test = paste0("Random Blood Glucose (AUC = ", round(first(auc), 2), ")"))

roc_both <- bind_rows(roc_hba1c, roc_rbg)

ggplot(roc_both, aes(x = fpr, y = tpr, colour = test)) +
  geom_path(linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              colour = "grey60", linewidth = 0.7) +
  scale_colour_manual(values = c("#27ae60", "#e67e22")) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Comparing Tests: HbA1c vs Random Blood Glucose for Diabetes",
    subtitle = "HbA1c hugs the top-left corner more closely → better discrimination",
    x = "False Positive Rate (1 − Specificity)",
    y = "True Positive Rate (Sensitivity)",
    colour = NULL
  ) +
  theme_clean() +
  theme(aspect.ratio = 1, legend.position = "bottom")
```

---

## Part 7: Choosing the Right Test — When Sensitivity vs Specificity Matters

Different clinical situations demand different test properties. The choice is not about which test is "better" in the abstract — it's about which test fits the **clinical question**.

```{r}
#| label: tbl-test-selection
#| echo: false
#| tbl-cap: "When to Prioritise Sensitivity vs Specificity"

choice_df <- data.frame(
  `Clinical Scenario` = c(
    "Community screening (healthy population)",
    "Confirmation after positive screen",
    "Ruling out a dangerous condition",
    "Disease very rare in population",
    "Disease very common (outbreak)",
    "Treatment is toxic or irreversible"
  ),
  Priority = c(
    "HIGH Sensitivity",
    "HIGH Specificity",
    "HIGH Sensitivity",
    "HIGH Specificity",
    "HIGH Sensitivity",
    "HIGH Specificity"
  ),
  Why = c(
    "Don't miss cases — false negatives are costly",
    "Reduce false positives before invasive treatment",
    "False negative = patient sent home with dangerous disease",
    "Low prevalence → FP vastly outnumber TP → need high Spec",
    "High prevalence → TP outnumber FP → Sens determines yield",
    "False positive → unnecessary harm from treatment"
  ),
  `Indian Example` = c(
    "VIA for cervical cancer; Rapid antigen for COVID",
    "Colposcopy + biopsy after VIA+; RT-PCR after antigen+",
    "D-dimer for PE; CT head for stroke",
    "HIV screening in low-prevalence population",
    "Malaria RDT during monsoon outbreak",
    "Starting chemotherapy based on biopsy"
  ),
  check.names = FALSE
)

kable(choice_df, format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE, width = "20%") %>%
  column_spec(2, bold = TRUE, background = "#ecf0f1")
```

### The Two-Step Strategy: Screen Then Confirm

Most diagnostic pathways use this approach:

1. **Step 1 — Screening test** (high sensitivity, moderate specificity): Cast a wide net, catch as many cases as possible. Accept some false positives.
2. **Step 2 — Confirmatory test** (high specificity, moderate sensitivity): Among those who screened positive, use a specific test to separate true positives from false positives.

```{r}
#| label: fig-two-step-strategy
#| fig-cap: "The Two-Step Diagnostic Strategy: Screen with a sensitive test, then confirm with a specific test. This approach maximises detection while minimising unnecessary treatment."
#| echo: false
#| fig-width: 12
#| fig-height: 6

# Cervical cancer screening: VIA → Colposcopy+Biopsy
step_data <- tibble(
  step = factor(c("Population\n(1000 women)", "VIA Positive\n(Screen)", "Biopsy Confirmed\n(Diagnose)"),
                levels = c("Population\n(1000 women)", "VIA Positive\n(Screen)", "Biopsy Confirmed\n(Diagnose)")),
  total = c(1000, tp_via + fp_via, tp_via),
  true_disease = c(diseased, tp_via, tp_via),
  false_positive = c(0, fp_via, 0)
)

p_funnel <- ggplot(step_data, aes(x = step, y = total)) +
  geom_col(fill = c("#3498db", "#f39c12", "#27ae60"), width = 0.6, alpha = 0.85) +
  geom_text(aes(label = total), vjust = -0.5, fontface = "bold", size = 6) +
  # Annotations inside bars
  geom_text(data = step_data[2, ],
            aes(x = step, y = total/2,
                label = paste0("TP=", tp_via, " + FP=", fp_via)),
            colour = "white", fontface = "bold", size = 4) +
  geom_text(data = step_data[3, ],
            aes(x = step, y = total/2, label = paste0("TP=", tp_via, "\n(all confirmed)")),
            colour = "white", fontface = "bold", size = 3.5) +
  labs(
    title = "Cervical Cancer Screening: The Funnel Approach",
    subtitle = paste0("Screen 1000 → VIA+ = ", tp_via + fp_via,
                      " → Biopsy confirms ", tp_via, " true CIN2+ cases"),
    x = NULL, y = "Number of Women"
  ) +
  theme_clean() +
  theme(axis.text.x = element_text(size = 11, lineheight = 1.0))

# Indian examples
examples_df <- tibble(
  Pathway = c("Cervical Cancer", "Tuberculosis", "HIV", "Breast Cancer"),
  `Screening Test` = c("VIA", "Symptom screening +\nsputum smear", "Rapid antibody test",
                        "Clinical breast exam"),
  `Confirmatory Test` = c("Colposcopy + Biopsy", "GeneXpert MTB/RIF +\nCulture",
                           "Western blot / NAAT", "Mammography + Biopsy"),
  `Screen: Sens / Spec` = c("80% / 85%", "70% / 90%", "99% / 99%", "60% / 95%"),
  `Confirm: Sens / Spec` = c("95% / 95%", "98% / 99%", "99.9% / 99.9%", "90% / 95%")
)

p_funnel
```

```{r}
#| label: tbl-two-step-india
#| echo: false
#| tbl-cap: "Two-Step Diagnostic Pathways Used in Indian Public Health"

kable(examples_df, format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, background = "#fef9e7") %>%
  column_spec(3, background = "#d5f5e3")
```

---

## Part 8: Indian Clinical Scenarios

### Scenario 1: TB Diagnostics — GeneXpert in a Rural PHC

**Setting:** Primary health centre in Bihar; TB prevalence among symptomatic patients ~3%.

**Test:** GeneXpert MTB/RIF — Sensitivity: 98%, Specificity: 99%

```{r}
#| label: fig-tb-scenario
#| fig-cap: "GeneXpert for TB: Excellent discrimination. At 3% prevalence, PPV = 75% and NPV = 99.9%. This is what happens when you have very high specificity — PPV stays useful even at low prevalence."
#| echo: false
#| fig-width: 11
#| fig-height: 6

tb_prev <- 0.03; tb_sens <- 0.98; tb_spec <- 0.99; tb_n <- 10000
tb_dis <- tb_n * tb_prev; tb_hel <- tb_n * (1 - tb_prev)
tb_tp <- round(tb_dis * tb_sens); tb_fn <- tb_dis - tb_tp
tb_fp <- round(tb_hel * (1 - tb_spec)); tb_tn <- tb_hel - tb_fp
tb_ppv <- tb_tp / (tb_tp + tb_fp)
tb_npv <- tb_tn / (tb_fn + tb_tn)
tb_lr_pos <- tb_sens / (1 - tb_spec)
tb_lr_neg <- (1 - tb_sens) / tb_spec

tb_metrics <- tibble(
  metric = factor(c("Sensitivity", "Specificity", "PPV\n(at 3%)", "NPV\n(at 3%)", "LR+", "LR−"),
                  levels = c("Sensitivity", "Specificity", "PPV\n(at 3%)", "NPV\n(at 3%)", "LR+", "LR−")),
  value = c(tb_sens*100, tb_spec*100, tb_ppv*100, tb_npv*100, tb_lr_pos, tb_lr_neg),
  display = c(paste0(round(tb_sens*100), "%"), paste0(round(tb_spec*100), "%"),
              paste0(round(tb_ppv*100), "%"), paste0(round(tb_npv*100, 1), "%"),
              round(tb_lr_pos), round(tb_lr_neg, 3)),
  bar_val = c(98, 99, 75, 99.9, 98, 0.02),
  fill = c("#3498db", "#3498db", "#27ae60", "#27ae60", "#e67e22", "#e67e22")
)

ggplot(tb_metrics %>% filter(metric %in% c("Sensitivity", "Specificity", "PPV\n(at 3%)", "NPV\n(at 3%)")),
       aes(x = metric, y = bar_val, fill = fill)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = display), vjust = -0.5, fontface = "bold", size = 5) +
  scale_fill_identity() +
  scale_y_continuous(limits = c(0, 115)) +
  labs(
    title = "GeneXpert MTB/RIF: Performance at 3% Prevalence",
    subtitle = paste0("LR+ = ", round(tb_lr_pos), " (near-diagnostic) | LR− = ",
                      round(tb_lr_neg, 3), " (strong rule-out)"),
    x = NULL, y = "%"
  ) +
  theme_clean()
```

**Why GeneXpert works so well even at low prevalence:** Its **99% specificity** is the key. At 3% prevalence, there are 9,700 non-diseased people. With 99% specificity, only 97 test falsely positive. Compare this to a test with 90% specificity — that would give 970 false positives, destroying the PPV. **High specificity protects PPV at low prevalence.**

### Scenario 2: Malaria RDT Across Seasons

**Setting:** Monsoon season in a tribal belt of Chhattisgarh — malaria prevalence ~25%.
Dry season in the same region — prevalence drops to ~2%.

**Test:** Rapid Diagnostic Test for *P. falciparum* — Sensitivity: 85%, Specificity: 95%

```{r}
#| label: fig-malaria-season
#| fig-cap: "Malaria RDT: Same test, different seasons, dramatically different PPV. During monsoon (25% prevalence), PPV = 85% — highly useful. During dry season (2%), PPV drops to 26% — most positives are false alarms."
#| echo: false
#| fig-width: 11
#| fig-height: 6

mal_sens <- 0.85; mal_spec <- 0.95

seasons <- tibble(
  season = c("Dry Season\n(2% prevalence)", "Post-Monsoon\n(10%)",
             "Monsoon Peak\n(25%)", "Outbreak\n(50%)"),
  prevalence = c(0.02, 0.10, 0.25, 0.50),
  ppv = sapply(c(0.02, 0.10, 0.25, 0.50), function(p) {
    tp <- p * mal_sens; fp <- (1-p) * (1-mal_spec)
    tp / (tp + fp) * 100
  }),
  npv = sapply(c(0.02, 0.10, 0.25, 0.50), function(p) {
    tn <- (1-p) * mal_spec; fn <- p * (1-mal_sens)
    tn / (tn + fn) * 100
  })
)

seasons_long <- seasons %>%
  pivot_longer(c(ppv, npv), names_to = "metric", values_to = "value") %>%
  mutate(metric = ifelse(metric == "ppv", "PPV", "NPV"))

ggplot(seasons_long, aes(x = season, y = value, fill = metric)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = paste0(round(value), "%")),
            position = position_dodge(width = 0.6),
            vjust = -0.5, fontface = "bold", size = 4) +
  scale_fill_manual(values = c("PPV" = "#e74c3c", "NPV" = "#3498db")) +
  scale_y_continuous(limits = c(0, 110)) +
  labs(
    title = "Malaria RDT Performance Across Seasons (Sens=85%, Spec=95%)",
    subtitle = "PPV swings from 26% to 94% depending on season. Same test, same lab, different clinical utility.",
    x = NULL, y = "%", fill = NULL
  ) +
  theme_clean()
```

**Clinical action:** During monsoon, a positive RDT is reliable enough to start treatment. During dry season, a positive RDT needs confirmation with peripheral smear microscopy.

### Scenario 3: Cervical Cancer Screening — VIA in the "See and Treat" Model

```{r}
#| label: fig-via-see-treat
#| fig-cap: "The 'See and Treat' Model: VIA's strength is not in PPV — it's in the ability to immediately treat positives with cryotherapy on the same visit, preventing loss to follow-up."
#| echo: false
#| fig-width: 12
#| fig-height: 5.5

see_treat <- tibble(
  step = factor(c("Women\nScreened\n(1000)", "VIA\nPositive\n(182)",
                  "Cryotherapy\nSame Day\n(182)", "Overtreatment\n(142 FP)",
                  "Cancers\nPrevented\n(~40 TP)"),
                levels = c("Women\nScreened\n(1000)", "VIA\nPositive\n(182)",
                  "Cryotherapy\nSame Day\n(182)", "Overtreatment\n(142 FP)",
                  "Cancers\nPrevented\n(~40 TP)")),
  count = c(1000, tp_via + fp_via, tp_via + fp_via, fp_via, tp_via),
  fill = c("#3498db", "#f39c12", "#e67e22", "#e74c3c", "#27ae60")
)

ggplot(see_treat, aes(x = step, y = count, fill = fill)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = count), vjust = -0.5, fontface = "bold", size = 5) +
  scale_fill_identity() +
  labs(
    title = "VIA 'See and Treat' Model for Rural Cervical Cancer Prevention",
    subtitle = "142 women overtreated (cryotherapy is safe) to prevent ~40 cancers. Acceptable trade-off in resource-limited settings.",
    x = NULL, y = "Number of Women"
  ) +
  theme_clean() +
  theme(axis.text.x = element_text(size = 9, lineheight = 1.0))
```

**The programme logic:** In rural India where follow-up is unreliable, VIA + immediate cryotherapy catches 80% of precancers in a single visit. The 142 women who receive unnecessary cryotherapy (false positives) experience minimal harm — cryotherapy is a simple, safe outpatient procedure. But the 40 women whose precancers are caught and treated are spared invasive cervical cancer. **The trade-off is overwhelmingly favourable.**

This is a clear example of why test selection depends on the **healthcare context**, not just the statistical properties.

---

## Summary and Key Takeaways

**What this module added to Module 4's foundation:**

1. **The 2×2 table** is the organising structure — index test in rows, gold standard in columns, FP in the top-right corner

2. **Sensitivity and specificity** are intrinsic test properties that don't change across populations. **PPV and NPV** do change with prevalence.

3. **Likelihood ratios** are the prevalence-independent bridge — LR+ > 10 is near-diagnostic, LR− < 0.1 is near-exclusion

4. **ROC curves** show the complete sensitivity-specificity trade-off for continuous tests. **AUC** summarises overall discrimination.

5. **Test selection** depends on clinical context: **screening → high sensitivity** (SnNOut); **confirmation → high specificity** (SpPIn)

6. **Two-step strategy** (screen then confirm) is the standard approach in public health and clinical practice

7. **Context matters enormously** — VIA was chosen for India not because it has the best PPV, but because it fits a "see and treat" model that prevents loss to follow-up

---

## Further Learning Resources

:::{.resources-box}
### Video Lectures

**StatQuest with Josh Starmer** — "Sensitivity and Specificity" (YouTube)
Clear visual explanation of the 2×2 table and its derivatives.

**StatQuest** — "ROC and AUC, Clearly Explained!" (YouTube)
The definitive visual guide to ROC curves.

**Zedstatistics (Andrew Mead)** — "Diagnostic Test Evaluation" playlist (YouTube)
Step-by-step walkthrough with clinical examples.

### Textbooks

**Sackett DL et al (2005). *Clinical Epidemiology: A Basic Science for Clinical Medicine.* 3rd ed. Ch 4: Diagnosis.**
Gold standard for diagnostic test evaluation in clinical context.

**Kirkwood BR & Sterne JAC (2003). *Essential Medical Statistics.* 2nd ed. Ch 36: Diagnostic tests and screening.**
Clear statistical treatment with worked examples.

### Indian Context

**NTEP (National TB Elimination Program) Guidelines** — GeneXpert sensitivity/specificity recommendations for Indian populations.

**Operational Framework for Cervical Cancer Screening and Management (MOHFW, India)** — VIA implementation guidelines, "see and treat" protocol.

**NVBDCP Guidelines** — Malaria RDT use across endemic and non-endemic zones.
:::

---

## Practice MCQs: NEET PG Level

:::{.neet-practice}

**Q1.** In the standard 2×2 table for diagnostic test evaluation with the index test in rows and gold standard in columns, which cell represents **false positives**?

```{r}
#| label: mcq-1
#| echo: false
#| results: asis
make_mcq(
  id = "m5q1",
  question = "",
  options = c(
    "Top-left (Test+, Disease+)" = "This is the True Positive cell — the test correctly identified a diseased person.",
    "answer:Top-right (Test+, Disease−)" = "Correct! In the standard layout (test in rows, gold standard in columns), the top-right cell is where the test says positive but the gold standard says no disease — a false positive (false alarm). This is the cell that inflates when specificity is low.",
    "Bottom-left (Test−, Disease+)" = "This is the False Negative cell — the test missed a diseased person. A very different kind of error from a false positive.",
    "Bottom-right (Test−, Disease−)" = "This is the True Negative cell — the test correctly identified a non-diseased person."
  )
)
```

---

**Q2.** VIA screening for cervical cancer has 80% sensitivity and 85% specificity. Pap smear has 60% sensitivity and 95% specificity. In a population screening programme where follow-up is unreliable, which test is preferred and why?

```{r}
#| label: mcq-2
#| echo: false
#| results: asis
make_mcq(
  id = "m5q2",
  question = "",
  options = c(
    "Pap smear because it has higher specificity" = "Higher specificity reduces false positives, but in a setting where follow-up is unreliable, missing cases (false negatives) is the greater danger. A woman who screens negative and is lost to follow-up may progress to invasive cancer.",
    "answer:VIA because higher sensitivity catches more cases and allows same-day treatment" = "Correct! VIA's higher sensitivity (80% vs 60%) means fewer missed precancers. Combined with the 'see and treat' model (cryotherapy on the same visit), VIA prevents loss to follow-up — a critical advantage in rural India. The higher false positive rate is acceptable because cryotherapy is safe and inexpensive.",
    "Pap smear because it has higher PPV" = "While Pap does have higher PPV due to its specificity, PPV alone doesn't determine the best screening strategy. The clinical context — unreliable follow-up, need for same-day treatment — makes sensitivity and logistics more important.",
    "Neither — both are equally good" = "The two tests have very different sensitivity-specificity profiles and serve different clinical roles. In resource-limited settings with unreliable follow-up, VIA is clearly preferred for population screening."
  )
)
```

---

**Q3.** A tuberculosis test has sensitivity 98% and specificity 99%. At 3% prevalence, 1,000 patients are tested. How many false positives will occur?

```{r}
#| label: mcq-3
#| echo: false
#| results: asis
make_mcq(
  id = "m5q3",
  question = "",
  options = c(
    "20" = "This would be the false negative count: 1000 × 0.03 × (1-0.98) = 0.6, rounded. But 20 comes from misapplying (1-sensitivity) to the wrong denominator.",
    "answer:About 10" = "Correct! False positives come from the non-diseased group. Non-diseased = 1000 × 0.97 = 970. FP = 970 × (1 − 0.99) = 970 × 0.01 = 9.7 ≈ 10. This is why 99% specificity is so valuable — even with 970 non-diseased patients, only ~10 test falsely positive.",
    "30" = "This treats 3% as the false positive rate, but specificity is 99%, making the false positive rate 1%. FP = non-diseased × (1-specificity) = 970 × 0.01 ≈ 10.",
    "97" = "This would be the result if specificity were 90% (970 × 0.10 = 97). With 99% specificity, the false positive rate is only 1%."
  )
)
```

---

**Q4.** A malaria RDT has sensitivity 85% and specificity 95%. During monsoon (prevalence 25%), LR+ = 17. During dry season (prevalence 2%), LR+ is still 17. Why does PPV change so dramatically between seasons if LR+ stays the same?

```{r}
#| label: mcq-4
#| echo: false
#| results: asis
make_mcq(
  id = "m5q4",
  question = "",
  options = c(
    "LR+ actually changes between seasons" = "LR+ is calculated from sensitivity and specificity, which are intrinsic properties of the test. They don't change between seasons. LR+ = Sensitivity / (1 − Specificity) = 0.85 / 0.05 = 17 regardless of prevalence.",
    "The lab technician performs differently in monsoon" = "The test characteristics (sensitivity and specificity) are determined by the test kit, not the season or technician. LR+ remains constant.",
    "answer:PPV depends on pre-test probability (prevalence), not just on LR+. LR+ multiplies the pre-test odds — different starting odds give different post-test odds even with the same LR+" = "Correct! Post-test odds = pre-test odds × LR+. During monsoon, pre-test odds = 0.25/0.75 = 0.33. Post-test odds = 0.33 × 17 = 5.67 → PPV = 85%. During dry season, pre-test odds = 0.02/0.98 = 0.02. Post-test odds = 0.02 × 17 = 0.35 → PPV = 26%. Same multiplication, different starting point, vastly different result.",
    "PPV doesn't actually change — only NPV does" = "Both PPV and NPV change with prevalence. PPV increases with higher prevalence (more true disease cases to find), and NPV decreases (more false negatives among the larger diseased group)."
  )
)
```

---

**Q5.** An ROC curve for a new biomarker has AUC = 0.55. What does this mean clinically?

```{r}
#| label: mcq-5
#| echo: false
#| results: asis
make_mcq(
  id = "m5q5",
  question = "",
  options = c(
    "The test has 55% sensitivity" = "AUC is not the same as sensitivity. AUC summarises the test's discrimination ability across ALL possible cutoffs, not at any single cutoff.",
    "The test correctly classifies 55% of patients" = "AUC is not an accuracy measure. It represents the probability that a randomly chosen diseased person has a higher test value than a randomly chosen non-diseased person.",
    "answer:The test has essentially no clinical value — it barely beats a coin flip (AUC = 0.5)" = "Correct! AUC = 0.5 means random guessing (no discrimination). AUC = 0.55 is barely above chance — if you randomly pick one diseased and one non-diseased person, there's only a 55% chance the test gives a higher value to the diseased person. This test should not be used for clinical decisions.",
    "The test is moderately good for diagnosis" = "Moderate discrimination typically requires AUC of 0.70–0.80. An AUC of 0.55 is in the 'poor to none' range — only 5% above random guessing."
  )
)
```

:::

---

## References

::: {#refs}
:::
