---
title: "Study Design, Bias, and Confounding"
subtitle: "Why Study Design Determines What You Can Conclude"
description: |
  Learn the spectrum of study designs from case reports to RCTs, understand
  selection bias, information bias, and confounding, and use DAGs to reason
  about causal inference. Indian clinical examples throughout.
categories:
  - Study Design
  - Bias
  - Confounding
  - Epidemiology
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(kableExtra)
source("../R/mcq.R")
library(patchwork)
library(scales)
library(DiagrammeR)

set.seed(42)

theme_clean <- function() {
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
}
```

::: {.callout-note appearance="minimal"}
**Lecture slides for this module:** [Open Slides](../slides/07-study-design-slides.html){target="_blank"}
:::

## Hook: Does Wine Prevent Heart Disease?

A landmark 10-year cohort study followed 20,000 adults and found that light wine drinkers had 30% lower risk of myocardial infarction compared to non-drinkers. Newspapers proclaimed: **"Wine is Cardioprotective!"** Cardiologists recommended moderate wine consumption as a preventive measure.

But careful reading reveals the study has three critical problems:

1. **Selection bias (healthy user bias)**: Non-drinkers include former alcoholics who quit *because* of prior health problems. They're inherently sicker.
2. **Confounding**: Wine drinkers have higher socioeconomic status (SES), better access to healthcare, exercise more regularly, and eat Mediterranean diets.
3. **Reverse causality**: Sick people quit drinking on doctor's orders. The low risk in "non-drinkers" reflects pre-existing health advantage.

**The conclusion the headlines drew is almost certainly wrong.** Yet the data itself is real. What went wrong? The study design didn't prevent the researchers from drawing false conclusions.

This module teaches you how to recognize which study designs allow which conclusions, and which biases can undermine even well-intentioned research.

> **Every clinical question can be studied with different designs. Each design has different strengths and weaknesses for establishing causality. This is the most important skill in evidence-based medicine.**

## Part 1: The Spectrum of Study Designs

The most fundamental classification of study designs relates to two dimensions:

1. **Direction of inquiry**: Does the study move forward in time (prospective) or backward (retrospective)?
2. **Control of assignment**: Are participants randomly assigned to exposure (experimental) or observed as they naturally occur (observational)?

### Visual Overview: Hierarchy and Spectrum

```{r}
#| echo: false
# Create a visual hierarchy
hierarchy_data <- tibble(
  design = factor(c("Case Series / Reports", "Cross-Sectional", "Case-Control",
                     "Cohort Study", "RCT", "Systematic Review"),
                  levels = c("Case Series / Reports", "Cross-Sectional", "Case-Control",
                            "Cohort Study", "RCT", "Systematic Review"),
                  ordered = TRUE),
  evidence_strength = c(1, 2, 3, 4, 5, 6),
  y_position = c(1, 2, 3, 4, 5, 6)
)

ggplot(hierarchy_data, aes(x = 0.5, y = y_position)) +
  geom_tile(aes(fill = evidence_strength, height = 0.8), width = 0.6, color = "white", linewidth = 1) +
  geom_text(aes(label = design), size = 4, color = "white", fontface = "bold") +
  scale_fill_gradient(low = "#FFA500", high = "#4CAF50", guide = "none") +
  xlim(0, 1) +
  ylim(0, 7) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 13),
    plot.subtitle = element_text(hjust = 0.5, size = 11, color = "gray40")
  ) +
  labs(
    title = "Hierarchy of Evidence for Causality",
    subtitle = "Green = Stronger evidence for causality | Orange = Weaker causal inference"
  )
```

### Timeline and Direction of Inquiry

```{r}
#| echo: false
# Create a timeline visualization showing different designs
timeline_data <- tibble(
  design = c("Case Report", "Cross-Sectional", "Case-Control", "Cohort", "RCT"),
  x_start = c(0, 0, 0.2, 0, 0),
  x_end = c(1, 1, 1, 1.2, 1.2),
  y_pos = c(5, 4, 3, 2, 1),
  color = c("gray60", "orange", "red", "blue", "green")
)

ggplot() +
  # Timeline axis
  geom_segment(x = 0, xend = 1.5, y = 0.3, yend = 0.3, color = "black", size = 1) +
  geom_text(x = c(0, 0.5, 1, 1.5), y = 0.05, label = c("PAST", "NOW", "FUTURE", ""), size = 3.5, hjust = 0.5) +

  # Design arrows
  geom_segment(aes(x = x_start, xend = x_end, y = y_pos, yend = y_pos, color = color),
               data = timeline_data, linewidth = 1.5, arrow = arrow(length = unit(0.2, "cm"))) +
  geom_text(aes(x = (x_start + x_end)/2, y = y_pos + 0.25, label = design, color = color),
            data = timeline_data, size = 3.5, fontface = "bold") +

  # Add direction labels
  geom_text(x = 0.1, y = 3.3, label = "← RETROSPECTIVE", size = 2.5, color = "red", angle = 0) +
  geom_text(x = 1.1, y = 1.3, label = "PROSPECTIVE →", size = 2.5, color = "green", angle = 0) +

  scale_color_identity() +
  xlim(-0.1, 1.6) +
  ylim(-0.3, 5.5) +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13)) +
  labs(title = "Direction of Inquiry in Different Study Designs")
```

### Key Distinctions

::: {.key-concept}
**Observational vs. Experimental Studies:**
- **Observational**: Researchers observe people as they naturally occur (exposed/unexposed). Cannot establish causality confidently because confounding is inevitable.
- **Experimental**: Researchers randomly assign people to exposure/no exposure. Randomization breaks confounding.
:::

## Part 2: Cross-Sectional Studies

### Design and Measures

A **cross-sectional study** is a snapshot. At one point in time, you measure both exposure and outcome in a defined population. Examples:
- NFHS survey measuring anemia prevalence by state
- A hospital survey asking patients about diet and cholesterol at the same visit
- Sentinel surveillance for COVID-19 prevalence

**Key limitation**: No temporal sequence. You don't know if exposure came before outcome.

```{r}
#| echo: false
grViz("
digraph cross_sectional {
  graph[rankdir = LR, bgcolor = 'transparent', nodesep = 0.6, ranksep = 0.8]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 11, margin = '0.2,0.1']

  A [label = 'Population', fillcolor = '#B8D4E3']
  B [label = 'Exposed', fillcolor = '#E8C07A']
  C [label = 'Non-exposed', fillcolor = '#A8D5A2']
  D [label = 'Presence of\\noutcome', fillcolor = '#E89595']
  E [label = 'Absence of\\noutcome', fillcolor = '#D4EAF7']
  F [label = 'Presence of\\noutcome', fillcolor = '#E89595']
  G [label = 'Absence of\\noutcome', fillcolor = '#D4EAF7']
  H [label = 'Prevalence of outcome\\nin Exposed', fillcolor = '#F0D4E8']
  I [label = 'Prevalence of outcome\\nin Non-exposed', fillcolor = '#F0D4E8']

  edge [label = '  classify  ', fontsize = 9, fontcolor = '#555555']
  A -> B; A -> C
  edge [label = '  assess  ', fontsize = 9, fontcolor = '#555555']
  B -> D; B -> E; C -> F; C -> G
  edge [arrowhead = none, arrowtail = none, style = tapered, label = '  estimate  ', fontsize = 9, fontcolor = '#555555']
  D -> H; E -> H; F -> I; G -> I
}
")
```

### Measures: Prevalence, Not Incidence

In cross-sectional studies, we calculate **prevalence** and the **prevalence odds ratio (POR)**—NOT relative risk.

::: {.formula-box}
**Prevalence** = (Number with outcome) / (Total population)

**Prevalence Odds Ratio (POR)** = (Odds of outcome in exposed) / (Odds of outcome in unexposed)

$$\text{POR} = \frac{a/b}{c/d} = \frac{ad}{bc}$$

where $a$ = exposed + outcome, $b$ = exposed – outcome, $c$ = unexposed + outcome, $d$ = unexposed – outcome
:::

::: {.common-mistake}
**Mistake**: Reporting "relative risk" from a cross-sectional study.

Cross-sectional studies measure **prevalence**, and the odds ratio approximates relative risk only when the outcome is rare (<10%). If anemia affects 40% of women, the POR is NOT a relative risk.
:::

### Example: Anemia in India (NFHS-like Data)

```{r}
#| echo: false
# Simulate NFHS-like anemia data by state
states <- c("Bihar", "Jharkhand", "Odisha", "Chhattisgarh", "Madhya Pradesh",
            "Rajasthan", "Uttar Pradesh", "Tamil Nadu", "Kerala")
anemia_prev <- c(65.3, 62.8, 60.2, 58.5, 55.1, 52.3, 50.5, 35.2, 28.4)

anemia_data <- tibble(
  state = factor(states, levels = states),
  prevalence = anemia_prev
)

ggplot(anemia_data, aes(x = reorder(state, -prevalence), y = prevalence, fill = prevalence)) +
  geom_col(color = "black", linewidth = 0.5) +
  geom_text(aes(label = paste0(round(prevalence, 1), "%")), vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_fill_gradient(low = "#FFE5B4", high = "#8B0000", guide = "none") +
  scale_y_continuous(limits = c(0, 75), labels = label_percent(scale = 1)) +
  labs(
    title = "Cross-Sectional Prevalence: Anemia in Indian States",
    subtitle = "Percentage of women with hemoglobin <12 g/dL",
    x = "State",
    y = "Prevalence of Anemia (%)"
  ) +
  theme_clean() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Strengths and Limitations

| Strength | Limitation |
|----------|-----------|
| Fast and inexpensive | No temporal sequence → cannot establish causality |
| Good for measuring burden of disease | Exposure may have changed since outcome occurred |
| Large sample sizes possible | Survivor bias (those with outcome who survived) |
| Useful for hypothesis generation | Recall bias in measuring past exposures |

## Part 3: Case-Control Studies

### Design: Starting with the Outcome

A **case-control study** reverses the usual causal pathway. You:
1. **Identify cases** (people with the outcome)
2. **Identify controls** (people without the outcome, from the same source)
3. **Look backward** to compare exposure frequency between groups

```{r}
#| echo: false
grViz("
digraph case_control {
  graph[rankdir = RL, bgcolor = 'transparent', nodesep = 0.6, ranksep = 0.8]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 11, margin = '0.2,0.1']

  A [label = 'Population', fillcolor = '#B8D4E3']
  B [label = 'Cases', fillcolor = '#E8C07A']
  C [label = 'Non-Cases', fillcolor = '#A8D5A2']
  D [label = 'Exposed', fillcolor = '#E89595']
  E [label = 'Non-Exposed', fillcolor = '#D4EAF7']
  F [label = 'Exposed', fillcolor = '#E89595']
  G [label = 'Non-Exposed', fillcolor = '#D4EAF7']
  H [label = 'Odds of Exposure\\nin Cases', fillcolor = '#F0D4E8']
  I [label = 'Odds of Exposure\\nin Non-Cases', fillcolor = '#F0D4E8']

  edge [label = '  select  ', fontsize = 9, fontcolor = '#555555']
  A -> B; A -> C
  edge [label = '  assess  ', fontsize = 9, fontcolor = '#555555']
  B -> D; B -> E; C -> F; C -> G
  edge [arrowhead = none, arrowtail = none, style = tapered, label = '  estimate  ', fontsize = 9, fontcolor = '#555555']
  D -> H; E -> H; F -> I; G -> I
}
")
```

### Measure: Odds Ratio

The **odds ratio** is the measure of association in case-control studies.

::: {.formula-box}
**Odds Ratio (OR)** = $\frac{a \times d}{b \times c}$

where:
- $a$ = cases with exposure
- $b$ = cases without exposure
- $c$ = controls with exposure
- $d$ = controls without exposure

**Interpretation**: An OR of 3.0 means the odds of exposure among cases is 3 times higher than among controls.
:::

### When to Use Case-Control Studies

Case-control studies are ideal when:

1. **Outcome is rare**: Following 10,000 people for 5 years to see if 3 get cancer is expensive. Starting with 200 cancer cases is much faster.
2. **Long latency period**: If a chemical exposure today causes cancer 20 years later, a cohort study would take 20 years. Case-control studies work with existing cases.
3. **Expensive exposure measurement**: If measuring blood PCB levels is costly, measure it only in cases and a matched sample of controls.

### Critical Issue: Control Selection

::: {.clinical-hook}
**The Oral Cancer Case-Control Study in North India**

A study aimed to identify risk factors for oral cancer, comparing 500 patients with oral cancer to 500 hospital controls. The control group consisted of patients admitted for other reasons (dental problems, gastroenteritis, orthopedic injuries).

**Problem**: The controls had different exposure patterns because they came from the same hospital. People admitted to a hospital are ill. The "healthy" reference we want doesn't match hospital patients.

**Result**: Chewing tobacco appeared less strongly associated with oral cancer than it should be, because the controls also had high tobacco use (common in North India).

**Solution**: Controls should come from the same source population as cases—for example, from the same neighborhoods where the cases were identified.
:::

::: {.key-concept}
**Berkson's Bias** (Hospital-Based Selection Bias)

When hospital-based controls are used, they don't represent the source population. Berkson's bias occurs because hospitalization itself is associated with certain exposures and conditions.

Hospital patients are:
- Sicker than the general population
- More likely to have multiple conditions
- Have different exposure patterns (more tobacco use, alcohol, occupational exposures)

**Always use population-based controls** whenever possible.
:::

### Other Biases in Case-Control Studies

**Recall bias**: Cases with disease often remember past exposures better (or blame past exposure for their illness), while controls forget. An oral cancer patient may more vividly recall 20 years of tobacco chewing than a healthy person recalls.

```{r}
#| echo: false
# Create a detailed 2x2 table visualization
cc_table_data <- tibble(
  group = c("Cases", "Cases", "Controls", "Controls", "Total", "Total"),
  exposure = c("Exposed (a)", "Unexposed (b)", "Exposed (c)", "Unexposed (d)", "Exposed", "Unexposed"),
  n = c(60, 90, 40, 110, 100, 200),
  row = c(1, 1, 2, 2, 3, 3),
  col = c(1, 2, 1, 2, 1, 2)
)

# Create 2x2 table with calculations
table_text <- "
2×2 TABLE: ORAL CANCER AND TOBACCO CHEWING

                    EXPOSED (a)    UNEXPOSED (b)    TOTAL
CASES                    60             90            150
CONTROLS                 40            110            150
TOTAL                   100            200            300

OR = (a × d) / (b × c) = (60 × 110) / (90 × 40) = 6600 / 3600 = 1.83

Interpretation: Cases are 1.83 times more likely to have been exposed to tobacco than controls.
"

ggplot() +
  annotate("text", x = 0.5, y = 0.5, label = table_text,
           size = 3.5, family = "mono", hjust = 0.5, vjust = 0.5,
           color = "black") +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = "black", linewidth = 1)
  )
```

### Example: Oral Cancer and Paan Chewing

```{r}
#| echo: false
# Simulate results from oral cancer study
or_data <- tibble(
  exposure_type = c("No paan", "Occasional", "Regular", "Heavy"),
  or_value = c(1.0, 2.5, 5.2, 8.7),
  ci_lower = c(0.8, 2.0, 4.1, 6.9),
  ci_upper = c(1.2, 3.2, 6.6, 11.2)
)

ggplot(or_data, aes(x = reorder(exposure_type, or_value), y = or_value)) +
  geom_point(size = 3.5, color = "darkred") +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2, color = "darkred", linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_text(aes(label = paste0("OR = ", round(or_value, 2))), hjust = -0.5, vjust = -0.5, size = 3) +
  scale_y_log10() +
  coord_flip() +
  labs(
    title = "Dose-Response: Oral Cancer and Paan Chewing",
    subtitle = "Case-control study in North India (n=500 cases, 500 controls)",
    x = "Paan Chewing Frequency",
    y = "Odds Ratio (log scale)",
    caption = "Error bars = 95% confidence interval"
  ) +
  theme_clean() +
  theme(axis.text.y = element_text(hjust = 1))
```

## Part 4: Cohort Studies

### Design: Following Exposed and Unexposed Forward

A **cohort study** follows the natural causal sequence:

1. **Identify exposed and unexposed** people (or different levels of exposure)
2. **Follow them forward** in time
3. **Measure outcome** incidence in each group

Unlike case-control studies, cohort studies measure **incidence** (new cases) and **relative risk**.

```{r}
#| echo: false
grViz("
digraph cohort {
  graph[rankdir = LR, bgcolor = 'transparent', nodesep = 0.6, ranksep = 0.8]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 11, margin = '0.2,0.1']

  A [label = 'Population', fillcolor = '#B8D4E3']
  B [label = 'Exposed', fillcolor = '#E8C07A']
  C [label = 'Non-exposed', fillcolor = '#A8D5A2']
  D [label = 'Incident Cases', fillcolor = '#E89595']
  E [label = 'Do not develop\\ndisease', fillcolor = '#D4EAF7']
  F [label = 'Incident Cases', fillcolor = '#E89595']
  G [label = 'Do not develop\\ndisease', fillcolor = '#D4EAF7']
  H [label = 'Incidence of disease\\nin Exposed', fillcolor = '#F0D4E8']
  I [label = 'Incidence of disease\\nin Non-exposed', fillcolor = '#F0D4E8']

  edge [label = '  ascertain  ', fontsize = 9, fontcolor = '#555555']
  A -> B; A -> C
  edge [label = '  follow-up  ', fontsize = 9, fontcolor = '#555555']
  B -> D; B -> E; C -> F; C -> G
  edge [arrowhead = none, arrowtail = none, style = tapered, label = '  estimate  ', fontsize = 9, fontcolor = '#555555']
  D -> H; E -> H; F -> I; G -> I
}
")
```

### Measures: Relative Risk and Absolute Risk

::: {.formula-box}
**Relative Risk (RR)** = $\frac{\text{Incidence in exposed}}{\text{Incidence in unexposed}} = \frac{a/(a+b)}{c/(c+d)}$

**Absolute Risk Reduction (ARR)** = Incidence(unexposed) − Incidence(exposed)

**Number Needed to Treat (NNT)** = $\frac{1}{\text{ARR}}$

where:
- $a$ = exposed with outcome
- $b$ = exposed without outcome
- $c$ = unexposed with outcome
- $d$ = unexposed without outcome
:::

### Prospective vs. Retrospective Cohorts

**Prospective cohorts**: Follow people forward from NOW into the FUTURE.
- Gold standard but slow and expensive
- Example: Framingham Heart Study (started 1948, still ongoing)

**Retrospective cohorts**: Use existing records (medical charts, insurance databases) to reconstruct exposure and outcome.
- Much faster and cheaper
- Relies on quality of historical records
- Example: Using hospital discharge records to follow patients with diabetes

### Example: Tobacco and Oral Cancer in a Cohort Study

```{r}
#| echo: false
# Simulate cohort data with follow-up
cohort_example <- tibble(
  exposure = c("Tobacco Users", "Non-Users"),
  total_n = c(1000, 1000),
  cases = c(45, 8),
  follow_up_years = c(10, 10)
)

cohort_example <- cohort_example %>%
  mutate(
    incidence_per_1000py = (cases / (total_n * follow_up_years)) * 1000,
    rr = incidence_per_1000py / incidence_per_1000py[2]
  )

ggplot(cohort_example, aes(x = exposure, y = incidence_per_1000py, fill = exposure)) +
  geom_col(color = "black", linewidth = 1, width = 0.6) +
  geom_text(aes(label = paste0(round(incidence_per_1000py, 1), " per 1000 PY")),
            vjust = -0.5, size = 3.5, fontface = "bold") +
  geom_text(aes(label = paste0("Cases: ", cases), y = incidence_per_1000py * 0.5),
            vjust = 0.5, size = 3, fontface = "bold", color = "white") +
  scale_fill_manual(values = c("darkred", "steelblue"), guide = "none") +
  labs(
    title = "Cohort Study: Incidence Rates and Relative Risk",
    subtitle = "10-year prospective follow-up of tobacco users vs. non-users (n=1000 each)",
    x = "Exposure Group",
    y = "Incidence Rate per 1000 Person-Years",
    caption = "RR = 45.0 / 8.0 = 5.6 (tobacco users have 5.6× higher risk)"
  ) +
  theme_clean() +
  theme(
    axis.text.x = element_text(size = 11)
  )
```

### Strengths and Limitations of Cohort Studies

| Strength | Limitation |
|----------|-----------|
| Clear temporal sequence (exposure before outcome) | Expensive and time-consuming |
| Can measure incidence and RR | Loss to follow-up can bias results |
| Can study multiple outcomes | Requires large sample sizes |
| Less prone to recall bias | Confounding is still possible |

## Part 5: Randomized Controlled Trials (RCTs)

### Why Randomization Breaks Confounding

The **randomized controlled trial** is the gold standard for establishing causality in medicine. Here's why:

Randomization *breaks the causal link between confounders and exposure*. It distributes all confounders (known and unknown) equally between the exposed and control groups.

```{r}
#| echo: false
# Simulate how randomization balances confounders
set.seed(42)
n <- 200

rct_sim <- tibble(
  id = 1:n,
  age = rnorm(n, 50, 15),
  ses = rnorm(n, 50, 20),
  group = rep(c("Drug", "Placebo"), each = n/2)
)

# Compare means before and after randomization
confounder_summary <- rct_sim %>%
  group_by(group) %>%
  summarise(
    mean_age = mean(age),
    mean_ses = mean(ses),
    sd_age = sd(age),
    sd_ses = sd(ses),
    .groups = "drop"
  )

# Plot comparison
p1 <- ggplot(rct_sim, aes(x = group, y = age, fill = group)) +
  geom_boxplot(alpha = 0.6, color = "black") +
  geom_jitter(width = 0.2, alpha = 0.3, size = 2) +
  scale_fill_manual(values = c("darkblue", "lightblue"), guide = "none") +
  labs(
    title = "Age Distribution by Group",
    x = "Group",
    y = "Age (years)"
  ) +
  theme_clean()

p2 <- ggplot(rct_sim, aes(x = group, y = ses, fill = group)) +
  geom_boxplot(alpha = 0.6, color = "black") +
  geom_jitter(width = 0.2, alpha = 0.3, size = 2) +
  scale_fill_manual(values = c("darkred", "lightcoral"), guide = "none") +
  labs(
    title = "Socioeconomic Status by Group",
    x = "Group",
    y = "SES Score"
  ) +
  theme_clean()

p1 + p2 +
  plot_annotation(
    title = "Randomization: Confounders Balanced Between Groups",
    subtitle = "After random assignment, treatment and control groups have similar age and SES distributions"
  )
```

### Key Features of RCTs

**1. Allocation Concealment**: Researchers don't know which participant will get drug vs. placebo until after enrollment. This prevents bias in who gets assigned to which group.

**2. Blinding**:
- **Single-blind**: Participants don't know if they're getting drug or placebo
- **Double-blind**: Neither participants nor researchers know
- **Triple-blind**: Participants, researchers, AND analysts don't know until after analysis

```{r}
#| echo: false
grViz("
digraph rct {
  graph[rankdir = LR, bgcolor = 'transparent', nodesep = 0.6, ranksep = 0.8]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 11, margin = '0.2,0.1']

  A [label = 'Population', fillcolor = '#B8D4E3']
  B [label = 'Intervention', fillcolor = '#E8C07A']
  C [label = 'Placebo or\\nStandard Care', fillcolor = '#A8D5A2']
  D [label = 'Meets end point\\nor Outcome', fillcolor = '#E89595']
  E [label = 'Does not meet\\nend point', fillcolor = '#D4EAF7']
  F [label = 'Meets end point\\nor Outcome', fillcolor = '#E89595']
  G [label = 'Does not meet\\nend point', fillcolor = '#D4EAF7']
  H [label = 'Rate of event\\nin Intervention', fillcolor = '#F0D4E8']
  I [label = 'Rate of event\\nin Placebo or SOC', fillcolor = '#F0D4E8']

  edge [label = '  randomization  ', fontsize = 9, fontcolor = '#555555']
  A -> B; A -> C
  edge [label = '  follow-up  ', fontsize = 9, fontcolor = '#555555']
  B -> D; B -> E; C -> F; C -> G
  edge [arrowhead = none, arrowtail = none, style = tapered, label = '  estimate  ', fontsize = 9, fontcolor = '#555555']
  D -> H; E -> H; F -> I; G -> I
}
")
```

### ITT vs. Per-Protocol Analysis

Two analysis approaches exist:

**Intention-to-Treat (ITT)**: Analyze all randomized participants in their assigned groups, whether they took the drug or not.
- Preserves randomization
- Reflects real-world effectiveness
- May underestimate true efficacy

**Per-Protocol**: Analyze only those who completed treatment as assigned.
- Measures efficacy under ideal conditions
- Can introduce bias (who drops out?)
- Less generalizable

::: {.key-concept}
Modern trials **always report ITT** as the primary analysis, because it maintains the validity of randomization. Per-protocol analysis is secondary.
:::

### Strengths and Limitations

| Strength | Limitation |
|----------|-----------|
| Gold standard for causality | Expensive and time-consuming |
| Randomization breaks confounding | May be unethical (can't randomize to harmful exposures) |
| Double-blinding reduces bias | Hawthorne effect (behavior changes when observed) |
| Clear temporal sequence | Limited to one outcome per trial |
| | Restricted to willing, healthy volunteers (low generalizability) |

### Example: Indian COVID Vaccine Trial

```{r}
#| echo: false
# Simulate RCT results
vaccine_trial <- tibble(
  group = c("Vaccine (n=500)", "Placebo (n=500)"),
  covid_cases = c(12, 48),
  total_n = c(500, 500),
  attack_rate = covid_cases / total_n,
  percent = attack_rate * 100
)

vaccine_trial <- vaccine_trial %>%
  mutate(rr = attack_rate / attack_rate[2])

ggplot(vaccine_trial, aes(x = group, y = percent, fill = group)) +
  geom_col(color = "black", linewidth = 1, width = 0.5) +
  geom_text(aes(label = paste0(covid_cases, " cases")), vjust = -0.5,
            size = 3.5, fontface = "bold") +
  geom_text(aes(label = paste0(round(percent, 1), "%"), y = percent/2),
            vjust = 0.5, size = 3.5, fontface = "bold", color = "white") +
  scale_fill_manual(values = c("darkgreen", "gray60"), guide = "none") +
  scale_y_continuous(limits = c(0, 12), labels = label_percent(scale = 1)) +
  labs(
    title = "RCT: COVID Vaccine Efficacy in India",
    subtitle = "Attack rate: Vaccine 2.4% vs. Placebo 9.6% | Efficacy = (1 - 2.4/9.6) × 100 = 75%",
    x = "Group",
    y = "Attack Rate (%)"
  ) +
  theme_clean()
```

## Part 6: Bias — Systematic Errors That Distort Truth

**Bias** is any systematic deviation from the true value. Unlike random error (which averages out), bias consistently pushes estimates in one direction.

### Selection Bias: Who Gets In?

**Selection bias** occurs when the way people are selected into the study is related to both exposure and outcome.

#### Healthy Volunteer Bias
People who volunteer for studies tend to be healthier. If comparing vitamin users (who volunteer) to non-users (population sample), vitamin users appear healthier—not because of the vitamin, but because healthier people volunteer.

#### Berkson's Bias (Hospital Selection Bias)
Already discussed in case-control section: Hospital-based controls don't represent the source population.

#### Loss to Follow-Up Bias
In cohort studies, if sicker people drop out preferentially, your final sample is artificially healthy.

```{r}
#| echo: false
# Simulate loss to follow-up
loss_data <- tibble(
  time = c("Baseline", "Year 1", "Year 2", "Year 3"),
  exposed_n = c(500, 450, 380, 290),
  unexposed_n = c(500, 480, 425, 360),
  exposed_pct = c(50, 50, 50, 50),
  unexposed_pct = c(50, 50, 50, 50)
)

# Create loss to follow-up visualization
loss_long <- loss_data %>%
  pivot_longer(cols = c(exposed_n, unexposed_n), names_to = "group", values_to = "n") %>%
  mutate(group = if_else(group == "exposed_n", "Exposed", "Unexposed"))

ggplot(loss_long, aes(x = time, y = n, color = group, group = group)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 4) +
  geom_label(aes(label = n), nudge_y = 15, size = 3) +
  scale_color_manual(values = c("darkred", "darkblue")) +
  scale_y_continuous(limits = c(250, 550)) +
  labs(
    title = "Loss to Follow-Up Over 3 Years",
    subtitle = "If sicker people drop out disproportionately, estimates become biased",
    x = "Follow-up Period",
    y = "Number Remaining in Study",
    color = "Group"
  ) +
  theme_clean() +
  theme(legend.position = "bottom")
```

### Information/Measurement Bias: How Accurately Measured?

**Recall bias**: Cases remember past exposures better than controls.
- Example: A mother of a child with birth defects might recall a fever in early pregnancy better than a mother of a healthy child.

**Interviewer bias**: Interviewer's knowledge of the outcome influences how questions are asked.
- Example: Knowing a patient has cancer, an interviewer might ask about pesticide exposure more probing questions.

**Misclassification**: Exposure or outcome is incorrectly classified.
- **Non-differential misclassification** (equal error in both groups): Biases toward the null (makes associations appear weaker)
- **Differential misclassification** (different error rates): Can bias in either direction

```{r}
#| echo: false
# Simulate effect of misclassification
misclass_data <- tibble(
  scenario = c("True association\n(no misclassification)", "With 10%\nmisclassification", "With 20%\nmisclassification"),
  true_or = c(3.0, 3.0, 3.0),
  observed_or = c(3.0, 2.5, 1.95),
  y = c(1, 2, 3)
)

ggplot(misclass_data, aes(x = observed_or, y = reorder(scenario, y))) +
  geom_segment(aes(xend = true_or, yend = y), color = "gray50", linewidth = 1) +
  geom_point(aes(color = "Observed OR"), size = 4) +
  geom_point(aes(x = true_or, color = "True OR"), size = 4) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("True OR" = "darkgreen", "Observed OR" = "orange")) +
  labs(
    title = "Non-Differential Misclassification Biases Toward the Null",
    x = "Odds Ratio",
    y = "",
    color = ""
  ) +
  theme_clean() +
  theme(legend.position = "bottom")
```

### Lead-Time Bias and Length Bias

**Lead-time bias**: If a disease is detected earlier (but prognosis doesn't change), survival appears longer.
- Example: Screening for cancer 6 months earlier doesn't extend life, just extends the "survival time" from diagnosis.

**Length bias**: Screening detects slower-growing, less fatal cancers preferentially.
- Example: Breast cancer screening finds more slow-growing cancers that have good prognosis anyway.

```{r}
#| echo: false
# Illustrate lead-time bias
lead_pts <- tibble(
  x = c(50, 65, 44, 65),
  y = c(2, 2, 1, 1),
  color = c("red", "black", "blue", "black")
)

lead_labels <- tibble(
  x = c(50, 65, 57.5, 44, 65, 54.5, 30),
  y = c(2.3, 2.3, 2.6, 1.3, 1.3, 1.6, 2.5),
  label = c("Diagnosis", "Death", "Survival = 15 years",
            "Diagnosis\n(Screening)", "Death", "Survival = 21 years",
            "Same death age\nBut screening makes it\nlook like survival improved!"),
  color = c("black", "black", "black", "black", "black", "blue", "red"),
  face = c("plain", "plain", "bold", "plain", "plain", "bold", "italic"),
  sz = c(2.5, 2.5, 3, 2.5, 2.5, 3, 3)
)

ggplot() +
  # Without screening
  geom_segment(aes(x = 50, xend = 65, y = 2, yend = 2), color = "black", linewidth = 3) +
  # With screening
  geom_segment(aes(x = 44, xend = 65, y = 1, yend = 1), color = "blue", linewidth = 3) +
  # Points
  geom_point(data = lead_pts, aes(x = x, y = y, color = color), size = 6) +
  scale_color_identity() +
  # Labels
  geom_text(data = lead_labels, aes(x = x, y = y, label = label, color = color),
            size = lead_labels$sz, fontface = lead_labels$face) +
  xlim(25, 75) +
  ylim(0.5, 3) +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 13)) +
  labs(title = "Lead-Time Bias: Earlier Detection ≠ Better Outcome")
```

## Part 7: Confounding — The Hidden Third Variable

A **confounder** is a variable that:
1. Is associated with the **exposure**
2. **Causes** the outcome
3. Is **not on the causal pathway** between exposure and outcome

```{r}
#| echo: false
grViz("
digraph confounder_dag {
  graph[rankdir = TB, bgcolor = 'transparent', nodesep = 1.0, ranksep = 0.8]
  node [fontname = 'Helvetica', shape = box, style = 'filled,rounded', fontsize = 12, margin = '0.25,0.15']

  C [label = 'CONFOUNDER\\n(e.g. Age, SES)', fillcolor = '#9C6ADE', fontcolor = 'white', width = 2.5]
  E [label = 'EXPOSURE', fillcolor = '#E89595', fontcolor = 'white', width = 2]
  O [label = 'OUTCOME', fillcolor = '#66BB6A', fontcolor = 'white', width = 2]

  edge [fontname = 'Helvetica', fontsize = 10, fontcolor = '#555555', penwidth = 2.0, color = '#555555']
  C -> E [label = ' associated with ']
  C -> O [label = ' causes ']

  edge [style = dashed, color = '#CC0000', fontcolor = '#CC0000', penwidth = 2.0]
  E -> O [label = ' observed association\\n(spurious or distorted) ']

  labelloc = t
  label = 'Confounding: A Third Variable Distorts the Exposure–Outcome Relationship\\n\\nThe confounder creates a backdoor path: Exposure ← Confounder → Outcome\\nThis makes it look like Exposure causes Outcome, even if it does not.'
  fontname = 'Helvetica'
  fontsize = 12
  fontcolor = '#333333'
}
")
```

### Classic Example: The Wine Paradox (Revisited)

Why does wine appear protective for heart disease in observational studies?

```{r}
#| echo: false
grViz("
digraph wine_confounding {
  graph[rankdir = TB, bgcolor = 'transparent', nodesep = 1.2, ranksep = 0.8]
  node [fontname = 'Helvetica', shape = box, style = 'filled,rounded', fontsize = 12, margin = '0.25,0.15']

  SES [label = 'Socioeconomic Status\\n& Healthy Lifestyle', fillcolor = '#9C6ADE', fontcolor = 'white', width = 3]
  WINE [label = 'Wine Drinking', fillcolor = '#C0392B', fontcolor = 'white', width = 2]
  MI [label = 'Lower MI Risk', fillcolor = '#27AE60', fontcolor = 'white', width = 2]

  edge [fontname = 'Helvetica', fontsize = 10, fontcolor = '#555555', penwidth = 2.0, color = '#555555']
  SES -> WINE [label = ' Higher SES → more\\n wine consumption ']
  SES -> MI [label = ' Higher SES → better diet,\\n exercise, healthcare ']

  edge [style = dashed, color = '#CC0000', fontcolor = '#CC0000', penwidth = 2.0]
  WINE -> MI [label = ' Observed: wine → lower MI\\n (spurious — actually SES!) ']

  labelloc = t
  label = 'Wine Paradox: SES Confounds the Wine–Heart Disease Association\\n\\nWine drinkers have lower MI not because of wine,\\nbut because they tend to have higher SES and healthier lifestyles.'
  fontname = 'Helvetica'
  fontsize = 12
  fontcolor = '#333333'
}
")
```

### Simpson's Paradox: Confounding in Action

A striking example of confounding is **Simpson's Paradox**: an association reverses when you account for a third variable.

```{r}
#| echo: false
# Simpson's paradox example: medication and recovery
# Hypothetical: medication appears harmful overall, but helpful in each gender
simpson_data <- tibble(
  group = c("Overall: Treatment", "Overall: Control",
            "Men: Treatment", "Men: Control",
            "Women: Treatment", "Women: Control"),
  recovered = c(250, 320, 200, 55, 50, 265),
  not_recovered = c(750, 80, 200, 45, 550, 35),
  total = recovered + not_recovered,
  recovery_rate = recovered / total * 100,
  stratum = c("Overall", "Overall", "Men", "Men", "Women", "Women"),
  treatment = c("Treatment", "Control", "Treatment", "Control", "Treatment", "Control")
)

p1 <- ggplot(filter(simpson_data, stratum == "Overall"),
             aes(x = treatment, y = recovery_rate, fill = treatment)) +
  geom_col(color = "black", linewidth = 1, width = 0.5) +
  geom_text(aes(label = paste0(round(recovery_rate, 1), "%")), vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("darkblue", "orange"), guide = "none") +
  scale_y_continuous(limits = c(0, 95), labels = label_percent(scale = 1)) +
  labs(
    title = "Overall: Control Looks Better",
    x = "",
    y = "Recovery Rate (%)"
  ) +
  theme_clean()

p2 <- ggplot(filter(simpson_data, stratum != "Overall"),
             aes(x = treatment, y = recovery_rate, fill = treatment)) +
  facet_wrap(~stratum, nrow = 1) +
  geom_col(color = "black", linewidth = 1, width = 0.5) +
  geom_text(aes(label = paste0(round(recovery_rate, 1), "%")), vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("darkblue", "orange"), guide = "none") +
  scale_y_continuous(limits = c(0, 95), labels = label_percent(scale = 1)) +
  labs(
    title = "Stratified by Gender: Treatment Is Better in BOTH Groups!",
    x = "",
    y = "Recovery Rate (%)"
  ) +
  theme_clean()

p1 / p2 +
  plot_annotation(
    title = "Simpson's Paradox: The Confounder (Gender) Reverses the Association",
    subtitle = "Control appears better overall, but treatment is better in men AND women. The confounding variable was gender distribution.",
    theme = theme(plot.title = element_text(face = "bold", size = 13))
  )
```

### Methods to Control Confounding

#### At Design Stage:
1. **Randomization**: RCTs randomize to balance all confounders
2. **Restriction**: Include only a specific stratum (e.g., only women) to eliminate confounding by gender
3. **Matching**: In case-control studies, match cases and controls on confounder

#### At Analysis Stage:
1. **Stratification**: Calculate association within each stratum of the confounder
2. **Regression adjustment**: Include confounder as a covariate in regression
3. **Propensity score**: Create a "score" of probability of exposure given confounders, then match/adjust

::: {.clinical-hook}
**Example: Chulha (Traditional Cooking Stove) and Respiratory Illness in Rural India**

A cross-sectional study in Rajasthan finds that women using traditional chulhas have 2.5 times higher risk of chronic respiratory illness compared to women using clean cookstoves.

**But is this causal?** Potential confounder: **Socioeconomic status (SES)**.
- Women in poorer households use chulhas (associated with exposure)
- Poorer women have worse overall health, crowded housing, and malnutrition (causes outcome)
- SES is not on the causal pathway from chulha to respiratory illness

**Confounding structure:**
- SES → Chulha use (poor families use chulhas)
- SES → Respiratory illness (poor health conditions)

**Solution**: Stratify by SES. Calculate the association within each SES stratum:
- Among poor women: chulha users have 2.0× risk (confounding adjusted)
- Among middle-class women: chulha users have 1.8× risk
- Adjusted estimate: ~1.9× (vs. crude 2.5×)

This adjusted estimate is closer to the true causal effect.
:::

## Part 8: Effect Modification (Interaction)

**Effect modification** is NOT confounding. It occurs when the effect of an exposure differs depending on the level of a third variable.

::: {.key-concept}
**Confounder vs. Effect Modifier:**

A **confounder** distorts the association and should be adjusted away.

An **effect modifier** indicates real biological heterogeneity. You don't adjust it away—you report the stratified results.

**Example**: A blood pressure medication works differently in men vs. women (effect modifier by gender). Report results for each gender separately. Don't adjust away the difference.
:::

```{r}
#| echo: false
# Simulate drug efficacy varying by baseline BP
em_data <- tibble(
  baseline_bp = c("Normal\n(110-120)", "Elevated\n(130-140)", "High\n(>140)"),
  treatment_effect_reduction = c(5, 15, 25),
  y = c(1, 2, 3)
)

ggplot(em_data, aes(x = reorder(baseline_bp, y), y = treatment_effect_reduction, fill = baseline_bp)) +
  geom_col(color = "black", linewidth = 1, width = 0.6) +
  geom_text(aes(label = paste0(treatment_effect_reduction, " mmHg")), vjust = -0.5, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("darkgreen", "orange", "darkred"), guide = "none") +
  scale_y_continuous(limits = c(0, 30)) +
  labs(
    title = "Effect Modification: Drug Efficacy Varies by Baseline BP",
    x = "Baseline Blood Pressure",
    y = "Average BP Reduction with Treatment (mmHg)",
    subtitle = "Real biological heterogeneity: the drug works better in those with higher baseline BP"
  ) +
  theme_clean()
```

## Part 9: Hierarchy of Evidence and Study Quality

The **hierarchy of evidence** ranks study designs by their ability to establish causality:

```{r}
#| echo: false
grViz("
digraph evidence_pyramid {
  graph[rankdir = TB, bgcolor = 'transparent', nodesep = 0.3, ranksep = 0.4]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 11, margin = '0.3,0.15']

  SR [label = 'Systematic Reviews &\\nMeta-Analyses', fillcolor = '#2E7D32', fontcolor = 'white', width = 2.5]
  RCT [label = 'Randomized Controlled Trials', fillcolor = '#43A047', fontcolor = 'white', width = 3.5]
  COH [label = 'Cohort Studies (Prospective)', fillcolor = '#66BB6A', fontcolor = 'white', width = 4.5]
  CC [label = 'Case-Control Studies', fillcolor = '#FFA726', fontcolor = 'white', width = 5.5]
  CS [label = 'Cross-Sectional Studies', fillcolor = '#FF7043', fontcolor = 'white', width = 6.5]
  CR [label = 'Case Reports & Case Series', fillcolor = '#EF5350', fontcolor = 'white', width = 7.5]

  edge [arrowhead = none, color = '#888888']
  SR -> RCT -> COH -> CC -> CS -> CR

  labelloc = t
  label = 'Evidence Pyramid: Hierarchy of Study Designs\\n(Top = Strongest evidence for causality)'
  fontname = 'Helvetica'
  fontsize = 14
  fontcolor = '#333333'
}
")
```

::: {.common-mistake}
**Important caveat**: A well-designed cohort study can provide stronger evidence than a poorly designed RCT with high dropout, outcome misclassification, or selection bias.

The hierarchy is a **guideline**, not a rule. Judge each study on methodological quality using the **GRADE framework**:
- **Risk of bias**: How likely is systematic error?
- **Consistency**: Do similar studies agree?
- **Directness**: Does the evidence apply to your patient?
- **Precision**: Are confidence intervals narrow or wide?
:::

## Part 10: Choosing the Right Study Design

### Decision Framework

Choosing a study design depends on your research question:

```{r}
#| echo: false
grViz("
digraph decision {
  graph[rankdir = TB, bgcolor = 'transparent', nodesep = 0.5, ranksep = 0.6]
  node [fontname = 'Helvetica', style = filled, fontsize = 10, margin = '0.2,0.1']

  Q [label = 'What is your\\nresearch question?', shape = diamond, fillcolor = '#B8D4E3', width = 2.5]

  Q1 [label = 'What is the\\nburden of disease?', shape = box, fillcolor = '#E8E8E8']
  Q2 [label = 'What factors are\\nassociated?', shape = box, fillcolor = '#E8E8E8']
  Q3 [label = 'Does X cause Y?\\n(Outcome is rare)', shape = box, fillcolor = '#E8E8E8']
  Q4 [label = 'Does X cause Y?\\n(Can randomize)', shape = box, fillcolor = '#E8E8E8']
  Q5 [label = 'Does X cause Y?\\n(Cannot randomize)', shape = box, fillcolor = '#E8E8E8']
  Q6 [label = 'Synthesize all\\nevidence?', shape = box, fillcolor = '#E8E8E8']

  A1 [label = 'Cross-Sectional\\nStudy', shape = box, fillcolor = '#FF7043', fontcolor = 'white']
  A2 [label = 'Cross-Sectional\\nor Cohort', shape = box, fillcolor = '#FFA726', fontcolor = 'white']
  A3 [label = 'Case-Control\\nStudy', shape = box, fillcolor = '#FFA726', fontcolor = 'white']
  A4 [label = 'Randomized\\nControlled Trial', shape = box, fillcolor = '#43A047', fontcolor = 'white']
  A5 [label = 'Cohort Study', shape = box, fillcolor = '#66BB6A', fontcolor = 'white']
  A6 [label = 'Systematic Review\\n& Meta-Analysis', shape = box, fillcolor = '#2E7D32', fontcolor = 'white']

  Q -> Q1; Q -> Q2; Q -> Q3; Q -> Q4; Q -> Q5; Q -> Q6

  edge [style = bold, color = '#555555']
  Q1 -> A1; Q2 -> A2; Q3 -> A3; Q4 -> A4; Q5 -> A5; Q6 -> A6

  labelloc = t
  label = 'Study Design Decision Framework'
  fontname = 'Helvetica'
  fontsize = 14
  fontcolor = '#333333'
}
")
```

### Comprehensive Comparison Table

```{r}
#| echo: false
design_table <- tibble(
  Aspect = c(
    "Temporal sequence",
    "Measures association",
    "Time required",
    "Cost",
    "Loss to follow-up",
    "Confounding risk",
    "Causality evidence",
    "Best for"
  ),
  "Cross-Sectional" = c(
    "No",
    "Prevalence/POR",
    "Months",
    "Low",
    "Not applicable",
    "High",
    "Weak",
    "Burden estimation"
  ),
  "Case-Control" = c(
    "Yes (backward)",
    "Odds Ratio",
    "Months",
    "Low-Medium",
    "Low (retrospective)",
    "High",
    "Moderate",
    "Rare outcomes"
  ),
  "Cohort" = c(
    "Yes (forward)",
    "Relative Risk",
    "Years",
    "High",
    "Moderate",
    "Moderate-High",
    "Moderate-Strong",
    "Incidence, multiple outcomes"
  ),
  "RCT" = c(
    "Yes (experimental)",
    "Relative Risk",
    "Years",
    "Very High",
    "Dropout bias",
    "Low",
    "Strong",
    "Causality, treatment efficacy"
  )
)

design_table %>%
  kbl(caption = "Comprehensive Comparison of Study Designs", booktabs = TRUE, linesep = "") %>%
  kable_styling(
    latex_options = c("scale_down", "striped"),
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = TRUE,
    font_size = 9
  ) %>%
  column_spec(1, bold = TRUE, width = "18%") %>%
  column_spec(2:5, width = "20.5%")
```

### Practical Scenarios: Which Design Would You Choose?

**Scenario 1**: A new antibiotic for tuberculosis is available. Does it work better than the standard regimen?
→ **RCT** (you can randomize treatment; need strong evidence for new therapy)

**Scenario 2**: There has been a cluster of birth defects in a rural village. What's the cause?
→ **Case-control** (rare outcome; need to identify exposure quickly) or **cross-sectional** (if ongoing cluster)

**Scenario 3**: What is the prevalence of diabetes in rural Odisha?
→ **Cross-sectional** (one-time survey; measure prevalence)

**Scenario 4**: Do indoor air pollutants cause lung cancer?
→ **Cohort** (long latency; outcome not rare enough for case-control to be efficient; can't randomize to pollutants)

### Exercise: Identify the Design

Study the blank schema below. Each study design uses the **same tree structure** — what changes are the **verbs on the edges** and the **labels on the nodes**. Can you identify which design this represents?

```{r}
#| echo: false
grViz("
digraph generic {
  graph[rankdir = LR, bgcolor = 'transparent', nodesep = 0.6, ranksep = 0.8]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 11, margin = '0.2,0.1']

  A [label = '?', fillcolor = '#B8D4E3']
  B [label = '?', fillcolor = '#E8C07A']
  C [label = '?', fillcolor = '#A8D5A2']
  D [label = '?', fillcolor = '#E89595']
  E [label = '?', fillcolor = '#D4EAF7']
  F [label = '?', fillcolor = '#E89595']
  G [label = '?', fillcolor = '#D4EAF7']
  H [label = '?', fillcolor = '#F0D4E8']
  I [label = '?', fillcolor = '#F0D4E8']

  A -> B; A -> C; B -> D; B -> E; C -> F; C -> G
  edge [arrowhead = none, arrowtail = none, style = tapered, label = '  estimate  ', fontsize = 9, fontcolor = '#555555']
  D -> H; E -> H; F -> I; G -> I
}
")
```

::: {.key-concept}
**The key to identifying any study design:**

| Design | First verb | Second verb | What is estimated? |
|--------|-----------|-------------|-------------------|
| Cross-Sectional | classify | assess | Prevalence |
| Case-Control | select | assess | Odds ratio |
| Cohort | ascertain | follow-up | Incidence / RR |
| RCT | randomization | follow-up | Rate of event |

**Note on case-control studies:** The diagram above shows the generic left-to-right structure. In teaching, case-control studies are often drawn **right-to-left** to emphasize their retrospective nature — you start with cases and look backward at exposure. Here is the same case-control design drawn in the conventional retrospective direction:
:::

```{r}
#| echo: false
grViz("
digraph case_control_rl {
  graph[rankdir = RL, bgcolor = 'transparent', nodesep = 0.6, ranksep = 0.8]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 11, margin = '0.2,0.1']

  A [label = 'Population', fillcolor = '#B8D4E3']
  B [label = 'Cases', fillcolor = '#E8C07A']
  C [label = 'Non-Cases', fillcolor = '#A8D5A2']
  D [label = 'Exposed', fillcolor = '#E89595']
  E [label = 'Non-Exposed', fillcolor = '#D4EAF7']
  F [label = 'Exposed', fillcolor = '#E89595']
  G [label = 'Non-Exposed', fillcolor = '#D4EAF7']
  H [label = 'Odds of Exposure\\nin Cases', fillcolor = '#F0D4E8']
  I [label = 'Odds of Exposure\\nin Non-Cases', fillcolor = '#F0D4E8']

  edge [label = '  select  ', fontsize = 9, fontcolor = '#555555']
  A -> B; A -> C
  edge [label = '  assess  ', fontsize = 9, fontcolor = '#555555']
  B -> D; B -> E; C -> F; C -> G
  edge [arrowhead = none, arrowtail = none, style = tapered, label = '  estimate  ', fontsize = 9, fontcolor = '#555555']
  D -> H; E -> H; F -> I; G -> I
}
")
```

## Summary: Choosing Study Design

The **right design** depends on:
- Your **research question** (what do you want to know?)
- The **disease rarity** (rare outcome → case-control; rare exposure → cohort)
- **Resources available** (time, money, personnel)
- **Ethical constraints** (can you randomize to the exposure?)
- **Practical considerations** (is follow-up feasible?)

## Practice MCQs: NEET PG Level

:::{.neet-practice}

**Q1.** A researcher measures tea consumption and anxiety levels in 2,000 college students at the same point in time. Which type of study design is this?

```{r}
#| echo: false
#| results: asis
make_mcq(
  id = "m7q1",
  question = "",
  options = c(
    "Case-control study" = "In a case-control study, you start with the outcome (anxious vs not anxious) and look backward at exposure. Here, both are measured simultaneously.",
    "answer:Cross-sectional study" = "Correct! Measuring both exposure (tea) and outcome (anxiety) at the same time point is the definition of a cross-sectional study. No temporal sequence is established — you cannot tell if tea caused anxiety or anxious people drink more tea.",
    "Prospective cohort study" = "A cohort study follows participants forward in time from exposure to outcome. Here there is no follow-up — everything is measured at one time point.",
    "Randomized controlled trial" = "An RCT would require randomly assigning students to drink or not drink tea. Here, the researcher simply observes existing habits."
  )
)
```

---

**Q2.** In a case-control study of oral cancer and paan chewing, 60 cases report paan exposure while 40 controls report exposure. Which measure of association is most appropriate?

```{r}
#| echo: false
#| results: asis
make_mcq(
  id = "m7q2",
  question = "",
  options = c(
    "Relative Risk (RR)" = "Relative risk requires knowing the incidence in exposed and unexposed groups. Case-control studies start with the outcome, so you cannot calculate incidence or RR directly.",
    "answer:Odds Ratio (OR)" = "Correct! Case-control studies measure the odds ratio because you start with the outcome and look backward at exposure. OR = (a/b) / (c/d). RR cannot be directly calculated from case-control designs because you select on the outcome, not the exposure.",
    "Prevalence Odds Ratio (POR)" = "POR is used in cross-sectional studies. Case-control studies use the standard odds ratio.",
    "Attributable Risk (AR)" = "Attributable risk (risk difference) requires incidence data, which case-control studies do not provide."
  )
)
```

---

**Q3.** A hospital-based case-control study found that MI patients were much more likely to have diabetes than hospital controls. However, diabetes is very common among hospital patients. What type of bias is this?

```{r}
#| echo: false
#| results: asis
make_mcq(
  id = "m7q3",
  question = "",
  options = c(
    "Recall bias" = "Recall bias occurs when cases remember exposures differently than controls. Here, the issue is that the controls don't represent the general population.",
    "answer:Berkson's bias (hospital selection bias)" = "Correct! Hospital-based controls don't represent the source population. Hospitalized patients have higher rates of many chronic diseases (including diabetes), so using them as controls inflates the apparent association between diabetes and MI. This is Berkson's bias.",
    "Lead-time bias" = "Lead-time bias relates to early detection appearing to improve survival. It's not relevant to control selection.",
    "Information bias" = "Information bias involves systematic errors in measurement. Here the problem is in who was selected as controls, not how data was measured."
  )
)
```

---

**Q4.** In a study of outdoor cooking (chulha use) and respiratory illness, the crude OR = 2.5. After stratifying by SES, the OR within each stratum is approximately 1.8–1.9. SES is likely:

```{r}
#| echo: false
#| results: asis
make_mcq(
  id = "m7q4",
  question = "",
  options = c(
    "An effect modifier (the association differs by SES)" = "Effect modification means the OR is substantially different across strata (e.g., 3.0 in low SES vs 1.2 in high SES). Here the ORs are similar across strata (~1.8–1.9), ruling out effect modification.",
    "answer:A confounder (SES is associated with both exposure and outcome)" = "Correct! When the association changes after stratification (crude 2.5 → adjusted ~1.8–1.9) but remains similar across strata, the stratifying variable is a confounder. SES is associated with chulha use (lower SES → more chulha use) AND with respiratory illness (lower SES → worse health conditions), inflating the crude OR.",
    "An intermediate variable (on the causal pathway)" = "If SES were on the pathway between chulha use and respiratory illness, adjusting for it would be inappropriate (overadjustment). SES precedes chulha use, not the other way around.",
    "Unrelated to the exposure-outcome association" = "If SES were unrelated, the crude and stratified ORs would be the same. The change from 2.5 to 1.8–1.9 shows SES is related."
  )
)
```

---

**Q5.** Why do RCTs control confounding better than observational cohort studies?

```{r}
#| echo: false
#| results: asis
make_mcq(
  id = "m7q5",
  question = "",
  options = c(
    "RCTs measure outcomes more accurately" = "Outcome measurement accuracy relates to blinding and measurement protocols, not to confounding control. A cohort study can measure outcomes just as accurately.",
    "answer:RCTs randomize assignment, breaking the association between confounders and exposure" = "Correct! Randomization distributes ALL confounders — both known and unknown — equally between treatment and control groups. This breaks the causal link between confounders and exposure assignment. No observational design can achieve this for unknown confounders.",
    "RCTs always have larger sample sizes" = "RCTs often have smaller sample sizes than observational studies. Sample size helps with precision but doesn't control confounding.",
    "RCTs use blinding to reduce confounders" = "Blinding reduces information bias (differential assessment of outcomes), not confounding. Randomization is what controls confounding."
  )
)
```

---

**Q6.** In Simpson's Paradox, an overall association reverses when stratified by a third variable. Why does this happen?

```{r}
#| echo: false
#| results: asis
make_mcq(
  id = "m7q6",
  question = "",
  options = c(
    "The third variable is an effect modifier" = "Effect modification means the effect varies across strata but doesn't reverse the overall direction. Simpson's Paradox is specifically about confounding reversing the apparent direction.",
    "answer:The third variable is a confounder that distorts the crude association" = "Correct! Simpson's Paradox is a dramatic example of confounding. The confounder is so strongly associated with both exposure and outcome that it reverses the apparent association when pooled. The stratified (adjusted) results reveal the true direction.",
    "The sample size is too small" = "Simpson's Paradox occurs regardless of sample size. It's a structural problem (confounding), not a precision problem.",
    "There is measurement error in the outcome" = "Measurement error typically biases toward the null — it doesn't reverse associations. Simpson's Paradox is caused by confounding, not measurement error."
  )
)
```

---

**Q7.** A drug reduces blood pressure by 15 mmHg in men but only 5 mmHg in women. Gender is a(n):

```{r}
#| echo: false
#| results: asis
make_mcq(
  id = "m7q7",
  question = "",
  options = c(
    "Confounder that should be adjusted in the analysis" = "If gender were a confounder, adjusting for it would give a single 'true' effect. But here the effect genuinely differs by gender — adjusting would hide real biological variation.",
    "answer:Effect modifier indicating real biological heterogeneity" = "Correct! When the effect of an exposure (drug) differs substantially by a third variable (gender), that variable is an effect modifier. This reflects real biology — different drug metabolism, hormonal effects, etc. You don't adjust away effect modification; you report stratified results (15 mmHg in men, 5 mmHg in women).",
    "Selection bias" = "Selection bias relates to who enters the study, not to how the drug works differently in subgroups.",
    "Recall bias" = "Recall bias involves differential memory of exposures. Blood pressure is measured objectively, not recalled."
  )
)
```

:::


---

## Further Learning

### Textbooks and References
- **Sackett DL, et al.** *Evidence-Based Medicine: How to Practice and Teach It*. Churchill Livingstone. (Gold standard on study design interpretation)
- **Greenland S, Rothman KJ.** Chapter 7 in *Epidemiology: An Introduction*, 3rd ed. Oxford University Press.
- **Polit DF, Beck CT.** *Nursing Research: Generating and Assessing Evidence for Nursing Practice*, 10th ed. (Excellent on study design and bias)

### Guidelines and Checklists
- **STROBE Statement** (STrengthening the Reporting Of Observational studies in Epidemiology): Guidelines for reporting observational studies → [www.strobe-statement.org](https://www.strobe-statement.org)
- **CONSORT Statement** (Consolidated Standards of Reporting Trials): Guidelines for reporting RCTs → [www.consort-statement.org](https://www.consort-statement.org)
- **GRADE** (Grading of Recommendations Assessment, Development and Evaluation): Framework for assessing evidence quality

### Indian Clinical Trial Resources
- **Clinical Trials Registry - India (CTRI)**: [ctri.nic.in](https://ctri.nic.in) — Browse Indian trials
- **ICMR Guidelines** on ethical standards for clinical research in India

### Video Resources
- Khan Academy: Evidence-based medicine and study design
- Cochrane Collaboration: How to critically appraise studies

---

*Module 7 completes the foundation of study design. Module 8 will address statistical inference, hypothesis testing, and p-values.*
