---
title: "Comparing Groups — t-tests, ANOVA, and Non-parametric Alternatives"
subtitle: "From Two Groups to Many — Choosing the Right Test and Interpreting Results"
description: "Master parametric tests (one-sample, paired, and independent t-tests; one-way ANOVA) for comparing means, and their non-parametric alternatives (Mann-Whitney U, Wilcoxon signed-rank, Kruskal-Wallis). Learn to check assumptions, choose the right test, perform post-hoc comparisons, and report results with effect sizes."
categories:
  - Biostatistics
  - t-tests
  - ANOVA
  - Non-parametric Tests
  - Group Comparisons
author: "AIIMS Bhopal Biostatistics Course"
date: "`r Sys.Date()`"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(kableExtra)
source("../R/mcq.R")
library(patchwork)
library(scales)

set.seed(42)

theme_clean <- function() {
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
}
```

::: {.callout-note appearance="minimal"}
**Lecture slides for this module:** [Open Slides](../slides/09-comparing-means-slides.html){target="_blank"}
:::

## Learning Objectives {#learning-objectives}

By the end of this module, you will be able to:

1. **Choose and perform** one-sample, paired, and independent-samples t-tests
2. **Interpret** t-test results including test statistic, degrees of freedom, p-value, and confidence interval
3. **Assess assumptions** for t-tests (normality, equal variance) and respond when they are violated
4. **Conduct one-way ANOVA** for comparing three or more groups
5. **Perform post-hoc comparisons** (Tukey, Bonferroni, Dunnett) and explain why they are needed
6. **Apply non-parametric alternatives** (Mann-Whitney U, Wilcoxon signed-rank, Kruskal-Wallis) when assumptions fail
7. **Navigate a decision flowchart** to select the correct test for any group comparison
8. **Report results** with effect sizes (Cohen's d), confidence intervals, and clinical interpretation

---

## Clinical Hook {.clinical-hook}

An ICMR-funded multicentre trial enrolls 90 women with iron-deficiency anaemia across three government hospitals in Jharkhand. They are randomized to three arms: **oral ferrous sulfate** (standard), **IV ferric carboxymaltose** (newer), and **placebo** (control), with n = 30 per group. After 12 weeks, mean haemoglobin (± SD) is:

- Oral iron: 10.8 ± 1.4 g/dL
- IV iron: 11.5 ± 1.6 g/dL
- Placebo: 9.6 ± 1.5 g/dL

The research questions cascade:

1. Is oral iron better than placebo? (**Two-group comparison** → t-test)
2. Are all three groups different? (**Three-group comparison** → ANOVA)
3. *Which* pairs differ, and by how much? (**Post-hoc comparisons**)
4. What if the data aren't normally distributed? (**Non-parametric alternatives**)

This module gives you the tools to answer each question correctly.

---

## Section 1: The One-Sample t-Test

### When to Use

You have **one group** and want to test whether its mean differs from a known or hypothesized value.

### Formula

$$t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}$$

where $\bar{x}$ = sample mean, $\mu_0$ = hypothesized population mean, $s$ = sample SD, $n$ = sample size. Degrees of freedom: df = n − 1.

### Worked Example: BMI of Medical Students in Lucknow

::::{.callout-note}
A study of 120 first-year MBBS students at KGMU, Lucknow measures BMI. The national average for 18–22 year-olds is 22.5 kg/m². Is this cohort different?

- $\bar{x}$ = 23.8 kg/m², $s$ = 4.2, $n$ = 120
::::

```{r}
#| label: fig-one-sample-t
#| fig-cap: "One-sample t-test: Is the mean BMI of medical students different from the national average of 22.5?"

# Simulate data matching the summary statistics
set.seed(42)
bmi <- rnorm(120, mean = 23.8, sd = 4.2)
bmi_mean <- mean(bmi)
bmi_sd <- sd(bmi)
mu0 <- 22.5

# Perform the test
t_result <- t.test(bmi, mu = mu0)

# Visualize
ggplot(tibble(BMI = bmi), aes(x = BMI)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "#3498db", alpha = 0.7, color = "white") +
  geom_vline(xintercept = mu0, linetype = "dashed", linewidth = 1.2, color = "#e74c3c") +
  geom_vline(xintercept = bmi_mean, linetype = "solid", linewidth = 1.2, color = "#2c3e50") +
  annotate("label", x = mu0 - 1, y = 0.09, label = paste0("μ₀ = ", mu0),
           fill = "#fadbd8", color = "#e74c3c", fontface = "bold", size = 4) +
  annotate("label", x = bmi_mean + 1.5, y = 0.09, label = paste0("x̄ = ", round(bmi_mean, 1)),
           fill = "#d6eaf8", color = "#2c3e50", fontface = "bold", size = 4) +
  labs(title = "BMI Distribution of 120 Medical Students (KGMU, Lucknow)",
       subtitle = paste0("t = ", round(t_result$statistic, 2),
                        ", df = ", t_result$parameter,
                        ", p = ", format.pval(t_result$p.value, digits = 3)),
       x = "BMI (kg/m²)", y = "Density") +
  theme_clean()
```

::::{.callout-tip}
**Interpretation:**

- t = `r round(t_result$statistic, 2)`, df = `r t_result$parameter`, p = `r format.pval(t_result$p.value, digits = 3)`
- 95% CI for true mean: (`r round(t_result$conf.int[1], 1)`, `r round(t_result$conf.int[2], 1)`) kg/m²
- Since the CI does not include 22.5, we reject H₀ — the students' mean BMI is significantly higher than the national average.
- **Clinical note:** A 1.3 kg/m² difference, while statistically significant, may not be clinically alarming. Always consider effect size and clinical context.
::::

---

## Section 2: The Paired t-Test

### When to Use

You have **two measurements on the same subjects** (before/after, left eye/right eye, two methods on same specimen). The paired design removes between-subject variability, making it more powerful.

### Formula

Compute the **difference** for each pair: $D_i = \text{After}_i - \text{Before}_i$, then apply a one-sample t-test to the differences:

$$t = \frac{\bar{D} - 0}{s_D / \sqrt{n}}$$

### Worked Example: BP Before and After Amlodipine

::::{.callout-note}
40 hypertensive patients at a PHC in Kolar (Karnataka) are started on amlodipine 5 mg daily. Systolic BP is measured before treatment and again at 8 weeks.
::::

```{r}
#| label: fig-paired-t
#| fig-cap: "Paired t-test: systolic BP before and after 8 weeks of amlodipine. Each line connects the same patient's two measurements."
#| fig-height: 8

set.seed(123)
n_paired <- 40
bp_before <- rnorm(n_paired, mean = 152, sd = 12)
bp_reduction <- rnorm(n_paired, mean = 14, sd = 8)  # Mean reduction ~14 mmHg
bp_after <- bp_before - bp_reduction

bp_data <- tibble(
  Patient = 1:n_paired,
  Before = bp_before,
  After = bp_after,
  Difference = bp_after - bp_before
)

# Paired t-test
paired_result <- t.test(bp_data$After, bp_data$Before, paired = TRUE)

# Spaghetti plot
p_spag <- bp_data %>%
  pivot_longer(cols = c(Before, After), names_to = "Time", values_to = "SBP") %>%
  mutate(Time = factor(Time, levels = c("Before", "After"))) %>%
  ggplot(aes(x = Time, y = SBP, group = Patient)) +
  geom_line(alpha = 0.3, color = "gray50") +
  geom_point(aes(color = Time), size = 2, alpha = 0.7) +
  stat_summary(aes(group = 1), fun = mean, geom = "line", linewidth = 2, color = "#e74c3c") +
  stat_summary(aes(group = 1), fun = mean, geom = "point", size = 4, color = "#e74c3c") +
  scale_color_manual(values = c("Before" = "#3498db", "After" = "#27ae60")) +
  labs(title = "SBP Before and After Amlodipine (n = 40)",
       subtitle = paste0("Mean difference = ", round(mean(bp_data$Difference), 1),
                        " mmHg | p ", ifelse(paired_result$p.value < 0.001, "< 0.001",
                                            paste0("= ", round(paired_result$p.value, 4)))),
       y = "Systolic BP (mmHg)", x = NULL) +
  theme_clean() + theme(legend.position = "none")

# Histogram of differences
p_diff <- ggplot(bp_data, aes(x = Difference)) +
  geom_histogram(aes(y = after_stat(density)), bins = 12, fill = "#8e44ad", alpha = 0.7, color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#e74c3c", linewidth = 1) +
  geom_vline(xintercept = mean(bp_data$Difference), linewidth = 1.2, color = "#2c3e50") +
  annotate("label", x = mean(bp_data$Difference) - 5, y = 0.05,
           label = paste0("Mean D = ", round(mean(bp_data$Difference), 1)),
           fill = "#e8daef", fontface = "bold", size = 3.5) +
  labs(title = "Distribution of Paired Differences (After − Before)",
       subtitle = "Negative values = BP decreased (desired outcome)",
       x = "Difference in SBP (mmHg)", y = "Density") +
  theme_clean()

p_spag / p_diff
```

::::{.callout-tip}
**Result:** Mean reduction = `r round(-mean(bp_data$Difference), 1)` mmHg (95% CI: `r round(-paired_result$conf.int[2], 1)` to `r round(-paired_result$conf.int[1], 1)`), t = `r round(paired_result$statistic, 2)`, p < 0.001.

**Why paired?** If we had used an independent t-test (ignoring pairing), we would include between-patient variability (some patients start at 140, others at 170). The paired design eliminates this, leaving only the treatment effect — making the test much more powerful.
::::

::::{.callout-warning}
**Common mistake:** Using an independent t-test when data are paired. This ignores the within-subject correlation, inflates the SE, and reduces power — you may miss a real treatment effect.
::::

---

## Section 3: The Independent Two-Sample t-Test

### When to Use

You have **two independent groups** and want to compare their means. The groups are different people (no pairing).

### Two Versions

```{r}
#| label: tbl-t-versions
#| tbl-cap: "Student's vs Welch's t-test"

tibble(
  Feature = c("Assumption", "Pooled variance?", "Degrees of freedom", "Default in R?", "Recommendation"),
  `Student's t` = c("Equal variances in both groups", "Yes (pooled SD)", "n₁ + n₂ − 2", "No", "Only if variances clearly equal"),
  `Welch's t` = c("No equal variance assumption", "No (separate SEs)", "Adjusted (Welch-Satterthwaite)", "YES (default)", "**Use this by default**")
) %>%
  kable(format = "html") %>%
  kable_styling(full_width = TRUE, font_size = 11)
```

### Worked Example: HbA1c in Two Diabetes Drug Arms

::::{.callout-note}
A trial at KEM Hospital, Pune compares metformin vs glimepiride in newly diagnosed Type 2 DM (n = 50 per group, 6 months). Outcome: HbA1c at 6 months.
::::

```{r}
#| label: fig-independent-t
#| fig-cap: "Independent t-test: HbA1c at 6 months, metformin vs glimepiride. Box plots with individual data points and group means."

set.seed(42)
metformin <- rnorm(50, mean = 7.1, sd = 0.9)
glimepiride <- rnorm(50, mean = 7.6, sd = 1.1)

drug_data <- tibble(
  HbA1c = c(metformin, glimepiride),
  Drug = rep(c("Metformin", "Glimepiride"), each = 50)
)

# Welch's t-test (default)
welch_result <- t.test(HbA1c ~ Drug, data = drug_data)

# Cohen's d
pooled_sd <- sqrt(((49 * sd(metformin)^2) + (49 * sd(glimepiride)^2)) / 98)
cohens_d <- (mean(glimepiride) - mean(metformin)) / pooled_sd

ggplot(drug_data, aes(x = Drug, y = HbA1c, fill = Drug)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA, width = 0.4) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "#e74c3c") +
  stat_summary(fun = mean, geom = "text", aes(label = paste0("x̄ = ", round(after_stat(y), 2))),
               vjust = -1.5, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Metformin" = "#3498db", "Glimepiride" = "#e74c3c")) +
  labs(title = "HbA1c at 6 Months: Metformin vs Glimepiride (KEM, Pune)",
       subtitle = paste0("Welch's t = ", round(welch_result$statistic, 2),
                        ", p = ", round(welch_result$p.value, 3),
                        " | Cohen's d = ", round(cohens_d, 2)),
       y = "HbA1c (%)", x = NULL) +
  theme_clean() + theme(legend.position = "none")
```

::::{.callout-tip}
**Result:**

- Mean difference: `r round(mean(glimepiride) - mean(metformin), 2)`% (95% CI: `r round(-welch_result$conf.int[2], 2)` to `r round(-welch_result$conf.int[1], 2)`)
- Welch's t = `r round(welch_result$statistic, 2)`, p = `r round(welch_result$p.value, 3)`
- Cohen's d = `r round(cohens_d, 2)` (medium effect)
- **Interpretation:** Metformin achieved statistically significantly lower HbA1c than glimepiride. The effect size (d ≈ 0.5) suggests a clinically meaningful difference. Compare the difference to the MCID for HbA1c (0.5%) before making treatment decisions.
::::

---

## Section 4: Checking Assumptions

Before applying any t-test, verify three assumptions:

### 4.1 Independence

This is a **study design** issue, not a statistical test. Each observation must come from a different, independent subject. Violated in clustered data (e.g., multiple teeth per patient, students within classrooms).

### 4.2 Normality

```{r}
#| label: fig-normality-check
#| fig-cap: "How to check normality: histogram (left) and Q-Q plot (right). Top row: approximately normal data. Bottom row: right-skewed data — consider non-parametric test if n is small."
#| fig-height: 8

set.seed(42)
normal_data <- rnorm(50, mean = 10, sd = 2)
skewed_data <- rgamma(50, shape = 2, rate = 0.5)

p1 <- ggplot(tibble(x = normal_data), aes(x)) +
  geom_histogram(bins = 12, fill = "#3498db", alpha = 0.7, color = "white") +
  labs(title = "Normal Data: Histogram", x = "Value", y = "Count") +
  theme_clean()

p2 <- ggplot(tibble(x = normal_data), aes(sample = x)) +
  stat_qq(color = "#3498db", size = 2) +
  stat_qq_line(color = "#e74c3c", linewidth = 1) +
  labs(title = "Normal Data: Q-Q Plot", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_clean()

p3 <- ggplot(tibble(x = skewed_data), aes(x)) +
  geom_histogram(bins = 12, fill = "#e74c3c", alpha = 0.7, color = "white") +
  labs(title = "Skewed Data: Histogram", x = "Value", y = "Count") +
  theme_clean()

p4 <- ggplot(tibble(x = skewed_data), aes(sample = x)) +
  stat_qq(color = "#e74c3c", size = 2) +
  stat_qq_line(color = "#3498db", linewidth = 1) +
  labs(title = "Skewed Data: Q-Q Plot", subtitle = "Points deviate from line → non-normal",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_clean()

(p1 + p2) / (p3 + p4)
```

::::{.callout-tip}
**Practical rules:**

- **n ≥ 30 per group:** t-test is robust to moderate non-normality (CLT from Module 8!)
- **n < 30 and data skewed:** Use a non-parametric alternative
- **Visual check** (histogram + Q-Q plot) is more useful than Shapiro-Wilk for practical decisions
- **Shapiro-Wilk test:** p > 0.05 → normality not rejected. But with large n, it detects trivial departures — rely on visuals.
::::

### 4.3 Equal Variances (for Independent t-test)

::::{.callout-important}
**Levene's test** tests H₀: σ₁² = σ₂². If p < 0.05, variances are unequal.

**Better approach:** Just use **Welch's t-test** by default — it performs well whether variances are equal or not, and avoids the preliminary test problem (testing assumptions with another test inflates overall error).
::::

---

## Section 5: One-Way ANOVA

### When to Use

You have **three or more independent groups** and want to test whether *any* group means differ.

### The Logic: Partitioning Variance

ANOVA asks: "Is the variability *between* groups larger than the variability *within* groups?"

$$F = \frac{MS_{\text{between}}}{MS_{\text{within}}} = \frac{\text{Variance due to treatment}}{\text{Variance due to random error}}$$

- **Large F:** Between-group variance >> within-group variance → groups are different
- **Small F (near 1):** Between-group ≈ within-group → no evidence of group differences

### Worked Example: Anaemia Trial (from Clinical Hook)

```{r}
#| label: fig-anova
#| fig-cap: "One-way ANOVA: Haemoglobin at 12 weeks across three treatment arms. Violin plots show distribution shape; diamonds mark group means."
#| fig-height: 6

set.seed(42)
oral_iron <- rnorm(30, mean = 10.8, sd = 1.4)
iv_iron <- rnorm(30, mean = 11.5, sd = 1.6)
placebo <- rnorm(30, mean = 9.6, sd = 1.5)

anaemia_data <- tibble(
  Hb = c(oral_iron, iv_iron, placebo),
  Treatment = factor(rep(c("Oral Iron", "IV Iron", "Placebo"), each = 30),
                     levels = c("Placebo", "Oral Iron", "IV Iron"))
)

# ANOVA
anova_result <- aov(Hb ~ Treatment, data = anaemia_data)
anova_summary <- summary(anova_result)
f_stat <- anova_summary[[1]]$`F value`[1]
p_val <- anova_summary[[1]]$`Pr(>F)`[1]

ggplot(anaemia_data, aes(x = Treatment, y = Hb, fill = Treatment)) +
  geom_violin(alpha = 0.4, width = 0.7) +
  geom_boxplot(width = 0.2, alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.4, size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, color = "#2c3e50") +
  scale_fill_manual(values = c("Placebo" = "#95a5a6", "Oral Iron" = "#3498db", "IV Iron" = "#e74c3c")) +
  labs(title = "Haemoglobin at 12 Weeks: Anaemia Treatment Trial (Jharkhand)",
       subtitle = paste0("One-way ANOVA: F(2, 87) = ", round(f_stat, 2),
                        ", p ", ifelse(p_val < 0.001, "< 0.001", paste0("= ", round(p_val, 4)))),
       y = "Haemoglobin (g/dL)", x = NULL) +
  theme_clean() + theme(legend.position = "none")
```

### The ANOVA Table

```{r}
#| label: tbl-anova
#| tbl-cap: "ANOVA table for the anaemia trial"

ss_between <- anova_summary[[1]]$`Sum Sq`[1]
ss_within <- anova_summary[[1]]$`Sum Sq`[2]
df_between <- anova_summary[[1]]$Df[1]
df_within <- anova_summary[[1]]$Df[2]
ms_between <- anova_summary[[1]]$`Mean Sq`[1]
ms_within <- anova_summary[[1]]$`Mean Sq`[2]

tibble(
  Source = c("Between Groups (Treatment)", "Within Groups (Error)", "Total"),
  df = c(df_between, df_within, df_between + df_within),
  `Sum of Squares` = c(round(ss_between, 2), round(ss_within, 2), round(ss_between + ss_within, 2)),
  `Mean Square` = c(round(ms_between, 2), round(ms_within, 2), NA),
  `F` = c(round(f_stat, 2), NA, NA),
  `p-value` = c(ifelse(p_val < 0.001, "< 0.001", round(p_val, 4)), NA, NA)
) %>%
  kable(format = "html") %>%
  kable_styling(full_width = TRUE, font_size = 11)
```

::::{.callout-important}
**What ANOVA tells you:** At least one group mean differs from the others.

**What ANOVA does NOT tell you:** *Which* groups differ. For that, you need **post-hoc comparisons** (next section).
::::

---

## Section 6: Post-Hoc Comparisons

### The Multiple Comparisons Problem

With k = 3 groups, there are $\binom{3}{2}$ = 3 pairwise comparisons. Doing three separate t-tests at α = 0.05 inflates the overall Type I error to 1 − (0.95)³ ≈ 14%. With k = 5 groups (10 comparisons), it reaches ~40%.

### Common Post-Hoc Methods

```{r}
#| label: tbl-posthoc-methods
#| tbl-cap: "Choosing a post-hoc method"

tibble(
  Method = c("Tukey HSD", "Bonferroni", "Dunnett", "Scheffé"),
  `Best for` = c("All pairwise comparisons", "Few pre-planned comparisons", "Multiple treatments vs one control", "Complex contrasts (e.g., Group 1 vs Groups 2+3)"),
  `Conservatism` = c("Moderate", "Conservative", "Moderate", "Most conservative"),
  `Power` = c("Good", "Lower", "Good for its purpose", "Lowest")
) %>%
  kable(format = "html") %>%
  kable_styling(full_width = TRUE, font_size = 11)
```

### Tukey HSD for the Anaemia Trial

```{r}
#| label: fig-tukey
#| fig-cap: "Tukey HSD pairwise comparisons with 95% family-wise confidence intervals. Intervals that don't cross zero indicate significant differences."

tukey_result <- TukeyHSD(anova_result)
tukey_df <- as.data.frame(tukey_result$Treatment)
tukey_df$Comparison <- rownames(tukey_df)

ggplot(tukey_df, aes(x = diff, y = reorder(Comparison, diff))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#e74c3c", linewidth = 1) +
  geom_point(size = 3, color = "#2c3e50") +
  geom_errorbarh(aes(xmin = lwr, xmax = upr), height = 0.2, linewidth = 1, color = "#3498db") +
  geom_text(aes(label = paste0("p = ", round(`p adj`, 4))), hjust = -0.3, vjust = -1, size = 3.5, fontface = "bold") +
  labs(title = "Tukey HSD: Pairwise Differences in Mean Hb (g/dL)",
       subtitle = "95% family-wise confidence intervals | Dashed line = no difference",
       x = "Difference in Mean Hb (g/dL)", y = NULL) +
  theme_clean()
```

::::{.callout-tip}
**Interpretation:**

```{r}
#| results: asis
cat("- **IV Iron vs Placebo:** Δ = ", round(tukey_df$diff[tukey_df$Comparison == "IV Iron-Placebo"], 2),
    " g/dL (p = ", round(tukey_df$`p adj`[tukey_df$Comparison == "IV Iron-Placebo"], 4), ") → Significant\n")
cat("- **Oral Iron vs Placebo:** Δ = ", round(tukey_df$diff[tukey_df$Comparison == "Oral Iron-Placebo"], 2),
    " g/dL (p = ", round(tukey_df$`p adj`[tukey_df$Comparison == "Oral Iron-Placebo"], 4), ") → Significant\n")
cat("- **IV Iron vs Oral Iron:** Δ = ", round(tukey_df$diff[tukey_df$Comparison == "IV Iron-Oral Iron"], 2),
    " g/dL (p = ", round(tukey_df$`p adj`[tukey_df$Comparison == "IV Iron-Oral Iron"], 4), ")\n")
```

Both iron formulations outperform placebo. Whether IV iron is significantly better than oral iron depends on the p-value and the clinical MCID for haemoglobin improvement.
::::

---

## Section 7: Non-Parametric Alternatives

When t-test/ANOVA assumptions are violated — data are heavily skewed, sample is small, or the outcome is ordinal — use non-parametric tests. These work with **ranks** rather than raw values, making them robust to outliers and non-normality.

### 7.1 Mann-Whitney U Test (Two Independent Groups)

The non-parametric alternative to the independent t-test. Compares whether one distribution tends to have larger values than the other.

::::{.callout-note}
**Example:** Hospital length of stay (LOS) for laparoscopic vs open cholecystectomy at Safdarjung Hospital, Delhi. LOS is typically right-skewed (most patients leave quickly, a few stay very long).
::::

```{r}
#| label: fig-mann-whitney
#| fig-cap: "Mann-Whitney U test: hospital length of stay is right-skewed, making a non-parametric test appropriate. Medians (red diamonds) are compared rather than means."

set.seed(42)
los_lap <- rgamma(35, shape = 2, rate = 0.8) + 1   # Laparoscopic: shorter stays
los_open <- rgamma(35, shape = 3, rate = 0.6) + 2   # Open: longer stays

los_data <- tibble(
  LOS = c(los_lap, los_open),
  Surgery = rep(c("Laparoscopic", "Open"), each = 35)
)

mw_result <- wilcox.test(LOS ~ Surgery, data = los_data)

ggplot(los_data, aes(x = Surgery, y = LOS, fill = Surgery)) +
  geom_boxplot(alpha = 0.5, width = 0.4, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 1.5) +
  stat_summary(fun = median, geom = "point", shape = 18, size = 5, color = "#e74c3c") +
  scale_fill_manual(values = c("Laparoscopic" = "#27ae60", "Open" = "#e74c3c")) +
  labs(title = "Hospital LOS: Laparoscopic vs Open Cholecystectomy (Safdarjung, Delhi)",
       subtitle = paste0("Mann-Whitney U, p = ", round(mw_result$p.value, 4),
                        " | Median: Lap = ", round(median(los_lap), 1),
                        " days, Open = ", round(median(los_open), 1), " days"),
       y = "Length of Stay (days)", x = NULL) +
  theme_clean() + theme(legend.position = "none")
```

### 7.2 Wilcoxon Signed-Rank Test (Paired Data)

The non-parametric alternative to the paired t-test. Useful for before-after comparisons with ordinal or skewed data.

::::{.callout-note}
**Example:** Pain scores (VAS 0–10) before and after physiotherapy in 25 knee OA patients at CMC Vellore. Pain scores are ordinal — a non-parametric paired test is appropriate.
::::

```{r}
#| label: fig-wilcoxon
#| fig-cap: "Wilcoxon signed-rank test: pain scores before and after physiotherapy. Most patients improve (dots below the diagonal)."

set.seed(42)
pain_before <- sample(3:9, 25, replace = TRUE, prob = c(0.05, 0.1, 0.15, 0.25, 0.2, 0.15, 0.1))
pain_after <- pmax(0, pain_before - sample(0:4, 25, replace = TRUE, prob = c(0.1, 0.2, 0.35, 0.25, 0.1)))

wilcox_result <- wilcox.test(pain_before, pain_after, paired = TRUE)

ggplot(tibble(Before = pain_before, After = pain_after), aes(x = Before, y = After)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#e74c3c", linewidth = 1) +
  geom_jitter(width = 0.15, height = 0.15, size = 3, alpha = 0.7, color = "#3498db") +
  annotate("text", x = 8, y = 3, label = "Improved\n(below line)", fontface = "bold", color = "#27ae60", size = 5) +
  annotate("text", x = 3, y = 8, label = "Worsened\n(above line)", fontface = "bold", color = "#e74c3c", size = 5) +
  labs(title = "Pain VAS Before vs After Physiotherapy (CMC Vellore, n = 25)",
       subtitle = paste0("Wilcoxon signed-rank, p = ", round(wilcox_result$p.value, 4),
                        " | Median before = ", median(pain_before),
                        ", after = ", median(pain_after)),
       x = "Pain VAS Before", y = "Pain VAS After") +
  coord_equal() +
  theme_clean()
```

### 7.3 Kruskal-Wallis Test (Three or More Groups)

The non-parametric alternative to one-way ANOVA. Compares distributions across three or more groups using ranks.

::::{.callout-note}
**Example:** Patient satisfaction scores (Likert 1–5) across three CHCs in Chhattisgarh. Likert data are ordinal — ANOVA assumptions don't hold.
::::

```{r}
#| label: fig-kruskal
#| fig-cap: "Kruskal-Wallis test: patient satisfaction (Likert 1–5) across three community health centres. Bar heights show count per satisfaction level."

set.seed(42)
sat_a <- sample(1:5, 40, replace = TRUE, prob = c(0.05, 0.10, 0.25, 0.35, 0.25))
sat_b <- sample(1:5, 40, replace = TRUE, prob = c(0.10, 0.20, 0.30, 0.25, 0.15))
sat_c <- sample(1:5, 40, replace = TRUE, prob = c(0.15, 0.25, 0.30, 0.20, 0.10))

sat_data <- tibble(
  Satisfaction = c(sat_a, sat_b, sat_c),
  CHC = rep(c("CHC Raipur", "CHC Bilaspur", "CHC Korba"), each = 40)
)

kw_result <- kruskal.test(Satisfaction ~ CHC, data = sat_data)

ggplot(sat_data, aes(x = factor(Satisfaction), fill = CHC)) +
  geom_bar(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("CHC Raipur" = "#3498db", "CHC Bilaspur" = "#e74c3c", "CHC Korba" = "#27ae60")) +
  labs(title = "Patient Satisfaction Across Three CHCs (Chhattisgarh)",
       subtitle = paste0("Kruskal-Wallis H = ", round(kw_result$statistic, 2),
                        ", df = ", kw_result$parameter,
                        ", p = ", round(kw_result$p.value, 3)),
       x = "Satisfaction Score (1 = Poor, 5 = Excellent)", y = "Count", fill = NULL) +
  theme_clean()
```

---

## Section 8: Decision Flowchart — Choosing the Right Test

```{r}
#| label: fig-decision-tree
#| fig-cap: "Decision flowchart for choosing the correct test for comparing group means."

decision <- tibble(
  Question = c(
    "How many groups?",
    "Two groups",
    "Three+ groups",
    "Are data paired?",
    "Are data paired?",
    "Normal data (or n ≥ 30)?",
    "Normal data (or n ≥ 30)?",
    "Normal data (or n ≥ 30)?",
    "Normal data (or n ≥ 30)?"
  ),
  Scenario = c(
    "", "Two", "Three+",
    "Paired, Two", "Independent, Two",
    "Paired, Normal", "Paired, Non-normal",
    "Independent, Normal", "Independent, Non-normal"
  ),
  Test = c(
    "", "", "",
    "", "",
    "Paired t-test", "Wilcoxon signed-rank",
    "Independent t-test (Welch)", "Mann-Whitney U"
  )
)

# Display as a structured table
tibble(
  `Number of Groups` = c("**One**", "**Two (Paired)**", "**Two (Paired)**",
                          "**Two (Independent)**", "**Two (Independent)**",
                          "**Three+ (Independent)**", "**Three+ (Independent)**"),
  `Data Type` = c("Continuous", "Normal / n ≥ 30", "Skewed / ordinal / n < 30",
                   "Normal / n ≥ 30", "Skewed / ordinal / n < 30",
                   "Normal / n ≥ 30", "Skewed / ordinal / n < 30"),
  `Parametric Test` = c("One-sample t-test", "Paired t-test", "—",
                         "Welch's t-test", "—",
                         "One-way ANOVA", "—"),
  `Non-parametric Alternative` = c("Wilcoxon signed-rank (vs μ₀)", "Wilcoxon signed-rank", "Wilcoxon signed-rank",
                                    "Mann-Whitney U", "Mann-Whitney U",
                                    "Kruskal-Wallis", "Kruskal-Wallis"),
  `Post-hoc (if significant)` = c("—", "—", "—", "—", "—",
                                    "Tukey HSD / Bonferroni", "Dunn's test with Bonferroni")
) %>%
  kable(format = "html", escape = FALSE) %>%
  kable_styling(full_width = TRUE, font_size = 11) %>%
  row_spec(0, background = "#2c3e50", color = "white", bold = TRUE) %>%
  column_spec(1, bold = TRUE, width = "18%")
```

::::{.callout-tip}
**Practical heuristic:**

1. **Visualize first** — always plot your data before choosing a test
2. **n ≥ 30 per group?** → t-test/ANOVA are robust even with mild non-normality
3. **Ordinal data?** (Likert scales, pain scores) → always non-parametric
4. **Outliers dominate?** → non-parametric or consider transformation
5. **When in doubt:** Welch's t-test and report the result. If conclusions differ between parametric and non-parametric, mention both.
::::

---

## Section 9: Reporting Results

::::{.callout-tip}
**Reporting template for two-group comparisons:**

"Groups differed significantly [or did not differ] in [outcome] (Welch's t = X.XX, df = Y, p = Z.ZZ). The [treatment] group had a mean of A ± B, compared to C ± D in the [control] group. The mean difference was E (95% CI: F to G), representing a [small/medium/large] effect (Cohen's d = H)."

**For ANOVA:**

"A one-way ANOVA revealed significant differences in [outcome] across [k] groups (F(df₁, df₂) = X.XX, p = Z.ZZ). Post-hoc Tukey HSD indicated that [Group A] was significantly higher than [Group B] (Δ = E, 95% CI: F to G, adjusted p = Z)."
::::

::::{.callout-warning}
**Common reporting mistakes:**

1. Reporting only the p-value without the test statistic, df, or effect size
2. Writing "the groups were not different" instead of "no significant difference was found" (absence of evidence ≠ evidence of absence — Module 8!)
3. Using "significant" without specifying statistical significance vs clinical significance
4. Performing multiple t-tests instead of ANOVA when comparing 3+ groups
5. Reporting means for ordinal data (report medians and IQR instead)
::::

---

## Section 10: Summary and Key Takeaways

::::{.callout-important}
**Test Selection Summary:**

| Situation | Test | Measure |
|-----------|------|---------|
| One sample vs known value | One-sample t | Mean, CI, t, p |
| Two paired groups | Paired t-test (or Wilcoxon signed-rank) | Mean diff, CI, d |
| Two independent groups | Welch's t-test (or Mann-Whitney U) | Mean diff, CI, d |
| Three+ independent groups | One-way ANOVA (or Kruskal-Wallis) | F, post-hoc CIs |
::::

::::{.callout-tip}
**Key principles:**

1. **Always visualize** before testing — boxplots, histograms, Q-Q plots
2. **Welch's t-test** should be your default for two independent groups
3. **ANOVA first, then post-hoc** — never multiple t-tests
4. **Non-parametric** when: n < 30 and skewed, ordinal data, or outliers dominate
5. **Report effect sizes** (Cohen's d) — a significant p-value with d = 0.1 is trivial
6. **Report CIs** — they show precision and allow clinical interpretation
7. **The CLT** (Module 8) protects you with n ≥ 30, but not against outliers
::::

---

## Practice Questions: NEET PG Style MCQs {.neet-practice}

### Q1: Choosing the Right Test

A researcher compares blood glucose levels before and after a yoga intervention in 30 diabetic patients. The data are approximately normally distributed. Which test is most appropriate?

```{r}
#| label: q1-mcq
#| results: asis

make_mcq(
  id = "m9q1",
  question = "Blood glucose before and after yoga in 30 patients (normal data). Which test?",
  options = c(
    "Independent-samples t-test" = "Wrong — the data are paired (same patients measured twice), not independent groups.",
    "answer:Paired t-test" = "Correct! Same patients measured before and after → paired design. Data are normal, so the parametric paired t-test is appropriate.",
    "Mann-Whitney U test" = "Wrong — Mann-Whitney is for independent groups with non-normal data. Here data are paired and normal.",
    "Chi-square test" = "Wrong — Chi-square is for categorical data. Blood glucose is continuous."
  )
)
```

### Q2: ANOVA Interpretation

An ANOVA comparing serum ferritin across four dietary groups yields F(3, 76) = 1.82, p = 0.15. Which conclusion is correct?

```{r}
#| label: q2-mcq
#| results: asis

make_mcq(
  id = "m9q2",
  question = "ANOVA: F(3,76) = 1.82, p = 0.15. Which conclusion is correct?",
  options = c(
    "All four groups have exactly the same mean ferritin" = "Wrong — failing to reject H₀ does not prove the means are equal. It means we lack sufficient evidence to conclude they differ.",
    "answer:There is no significant evidence that the group means differ" = "Correct! p = 0.15 > 0.05, so we fail to reject H₀. This does NOT prove equality — the study may be underpowered.",
    "Post-hoc tests should be performed to find which pairs differ" = "Wrong — post-hoc tests are only performed after a significant ANOVA (p < 0.05). With p = 0.15, there is no basis for pairwise comparisons.",
    "The result proves the dietary interventions are all equally effective" = "Wrong — absence of evidence is not evidence of absence (Module 8). The trial may simply be underpowered."
  )
)
```

### Q3: Non-parametric Test Selection

A researcher compares hospital length of stay between three surgical techniques. LOS is heavily right-skewed with several extreme outliers, and n = 15 per group. Which test is most appropriate?

```{r}
#| label: q3-mcq
#| results: asis

make_mcq(
  id = "m9q3",
  question = "LOS across 3 surgical groups, skewed data, n=15/group. Which test?",
  options = c(
    "One-way ANOVA" = "Wrong — ANOVA assumes approximate normality. With n = 15 and heavy skew + outliers, the CLT doesn't protect you sufficiently.",
    "Three separate Mann-Whitney U tests" = "Wrong — this inflates Type I error (multiple comparisons problem). Use a single omnibus test first.",
    "answer:Kruskal-Wallis test" = "Correct! Three independent groups with non-normal data and small n → Kruskal-Wallis is the non-parametric equivalent of ANOVA. Follow with Dunn's post-hoc if significant.",
    "Paired t-test" = "Wrong — the groups are independent (different patients in each technique), not paired."
  )
)
```

### Q4: Multiple Comparisons

After a significant one-way ANOVA with 5 groups, a researcher performs 10 pairwise t-tests at α = 0.05 without correction. What is the approximate family-wise error rate?

```{r}
#| label: q4-mcq
#| results: asis

make_mcq(
  id = "m9q4",
  question = "5 groups, 10 pairwise t-tests at α = 0.05, no correction. Family-wise error rate?",
  options = c(
    "0.05 (5%)" = "Wrong — that would be true for a single test. With 10 independent tests, error accumulates.",
    "0.10 (10%)" = "Wrong — too low. The formula is 1 − (1 − α)^k.",
    "answer:Approximately 0.40 (40%)" = "Correct! P(≥1 false positive) = 1 − (0.95)^10 ≈ 0.40. This is why post-hoc corrections (Tukey, Bonferroni) are essential.",
    "0.50 (50%)" = "Wrong — close but not accurate. 1 − (0.95)^10 = 0.40, not 0.50."
  )
)
```

### Q5: Effect Size Interpretation

Two antihypertensives are compared in an RCT (n = 500 per group). Drug A reduces SBP by 0.5 mmHg more than Drug B (p = 0.04, Cohen's d = 0.08). What should the clinician conclude?

```{r}
#| label: q5-mcq
#| results: asis

make_mcq(
  id = "m9q5",
  question = "Drug A beats Drug B by 0.5 mmHg (p = 0.04, d = 0.08). Clinician's conclusion?",
  options = c(
    "Drug A is significantly better and should replace Drug B" = "Wrong — 0.5 mmHg is far below the MCID (5–10 mmHg). Statistical significance with a trivial effect size should not change practice.",
    "The p-value of 0.04 proves Drug A is clinically superior" = "Wrong — p < 0.05 means the difference is unlikely due to chance, but says nothing about clinical importance. Cohen's d = 0.08 is negligible.",
    "answer:The difference is statistically significant but clinically irrelevant; the effect size is negligible" = "Correct! This is the classic 'statistical vs clinical significance' trap (Module 8). With n = 500/group, even tiny differences reach p < 0.05. d = 0.08 is negligible, and 0.5 mmHg is far below MCID.",
    "The sample size was too small to detect a meaningful difference" = "Wrong — the opposite! n = 500/group is very large. The problem is that the true effect is trivially small, not that the study is underpowered."
  )
)
```

### Q6: Paired vs Independent

A researcher wants to compare the effectiveness of two blood glucose meters. She measures glucose on 40 blood samples, testing each sample on both meters. Which statistical test should she use?

```{r}
#| label: q6-mcq
#| results: asis

make_mcq(
  id = "m9q6",
  question = "Same 40 blood samples tested on two glucose meters. Which test?",
  options = c(
    "Independent t-test" = "Wrong — the measurements are paired (same sample on both meters). An independent t-test would ignore this pairing and lose power.",
    "answer:Paired t-test" = "Correct! Each blood sample is measured by both meters → naturally paired data. The paired t-test accounts for within-sample correlation and is the correct choice (assuming differences are approximately normal).",
    "Chi-square test" = "Wrong — glucose values are continuous, not categorical. Chi-square is for frequency data.",
    "One-way ANOVA" = "Wrong — ANOVA is for 3+ independent groups. Here we have 2 paired measurements."
  )
)
```

---

## Resources {.resources-box}

### Video Resources

**StatQuest with Josh Starmer (YouTube):**

- [t-tests](https://www.youtube.com/watch?v=0Pd3dc1GcHc)
- [ANOVA](https://www.youtube.com/watch?v=oOuu8IcsPJo)
- [Mann-Whitney U Test](https://www.youtube.com/watch?v=BT1FKd1Qzjw)

**Zed Statistics (YouTube):**

- [Paired t-test](https://www.youtube.com/watch?v=D43FVENPces)
- [ANOVA Explained](https://www.youtube.com/watch?v=0Vj2V2qRU10)

### Recommended Reading

1. **Bland M (2015).** *An Introduction to Medical Statistics (4th ed).* Oxford University Press. Chapters 9–10: t-tests and ANOVA.

2. **Kirkwood BR, Sterne JAC (2003).** *Essential Medical Statistics (2nd ed).* Blackwell Science. Chapter 7: Comparing Groups.

3. **Altman DG (1991).** *Practical Statistics for Medical Research.* Chapman & Hall. Chapter 9: Comparing Means.

### Key Papers

- **Bland JM, Altman DG (1995).** Multiple Significance Tests: the Bonferroni Method. *BMJ*, 310, 170.

- **de Winter JCF (2013).** Using the Student's t-test with extremely small sample sizes. *Practical Assessment, Research & Evaluation*, 18(10).

---

## References {#refs}

Altman DG. (1991). *Practical Statistics for Medical Research.* Chapman & Hall.

Bland M. (2015). *An Introduction to Medical Statistics* (4th ed). Oxford University Press.

Cohen J. (1988). *Statistical Power Analysis for the Behavioral Sciences* (2nd ed). Lawrence Erlbaum Associates.

Kirkwood BR, Sterne JAC. (2003). *Essential Medical Statistics* (2nd ed). Blackwell Science.

Kruskal WH, Wallis WA. (1952). Use of Ranks in One-Criterion Variance Analysis. *Journal of the American Statistical Association*, 47(260), 583–621.

Mann HB, Whitney DR. (1947). On a Test of Whether One of Two Random Variables Is Stochastically Larger Than the Other. *Annals of Mathematical Statistics*, 18(1), 50–60.

Wilcoxon F. (1945). Individual Comparisons by Ranking Methods. *Biometrics Bulletin*, 1(6), 80–83.

---

**Last Updated:** `r Sys.Date()`
**Module Authors:** Dr. Abhijit Pakhare & Dr. Ankur Joshi, AIIMS Bhopal
