---
title: "Categorical Data — Chi-square, Fisher's, and Risk Measures"
subtitle: "Statistical Analysis of Exposure-Disease Relationships"
description: |
  Master contingency tables, chi-square testing, and measures of association for categorical data.
  From outbreak investigation to epidemiological study design, learn when and how to use each test.
  Understand the critical difference between Relative Risk and Odds Ratio—a NEET PG favourite.
categories:
  - Biostatistics
  - Categorical Data
  - Chi-square
  - Odds Ratio
  - Risk Measures
author: "AIIMS Bhopal Biostatistics Course"
date: "`r Sys.Date()`"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(kableExtra)
source("../R/mcq.R")
library(patchwork)
library(scales)
library(vcd)

set.seed(42)

# Suppress warnings for cleaner output
options(warn = -1)

theme_clean <- function() {
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
}
```

::: {.callout-note appearance="minimal"}
**Lecture slides for this module:** [Open Slides](../slides/10-categorical-data-slides.html){target="_blank"}
:::

## Learning Objectives {#learning-objectives}

By the end of this module, you will be able to:

1. **Construct and interpret contingency tables** (2×2 and r×c) for categorical data
2. **Apply chi-square test of independence** and understand expected frequencies and degrees of freedom
3. **Distinguish between Fisher's exact test and Yates' correction** and when each is appropriate
4. **Explain McNemar's test** for paired categorical data in before-after studies
5. **Calculate and interpret measures of association**: Relative Risk (RR), Odds Ratio (OR), ARR, RRR, NNT, NNH, AR, and PAR
6. **Understand when OR ≈ RR** and the critical difference between them at high disease prevalence
7. **Develop a decision tree** for choosing the correct test for categorical data analysis

---

## Clinical Hook

:::{.callout-tip title="An Outbreak Investigation in Rural Madhya Pradesh"}
It's monsoon season in a village near Sagar, Madhya Pradesh. On June 15, 200 residents attend a community feast organised for a wedding. By June 17, local ASHA workers report that 80 villagers have developed acute gastroenteritis (abdominal cramps, diarrhoea, vomiting). The district surveillance officer arrives with her team.

She suspects three main dishes: **biryani, paneer tikka, and samosas**. She conducts a **case-control study** (matching 80 affected individuals with 80 unaffected attendees) and constructs 2×2 contingency tables.

For **biryani exposure**:

|  | Ate Biryani | Did Not Eat Biryani | Total |
|--|---|---|---|
| **Ill** | 65 | 15 | 80 |
| **Well** | 30 | 50 | 80 |
| **Total** | 95 | 65 | 160 |

**Your Tasks:**

1. Was the biryani the culprit? (Chi-square test)
2. How much more likely are people who ate biryani to get sick? (Odds Ratio)
3. If you were the food safety officer, would you shut down the catering business?

By the end of this module, you'll have the statistical tools to answer these questions with confidence.
:::

---

## Section 1: Contingency Tables and Chi-square Test

### 1.1 The 2×2 Contingency Table

A **2×2 contingency table** organizes data where two categorical variables, each with two levels, are cross-tabulated:

```{r}
#| label: generic-2x2
#| echo: false

cont_table <- data.frame(
  ` ` = c("**Exposed**", "**Not Exposed**", "**Total**"),
  `**Disease**` = c("a", "c", "a+c"),
  `**No Disease**` = c("b", "d", "b+d"),
  `**Total**` = c("a+b", "c+d", "N")
)

kable(cont_table, format = "html", escape = FALSE) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("bordered", "hover"))
```

Where:

- **a** = exposed with disease
- **b** = exposed without disease
- **c** = not exposed with disease
- **d** = not exposed without disease
- **N** = total sample size

### Outbreak Data (Biryani Exposure)

```{r}
#| label: biryani-table
#| echo: false

biryani_data <- data.frame(
  ` ` = c("**Ill (Gastroenteritis)**", "**Well**", "**Total**"),
  `**Ate Biryani**` = c("65", "30", "95"),
  `**Did Not Eat Biryani**` = c("15", "50", "65"),
  `**Total**` = c("80", "80", "160")
)

kable(biryani_data, format = "html", escape = FALSE, align = "c") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("bordered", "hover")) %>%
  column_spec(1, bold = TRUE)
```

In our outbreak data: a = 65, b = 30, c = 15, d = 50.

---

### 1.2 Chi-square Test of Independence

The **chi-square test** determines whether two categorical variables are independent (unrelated) or associated.

**Hypotheses:**

- **H₀**: Exposure and disease are **independent** (not associated)
- **H₁**: Exposure and disease are **associated** (dependent)

**Chi-square formula (in words):** For each cell, take the difference between what you *observed* and what you *expected* under independence, square it, and divide by the expected. Sum across all cells.

$$\chi^2 = \sum \frac{(\text{Observed} - \text{Expected})^2}{\text{Expected}} = \sum \frac{(O_i - E_i)^2}{E_i}$$

**Expected frequency** for cell (i, j):

$$E_{ij} = \frac{(\text{Row Total}_i) \times (\text{Column Total}_j)}{\text{Grand Total}}$$

For a 2×2 table, the **simplified formula** is:

$$\chi^2 = \frac{N(ad - bc)^2}{(a+b)(c+d)(a+c)(b+d)}$$

**Degrees of freedom:** $df = (r - 1) \times (c - 1)$. For a 2×2 table: df = 1.

---

### 1.3 Worked Example: Biryani Outbreak

**Step 1: Calculate expected frequencies**

```{r}
#| label: chi-sq-calc
#| echo: true

# Observed data
a <- 65  # ill, ate biryani
b <- 30  # well, ate biryani
c <- 15  # ill, didn't eat biryani
d <- 50  # well, didn't eat biryani
N <- a + b + c + d

# Expected frequencies
E_ill_biryani     <- ((a + c) * (a + b)) / N
E_well_biryani    <- ((b + d) * (a + b)) / N
E_ill_no_biryani  <- ((a + c) * (c + d)) / N
E_well_no_biryani <- ((b + d) * (c + d)) / N

cat("Expected Frequencies:\n")
cat("Ill, Ate Biryani:", round(E_ill_biryani, 2), "\n")
cat("Well, Ate Biryani:", round(E_well_biryani, 2), "\n")
cat("Ill, No Biryani:", round(E_ill_no_biryani, 2), "\n")
cat("Well, No Biryani:", round(E_well_no_biryani, 2), "\n")
```

**Step 2: Calculate chi-square statistic**

```{r}
#| label: chi-sq-stat
#| echo: true

chi_sq <- (N * (a*d - b*c)^2) / ((a+b) * (c+d) * (a+c) * (b+d))

cat("Chi-square statistic:", round(chi_sq, 4), "\n")
cat("Degrees of freedom: 1\n")

p_value <- 1 - pchisq(chi_sq, df = 1)
cat("p-value:", format(p_value, scientific = TRUE), "\n")
cat("\nInterpretation: p < 0.05, so we REJECT H₀.\n")
cat("The biryani exposure and illness ARE significantly associated.\n")
```

:::{.callout-note title="Key Insight"}
A small p-value (< 0.05) indicates that the observed pattern is unlikely under the assumption of independence. This strongly suggests that eating biryani and developing gastroenteritis are **not independent** — the biryani is likely the culprit.
:::

---

### 1.4 Visualising the Outbreak Data

```{r}
#| label: mosaic-plot
#| echo: true
#| fig-width: 8
#| fig-height: 6

# Create contingency table
outbreak_table <- matrix(c(65, 30, 15, 50), nrow = 2, byrow = TRUE)
dimnames(outbreak_table) <- list(
  Biryani = c("Ate", "Did Not Eat"),
  Health = c("Ill", "Well")
)

# Mosaic plot
mosaic(outbreak_table,
       shade = TRUE,
       legend = TRUE,
       main = "Biryani Exposure and Gastroenteritis\nOutbreak Investigation, Sagar District",
       xlab = "Biryani Exposure",
       ylab = "Health Outcome",
       gp = shading_Friendly)
```

The **size of each rectangle** represents the proportion of people in that category. The **shading** indicates whether the observed count is more (blue) or less (red) than expected under independence.

### 1.5 Observed vs. Expected Frequencies

```{r}
#| label: freq-comparison
#| echo: true
#| fig-width: 10
#| fig-height: 6

freq_data <- data.frame(
  Category = c("Ate Biryani, Ill", "Ate Biryani, Well",
               "No Biryani, Ill", "No Biryani, Well"),
  Observed = c(65, 30, 15, 50),
  Expected = c(round(E_ill_biryani, 1), round(E_well_biryani, 1),
               round(E_ill_no_biryani, 1), round(E_well_no_biryani, 1))
)

freq_long <- freq_data %>%
  pivot_longer(cols = c("Observed", "Expected"),
               names_to = "Type", values_to = "Frequency")

ggplot(freq_long, aes(x = Category, y = Frequency, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = round(Frequency, 1)),
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Observed vs. Expected Frequencies",
       subtitle = "Biryani outbreak 2×2 table",
       x = "Category", y = "Frequency",
       fill = "Frequency Type") +
  theme_clean() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
```

:::{.callout-note title="The Chi-square Formula at a Glance"}
For a 2×2 table:

$$\chi^2 = \frac{N(ad - bc)^2}{(a+b)(c+d)(a+c)(b+d)}$$

The numerator is the **cross-product** difference. The larger the discrepancy between $ad$ and $bc$, the larger chi-square, and the stronger the evidence against independence.
:::

---

## Section 2: Assumptions and Alternatives

### 2.1 Assumptions of the Chi-square Test

The chi-square test requires:

1. **Random sample** from the population
2. **Independence of observations** (each person counted only once)
3. **Adequate sample size** — all expected frequencies ≥ 5 (for a 2×2 table)

When these assumptions are violated, we turn to alternatives.

---

### 2.2 Fisher's Exact Test

When **expected cell frequency < 5** in any cell of a 2×2 table, use **Fisher's exact test**. It computes exact probabilities rather than relying on the chi-square approximation.

**When to use Fisher's:**

- Any expected frequency < 5 in a 2×2 table
- Small sample sizes (N < 20–30)
- Rare diseases or rare exposures

**Worked example:** A small cluster of leptospirosis cases after flooding in Alappuzha, Kerala. Only 25 individuals are studied:

```{r}
#| label: fisher-test
#| echo: true

small_outbreak <- matrix(c(8, 2, 3, 12), nrow = 2, byrow = TRUE)
dimnames(small_outbreak) <- list(
  Exposure = c("Waded in floodwater", "Did not wade"),
  Disease = c("Leptospirosis", "Healthy")
)

cat("Small Outbreak 2×2 Table:\n")
print(small_outbreak)
cat("\n")

fisher_result <- fisher.test(small_outbreak)
cat("Fisher's Exact Test Results:\n")
cat("p-value:", round(fisher_result$p.value, 4), "\n")
cat("Odds Ratio:", round(fisher_result$estimate, 2), "\n")
```

:::{.callout-warning title="Common Mistake"}
Using chi-square when expected frequencies are < 5. The p-value becomes unreliable. **Always check expected frequencies first.**

Rule of thumb: For a 2×2 table with N < 30 or any expected frequency < 5, default to **Fisher's exact test**.
:::

---

### 2.3 Yates' Continuity Correction

For moderate sample sizes (20 < N < 50), some texts recommend **Yates' correction**:

$$\chi^2_{\text{Yates}} = \frac{N(|ad - bc| - N/2)^2}{(a+b)(c+d)(a+c)(b+d)}$$

**Modern perspective:** Yates' correction is often considered **overly conservative** and not routinely recommended. Fisher's exact test is preferred for small samples, and the standard chi-square is fine for large samples.

---

### 2.4 McNemar's Test for Paired Categorical Data

When the **same subjects are measured twice** (before–after or paired design), use **McNemar's test**.

**Scenario:** A TB control programme in Patna, Bihar, tests 100 contacts of TB patients for knowledge about transmission — before and after a health education session.

The McNemar table records **who changed their answer**:

|  | **After: Correct** | **After: Incorrect** |
|--|---|---|
| **Before: Correct** | 55 (concordant) | 5 (lost knowledge) |
| **Before: Incorrect** | 25 (gained knowledge) | 15 (concordant) |

Only the **discordant pairs** (b = 5, c = 25) contribute to the test:

$$\chi^2_{\text{McNemar}} = \frac{(b - c)^2}{b + c}$$

```{r}
#| label: mcnemar-example
#| echo: true

b <- 5   # Before correct, After incorrect
c <- 25  # Before incorrect, After correct

chi_sq_mcnemar <- (b - c)^2 / (b + c)
p_mcnemar <- 1 - pchisq(chi_sq_mcnemar, df = 1)

cat("McNemar's Test for TB Health Education Programme\n")
cat("Discordant pairs: b =", b, ", c =", c, "\n")
cat("Chi-square statistic:", round(chi_sq_mcnemar, 4), "\n")
cat("p-value:", round(p_mcnemar, 4), "\n")
cat("\nThe proportion with correct knowledge significantly\n")
cat("increased after health education (p < 0.05).\n")
```

:::{.callout-note title="McNemar's Test"}
Use when comparing **paired categorical outcomes** (same subject, before–after). It is the categorical equivalent of the paired t-test. Concordant pairs are ignored — only those who *changed* contribute to the test statistic.
:::

---

## Section 3: Measures of Association

We move beyond "is there an association?" to **"how strong is the association?"** and **"what is the clinical impact?"**

### 3.1 Risk and Probability

From a 2×2 table:

- **Risk in Exposed:** $P(\text{Disease | Exposed}) = \frac{a}{a+b}$
- **Risk in Unexposed:** $P(\text{Disease | Unexposed}) = \frac{c}{c+d}$

```{r}
#| label: risk-calculation
#| echo: true

# Reset outbreak data
a <- 65; b <- 30; c <- 15; d <- 50

risk_exposed   <- a / (a + b)
risk_unexposed <- c / (c + d)
cat("Risk in Exposed (ate biryani):", round(risk_exposed, 3),
    paste0("(", round(risk_exposed * 100, 1), "%)"), "\n")
cat("Risk in Unexposed (no biryani):", round(risk_unexposed, 3),
    paste0("(", round(risk_unexposed * 100, 1), "%)"), "\n")
```

---

### 3.2 Relative Risk (RR)

**Relative Risk** compares disease incidence between exposed and unexposed groups.

**In words:**

$$RR = \frac{\text{Risk of disease in Exposed}}{\text{Risk of disease in Unexposed}}$$

**In cell notation:**

$$RR = \frac{a/(a+b)}{c/(c+d)}$$

Interpretation: RR = 1 means no association; RR > 1 means increased risk; RR < 1 means protective.

```{r}
#| label: rr-calculation
#| echo: true

RR <- risk_exposed / risk_unexposed
cat("Relative Risk (RR):", round(RR, 3), "\n")
cat("\nPeople who ate biryani were", round(RR, 1), "times more likely\n")
cat("to develop gastroenteritis compared to those who didn't.\n")
```

:::{.callout-note title="Relative Risk (RR)"}
$$RR = \frac{\text{Incidence in Exposed}}{\text{Incidence in Unexposed}} = \frac{a/(a+b)}{c/(c+d)}$$

**Used in:** Cohort studies (follow exposed and unexposed groups forward in time).

**Advantage:** Directly estimates the factor by which risk changes with exposure.

**Limitation:** Can only be calculated from cohort studies where you can compute incidence; case-control studies must use the Odds Ratio.
:::

---

### 3.3 Odds Ratio (OR)

The **odds** of disease given exposure = number with disease divided by number without disease in the exposed group = $a/b$.

**In words:**

$$OR = \frac{\text{Odds of disease in Exposed}}{\text{Odds of disease in Unexposed}}$$

**In cell notation (two equivalent forms):**

$$OR = \frac{a/b}{c/d} = \frac{ad}{bc} \quad \text{(cross-product ratio)}$$

The cross-product form ($ad/bc$) is the quickest way to compute OR from a 2×2 table — multiply the diagonal cells and take the ratio.

```{r}
#| label: or-calculation
#| echo: true

odds_exposed   <- a / b
odds_unexposed <- c / d
OR <- odds_exposed / odds_unexposed

cat("Odds of illness in exposed:", round(odds_exposed, 3), "\n")
cat("Odds of illness in unexposed:", round(odds_unexposed, 3), "\n")
cat("\nOdds Ratio (OR):", round(OR, 3), "\n")
cat("\nThe odds of gastroenteritis are", round(OR, 2),
    "times higher\nin those who ate biryani.\n")
```

---

### 3.4 When Does OR ≈ RR? The Critical Relationship

:::{.callout-important title="NEET PG High-Yield"}
$$OR \approx RR \text{ when disease is RARE (incidence < 5–10%)}$$

When disease is **common** (incidence > 20%), **OR greatly overestimates RR**.
:::

```{r}
#| label: or-rr-comparison
#| echo: true

scenarios <- expand_grid(
  Prevalence = c(0.01, 0.05, 0.10, 0.20, 0.40),
  RR = c(1.5, 2.0, 3.0)
)

scenarios <- scenarios %>%
  mutate(
    baseline_risk = Prevalence,
    risk_exposed = baseline_risk * RR,
    OR_est = RR * ((1 - baseline_risk) / (1 - risk_exposed))
  ) %>%
  select(Prevalence, RR, OR_est) %>%
  mutate(
    across(c(Prevalence, OR_est), ~ round(., 3)),
    Prevalence = paste0(Prevalence * 100, "%")
  ) %>%
  arrange(desc(Prevalence))

kable(scenarios, col.names = c("Disease Prevalence", "Relative Risk", "Odds Ratio (Est.)"),
      format = "html") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("bordered", "hover")) %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3")
```

At 1% prevalence the OR closely matches the RR. At 40% prevalence the OR overshoots by 50% or more.

```{r}
#| label: or-vs-rr-plot
#| echo: true
#| fig-width: 10
#| fig-height: 6

prev_range <- seq(0.01, 0.60, by = 0.02)
rr_val <- 2.5

or_vals <- rr_val * ((1 - prev_range) / (1 - prev_range * rr_val))

df_or_rr <- data.frame(
  Prevalence = prev_range * 100,
  RR = rr_val,
  OR = or_vals
) %>%
  pivot_longer(cols = c("RR", "OR"), names_to = "Measure", values_to = "Value")

ggplot(df_or_rr, aes(x = Prevalence, y = Value, color = Measure,
                      linetype = Measure)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = rr_val, linetype = "dashed", color = "gray",
             linewidth = 0.5) +
  scale_color_manual(values = c("RR" = "#1f77b4", "OR" = "#ff7f0e")) +
  scale_linetype_manual(values = c("RR" = "solid", "OR" = "dashed")) +
  annotate("text", x = 35, y = rr_val + 0.2, label = "RR constant = 2.5",
           color = "#1f77b4", fontface = "bold") +
  annotate("text", x = 50, y = 5, label = "OR diverges\nfrom RR at\nhigh prevalence",
           color = "#ff7f0e", fontface = "bold") +
  labs(title = "Odds Ratio vs. Relative Risk",
       subtitle = "RR = 2.5 fixed; disease prevalence varies",
       x = "Disease Prevalence (%)",
       y = "Measure Value") +
  theme_clean()
```

:::{.callout-warning title="NEET PG Trap"}
A case-control study of melanoma reports OR = 3.2 for sun exposure. A student incorrectly states: "Sun exposure increases melanoma risk by 3.2 times."

**The error:** Case-control studies cannot directly estimate RR. The OR = 3.2 approximates RR **only if melanoma is rare**. If prevalence is high, OR overestimates the true RR.

**Correct statement:** "The odds of past sun exposure are 3.2 times higher in cases than controls. Under the rare disease assumption, RR ≈ 3.2."
:::

---

### 3.5 Confidence Intervals for OR

$$OR_{95\%CI} = OR \times e^{\pm 1.96 \times SE(\ln OR)}$$

where $SE(\ln OR) = \sqrt{\frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}}$

```{r}
#| label: or-ci
#| echo: true

ln_OR <- log(OR)
se_ln_OR <- sqrt(1/a + 1/b + 1/c + 1/d)
margin_error <- 1.96 * se_ln_OR

or_ci_lower <- exp(ln_OR - margin_error)
or_ci_upper <- exp(ln_OR + margin_error)

cat("Odds Ratio and 95% Confidence Interval:\n")
cat("OR =", round(OR, 3), "\n")
cat("95% CI:", round(or_ci_lower, 3), "–", round(or_ci_upper, 3), "\n")
cat("\nSince the CI does NOT include 1.0, the association\n")
cat("is statistically significant at the 0.05 level.\n")
```

:::{.callout-note title="Confidence Interval Rule"}
- **CI includes 1.0:** Not statistically significant (p > 0.05)
- **CI excludes 1.0:** Statistically significant (p < 0.05)
:::

---

## Section 4: Measures of Impact and Clinical Significance

We shift from association (RR, OR) to **impact measures** that directly inform clinical decisions.

### 4.1 Absolute Risk Reduction (ARR) and Relative Risk Reduction (RRR)

These measures translate association into clinical impact.

**In words:**

- **ARR** = How much does treatment lower the absolute risk? = Risk in Control group − Risk in Treatment group
- **RRR** = What fraction of the baseline risk is removed by treatment? = ARR ÷ Risk in Control group

**Scenario:** A TB vaccine trial conducted across PHCs in Varanasi, Uttar Pradesh — 500 vaccinated vs. 500 unvaccinated individuals:

- Unvaccinated: 20/500 developed TB (4.0%)
- Vaccinated: 5/500 developed TB (1.0%)

```{r}
#| label: arr-rrr-calc
#| echo: true

risk_control       <- 20 / 500
risk_intervention  <- 5 / 500

ARR <- risk_control - risk_intervention
RRR <- ARR / risk_control

cat("Risk in Unvaccinated:", round(risk_control * 100, 2), "%\n")
cat("Risk in Vaccinated:", round(risk_intervention * 100, 2), "%\n")
cat("\nAbsolute Risk Reduction (ARR):", round(ARR * 100, 2), "%\n")
cat("Vaccinating prevents illness in 3 per 100 people.\n")
cat("\nRelative Risk Reduction (RRR):", round(RRR * 100, 2), "%\n")
cat("Vaccination reduces TB risk by 75% relative to baseline.\n")
```

:::{.callout-note title="ARR vs. RRR"}
$$ARR = \text{Risk}_{\text{Control}} - \text{Risk}_{\text{Intervention}}$$

$$RRR = \frac{ARR}{\text{Risk}_{\text{Control}}}$$

**Critical difference:** ARR is what matters clinically; RRR can sound impressive but overstate impact when baseline risk is low.

Example: If baseline risk is 1%, an intervention that halves it gives RRR = 50% (sounds great) but ARR = 0.5% (modest impact).
:::

---

### 4.2 Number Needed to Treat (NNT)

**NNT** = how many patients you must treat to prevent **one** adverse outcome.

**In words:** If the ARR is 3% (i.e., treatment prevents disease in 3 out of every 100), then you need to treat 100/3 ≈ 33 people to prevent one case.

$$NNT = \frac{1}{ARR}$$

```{r}
#| label: nnt-calc
#| echo: true

NNT <- 1 / ARR
cat("Number Needed to Treat (NNT):", round(NNT, 1), "\n")
cat("\nYou must vaccinate", round(NNT, 1),
    "people to prevent one case of TB.\n")
```

```{r}
#| label: nnt-icon-array
#| echo: true
#| fig-width: 10
#| fig-height: 4

nnt_value <- round(NNT, 0)

icon_data <- data.frame(
  x = rep(1:nnt_value, length.out = nnt_value),
  y = rep(1, nnt_value),
  Status = c("Protected by vaccine", rep("No change", nnt_value - 1))
)

ggplot(icon_data, aes(x = x, y = y, fill = Status)) +
  geom_tile(color = "white", linewidth = 0.8, height = 0.8) +
  scale_fill_manual(values = c("Protected by vaccine" = "#2ecc71",
                                "No change" = "#bdc3c7")) +
  coord_equal() +
  labs(title = paste0("NNT = ", nnt_value,
                      " — Vaccinate ", nnt_value,
                      " to prevent 1 case of TB"),
       x = "", y = "", fill = "Outcome") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
        legend.position = "bottom")
```

:::{.callout-note title="Interpreting NNT"}
- **NNT ≈ 10:** Excellent (treat 10 to save one)
- **NNT ≈ 50:** Moderate benefit
- **NNT > 200:** Modest benefit

Compare across interventions: an NNT of 15 is superior to an NNT of 40.
:::

---

### 4.3 Number Needed to Harm (NNH)

$$NNH = \frac{1}{\text{Risk}_{\text{Treatment}}^{\text{harm}} - \text{Risk}_{\text{Control}}^{\text{harm}}}$$

Suppose in the same vaccine trial, severe local reactions occur in 8/500 vaccinated vs. 2/500 unvaccinated:

```{r}
#| label: nnh-calc
#| echo: true

risk_harm_vaccine  <- 8 / 500
risk_harm_control  <- 2 / 500

ARH <- risk_harm_vaccine - risk_harm_control
NNH <- 1 / ARH

cat("Risk of severe reaction in Vaccinated:", round(risk_harm_vaccine * 100, 2), "%\n")
cat("Risk of severe reaction in Unvaccinated:", round(risk_harm_control * 100, 2), "%\n")
cat("Absolute Risk of Harm:", round(ARH * 100, 2), "%\n")
cat("\nNNH:", round(NNH, 1), "\n")
cat("For every", round(NNH, 0), "vaccinated, 1 has a severe reaction.\n")
```

**Therapeutic index:** NNH / NNT tells you how the benefit compares to harm:

```{r}
#| label: therapeutic-index
#| echo: true

TI <- NNH / NNT
cat("Therapeutic Index (NNH/NNT):", round(TI, 2), "\n")
if (TI > 1) {
  cat("Benefit clearly outweighs harm — VACCINATION RECOMMENDED.\n")
} else {
  cat("Harm exceeds benefit — reconsider intervention.\n")
}
```

---

### 4.4 Attributable Risk (AR) and Population Attributable Risk (PAR)

**Attributable Risk (AR)** answers: *"How much extra disease does the exposure cause in exposed individuals?"*

**In words:**

- **AR** = Risk in Exposed − Risk in Unexposed (the excess risk attributable to the exposure)
- **AR%** = What percentage of disease among the exposed is caused by the exposure? = (AR ÷ Risk in Exposed) × 100

**Population Attributable Risk (PAR)** answers: *"If we could eliminate the exposure entirely, what fraction of disease in the whole population would disappear?"*

- **PAR** depends on both the strength of association (RR) *and* how common the exposure is in the population ($P_e$)

**Scenario:** A cohort study of smoking and lung cancer:

- Risk in smokers: 8%
- Risk in non-smokers: 1%

```{r}
#| label: ar-calculation
#| echo: true

risk_smokers    <- 0.08
risk_nonsmokers <- 0.01

AR <- risk_smokers - risk_nonsmokers
AR_percent <- (AR / risk_smokers) * 100

cat("Attributable Risk (AR):", round(AR * 100, 2), "%\n")
cat("AR%:", round(AR_percent, 1), "%\n")
cat("\n", round(AR_percent, 1), "% of lung cancer in smokers\n")
cat("is attributable to (caused by) smoking.\n")
```

**Population Attributable Risk (PAR)** answers: *"What proportion of disease in the entire population is due to this exposure?"*

$$PAR\% = \frac{P_e(RR - 1)}{P_e(RR - 1) + 1} \times 100$$

where $P_e$ = prevalence of exposure.

```{r}
#| label: par-calculation
#| echo: true

P_e <- 0.40  # 40% of population smokes
RR_smoking <- risk_smokers / risk_nonsmokers

PAR_percent <- (P_e * (RR_smoking - 1)) / (P_e * (RR_smoking - 1) + 1) * 100

cat("Smoking prevalence:", P_e * 100, "%\n")
cat("RR:", round(RR_smoking, 2), "\n")
cat("\nPAR%:", round(PAR_percent, 1), "%\n")
cat(round(PAR_percent, 1), "% of ALL lung cancer in the population\n")
cat("could be prevented if smoking were eliminated.\n")
```

:::{.callout-note title="AR vs. PAR — Clinical vs. Public Health"}
$$AR = \text{Risk}_{\text{Exposed}} - \text{Risk}_{\text{Unexposed}}$$

$$PAR\% = \frac{P_e(RR - 1)}{P_e(RR - 1) + 1} \times 100$$

- **AR:** Relevant for individual patient counselling ("your personal risk from this exposure")
- **PAR%:** Relevant for public health policy ("population-level benefit of removing exposure")
:::

---

## Section 5: Decision Guide for Test Selection

```{r}
#| label: decision-guide
#| echo: false

decision_data <- tibble(
  Scenario = c(
    "Two independent groups, all expected freq ≥ 5",
    "Two independent groups, any expected freq < 5",
    "Paired / before–after categorical data",
    "Ordered categories, testing for trend",
    "Cohort study, 2×2 table",
    "Case-control study, 2×2 table",
    "Measuring clinical impact of treatment"
  ),
  Test = c(
    "Chi-square test",
    "Fisher's exact test",
    "McNemar's test",
    "Chi-square test for trend",
    "Relative Risk (RR)",
    "Odds Ratio (OR)",
    "ARR, RRR, NNT, NNH"
  ),
  Note = c(
    "Most common; requires adequate sample size",
    "Exact probability; no sample size restriction",
    "Uses only discordant pairs",
    "Extension for ordinal categories",
    "Direct risk measure; OR ≈ RR when disease is rare",
    "Only valid measure from case-control design",
    "Translate statistical results into clinical decisions"
  )
)

kable(decision_data, col.names = c("Scenario", "Test / Measure", "Key Note"),
      format = "html", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  column_spec(1, bold = TRUE, width = "30%") %>%
  column_spec(2, bold = TRUE, color = "#2c3e50", width = "25%") %>%
  column_spec(3, width = "45%")
```

### Quick Reference Table

```{r}
#| label: quick-reference
#| echo: false

decision_df <- data.frame(
  Question = c(
    "Are two categorical variables associated?",
    "Small sample / expected freq < 5?",
    "Same subjects, before–after?",
    "Cohort study (can calculate incidence)?",
    "Case-control study?",
    "How many to treat to prevent one outcome?",
    "Proportion of disease due to exposure?",
    "Population benefit of removing exposure?"
  ),
  Measure = c(
    "Chi-square test",
    "Fisher's exact test",
    "McNemar's test",
    "Relative Risk (RR)",
    "Odds Ratio (OR)",
    "NNT = 1/ARR",
    "Attributable Risk (AR)",
    "Population Attributable Risk (PAR)"
  ),
  Null = c(
    "Variables are independent",
    "Variables are independent (exact)",
    "No change in paired proportions",
    "RR = 1",
    "OR = 1",
    "ARR = 0 (no benefit)",
    "Risk same in exposed/unexposed",
    "Population at baseline unexposed risk"
  )
)

kable(decision_df, col.names = c("Study Design / Question", "Test / Measure",
                                  "Null Hypothesis / Assumption"),
      format = "html", escape = FALSE) %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("bordered", "hover")) %>%
  row_spec(0, bold = TRUE, background = "#4472C4", color = "white")
```

:::{.callout-tip title="Core Formulae — In Words and Notation"}

| Measure | In Words | Cell Notation |
|---------|----------|---------------|
| **Chi-square** | Sum of (Observed − Expected)² ÷ Expected | $\sum (O - E)^2 / E$ |
| **RR** | Risk in Exposed ÷ Risk in Unexposed | $\frac{a/(a+b)}{c/(c+d)}$ |
| **OR** | Odds in Exposed ÷ Odds in Unexposed = Cross-product ratio | $\frac{ad}{bc}$ |
| **ARR** | Risk in Control − Risk in Treatment | $R_c - R_t$ |
| **RRR** | ARR ÷ Risk in Control | $ARR / R_c$ |
| **NNT** | 1 ÷ ARR (how many to treat to save one) | $1 / ARR$ |
| **NNH** | 1 ÷ Absolute Risk of Harm | $1 / ARH$ |
| **AR** | Risk in Exposed − Risk in Unexposed (excess risk from exposure) | $R_e - R_u$ |
| **AR%** | (AR ÷ Risk in Exposed) × 100 | $(AR / R_e) \times 100$ |
| **PAR%** | Fraction of disease in population attributable to exposure | $\frac{P_e(RR-1)}{P_e(RR-1)+1} \times 100$ |
:::

---

## Section 6: Complete R Workflow

```{r}
#| label: complete-workflow
#| echo: true

# ==========================================
# COMPLETE OUTBREAK ANALYSIS WORKFLOW
# ==========================================

# Step 1: Input data
a <- 65; b <- 30; c <- 15; d <- 50
N <- a + b + c + d

outbreak <- data.frame(
  Exposure = c("Ate Biryani", "Ate Biryani", "No Biryani", "No Biryani"),
  Health = c("Ill", "Well", "Ill", "Well"),
  Count = c(a, b, c, d)
)

cont_table <- xtabs(Count ~ Exposure + Health, data = outbreak)
print(cont_table)

# Step 2: Chi-square test
chi_test <- chisq.test(cont_table)
cat("\n--- CHI-SQUARE TEST ---\n")
print(chi_test)

# Step 3: Measures of association
risk_exp   <- a / (a + b)
risk_unexp <- c / (c + d)
RR <- risk_exp / risk_unexp
OR <- (a * d) / (b * c)

ln_OR    <- log(OR)
se_ln_OR <- sqrt(1/a + 1/b + 1/c + 1/d)
or_ci    <- exp(ln_OR + c(-1, 1) * 1.96 * se_ln_OR)

ARR <- risk_exp - risk_unexp

cat("\n--- MEASURES OF ASSOCIATION ---\n")
cat("Risk in Exposed:", round(risk_exp, 3), "\n")
cat("Risk in Unexposed:", round(risk_unexp, 3), "\n")
cat("Relative Risk (RR):", round(RR, 3), "\n")
cat("Odds Ratio (OR):", round(OR, 3), "\n")
cat("95% CI for OR:", round(or_ci[1], 3), "–", round(or_ci[2], 3), "\n")
cat("ARR:", round(ARR * 100, 2), "%\n")

cat("\n--- INTERPRETATION ---\n")
cat("Biryani exposure is significantly associated with\n")
cat("gastroenteritis (χ² =", round(chi_test$statistic, 3),
    ", p =", round(chi_test$p.value, 4), ").\n")
cat("People eating biryani were", round(RR, 1),
    "times more likely to get ill.\n")
cat("The odds of illness were", round(OR, 1),
    "times higher in those who ate biryani.\n")
```

---

## Summary

:::{.callout-tip title="Key Takeaways"}
1. **Chi-square test** determines whether two categorical variables are associated. Check expected frequencies ≥ 5; use Fisher's exact test for small samples.

2. **McNemar's test** handles paired categorical data (before–after). It only uses discordant pairs.

3. **Relative Risk (RR)** compares incidence in exposed vs. unexposed — used in **cohort studies**.

4. **Odds Ratio (OR)** compares odds — used in **case-control studies**. When disease is rare (< 5–10%), OR ≈ RR. At high prevalence, OR overestimates RR.

5. **NNT = 1/ARR** answers "How many must I treat to prevent one outcome?" Lower NNT = greater benefit.

6. **AR and PAR** quantify individual and population-level impact, guiding public health policy.

7. **Always report 95% CI** for OR and RR. If the CI excludes 1.0, the result is statistically significant.
:::

---

## Practice Questions

### Q1: Number Needed to Treat

```{r}
#| label: mcq1
#| echo: false
#| results: asis

make_mcq(
  "m10q1",
  "A vaccine trial enrolls 500 vaccinated and 500 unvaccinated individuals. Over 2 years, 10 vaccinated and 40 unvaccinated develop typhoid. What is the NNT?",
  c(
    "6.7" = "Incorrect. This uses the wrong denominator — check your ARR calculation.",
    "answer:16.7" = "Correct! ARR = 40/500 − 10/500 = 0.06. NNT = 1/0.06 = 16.7. About 17 people must be vaccinated to prevent one case.",
    "12.5" = "Incorrect. This confuses the ARR with a different quantity.",
    "25.0" = "Incorrect. This is 1/(risk in unvaccinated), not 1/ARR."
  )
)
```

---

### Q2: OR Approximates RR

```{r}
#| label: mcq2
#| echo: false
#| results: asis

make_mcq(
  "m10q2",
  "In a case-control study, OR for processed meat and colorectal cancer is 2.5. Population prevalence of colorectal cancer is 3%. Which statement is MOST accurate?",
  c(
    "Processed meat increases cancer risk by 2.5 times (RR = 2.5)" = "Incorrect. Case-control studies yield OR, not RR directly. OR = RR only under rare disease assumption.",
    "answer:OR ≈ RR because colorectal cancer is rare (< 5%)" = "Correct! At 3% prevalence, the rare disease assumption holds and OR closely approximates RR.",
    "OR overstates RR because cancer prevalence is high" = "Incorrect. 3% is below the 5–10% threshold — the rare disease assumption applies here.",
    "OR and RR are always identical in case-control studies" = "Incorrect. They converge only when disease prevalence is low."
  )
)
```

---

### Q3: Chi-Square Interpretation

```{r}
#| label: mcq3
#| echo: false
#| results: asis

make_mcq(
  "m10q3",
  "A foodborne outbreak table shows: Exposed+Ill = 45, Exposed+Well = 20, Unexposed+Ill = 10, Unexposed+Well = 75. Chi-square = 34.5, p < 0.001. Which interpretation is CORRECT?",
  c(
    "The food definitely caused the illness" = "Incorrect. Chi-square tests association, not causation. Causation requires Bradford Hill criteria.",
    "answer:There is a statistically significant association between food exposure and illness" = "Correct! We reject H₀ of independence — exposure and illness are significantly associated. Association ≠ causation.",
    "The contaminated food is the only cause" = "Incorrect. Chi-square cannot establish sole causation.",
    "45 people were definitely poisoned by this food" = "Incorrect. The test addresses the population-level pattern, not individual-level certainty."
  )
)
```

---

### Q4: Relative Risk Calculation

```{r}
#| label: mcq4
#| echo: false
#| results: asis

make_mcq(
  "m10q4",
  "In a cohort study, 120 of 1000 smokers and 10 of 1000 non-smokers develop lung cancer over 10 years. What is the Relative Risk?",
  c(
    "answer:12.0" = "Correct! RR = (120/1000) / (10/1000) = 0.12/0.01 = 12.0. Smokers are 12 times more likely to develop lung cancer.",
    "11.0" = "Incorrect. This confuses the risk difference (ARR = 0.11) with the risk ratio.",
    "110/1000" = "Incorrect. This is the absolute risk difference, not the relative risk.",
    "120/130" = "Incorrect. This divides cases in smokers by total cases — that is not RR."
  )
)
```

---

### Q5: Absolute Risk Reduction

```{r}
#| label: mcq5
#| echo: false
#| results: asis

make_mcq(
  "m10q5",
  "A vaccine trial: 5/500 vaccinated and 20/500 placebo develop disease. What is the ARR and its clinical meaning?",
  c(
    "ARR = 0.03; for every 3 vaccinated, 1 is protected" = "Incorrect. ARR = 0.03 means 3 per 100, not 1 per 3.",
    "answer:ARR = 0.03; for every 100 vaccinated, 3 are protected" = "Correct! ARR = 4% − 1% = 3% = 0.03. Per 100 vaccinated, 3 are protected. NNT = 1/0.03 ≈ 33.",
    "ARR = 15; vaccine is 15% effective" = "Incorrect. ARR is a proportion (0.03), not a count.",
    "ARR = 0.05; for every 20 vaccinated, 1 is protected" = "Incorrect. Check the risk calculation: 20/500 = 0.04, 5/500 = 0.01, ARR = 0.03."
  )
)
```

---

### Q6: McNemar's Test

```{r}
#| label: mcq6
#| echo: false
#| results: asis

make_mcq(
  "m10q6",
  "100 patients are tested before and after a health education session. Results: 55 correct→correct, 5 correct→incorrect, 25 incorrect→correct, 15 incorrect→incorrect. Which test is appropriate and what are the discordant pairs?",
  c(
    "Chi-square test; all 100 observations" = "Incorrect. Chi-square assumes independent observations — these are paired (same subjects measured twice).",
    "Paired t-test; mean scores before and after" = "Incorrect. The outcome is categorical (correct/incorrect), not continuous.",
    "answer:McNemar's test; b = 5 and c = 25 (discordant pairs)" = "Correct! McNemar's test is for paired categorical data. Only those who changed (5 + 25 = 30 discordant pairs) contribute to the test.",
    "Fisher's exact test; small expected frequencies" = "Incorrect. Fisher's is for unpaired 2×2 tables with small samples, not paired designs."
  )
)
```

---

## Appendix: R Functions Quick Reference

```{r}
#| label: appendix
#| echo: true
#| eval: false

# ======================================
# QUICK R REFERENCE — CATEGORICAL ANALYSIS
# ======================================

# Create 2×2 contingency table
cont_table <- matrix(c(a, b, c, d), nrow = 2, byrow = TRUE)

# Chi-square test
chisq.test(cont_table)

# Fisher's exact test (small samples)
fisher.test(cont_table)

# McNemar's test (paired data)
mcnemar.test(cont_table)

# Relative Risk
RR <- (a/(a+b)) / (c/(c+d))

# Odds Ratio
OR <- (a*d) / (b*c)

# 95% CI for OR
ln_OR    <- log(OR)
se_ln_OR <- sqrt(1/a + 1/b + 1/c + 1/d)
ci_or    <- exp(ln_OR + c(-1.96, 1.96) * se_ln_OR)

# ARR and NNT
ARR <- risk_control - risk_treatment
NNT <- 1 / ARR

# Attributable Risk Percent
AR_percent <- ((risk_exposed - risk_unexposed) / risk_exposed) * 100

# Population Attributable Risk
PAR_pct <- (Pe * (RR - 1)) / (Pe * (RR - 1) + 1) * 100
```

---

**End of Module 10**

*AIIMS Bhopal Biostatistics Course | Updated: `r Sys.Date()`*
