---
title: "Correlation and Regression"
subtitle: "From Measuring Association to Predicting Outcomes"
description: |
  Quantify relationships between continuous variables using Pearson and Spearman
  correlation. Build simple linear regression models, check assumptions, and interpret
  R². Extend to multiple regression for confounding adjustment and logistic regression
  for binary outcomes. Indian clinical examples throughout.
categories:
  - Biostatistics
  - Correlation
  - Regression
  - Linear Models
  - Logistic Regression
author: "AIIMS Bhopal Biostatistics Course"
date: "`r Sys.Date()`"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(kableExtra)
source("../R/mcq.R")
library(patchwork)
library(scales)
library(broom)

set.seed(42)

theme_clean <- function() {
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
}
```

::: {.callout-note appearance="minimal"}
**Lecture slides for this module:** [Open Slides](../slides/11-correlation-regression-slides.html){target="_blank"}
:::

## Learning Objectives {#learning-objectives}

By the end of this module, you will be able to:

1. **Calculate and interpret** Pearson and Spearman correlation coefficients
2. **Distinguish** correlation from causation and identify lurking variables
3. **Fit and interpret** a simple linear regression model (slope, intercept, R²)
4. **Check regression assumptions** using residual plots and respond to violations
5. **Explain** confidence intervals vs. prediction intervals for regression
6. **Understand multiple regression** as a tool for confounding adjustment
7. **Interpret logistic regression** output (odds ratios, predicted probabilities) for binary outcomes
8. **Navigate a decision framework** for choosing the right regression model

---

## Clinical Hook

A research team at JIPMER, Puducherry, is studying whether higher BMI is associated with higher fasting blood sugar (FBS) in 200 adults attending a health screening camp. They plot the data and see what looks like a positive trend. But the questions cascade:

1. **How strong** is the association? (Correlation)
2. **Can we predict** FBS from BMI? (Simple linear regression)
3. **Does the association persist** after adjusting for age and sex? (Multiple regression)
4. If the outcome were diabetes (yes/no) rather than FBS, **what model should they use?** (Logistic regression)

This module builds each tool in sequence.

---

## Section 1: Correlation

### 1.1 What Correlation Measures

Correlation quantifies the **strength and direction** of the *linear* relationship between two continuous variables.

**In words:** As one variable goes up, does the other tend to go up (positive), go down (negative), or show no pattern (zero)?

$$r = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum (x_i - \bar{x})^2 \cdot \sum (y_i - \bar{y})^2}}$$

**Conceptually:** r = average of the products of standardised deviations = Covariance(x, y) / (SD_x × SD_y).

Range: −1 (perfect negative) to +1 (perfect positive). r = 0 means no *linear* association.

---

### 1.2 Generating the Data

```{r}
#| label: generate-data
#| echo: true

# 200 adults at JIPMER health camp
n <- 200
age <- round(rnorm(n, mean = 45, sd = 12))
bmi <- round(rnorm(n, mean = 26, sd = 4.5), 1)
# FBS partly driven by BMI + age + noise
fbs <- round(80 + 1.8 * bmi + 0.5 * age + rnorm(n, sd = 15))

health <- data.frame(
  ID = 1:n, Age = age, BMI = bmi, FBS = fbs,
  Sex = sample(c("Male", "Female"), n, replace = TRUE)
)

cat("First 6 rows:\n")
head(health)
```

---

### 1.3 Pearson Correlation: Worked Example

```{r}
#| label: pearson-scatter
#| echo: true
#| fig-width: 9
#| fig-height: 6

r_bmi_fbs <- cor(health$BMI, health$FBS)
cor_test <- cor.test(health$BMI, health$FBS)

ggplot(health, aes(x = BMI, y = FBS)) +
  geom_point(alpha = 0.5, color = "#2c3e50") +
  geom_smooth(method = "lm", se = TRUE, color = "#e74c3c", linewidth = 1) +
  annotate("text", x = 35, y = 85,
           label = paste0("r = ", round(r_bmi_fbs, 3),
                         "\np < 0.001"),
           fontface = "bold", size = 5, color = "#e74c3c") +
  labs(title = "BMI vs. Fasting Blood Sugar",
       subtitle = "JIPMER Health Screening Camp, n = 200",
       x = "BMI (kg/m²)", y = "FBS (mg/dL)") +
  theme_clean()
```

**Interpretation:** r = `r round(r_bmi_fbs, 3)` — a moderate positive correlation. As BMI increases, FBS tends to increase. The p-value tests H₀: ρ = 0 (no linear association in the population).

:::{.callout-note title="Interpreting Correlation Strength"}

| |r| | Interpretation |
|------|---------------|
| 0.00 – 0.19 | Negligible |
| 0.20 – 0.39 | Weak |
| 0.40 – 0.59 | Moderate |
| 0.60 – 0.79 | Strong |
| 0.80 – 1.00 | Very strong |

These are rough guidelines — clinical context matters. A correlation of 0.3 between a biomarker and disease may be clinically important.
:::

---

### 1.4 Spearman Rank Correlation

When the relationship is **monotonic but not linear**, or the data are **ordinal**, use Spearman's rₛ (rank-based):

```{r}
#| label: spearman-example
#| echo: true

# Severity score (ordinal 1-5) vs. hospital LOS
severity <- sample(1:5, 60, replace = TRUE, prob = c(0.15, 0.25, 0.3, 0.2, 0.1))
los <- severity * 2 + rpois(60, lambda = 3) + round(rexp(60, rate = 0.3))

rs <- cor(severity, los, method = "spearman")
cat("Spearman rₛ =", round(rs, 3), "\n")
```

:::{.callout-tip title="Pearson vs. Spearman"}

| Feature | Pearson (r) | Spearman (rₛ) |
|---------|------------|---------------|
| Relationship type | Linear | Monotonic (any shape) |
| Data type | Continuous, roughly normal | Ordinal or skewed continuous |
| Sensitive to outliers? | Yes — a single outlier can flip r | No — works on ranks |
| When to use | Both variables continuous, linear scatter | Ordinal data, skewed, or outliers present |
:::

---

### 1.5 Correlation ≠ Causation

:::{.callout-warning title="The Ice-Cream-Drowning Fallacy"}
In summer months, ice cream sales and drowning deaths both rise. The correlation is strong (r ≈ 0.8). Should we ban ice cream?

**No.** The lurking variable is **temperature** — warm weather drives both.

Sources of spurious association include confounding (a third variable drives both), reverse causality (y causes x), selection bias (how the sample was chosen), and measurement error.

A strong r tells you the variables *move together*, not that one *causes* the other.
:::

---

## Section 2: Simple Linear Regression

### 2.1 The Model

We want to **predict** FBS from BMI. Simple linear regression fits the equation:

**In words:** Predicted outcome = Baseline level + (Rate of change × Predictor) + Random error

$$y = \beta_0 + \beta_1 x + \varepsilon$$

where:

- $\beta_0$ = **intercept** (predicted y when x = 0)
- $\beta_1$ = **slope** (change in y for each 1-unit increase in x)
- $\varepsilon$ = random error (what the model can't explain)

The method of **least squares** finds $\beta_0$ and $\beta_1$ that minimise $\sum (y_i - \hat{y}_i)^2$.

---

### 2.2 Fitting the Model in R

```{r}
#| label: fit-regression
#| echo: true

model <- lm(FBS ~ BMI, data = health)
summary(model)
```

```{r}
#| label: extract-coefs
#| echo: true

coefs <- tidy(model, conf.int = TRUE)
b0 <- coefs$estimate[1]
b1 <- coefs$estimate[2]

cat("Regression equation:\n")
cat("FBS =", round(b0, 2), "+", round(b1, 2), "× BMI\n")
cat("\nInterpretation:\n")
cat("For each 1 kg/m² increase in BMI, FBS increases by",
    round(b1, 2), "mg/dL on average.\n")
cat("95% CI for slope:", round(coefs$conf.low[2], 2),
    "to", round(coefs$conf.high[2], 2), "\n")
```

:::{.callout-note title="Interpreting the Slope"}
The slope $\beta_1$ = `r round(b1, 2)` means: **for every additional 1 kg/m² of BMI, fasting blood sugar increases by about `r round(b1, 1)` mg/dL**, holding nothing else constant.

The intercept $\beta_0$ = `r round(b0, 1)` is the predicted FBS when BMI = 0 — biologically meaningless here, but mathematically necessary to position the line.
:::

---

### 2.3 Visualising the Regression

```{r}
#| label: regression-plot
#| echo: true
#| fig-width: 9
#| fig-height: 6

ggplot(health, aes(x = BMI, y = FBS)) +
  geom_point(alpha = 0.4, color = "#2c3e50") +
  geom_smooth(method = "lm", se = TRUE, color = "#e74c3c", fill = "#e74c3c",
              alpha = 0.15, linewidth = 1) +
  annotate("text", x = 36, y = 90,
           label = paste0("FBS = ", round(b0, 1), " + ",
                         round(b1, 1), " × BMI\nR² = ",
                         round(summary(model)$r.squared, 3)),
           fontface = "bold", color = "#e74c3c", size = 4.5) +
  labs(title = "Simple Linear Regression: BMI → FBS",
       subtitle = "Shaded band = 95% confidence interval for the mean",
       x = "BMI (kg/m²)", y = "Fasting Blood Sugar (mg/dL)") +
  theme_clean()
```

---

### 2.4 R²: How Much Variance Is Explained?

**In words:** R² tells you what fraction of the total scatter in y is "explained" by the regression line.

$$R^2 = 1 - \frac{\text{Residual SS}}{\text{Total SS}} = 1 - \frac{\sum(y_i - \hat{y}_i)^2}{\sum(y_i - \bar{y})^2}$$

For simple regression, $R^2 = r^2$ (the square of the Pearson correlation).

```{r}
#| label: r-squared
#| echo: true

r_sq <- summary(model)$r.squared
cat("R² =", round(r_sq, 4), "\n")
cat("BMI explains", round(r_sq * 100, 1), "% of the variation in FBS.\n")
cat("The remaining", round((1 - r_sq) * 100, 1),
    "% is due to age, diet, genetics, measurement error, etc.\n")
```

:::{.callout-tip title="What Is a 'Good' R²?"}
There is no universal threshold. In clinical prediction models, R² = 0.3–0.5 may be useful. In lab assays, R² > 0.95 is expected. Always ask: "Is this model useful for the clinical question?"
:::

---

### 2.5 Confidence Interval vs. Prediction Interval

```{r}
#| label: ci-vs-pi
#| echo: true
#| fig-width: 9
#| fig-height: 6

new_data <- data.frame(BMI = seq(16, 40, by = 0.5))
ci_pred <- predict(model, newdata = new_data, interval = "confidence")
pi_pred <- predict(model, newdata = new_data, interval = "prediction")

plot_data <- data.frame(
  BMI = new_data$BMI,
  Fit = ci_pred[, "fit"],
  CI_lower = ci_pred[, "lwr"], CI_upper = ci_pred[, "upr"],
  PI_lower = pi_pred[, "lwr"], PI_upper = pi_pred[, "upr"]
)

ggplot() +
  geom_ribbon(data = plot_data, aes(x = BMI, ymin = PI_lower, ymax = PI_upper),
              fill = "#3498db", alpha = 0.15) +
  geom_ribbon(data = plot_data, aes(x = BMI, ymin = CI_lower, ymax = CI_upper),
              fill = "#e74c3c", alpha = 0.25) +
  geom_line(data = plot_data, aes(x = BMI, y = Fit), color = "#2c3e50",
            linewidth = 1) +
  geom_point(data = health, aes(x = BMI, y = FBS), alpha = 0.3) +
  annotate("text", x = 37, y = 170, label = "Prediction interval\n(single new patient)",
           color = "#3498db", fontface = "bold", size = 3.5) +
  annotate("text", x = 37, y = 135, label = "Confidence interval\n(population mean)",
           color = "#e74c3c", fontface = "bold", size = 3.5) +
  labs(title = "Confidence Interval vs. Prediction Interval",
       subtitle = "CI = precision of estimated mean; PI = range for a new individual",
       x = "BMI (kg/m²)", y = "FBS (mg/dL)") +
  theme_clean()
```

:::{.callout-note title="CI vs. PI"}
- **Confidence interval** (narrow, red): Where we expect the *average* FBS to lie for all people with a given BMI. It reflects uncertainty in the *regression line itself*.
- **Prediction interval** (wide, blue): Where we expect a *single new patient's* FBS to lie. It includes both line uncertainty *and* individual variation.

Clinical prediction models report PIs. Research papers typically report CIs.
:::

---

## Section 3: Regression Assumptions and Diagnostics

The four key assumptions (**LINE**):

1. **L**inearity — the relationship between x and y is linear
2. **I**ndependence — observations are independent
3. **N**ormality — residuals are approximately normally distributed
4. **E**qual variance (homoscedasticity) — spread of residuals is constant across fitted values

### 3.1 Diagnostic Plots

```{r}
#| label: diagnostic-plots
#| echo: true
#| fig-width: 10
#| fig-height: 8

par(mfrow = c(2, 2))
plot(model)
```

**How to read these:**

- **Residuals vs. Fitted:** Should be a random cloud. Patterns (curves, funnels) indicate non-linearity or heteroscedasticity.
- **Q-Q plot:** Points should follow the diagonal. Deviations at the tails suggest non-normal residuals.
- **Scale-Location:** Should show a roughly flat red line. An upward trend suggests increasing variance.
- **Residuals vs. Leverage:** Points in the upper/lower right (high leverage, large residual) may be influential — check Cook's distance > 0.5.

---

### 3.2 What to Do When Assumptions Fail

```{r}
#| label: assumptions-table
#| echo: false

assumptions_df <- tibble(
  Problem = c("Non-linear relationship",
              "Non-normal residuals",
              "Heteroscedasticity (funnel shape)",
              "Influential outliers"),
  Detection = c("Curved pattern in residuals vs. fitted",
                "Deviation from diagonal on Q-Q plot",
                "Spread increases with fitted values",
                "Cook's distance > 0.5 or 1"),
  Solution = c("Transform x or y (log, sqrt); add polynomial term",
               "Transform y; use robust regression; often OK if n > 30",
               "Transform y (log); use weighted least squares; robust SE",
               "Investigate data entry errors; report results with and without")
)

kable(assumptions_df, format = "html") %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("bordered", "hover")) %>%
  row_spec(0, bold = TRUE, background = "#1e3a5f", color = "white") %>%
  column_spec(1, bold = TRUE)
```

---

## Section 4: Multiple Regression

### 4.1 Why Adjust for Confounders?

In our JIPMER data, BMI and FBS are associated — but **age** also influences FBS. Is the BMI–FBS association genuine, or is it partly (or entirely) confounded by age?

**Multiple regression** lets us ask: *"What is the effect of BMI on FBS, holding age and sex constant?"*

$$\text{FBS} = \beta_0 + \beta_1 \times \text{BMI} + \beta_2 \times \text{Age} + \beta_3 \times \text{Sex} + \varepsilon$$

**In words:** Each coefficient tells you the effect of that variable *while all others are held fixed*.

---

### 4.2 Fitting the Multiple Regression

```{r}
#| label: multiple-regression
#| echo: true

model_multi <- lm(FBS ~ BMI + Age + Sex, data = health)
summary(model_multi)
```

```{r}
#| label: compare-models
#| echo: true

coefs_simple <- tidy(model, conf.int = TRUE)
coefs_multi  <- tidy(model_multi, conf.int = TRUE)

comparison <- tibble(
  Variable = c("BMI (simple)", "BMI (adjusted)"),
  Estimate = c(coefs_simple$estimate[2], coefs_multi$estimate[2]),
  `95% CI` = c(
    paste0(round(coefs_simple$conf.low[2], 2), " to ", round(coefs_simple$conf.high[2], 2)),
    paste0(round(coefs_multi$conf.low[2], 2), " to ", round(coefs_multi$conf.high[2], 2))
  ),
  p = c(format(coefs_simple$p.value[2], digits = 3),
        format(coefs_multi$p.value[2], digits = 3))
)

kable(comparison, format = "html") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("bordered"))
```

:::{.callout-tip title="Reading Multiple Regression Output"}
- The BMI coefficient in the adjusted model tells you the effect of BMI *independent of* age and sex.
- If the coefficient changes substantially after adjustment, **confounding was present**.
- **Adjusted R²** penalises for the number of predictors and is more honest than raw R² for comparing models.
:::

---

### 4.3 Adjusted R² and Model Comparison

```{r}
#| label: adjusted-r2
#| echo: true

cat("Simple model R²:", round(summary(model)$r.squared, 4),
    "| Adj R²:", round(summary(model)$adj.r.squared, 4), "\n")
cat("Multiple model R²:", round(summary(model_multi)$r.squared, 4),
    "| Adj R²:", round(summary(model_multi)$adj.r.squared, 4), "\n")
cat("\nAdding age and sex explains an additional",
    round((summary(model_multi)$r.squared - summary(model)$r.squared) * 100, 1),
    "% of FBS variance.\n")
```

:::{.callout-warning title="Overfitting"}
Adding more predictors always increases R² — even if the new variables are noise. **Adjusted R²** corrects for this. If adjusted R² drops when you add a variable, that variable is not improving the model.
:::

---

## Section 5: Logistic Regression

### 5.1 When the Outcome Is Binary

What if the research question is: *"Does higher BMI predict diabetes (yes/no)?"*

FBS is continuous — but diabetes status is **binary**. Linear regression can't model binary outcomes properly (it predicts values outside 0–1). We need **logistic regression**.

**In words:** Logistic regression models the **log-odds** of the outcome:

$$\ln\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x$$

where $p$ = probability of the event (diabetes = yes).

The **odds ratio** for a 1-unit increase in x is:

$$OR = e^{\beta_1}$$

---

### 5.2 Generating Binary Outcome Data

```{r}
#| label: logistic-data
#| echo: true

# Create diabetes outcome (higher BMI and age → higher probability)
health$diabetes_prob <- plogis(-8 + 0.15 * health$BMI + 0.05 * health$Age)
health$Diabetes <- rbinom(n, 1, health$diabetes_prob)
health$Diabetes_label <- ifelse(health$Diabetes == 1, "Diabetic", "Non-diabetic")

cat("Diabetes prevalence:", round(mean(health$Diabetes) * 100, 1), "%\n")
```

---

### 5.3 Fitting the Logistic Regression

```{r}
#| label: logistic-fit
#| echo: true

logit_model <- glm(Diabetes ~ BMI + Age + Sex, data = health, family = binomial)
summary(logit_model)
```

```{r}
#| label: logistic-or
#| echo: true

logit_coefs <- tidy(logit_model, conf.int = TRUE, exponentiate = TRUE)

or_table <- logit_coefs %>%
  filter(term != "(Intercept)") %>%
  select(Variable = term, OR = estimate, CI_low = conf.low, CI_high = conf.high,
         p = p.value) %>%
  mutate(across(c(OR, CI_low, CI_high), ~ round(., 3)),
         p = format(p, digits = 3))

kable(or_table, col.names = c("Variable", "OR", "95% CI Lower", "95% CI Upper", "p-value"),
      format = "html") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("bordered", "hover"))
```

:::{.callout-note title="Interpreting Logistic Regression ORs"}
- **OR for BMI:** For each 1 kg/m² increase in BMI, the odds of diabetes change by a factor of `r round(logit_coefs$estimate[logit_coefs$term == "BMI"], 2)` (holding age and sex constant).
- **OR > 1:** Higher BMI increases the odds of diabetes.
- **OR < 1:** The variable is protective.
- **95% CI excluding 1.0:** Statistically significant.

This connects directly to the Odds Ratio from Module 10 — logistic regression is the **multivariable extension** of the 2×2 table OR.
:::

---

### 5.4 Predicted Probabilities

```{r}
#| label: predicted-probs
#| echo: true
#| fig-width: 9
#| fig-height: 6

bmi_seq <- data.frame(BMI = seq(16, 40, by = 0.5), Age = 45, Sex = "Male")
bmi_seq$pred_prob <- predict(logit_model, newdata = bmi_seq, type = "response")

ggplot(bmi_seq, aes(x = BMI, y = pred_prob)) +
  geom_line(color = "#e74c3c", linewidth = 1.2) +
  geom_rug(data = health %>% filter(Diabetes == 1), aes(x = BMI, y = NULL),
           sides = "t", alpha = 0.3, color = "#e74c3c") +
  geom_rug(data = health %>% filter(Diabetes == 0), aes(x = BMI, y = NULL),
           sides = "b", alpha = 0.3, color = "#2ecc71") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  annotate("text", x = 37, y = 0.52, label = "50% threshold",
           color = "gray50", fontface = "italic") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(title = "Predicted Probability of Diabetes by BMI",
       subtitle = "Logistic regression (Age = 45, Male) | Rug marks = actual cases",
       x = "BMI (kg/m²)", y = "Predicted probability of diabetes") +
  theme_clean()
```

The S-shaped (sigmoid) curve is the hallmark of logistic regression. The probability is bounded between 0 and 1 — unlike linear regression, which would predict impossible values.

---

## Section 6: The Correlation–Causation–Regression Framework

:::{.callout-important title="Correlation → Regression → Causation?"}
| Tool | What it tells you | What it does NOT tell you |
|------|-------------------|--------------------------|
| Correlation (r) | Strength and direction of linear association | Whether x causes y |
| Simple regression | How y changes per unit of x; prediction | Whether the relationship is causal |
| Multiple regression | Effect of x *adjusted for confounders* | Whether all confounders are measured |
| Logistic regression | OR for binary outcomes, adjusted | Causation without proper study design |

**Regression adjusts for measured confounders** — but unmeasured confounders can still bias results. Only a well-designed **RCT** (Module 7) can establish causation. Regression in observational data provides *adjusted associations*, not proof of causation.
:::

---

## Section 7: Decision Framework

```{r}
#| label: decision-table
#| echo: false

decision <- tibble(
  Scenario = c(
    "Two continuous variables — strength of association?",
    "Ordinal data or outliers present?",
    "Predict continuous outcome from one predictor?",
    "Predict continuous outcome, adjust for confounders?",
    "Predict binary outcome (yes/no)?",
    "Predict time-to-event (survival)?"
  ),
  Method = c(
    "Pearson correlation (r)",
    "Spearman rank correlation (rₛ)",
    "Simple linear regression",
    "Multiple linear regression",
    "Logistic regression",
    "Cox regression (Module 12)"
  ),
  Output = c(
    "r, p-value, 95% CI",
    "rₛ, p-value",
    "Slope, R², equation, CI/PI",
    "Adjusted slopes, adjusted R²",
    "Odds ratios, predicted probabilities",
    "Hazard ratios"
  )
)

kable(decision, format = "html") %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("bordered", "hover")) %>%
  row_spec(0, bold = TRUE, background = "#1e3a5f", color = "white") %>%
  column_spec(1, bold = TRUE, width = "40%")
```

---

## Summary

:::{.callout-tip title="Key Takeaways"}
1. **Pearson r** measures linear association (−1 to +1). **Spearman rₛ** is the rank-based alternative for ordinal or skewed data.

2. **Correlation ≠ causation.** Confounding, reverse causality, and selection bias can create spurious associations.

3. **Simple linear regression** fits y = β₀ + β₁x. The slope is the change in y per 1-unit increase in x. R² = fraction of variance explained.

4. **Always check LINE assumptions** — linearity, independence, normality of residuals, equal variance. Use diagnostic plots.

5. **Confidence intervals** (for the mean) are narrower than **prediction intervals** (for a new individual). Always report uncertainty.

6. **Multiple regression** adjusts for confounders: "effect of x₁ holding x₂, x₃, ... constant." Use adjusted R² for model comparison.

7. **Logistic regression** models binary outcomes. Coefficients exponentiate to **odds ratios** — connecting directly to Module 10.

8. **No regression model proves causation** in observational data. Regression adjusts for *measured* confounders; RCTs are needed for causal claims.
:::

---

## Practice Questions

### Q1: Correlation vs Causation

```{r}
#| label: mcq1
#| echo: false
#| results: asis

make_mcq(
  "m11q1",
  "A study finds r = 0.85 between hours of TV watched per day and BMI in 500 adults. Which interpretation is MOST appropriate?",
  c(
    "Watching more TV causes weight gain" = "Incorrect. Correlation does not establish causation — confounders like sedentary lifestyle, diet, and socioeconomic status may explain the association.",
    "answer:There is a strong positive linear association between TV hours and BMI" = "Correct! r = 0.85 indicates a strong positive linear association. We cannot infer causation from correlation alone.",
    "85% of the variation in BMI is explained by TV watching" = "Incorrect. R² (not r) gives the proportion of variance explained. R² = 0.85² = 0.72, meaning 72% — but this still doesn't prove causation.",
    "The relationship is statistically significant but clinically irrelevant" = "Incorrect. r = 0.85 is both statistically and practically strong. Clinical relevance cannot be dismissed without context."
  )
)
```

---

### Q2: Regression Prediction

```{r}
#| label: mcq2
#| echo: false
#| results: asis

make_mcq(
  "m11q2",
  "A regression model gives: SBP = 90 + 0.6 × Age (years). For a 60-year-old patient, what is the predicted systolic BP?",
  c(
    "90 mmHg" = "Incorrect. That is the intercept only — you need to add the age contribution: 0.6 × 60 = 36.",
    "answer:126 mmHg" = "Correct! SBP = 90 + 0.6 × 60 = 90 + 36 = 126 mmHg.",
    "150 mmHg" = "Incorrect. Check the arithmetic: 90 + (0.6 × 60) = 126, not 150.",
    "36 mmHg" = "Incorrect. 36 is the age contribution (0.6 × 60), but you must add the intercept (90)."
  )
)
```

---

### Q3: R-Squared Interpretation

```{r}
#| label: mcq3
#| echo: false
#| results: asis

make_mcq(
  "m11q3",
  "In a simple linear regression, R² = 0.64. Which statement is correct?",
  c(
    "The correlation coefficient r = 0.64" = "Incorrect. R² = r², so r = √0.64 = 0.80 (assuming positive association).",
    "answer:The predictor explains 64% of the variation in the outcome" = "Correct! R² = 0.64 means 64% of outcome variance is explained by the model. The remaining 36% is unexplained.",
    "The model correctly classifies 64% of observations" = "Incorrect. R² is not a classification metric — it measures proportion of variance explained.",
    "The slope of the regression line is 0.64" = "Incorrect. R² and the slope are different quantities. The slope depends on the units of x and y."
  )
)
```

---

### Q4: Logistic Regression OR

```{r}
#| label: mcq4
#| echo: false
#| results: asis

make_mcq(
  "m11q4",
  "A logistic regression for diabetes (yes/no) gives OR = 1.12 (95% CI: 1.04–1.21) for BMI. What is the correct interpretation?",
  c(
    "For each 1 kg/m² increase in BMI, the probability of diabetes increases by 12%" = "Incorrect. OR is not the same as probability increase. The OR says the odds multiply by 1.12 per unit BMI increase.",
    "answer:For each 1 kg/m² increase in BMI, the odds of diabetes increase by 12%, and this is statistically significant" = "Correct! OR = 1.12 means the odds multiply by 1.12 per unit BMI increase (12% increase). The CI excludes 1.0, so p < 0.05.",
    "BMI causes diabetes with 12% certainty" = "Incorrect. An OR does not express certainty of causation. It quantifies the association between BMI and odds of diabetes.",
    "The model explains 12% of the variance in diabetes" = "Incorrect. OR is not R². The 1.12 refers to the odds ratio per unit change in the predictor."
  )
)
```

---

### Q5: Regression Assumptions

```{r}
#| label: mcq5
#| echo: false
#| results: asis

make_mcq(
  "m11q5",
  "Which of the following is NOT an assumption of simple linear regression?",
  c(
    "Linearity of the relationship between x and y" = "Incorrect — this IS a required assumption (the L in LINE).",
    "Independence of observations" = "Incorrect — this IS a required assumption (the I in LINE).",
    "answer:The predictor variable x must be normally distributed" = "Correct! Regression assumes normality of *residuals*, not of x. The predictor can have any distribution.",
    "Homoscedasticity (constant variance of residuals)" = "Incorrect — this IS a required assumption (the E in LINE)."
  )
)
```

---

### Q6: Confounding in Multiple Regression

```{r}
#| label: mcq6
#| echo: false
#| results: asis

make_mcq(
  "m11q6",
  "A multiple regression shows that the association between BMI and FBS drops from β = 2.1 (simple) to β = 1.4 (after adjusting for age and sex). What does this suggest?",
  c(
    "The original association was entirely spurious" = "Incorrect. The adjusted coefficient is still 1.4 and presumably still significant — the association remains, just weaker.",
    "Age and sex are mediators in the causal pathway" = "Incorrect. Without a causal diagram, we cannot distinguish confounding from mediation based on coefficient change alone.",
    "answer:Age and/or sex were confounders — they partly explained the BMI–FBS association" = "Correct! The coefficient dropped by ~33% after adjustment, suggesting age/sex confounded part (but not all) of the association.",
    "The multiple regression model is invalid" = "Incorrect. A coefficient change after adjustment is expected and informative — it means confounding was present and has been partially addressed."
  )
)
```

---

## Appendix: R Functions Quick Reference

```{r}
#| label: appendix
#| echo: true
#| eval: false

# ======================================
# QUICK R REFERENCE — CORRELATION & REGRESSION
# ======================================

# Pearson correlation
cor(x, y)
cor.test(x, y)  # with p-value and 95% CI

# Spearman correlation
cor(x, y, method = "spearman")

# Simple linear regression
model <- lm(y ~ x, data = df)
summary(model)                       # coefficients, R², p-values
confint(model)                       # 95% CI for coefficients
predict(model, newdata, interval = "confidence")  # CI for mean
predict(model, newdata, interval = "prediction")  # PI for individual

# Diagnostic plots
par(mfrow = c(2,2)); plot(model)

# Multiple regression
model_multi <- lm(y ~ x1 + x2 + x3, data = df)
summary(model_multi)

# Logistic regression
logit <- glm(y ~ x1 + x2, data = df, family = binomial)
summary(logit)
exp(coef(logit))          # odds ratios
exp(confint(logit))       # 95% CI for ORs
predict(logit, type = "response")  # predicted probabilities

# Compare models
anova(model_simple, model_multi)
AIC(model_simple, model_multi)
```

---

**End of Module 11**

*AIIMS Bhopal Biostatistics Course | Updated: `r Sys.Date()`*
