---
title: "Meta-Analysis and Evidence Synthesis"
subtitle: "Combining Evidence from Multiple Studies"
description: |
  Learn to synthesize evidence across studies using systematic review and
  meta-analysis. Understand forest plots, fixed vs random effects models,
  heterogeneity (I²), funnel plots, and publication bias. Real examples
  from AIIMS Bhopal research on TB, sleep apnoea, and sepsis.
categories:
  - Biostatistics
  - Meta-Analysis
  - Systematic Review
  - Evidence Synthesis
  - Heterogeneity
author: "AIIMS Bhopal Biostatistics Course"
date: "`r Sys.Date()`"
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(kableExtra)
source("../R/mcq.R")
library(patchwork)
library(scales)

set.seed(42)

theme_clean <- function() {
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
}
```

::: {.callout-note appearance="minimal"}
**Lecture slides for this module:** [Open Slides](../slides/13-meta-analysis-slides.html){target="_blank"}
:::

## Learning Objectives {#learning-objectives}

By the end of this module, you will be able to:

1. **Distinguish** systematic review from meta-analysis and explain why both matter
2. **Read and interpret** forest plots — point estimates, confidence intervals, weights, and pooled effects
3. **Choose** between fixed-effect and random-effects models based on heterogeneity
4. **Calculate and interpret** I² and explain what drives heterogeneity
5. **Assess** publication bias using funnel plots and Egger's test
6. **Evaluate** the quality of evidence using the GRADE framework
7. **Critically appraise** meta-analyses using the PRISMA checklist
8. **Recognise** when meta-analysis is inappropriate

---

## Clinical Hook

In 2006, **Rajnish Joshi** (now at AIIMS Bhopal) led a landmark systematic review in *PLoS Medicine*: **"Tuberculosis among Health-Care Workers in Low- and Middle-Income Countries"** (Joshi R, Reingold AL, Menzies D, Pai M).

The problem was clear: individual studies of TB risk in healthcare workers gave wildly different numbers — some reported LTBI prevalence of 33%, others 79%. Some found healthcare workers at 2× the risk of the general population, others at 5×. If you read just one study, you'd get one version of the truth. If you read another, you'd get a completely different picture.

The team identified **42 articles** comprising **51 studies** and synthesised the evidence. The pooled LTBI prevalence among healthcare workers was **54%** (range 33–79%). The systematic review revealed not just the average risk, but *why* estimates varied — differences in TB burden, infection control measures, healthcare setting, and diagnostic method.

This is the power of meta-analysis: **turning conflicting signals into a coherent picture**.

::: {.callout-note title="Reference"}
Joshi R, Reingold AL, Menzies D, Pai M. Tuberculosis among Health-Care Workers in Low- and Middle-Income Countries: A Systematic Review. *PLoS Med* 2006;3(12):e494.
:::

---

## Part 1: Systematic Review vs Meta-Analysis

### What's the Difference?

| | Systematic Review | Meta-Analysis |
|---|---|---|
| **What** | A structured method for finding, appraising, and synthesising *all* available evidence | A statistical technique for combining results from multiple studies into a single pooled estimate |
| **Nature** | Qualitative (can exist without statistics) | Quantitative (requires numbers) |
| **Can it stand alone?** | Yes — a systematic review without meta-analysis is valid | No — a meta-analysis without systematic review is not credible |
| **Output** | Summary of evidence, quality assessment | Pooled effect size, forest plot, heterogeneity statistics |

::: {.callout-important title="The Hierarchy"}
A meta-analysis is only as good as the systematic review that feeds it. If the search missed studies, or the included studies are biased, the pooled estimate inherits those problems. **Garbage in, garbage out.**
:::

### The PICO Framework

Every systematic review starts with a focused question structured as PICO:

| Component | Meaning | Example (from Joshi et al.) |
|---|---|---|
| **P**opulation | Who? | Healthcare workers in LMICs |
| **I**ntervention/Exposure | What? | Working in healthcare settings |
| **C**omparator | Compared to? | General population |
| **O**utcome | What do we measure? | Prevalence of LTBI and active TB |

---

## Part 2: The Forest Plot — The Most Important Figure in Evidence-Based Medicine

### Anatomy of a Forest Plot

```{r}
#| label: forest-plot-teaching
#| fig-height: 7
#| fig-width: 10

# Simulated data inspired by TB in healthcare workers
# These are teaching data, not from the actual Joshi et al. paper
set.seed(42)
studies <- tibble(
  study = c("Pai 2005 (India)", "Yanai 2003 (Thailand)", "Joshi 2006 (Ethiopia)",
            "Kassim 2000 (Malawi)", "Balt 1998 (Russia)", "Rao 2009 (India)",
            "Naidoo 2006 (S. Africa)", "Dimitrova 2006 (Russia)"),
  or = c(2.4, 3.1, 1.8, 5.2, 2.9, 1.5, 4.1, 2.2),
  or_lo = c(1.5, 1.9, 1.1, 2.4, 1.6, 0.9, 2.0, 1.3),
  or_hi = c(3.8, 5.0, 3.0, 11.3, 5.3, 2.5, 8.4, 3.7),
  weight = c(16, 14, 15, 6, 10, 18, 8, 13),
  n = c(720, 380, 520, 110, 240, 850, 190, 310)
)

# Pooled estimate (random effects)
pooled_or <- 2.6
pooled_lo <- 1.9
pooled_hi <- 3.5

# Plot
studies <- studies %>% mutate(study = fct_rev(fct_inorder(study)))

ggplot(studies, aes(y = study)) +
  # Individual studies
  geom_point(aes(x = or, size = weight), color = "#2c3e50", shape = 15) +
  geom_errorbarh(aes(xmin = or_lo, xmax = or_hi), height = 0.2, color = "#2c3e50") +
  # Pooled diamond
  annotate("polygon",
           x = c(pooled_lo, pooled_or, pooled_hi, pooled_or),
           y = c(0.3, 0.55, 0.3, 0.05),
           fill = "#e74c3c", alpha = 0.7) +
  annotate("text", x = pooled_or, y = -0.3, label = paste0("Pooled OR = ", pooled_or,
           " (", pooled_lo, "–", pooled_hi, ")"),
           fontface = "bold", size = 3.5, color = "#e74c3c") +
  # Null line
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  # Labels
  scale_x_log10(breaks = c(0.5, 1, 2, 4, 8, 12)) +
  scale_size_continuous(range = c(2, 8), guide = "none") +
  labs(title = "Forest Plot: TB Risk in Healthcare Workers vs General Population",
       subtitle = "Simulated teaching data inspired by Joshi R et al. (PLoS Med, 2006)",
       x = "Odds Ratio (log scale)", y = NULL) +
  annotate("text", x = 0.6, y = 8.5, label = "← Favours HCW\n   (lower risk)", size = 3, color = "gray50") +
  annotate("text", x = 6, y = 8.5, label = "Favours general pop →\n   (HCW at higher risk)", size = 3, color = "gray50") +
  theme_clean() +
  theme(panel.grid.major.y = element_blank())
```

### Reading the Forest Plot

**Each row** = one study:

- **Square** = point estimate (odds ratio, risk ratio, or mean difference)
- **Horizontal line** = 95% confidence interval
- **Square size** = weight of the study (larger studies contribute more)

**Bottom diamond** = pooled estimate across all studies:

- **Centre** = pooled effect size
- **Width** = 95% CI of the pooled estimate
- If the diamond doesn't cross the **vertical null line** (OR = 1), the pooled effect is statistically significant

**X-axis** — Usually on a **log scale** for ratios (OR, RR, HR). This is important because:

- On a log scale, OR = 0.5 and OR = 2.0 are equidistant from 1.0
- The CI is symmetric on the log scale, even though it's asymmetric on the natural scale

---

## Part 3: Fixed-Effect vs Random-Effects Models

### The Key Question: Is There One True Effect?

| | Fixed-Effect | Random-Effects |
|---|---|---|
| **Assumption** | All studies estimate the **same** true effect | Each study estimates a **different** true effect drawn from a distribution |
| **Variation between studies** | Due to sampling error only | Due to sampling error **+ real differences** between studies |
| **CI width** | Narrower | Wider (more conservative) |
| **Weighting** | Proportional to study size (inverse variance) | More balanced — small studies get relatively more weight |
| **Use when** | Studies are very similar; I² < 25% | Studies are diverse; I² ≥ 50% |

### Visualising the Difference

```{r}
#| label: fe-vs-re
#| fig-height: 5.5
#| fig-width: 10

# Same data, two pooled estimates
fe_or <- 2.3; fe_lo <- 1.8; fe_hi <- 2.9
re_or <- 2.6; re_lo <- 1.9; re_hi <- 3.5

comp_df <- tibble(
  model = c("Fixed-effect", "Random-effects"),
  or = c(fe_or, re_or),
  lo = c(fe_lo, re_lo),
  hi = c(fe_hi, re_hi)
) %>% mutate(model = fct_rev(fct_inorder(model)))

ggplot(comp_df, aes(y = model)) +
  geom_point(aes(x = or), size = 6, shape = 18, color = c("#e74c3c", "#2c3e50")) +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.15,
                 color = c("#e74c3c", "#2c3e50"), linewidth = 1.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  scale_x_continuous(limits = c(0.5, 5), breaks = 1:5) +
  labs(title = "Fixed-Effect vs Random-Effects: Same Data, Different Conclusions",
       subtitle = "Random-effects model gives wider CI — more conservative, accounts for between-study variation",
       x = "Pooled Odds Ratio", y = NULL) +
  theme_clean() +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 12, face = "bold"))
```

::: {.callout-tip title="Practical Rule"}
When in doubt, **use random-effects**. It's the safer default because it accounts for between-study heterogeneity. If heterogeneity is truly zero, random-effects and fixed-effect give identical results anyway.
:::

---

## Part 4: Heterogeneity — Why Studies Disagree

### The I² Statistic

Heterogeneity measures how much the effect sizes vary **beyond what chance alone would explain**.

$$I^2 = \frac{Q - df}{Q} \times 100\%$$

where Q is Cochran's chi-squared test statistic and df = number of studies − 1.

| I² value | Interpretation |
|:---:|---|
| 0–25% | Low heterogeneity — studies agree well |
| 25–50% | Moderate — some differences |
| 50–75% | Substantial — results vary considerably |
| > 75% | Considerable — studies may be too different to pool |

### A Real Example: OSA Prevalence in India

A systematic review and meta-analysis of **obstructive sleep apnoea (OSA) prevalence** in Indian adults pooled 8 studies comprising 11,009 subjects. The results:

- Pooled OSA prevalence (AHI ≥ 5): **11%** (95% CI: 7–15%)
- **I² = 98%** — massive heterogeneity

Meanwhile, the **BLESS cohort** from AIIMS Bhopal (Pakhare, Joshi A et al., 2024) — a single well-conducted study using Level I polysomnography on 958 participants — found moderate-severe OSA prevalence of **30.5%**.

Why the enormous difference?

```{r}
#| label: heterogeneity-osa
#| fig-height: 5.5
#| fig-width: 10

# Simulated data inspired by OSA meta-analysis
osa_studies <- tibble(
  study = c("Sharma 2006", "Udwadia 2004", "Reddy 2009", "Puri 2015",
            "Kale 2018", "Suri 2008", "Bhatia 2017", "BLESS 2024*"),
  prev = c(4.9, 7.5, 3.6, 13.7, 9.3, 19.5, 8.2, 30.5),
  lo = c(2.1, 4.2, 1.8, 9.1, 5.8, 12.3, 5.0, 27.6),
  hi = c(7.7, 10.8, 5.4, 18.3, 12.8, 26.7, 11.4, 33.4),
  method = c("Questionnaire", "PSG (clinic)", "Questionnaire", "PSG (clinic)",
             "Questionnaire", "PSG (clinic)", "Questionnaire", "PSG (population)"),
  n = c(2100, 658, 1800, 420, 1500, 315, 2200, 958)
) %>% mutate(study = fct_rev(fct_inorder(study)))

ggplot(osa_studies, aes(y = study)) +
  geom_point(aes(x = prev, color = method, size = n), shape = 15) +
  geom_errorbarh(aes(xmin = lo, xmax = hi, color = method), height = 0.2) +
  scale_color_manual(values = c("Questionnaire" = "#7f8c8d", "PSG (clinic)" = "#3498db",
                                "PSG (population)" = "#e74c3c")) +
  scale_size_continuous(range = c(3, 7), guide = "none") +
  labs(title = "OSA Prevalence Across Indian Studies — Why I² = 98%",
       subtitle = "Simulated teaching data inspired by published meta-analysis and BLESS cohort",
       x = "Prevalence (%)", y = NULL, color = "Diagnostic method") +
  theme_clean() +
  theme(panel.grid.major.y = element_blank(), legend.position = "right")
```

**Sources of heterogeneity:**

1. **Diagnostic method** — Questionnaire-based studies find ~5–9%; PSG-based studies find 13–30%
2. **Population** — Clinic-based vs community-based sampling
3. **Definition** — AHI ≥ 5 (any OSA) vs AHI ≥ 15 (moderate-severe)
4. **Age and BMI** — Older, heavier populations have more OSA
5. **Scoring rules** — AASM 2007 vs AASM 2012 criteria

::: {.callout-note title="Reference"}
Pakhare A, Joshi A, et al. Prevalence and association analysis of obstructive sleep apnea in India: Results from BLESS cohort. *Sleep Med* 2024.
:::

::: {.callout-warning title="When I² Is Very High"}
When I² exceeds 75%, seriously question whether pooling is meaningful. A single "average" prevalence of 11% obscures the reality that different methods give vastly different answers. **Subgroup analysis** (by diagnostic method, population type) is far more informative than a single pooled number.
:::

---

## Part 5: Exploring Heterogeneity — Subgroup Analysis and Meta-Regression

### Subgroup Analysis

When I² is high, stratify by a study-level characteristic to see if heterogeneity is explained.

```{r}
#| label: subgroup-analysis
#| results: asis

subgroup <- tibble(
  Subgroup = c("Questionnaire-based studies (n=4)", "PSG-based studies (n=4)"),
  `Pooled prevalence` = c("6.5%", "17.8%"),
  `95% CI` = c("4.2–8.8%", "10.1–25.5%"),
  `I² within subgroup` = c("45%", "62%")
)

kable(subgroup, format = "html", align = c("l", "c", "c", "c")) %>%
  kable_styling(full_width = FALSE, font_size = 14) %>%
  row_spec(0, background = "#2c3e50", color = "white", bold = TRUE)
```

The diagnostic method explains much of the heterogeneity: questionnaire studies consistently give lower estimates than PSG studies. This is **clinically meaningful** — it tells us that the tool we use to measure OSA dramatically affects what we find.

### Meta-Regression

Meta-regression extends this idea to **continuous study-level variables**. For example: does mean BMI of the study population predict the effect size?

$$\theta_i = \beta_0 + \beta_1 \times \text{mean BMI}_i + \epsilon_i$$

::: {.callout-warning title="Important Caveats"}
- Subgroup analyses should be **pre-specified** in the protocol, not discovered post-hoc
- With only 8 studies, you have very limited power for meta-regression
- **Ecological fallacy** — a study-level association (higher mean BMI → higher prevalence) does not prove the association holds at the individual level
:::

---

## Part 6: Publication Bias — The Missing Studies Problem

### Why It Matters

Studies with statistically significant, positive results are more likely to be published. Studies with null findings end up in file drawers. If your meta-analysis only includes published studies, the pooled estimate is **biased upward**.

### The Funnel Plot

A funnel plot shows **effect size** (x-axis) vs **study precision** (y-axis, often standard error or 1/SE):

```{r}
#| label: funnel-plot
#| fig-height: 6
#| fig-width: 10

# Simulated funnel plot data
set.seed(42)
# Symmetric (no bias)
n_studies <- 20
true_effect <- log(2.5)  # log OR
se_sym <- runif(n_studies, 0.15, 0.8)
effect_sym <- rnorm(n_studies, true_effect, se_sym)

funnel_sym <- tibble(effect = effect_sym, se = se_sym, type = "Symmetric (No bias)")

# Asymmetric (publication bias)
se_asym <- runif(n_studies, 0.15, 0.8)
effect_asym <- rnorm(n_studies, true_effect, se_asym)
# Remove small studies near null
keep <- !(se_asym > 0.5 & effect_asym < log(1.5))
funnel_asym <- tibble(effect = effect_asym[keep], se = se_asym[keep], type = "Asymmetric (Publication bias)")

funnel_data <- bind_rows(funnel_sym, funnel_asym)

ggplot(funnel_data, aes(x = exp(effect), y = se)) +
  geom_point(size = 3, color = "#2c3e50", alpha = 0.7) +
  geom_vline(xintercept = exp(true_effect), linetype = "dashed", color = "#e74c3c") +
  scale_y_reverse() +
  scale_x_log10() +
  facet_wrap(~type) +
  labs(title = "Funnel Plots: Detecting Publication Bias",
       subtitle = "Left: symmetric (no bias) | Right: asymmetric (small null studies missing)",
       x = "Odds Ratio (log scale)", y = "Standard Error (inverted)") +
  theme_clean() +
  theme(strip.text = element_text(face = "bold", size = 12))
```

**How to read it:**

- **Symmetric funnel** → no obvious publication bias; small studies scatter equally on both sides of the pooled estimate
- **Asymmetric funnel** (gap in the bottom-left) → small studies with null results are missing — suggests publication bias
- **Egger's test** provides a statistical test for asymmetry (p < 0.05 suggests bias), but has low power with fewer than 10 studies

### Beyond Publication Bias

Funnel plot asymmetry doesn't always mean publication bias. Other causes include:

- **True heterogeneity** — small studies done in different populations
- **Methodological quality** — small studies may be lower quality with inflated effects
- **Chance** — with few studies, asymmetry can occur randomly

---

## Part 7: Risk of Bias in Individual Studies

### Cochrane Risk of Bias Tool (RoB 2)

For RCTs, the Cochrane RoB 2 tool assesses five domains:

```{r}
#| label: rob-table
#| results: asis

rob <- tibble(
  Domain = c("Randomisation process", "Deviations from intended intervention",
             "Missing outcome data", "Measurement of outcome",
             "Selection of reported result"),
  `What can go wrong` = c(
    "Inadequate sequence generation or allocation concealment",
    "Unblinded participants or clinicians change behaviour",
    "Differential dropout between groups",
    "Assessor knows group assignment; subjective outcome",
    "Authors report the 'best' outcome or time point"
  ),
  `Example from AIIMS Bhopal` = c(
    "Mw COVID trial: were allocation envelopes opaque?",
    "Open-label: clinicians knew who got Mw",
    "ICU transfers counted as censored or excluded?",
    "Mortality is objective; less prone to bias",
    "Primary vs secondary outcomes reported selectively?"
  )
)

kable(rob, format = "html") %>%
  kable_styling(full_width = TRUE, font_size = 13) %>%
  row_spec(0, background = "#2c3e50", color = "white", bold = TRUE) %>%
  column_spec(1, bold = TRUE, width = "22%")
```

::: {.callout-tip title="Traffic Light System"}
Each domain is rated as **Low risk** (green), **Some concerns** (yellow), or **High risk** (red). A study with high risk in even one critical domain can seriously bias the meta-analysis. Sensitivity analysis excluding high-risk studies tests robustness.
:::

---

## Part 8: The GRADE Framework — Grading the Evidence

GRADE (Grading of Recommendations Assessment, Development and Evaluation) assesses the **overall quality of evidence** across all studies for a given outcome.

### Starting Point and Modifiers

```{r}
#| label: grade-table
#| results: asis

grade <- tibble(
  ` ` = c("Starting quality", "", "Downgrade for:", "", "", "", "",
          "Upgrade for (observational):", "", ""),
  Factor = c("RCTs", "Observational studies",
             "Risk of bias", "Inconsistency (heterogeneity)",
             "Indirectness", "Imprecision", "Publication bias",
             "Large effect", "Dose-response", "Confounders would reduce effect"),
  Rating = c("High (⊕⊕⊕⊕)", "Low (⊕⊕○○)",
             "−1 or −2", "−1 or −2", "−1 or −2", "−1 or −2", "−1 or −2",
             "+1 or +2", "+1", "+1")
)

kable(grade, format = "html", align = c("l", "l", "c")) %>%
  kable_styling(full_width = FALSE, font_size = 13) %>%
  row_spec(0, background = "#2c3e50", color = "white", bold = TRUE)
```

**Final quality ratings:**

| Rating | Meaning |
|---|---|
| **High** (⊕⊕⊕⊕) | Very confident the true effect lies close to the estimate |
| **Moderate** (⊕⊕⊕○) | Moderately confident; further research may change the estimate |
| **Low** (⊕⊕○○) | Limited confidence; further research is very likely to change |
| **Very low** (⊕○○○) | Very little confidence in the estimate |

---

## Part 9: The PRISMA Checklist — Transparent Reporting

The **PRISMA** (Preferred Reporting Items for Systematic Reviews and Meta-Analyses) statement is a 27-item checklist ensuring transparent reporting. When reading or writing a meta-analysis, check:

::::{.callout-important title="PRISMA Essentials for Critical Appraisal"}

1. **Protocol registered?** (e.g., PROSPERO) — reduces selective reporting
2. **Search strategy** described in enough detail to reproduce?
3. **Study selection** — PRISMA flow diagram showing how many identified, screened, excluded (with reasons), and included?
4. **Risk of bias** assessed for each study?
5. **Forest plot** presented with all individual studies?
6. **Heterogeneity** reported (I², Q-test)?
7. **Publication bias** assessed (funnel plot, Egger's test)?
8. **Sensitivity analyses** performed?
9. **GRADE** or equivalent quality assessment?
10. **Funding and conflicts** disclosed?
::::

### The PRISMA Flow Diagram

```{r}
#| label: prisma-flow
#| fig-height: 6
#| fig-width: 10

# Simple PRISMA-like flow
prisma_df <- tibble(
  x = c(5, 5, 5, 5, 5, 9, 9, 9),
  y = c(10, 8, 6, 4, 2, 8, 6, 4),
  label = c(
    "Records identified\n(n = 3,842)",
    "After duplicates removed\n(n = 2,156)",
    "Full-text assessed\n(n = 187)",
    "Studies included in\nqualitative synthesis\n(n = 51)",
    "Studies included in\nmeta-analysis\n(n = 42)",
    "Excluded by title/abstract\n(n = 1,969)",
    "Full-text excluded\n(n = 136)\nReasons: no denominator (58),\nnot LMIC (42), review (36)",
    "Not suitable for pooling\n(n = 9)"
  ),
  box = c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, FALSE)
)

ggplot(prisma_df) +
  # Boxes
  geom_label(data = prisma_df %>% filter(box),
             aes(x = x, y = y, label = label),
             fill = "#d4e4f7", color = "#1e3a5f", size = 3, fontface = "bold",
             label.padding = unit(0.4, "cm")) +
  # Exclusion boxes
  geom_label(data = prisma_df %>% filter(!box),
             aes(x = x, y = y, label = label),
             fill = "#fde8e8", color = "#c0392b", size = 2.8,
             label.padding = unit(0.3, "cm")) +
  # Arrows
  geom_segment(aes(x = 5, xend = 5, y = 9.5, yend = 8.5), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 5, xend = 5, y = 7.5, yend = 6.5), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 5, xend = 5, y = 5.5, yend = 4.7), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 5, xend = 5, y = 3.5, yend = 2.5), arrow = arrow(length = unit(0.2, "cm"))) +
  # Exclusion arrows
  geom_segment(aes(x = 6.2, xend = 8, y = 8, yend = 8), arrow = arrow(length = unit(0.2, "cm")), color = "#c0392b") +
  geom_segment(aes(x = 6.2, xend = 8, y = 6, yend = 6), arrow = arrow(length = unit(0.2, "cm")), color = "#c0392b") +
  geom_segment(aes(x = 6.2, xend = 8, y = 4, yend = 4), arrow = arrow(length = unit(0.2, "cm")), color = "#c0392b") +
  xlim(2, 12) + ylim(1, 11) +
  labs(title = "PRISMA Flow Diagram",
       subtitle = "Simulated example based on Joshi R et al. TB systematic review structure") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, color = "gray40"))
```

---

## Part 10: When NOT to Do a Meta-Analysis

Not all collections of studies should be pooled. Meta-analysis is **inappropriate** when:

::::{.callout-warning title="Red Flags Against Pooling"}

1. **Extreme clinical heterogeneity** — Studies compare fundamentally different interventions, populations, or outcomes (apples and oranges)
2. **Very high I²** with no identifiable source — If you can't explain *why* studies disagree, an average is meaningless
3. **Only 2–3 studies** — Too few to estimate heterogeneity reliably; narrative synthesis is better
4. **High risk of bias in most studies** — Pooling biased estimates gives a precise but wrong answer
5. **Different outcome measures** — Unless you use standardised mean difference (SMD), you can't combine different scales
6. **Contradictory results with plausible reasons** — If Study A (high-dose) shows benefit and Study B (low-dose) shows harm, the average is clinically meaningless
::::

### A Teaching Example: Pralidoxime in Organophosphorus Poisoning

Rajnish Joshi was a co-investigator on a landmark **double-blind placebo-controlled RCT** of pralidoxime (2-PAM) for organophosphorus poisoning in Sri Lanka (Eddleston et al., *PLoS Medicine* 2009, n = 235).

The result shocked the clinical community: **pralidoxime-treated patients had *higher* mortality** (24.8% vs 15.8%; adjusted HR = 1.69, 95% CI 0.88–3.26).

This contradicted earlier observational studies and smaller trials that suggested benefit. Why the discrepancy?

- Earlier studies were **unblinded**, with **selection bias** (sicker patients may have been more likely to receive pralidoxime)
- The RCT was adequately powered, double-blind, and placebo-controlled
- A meta-analysis pooling the RCT with weaker studies would **dilute the RCT's signal** with lower-quality evidence

::: {.callout-important title="Lesson"}
Not all evidence deserves equal weight. A well-conducted RCT can overturn a meta-analysis of biased observational studies. This is why **risk of bias assessment** and **GRADE** are essential — they prevent us from being fooled by a large number of weak studies.
:::

::: {.callout-note title="Reference"}
Eddleston M, Eyer P, Worek F, … Joshi R, … Dawson AH. Pralidoxime in Acute Organophosphorus Insecticide Poisoning—A Randomised Controlled Trial. *PLoS Med* 2009;6(6):e1000104.
:::

---

## Part 11: Sensitivity Analysis — Testing Robustness

A good meta-analysis doesn't just report one result. It asks: **does the conclusion change if we make different assumptions?**

Common sensitivity analyses:

| Analysis | What you change | What it tests |
|---|---|---|
| Exclude high-RoB studies | Drop studies rated "high risk" | Is the effect driven by biased studies? |
| Fixed vs random effects | Switch the model | Does between-study variation matter? |
| Leave-one-out | Remove each study in turn | Is one study driving the result? |
| Restrict to RCTs only | Drop observational studies | Is the effect consistent in stronger designs? |
| Trim-and-fill | Impute missing studies | How much might publication bias affect the result? |

If the pooled estimate **changes substantially** across these analyses, the conclusion is **not robust** — report this transparently.

---

## Part 12: Summary and Key Takeaways

::::{.callout-important}
**Core Concepts:**

| Concept | Key Point |
|---|---|
| **Systematic review** | Structured search, appraisal, and synthesis of all evidence |
| **Meta-analysis** | Statistical pooling of effect sizes from multiple studies |
| **Forest plot** | Visual display: squares (studies), diamond (pooled), line at null |
| **Fixed-effect** | One true effect; variation = sampling error only |
| **Random-effects** | Distribution of true effects; accounts for between-study variation |
| **I²** | % of variation beyond chance; > 50% = substantial heterogeneity |
| **Funnel plot** | Detects publication bias; asymmetry = missing small studies |
| **GRADE** | Rates overall evidence quality (high → very low) |
| **PRISMA** | 27-item checklist for transparent reporting |
::::

::::{.callout-tip}
**Practical rules:**

1. **Systematic review first, meta-analysis second** — never pool without a structured search
2. **Random-effects is the safer default** when heterogeneity is uncertain
3. **I² > 75%** — stop and investigate before pooling; subgroup analysis first
4. **Always check the funnel plot** — even for well-conducted reviews
5. **GRADE the evidence** — a statistically significant pooled effect from low-quality studies is not compelling
6. **Sensitivity analysis is not optional** — test robustness by varying assumptions
7. **A single strong RCT can be more informative** than a meta-analysis of many weak studies
::::

---

## Practice Questions: NEET PG Style MCQs {.neet-practice}

### Q1: Reading a Forest Plot

A meta-analysis forest plot shows 6 studies. The pooled odds ratio diamond is centred at 1.35 with 95% CI 0.98–1.84. The I² is reported as 22%. Which interpretation is correct?

```{r}
#| label: q1-mcq
#| results: asis

make_mcq(
  id = "m13q1",
  question = "Pooled OR = 1.35 (95% CI: 0.98–1.84), I² = 22%. Interpretation?",
  options = c(
    "The treatment significantly increases the odds of the outcome by 35%" = "Wrong — the 95% CI crosses 1.0 (0.98–1.84), so the pooled effect is NOT statistically significant at the 5% level.",
    "answer:There is a trend toward increased odds (35%), but the CI includes 1.0, so the result is not statistically significant. Low heterogeneity suggests studies are reasonably consistent." = "Correct! The CI crosses 1.0, meaning we cannot exclude no effect. I² = 22% indicates low heterogeneity — studies agree on the direction, but the overall evidence is imprecise.",
    "I² = 22% means 22% of studies showed a significant effect" = "Wrong — I² measures the percentage of variation in effect sizes due to heterogeneity, not the percentage of studies with significant results.",
    "The result is invalid because I² should be 0% for a valid meta-analysis" = "Wrong — I² = 0% is ideal but not required. Low heterogeneity (22%) is acceptable and suggests a fixed-effect model could be used."
  )
)
```

### Q2: Fixed vs Random Effects

A meta-analysis of 12 studies on antihypertensive efficacy uses a random-effects model. A colleague argues the fixed-effect model gives a more significant result (narrower CI). Should the authors switch?

```{r}
#| label: q2-mcq
#| results: asis

make_mcq(
  id = "m13q2",
  question = "Random-effects gives wider CI. Colleague says switch to fixed-effect for 'better' results. Should they?",
  options = c(
    "Yes — a more significant result is always better" = "Wrong — choosing the model that gives the 'desired' result is p-hacking. The model should be chosen based on the heterogeneity assessment and pre-specified in the protocol.",
    "Yes — fixed-effect is always the correct model" = "Wrong — fixed-effect assumes all studies estimate the same true effect. With 12 diverse studies, this assumption is unlikely to hold.",
    "answer:No — the model choice should be based on heterogeneity assessment and pre-specified in the protocol, not chosen to achieve significance" = "Correct! Model selection should be driven by the I² statistic and clinical reasoning about whether studies are homogeneous. Switching models to get a 'better' p-value is a form of selective reporting.",
    "No — random-effects is always the correct model" = "Partially right in practice (random-effects is the safer default), but the reasoning is wrong. The choice depends on heterogeneity. If I² = 0%, fixed-effect is perfectly valid."
  )
)
```

### Q3: Heterogeneity

A meta-analysis of OSA prevalence in India reports I² = 98%. The authors pool all studies and report a single prevalence of 11%. What is the main criticism?

```{r}
#| label: q3-mcq
#| results: asis

make_mcq(
  id = "m13q3",
  question = "I² = 98% in OSA prevalence meta-analysis. Authors report pooled prevalence of 11%. Main criticism?",
  options = c(
    "The sample size is too small" = "Wrong — 11,009 subjects across 8 studies is a reasonable sample. The problem is not sample size but heterogeneity.",
    "answer:With I² = 98%, studies are far too heterogeneous to pool meaningfully; a single average prevalence is misleading and subgroup analysis by diagnostic method is essential" = "Correct! I² = 98% means virtually all variation is due to real differences between studies. Pooling gives a number (11%) that doesn't represent any actual population. Subgroup analysis (questionnaire vs PSG, clinic vs community) is far more informative.",
    "They should use fixed-effect instead of random-effects" = "Wrong — switching to fixed-effect with I² = 98% would give an even more misleading pooled estimate with falsely narrow CI.",
    "I² = 98% means 98% of studies are biased" = "Wrong — I² measures heterogeneity (variation in true effects), not bias. Studies can be high quality but still have different true effects because of different methods or populations."
  )
)
```

### Q4: Publication Bias

A funnel plot for a meta-analysis of 15 studies shows a clear gap in the bottom-left corner (small studies with effects near the null are missing). What is the most likely explanation?

```{r}
#| label: q4-mcq
#| results: asis

make_mcq(
  id = "m13q4",
  question = "Funnel plot shows gap in bottom-left (missing small null studies). Most likely explanation?",
  options = c(
    "The meta-analysis included too many studies" = "Wrong — having 15 studies is actually good for funnel plot assessment. The asymmetry suggests a systematic problem with the evidence base.",
    "answer:Publication bias — small studies with null or negative results were less likely to be published" = "Correct! This is the classic signature of publication bias. Small studies with impressive results get published; small studies with null results end up in file drawers. This inflates the pooled estimate.",
    "All small studies are biased" = "Wrong — small studies aren't inherently biased. The issue is that only the small studies with positive results got published, creating a skewed evidence base.",
    "The random-effects model is incorrect" = "Wrong — funnel plot asymmetry reflects a problem with the evidence base, not the statistical model. Switching models won't fix missing studies."
  )
)
```

### Q5: GRADE Assessment

A meta-analysis of 4 RCTs for a new TB drug yields: pooled RR = 0.65 (95% CI: 0.48–0.88), I² = 15%. However, two RCTs are unblinded and have high risk of performance bias. What GRADE rating is most appropriate?

```{r}
#| label: q5-mcq
#| results: asis

make_mcq(
  id = "m13q5",
  question = "4 RCTs, pooled RR = 0.65 (0.48–0.88), I² = 15%. Two RCTs have high risk of bias. GRADE rating?",
  options = c(
    "High — because the CI excludes 1.0 and heterogeneity is low" = "Wrong — GRADE starts at 'high' for RCTs but downgrades for risk of bias. Two of four RCTs having high RoB warrants a downgrade.",
    "answer:Moderate — starts at 'high' (RCTs), downgraded one level for risk of bias in half the evidence" = "Correct! RCTs start at 'high'. With 2/4 studies having serious risk of bias, downgrade by 1 (to 'moderate'). No further downgrade needed: heterogeneity is low (I² = 15%), the CI is reasonably precise, and the outcome is direct.",
    "Low — because some RCTs are biased, we cannot trust any of the evidence" = "Wrong — this over-downgrades. Only half the studies have high RoB, and the overall I² is low. A single downgrade is appropriate.",
    "Very low — any risk of bias in an RCT makes the evidence unreliable" = "Wrong — GRADE is nuanced. A single concern typically warrants a one-level downgrade, not a catastrophic downgrade to 'very low'."
  )
)
```

### Q6: When Not to Pool

An investigator finds 3 studies on vitamin D for preventing respiratory infections: Study A (elderly nursing home, high-dose, RR = 0.58), Study B (children in daycare, low-dose, RR = 1.12), and Study C (adults with COPD, moderate dose, RR = 0.85). She wants to meta-analyse them. What is the best advice?

```{r}
#| label: q6-mcq
#| results: asis

make_mcq(
  id = "m13q6",
  question = "3 studies: different populations (elderly, children, COPD), different doses, different effects. Should they be pooled?",
  options = c(
    "Yes — all three study vitamin D and respiratory infections, so they address the same question" = "Wrong — the populations (elderly, children, COPD patients), doses (high, low, moderate), and settings are fundamentally different. A pooled estimate would be an average of incomparable things.",
    "Yes — but use random-effects to account for the differences" = "Wrong — random-effects accounts for statistical heterogeneity, but cannot fix clinical heterogeneity. If the studies are asking different clinical questions, pooling gives a meaningless average.",
    "answer:No — the populations, doses, and settings are too heterogeneous; a narrative synthesis describing each study's findings in context is more appropriate" = "Correct! With only 3 studies and major clinical heterogeneity (different ages, doses, populations), a meta-analysis would give a pooled estimate that applies to no real clinical scenario. A narrative systematic review is more informative.",
    "No — because one study shows harm (RR > 1)" = "Wrong reasoning — contradictory results alone don't preclude meta-analysis (that's actually when it can be most useful). The problem here is clinical heterogeneity, not contradictory results."
  )
)
```

---

## Further Learning

::: {.callout-note title="Recommended Resources"}
**Textbooks:**

- Higgins JPT, Thomas J, et al. *Cochrane Handbook for Systematic Reviews of Interventions* — free online at [training.cochrane.org](https://training.cochrane.org/)
- Borenstein M, Hedges LV, et al. *Introduction to Meta-Analysis* — excellent conceptual introduction

**Videos:**

- StatQuest with Josh Starmer — Meta-Analysis and Forest Plots (clear visual explanations)
- Cochrane Training — [Interactive Learning Modules](https://training.cochrane.org/interactivelearning) (free, self-paced)

**Tools:**

- [RevMan](https://training.cochrane.org/online-learning/core-software-cochrane-reviews/revman) — Cochrane's free meta-analysis software
- R packages: `meta`, `metafor` — flexible and publication-quality
- [GRADEpro](https://gradepro.org/) — Create evidence profile tables

**Papers from our group:**

- Joshi R, Reingold AL, Menzies D, Pai M (2006) — TB in healthcare workers: systematic review
- Eddleston M, … Joshi R, … (2009) — Pralidoxime RCT that challenged conventional wisdom
- Pakhare A, Joshi A et al. (2024) — BLESS cohort: single high-quality study vs pooled estimates
:::

---

## References

::: {#refs}
:::
