---
title: "Why Biostatistics Matters"
subtitle: "The Clinician's Statistical Toolkit"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    footer: "Biostatistics for Clinicians | Module 1"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
set.seed(42)
```

## Learning Objectives

::: {.incremental}
1. Explain why statistical literacy is essential for clinical practice
2. Distinguish between qualitative (categorical) and quantitative (numerical) data
3. Classify variables using **NOIR**: Nominal, Ordinal, Interval, Ratio
4. Choose appropriate statistics based on measurement scale
5. Distinguish descriptive from inferential statistics
6. Identify independent and dependent variables
7. Recognise how statistical errors cause real patient harm
:::

## Clinical Hook: The CAST Trial {.smaller}

::: {.columns}
::: {.column width="60%"}
**The Deadly Assumption (1980s)**

- Post-MI patients have PVCs
- PVCs predict sudden cardiac death
- **Assumption:** suppress PVCs → reduce mortality

**The CAST Result (1989)**

- Antiarrhythmics **doubled** mortality (RR = 2.5)
- Trial stopped early
- Estimated **tens of thousands** of excess deaths per year in the US alone
:::

::: {.column width="40%"}
::: {.callout-warning}
## Surrogate Endpoint Fallacy
Suppressing a risk marker is not the same as preventing the disease. The marker (PVCs) was treated instead of the patient.
:::
:::
:::

## More Examples of Statistics Gone Wrong

| Case | What happened | Module |
|:-----|:-------------|:-------|
| **HRT** | Observational studies said it prevented heart disease. RCT (WHI) showed it caused it. | 6, 14 |
| **Hydroxychloroquine** | Preliminary data drove global prescribing. RCTs showed no benefit. | 8 |
| **Vioxx** | Cardiovascular risk hidden in subgroup analyses. Withdrawn after ~88,000 excess MI. | 14 |

. . .

> **As a clinician, you will read research every week. This course gives you the tools to read it critically.**

## What Is Biostatistics? {.smaller}

Three practical functions for clinicians:

::: {.columns}
::: {.column width="33%"}
### Summarise
When you look at lab reports over time or compare ward outcomes — that's **descriptive statistics**.
:::

::: {.column width="33%"}
### Conclude
A BP of 142/90 in one reading doesn't confirm HTN. A p = 0.04 doesn't prove a drug works. Statistics handles **uncertainty**.
:::

::: {.column width="33%"}
### Evaluate
Understanding study design, bias, and analysis lets you separate **reliable evidence from noise**.
:::
:::

## Data Types in Medicine

```{r}
#| label: fig-data-types
#| fig-height: 5.5
#| fig-width: 10

nodes <- tibble(
  label = c(
    "Data Types",
    "Qualitative\n(Categorical)", "Quantitative\n(Numerical)",
    "Nominal", "Ordinal", "Discrete", "Continuous",
    "Blood group\nSex (M, F)",
    "Pain scale\nASA grade",
    "No. of pregnancies\nPulse rate",
    "Height (cm)\nSerum creatinine"
  ),
  x = c(4, 2, 6, 1, 3, 5, 7, 1, 3, 5, 7),
  y = c(5, 3.5, 3.5, 2, 2, 2, 2, 0.7, 0.7, 0.7, 0.7),
  fill = c("#2c3e50", "#3498db", "#e74c3c",
           "#85c1e9", "#85c1e9", "#f1948a", "#f1948a",
           "#d5e8f0", "#d5e8f0", "#fad7d3", "#fad7d3"),
  text_col = c("white", "white", "white",
               "black", "black", "black", "black",
               "black", "black", "black", "black"),
  fontsize = c(5, 4, 4, 3.5, 3.5, 3.5, 3.5, 3, 3, 3, 3),
  width = c(2.5, 2.2, 2.2, 1.5, 1.5, 1.5, 1.5, 1.8, 1.8, 1.8, 1.8),
  height = c(0.7, 0.7, 0.7, 0.6, 0.6, 0.6, 0.6, 0.8, 0.8, 0.8, 0.8)
)

edges <- tibble(
  x = c(4, 4, 2, 2, 6, 6, 1, 3, 5, 7),
  xend = c(2, 6, 1, 3, 5, 7, 1, 3, 5, 7),
  y = c(4.65, 4.65, 3.15, 3.15, 3.15, 3.15, 1.7, 1.7, 1.7, 1.7),
  yend = c(3.85, 3.85, 2.3, 2.3, 2.3, 2.3, 1.15, 1.15, 1.15, 1.15)
)

ggplot() +
  geom_segment(data = edges, aes(x = x, y = y, xend = xend, yend = yend),
               colour = "#7f8c8d", linewidth = 0.8) +
  geom_tile(data = nodes, aes(x = x, y = y, width = width, height = height, fill = fill),
            colour = "white", linewidth = 0.3) +
  geom_text(data = nodes, aes(x = x, y = y, label = label, colour = text_col, size = fontsize),
            lineheight = 0.9) +
  scale_fill_identity() + scale_colour_identity() + scale_size_identity() +
  coord_cartesian(xlim = c(-0.5, 8.5), ylim = c(0, 5.7)) +
  theme_void()
```

## The NOIR Classification {.smaller}

**Stevens' Four Levels of Measurement**

| Scale | Order? | Equal intervals? | True zero? | Example |
|:------|:------:|:---------------:|:---------:|:--------|
| **N**ominal | No | No | No | Blood group, Sex |
| **O**rdinal | Yes | No | No | NYHA class, GCS |
| **I**nterval | Yes | Yes | No | Temperature (°C) |
| **R**atio | Yes | Yes | Yes | Height, Weight, BP |

. . .

::: {.callout-tip}
## The Ratio Test
Can you meaningfully say "twice as much"? If yes → **Ratio**. If no → **Interval** or lower.

- "14 g/dL Hb is twice 7 g/dL" ✓ → Ratio
- "40°C is twice as hot as 20°C" ✗ → Interval
:::

## NOIR: Permissible Operations

```{r}
#| label: fig-noir
#| fig-height: 4.5
#| fig-width: 9

scales_df <- tibble(
  scale = factor(c("Nominal", "Ordinal", "Interval", "Ratio"),
                 levels = c("Nominal", "Ordinal", "Interval", "Ratio")),
  equality = c(TRUE, TRUE, TRUE, TRUE),
  ordering = c(FALSE, TRUE, TRUE, TRUE),
  differences = c(FALSE, FALSE, TRUE, TRUE),
  ratios = c(FALSE, FALSE, FALSE, TRUE)
)

cap_long <- scales_df %>%
  pivot_longer(cols = c(equality, ordering, differences, ratios),
               names_to = "operation", values_to = "allowed") %>%
  mutate(operation = factor(operation,
    levels = c("equality", "ordering", "differences", "ratios"),
    labels = c("= / \u2260\n(classify)", "> / <\n(rank)",
               "+ / \u2212\n(differences)", "\u00d7 / \u00f7\n(ratios)")))

ggplot(cap_long, aes(x = operation, y = scale)) +
  geom_tile(aes(fill = allowed), colour = "white", linewidth = 2) +
  geom_text(aes(label = ifelse(allowed, "\u2713", "\u2717"), colour = allowed),
            size = 8, fontface = "bold") +
  scale_fill_manual(values = c("TRUE" = "#d4edda", "FALSE" = "#f8d7da"), guide = "none") +
  scale_colour_manual(values = c("TRUE" = "#155724", "FALSE" = "#721c24"), guide = "none") +
  labs(title = "What Operations Are Permissible?",
       x = "Mathematical Operation", y = "Measurement Scale") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"), panel.grid = element_blank(),
        axis.text = element_text(size = 12))
```

## Choosing the Right Statistics

| Scale | Central Tendency | Spread | Tests |
|:------|:----------------|:-------|:------|
| Nominal | Mode | Frequency | Chi-square, Fisher |
| Ordinal | Median | IQR | Mann-Whitney, Spearman |
| Interval | Mean, Median | SD, IQR | t-test, ANOVA, Pearson |
| Ratio | Mean, Geo. mean | SD, CV | All parametric tests |

. . .

::: {.callout-warning}
## Common Mistake
Computing the **mean** of NYHA class or Likert scale data. These are ordinal — use the **median** and non-parametric tests.
:::

## Descriptive vs Inferential Statistics

```{r}
#| label: fig-desc-inf
#| fig-height: 4.5
#| fig-width: 10

population <- tibble(bp = rnorm(10000, 130, 18), group = "Population (N = 10,000)")
sample_data <- tibble(bp = sample(population$bp, 80), group = "Your Sample (n = 80)")

p1 <- ggplot(population, aes(x = bp)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "#3498db", alpha = 0.6, colour = "white") +
  geom_vline(xintercept = mean(population$bp), colour = "#e74c3c", linewidth = 1.2, linetype = "dashed") +
  labs(title = "Population (Unknown)", x = "Systolic BP (mmHg)", y = "Density") +
  theme_minimal(base_size = 13) + theme(plot.title = element_text(face = "bold"))

p2 <- ggplot(sample_data, aes(x = bp)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "#e67e22", alpha = 0.7, colour = "white") +
  geom_vline(xintercept = mean(sample_data$bp), colour = "#e74c3c", linewidth = 1.2, linetype = "dashed") +
  labs(title = "Your Sample (What you observe)", x = "Systolic BP (mmHg)", y = "Density") +
  theme_minimal(base_size = 13) + theme(plot.title = element_text(face = "bold"))

p1 + p2 + plot_annotation(
  title = "Descriptive (summarise your data) \u2192 Inferential (generalise to the population)",
  theme = theme(plot.title = element_text(face = "bold", size = 15)))
```

## The Research-to-Bedside Pipeline

```{r}
#| label: fig-pipeline
#| fig-height: 3.5
#| fig-width: 11

pipeline <- tibble(
  stage = factor(c("1. Clinical\nQuestion", "2. Study\nDesign", "3. Data\nCollection",
                    "4. Statistical\nAnalysis", "5. Interpretation", "6. Clinical\nDecision"),
                 levels = c("1. Clinical\nQuestion", "2. Study\nDesign", "3. Data\nCollection",
                            "4. Statistical\nAnalysis", "5. Interpretation", "6. Clinical\nDecision")),
  stats_role = c("Testable\nhypothesis", "Sample size,\nrandomisation", "Sampling,\ndata quality",
                 "Correct tests,\nestimation", "p-values, CIs,\nsignificance", "NNT,\nrisk-benefit"),
  y = 1)

ggplot(pipeline, aes(x = stage, y = y)) +
  geom_tile(aes(fill = stage), height = 0.6, width = 0.9, alpha = 0.85) +
  geom_text(aes(label = stats_role), size = 3.5, lineheight = 0.9) +
  scale_fill_brewer(palette = "Blues", direction = 1) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", axis.text.y = element_blank(),
        axis.text.x = element_text(face = "bold", size = 10), panel.grid = element_blank())
```

## Variables and Their Roles

```{r}
#| label: fig-vars
#| fig-height: 3
#| fig-width: 10

roles <- tibble(
  box = c("Independent Variable\n(Exposure / Predictor)", "Dependent Variable\n(Outcome)"),
  example = c("Drug A vs Placebo", "Mortality at 28 days"),
  x = c(1, 3), y = 1)

ggplot(roles) +
  geom_tile(aes(x = x, y = y, width = 1.6, height = 0.6),
            fill = c("#2980b9", "#c0392b"), alpha = 0.85) +
  geom_text(aes(x = x, y = y + 0.05, label = box),
            colour = "white", fontface = "bold", size = 4.5, lineheight = 0.9) +
  geom_text(aes(x = x, y = y - 0.45, label = example),
            colour = "grey30", size = 3.5) +
  annotate("segment", x = 1.85, xend = 2.15, y = 1, yend = 1,
           arrow = arrow(length = unit(0.3, "cm")), colour = "#7f8c8d", linewidth = 1.5) +
  annotate("text", x = 2, y = 1.15, label = "causes / predicts?",
           colour = "#7f8c8d", fontface = "italic", size = 4) +
  theme_void() + coord_cartesian(xlim = c(0, 4), ylim = c(0.2, 1.5))
```

. . .

::: {.callout-note}
## Confounders
A **confounder** is associated with both exposure and outcome, distorting the apparent relationship. Coffee → Lung cancer? No — **smoking** is the confounder.
:::

## Key Takeaways

1. **Statistics is not optional** — flawed reasoning has killed patients (CAST, HRT, HCQ)

2. **NOIR classification** determines which statistics and tests are valid

3. **Interval ≠ Ratio** — the "true zero" test (Temperature °C vs Height cm)

4. **Descriptive → Inferential** — from "what I see" to "what is true"

5. **Independent → Dependent** — beware of confounders

## Further Learning

**Videos:**

- [StatQuest: Statistics Fundamentals](https://www.youtube.com/playlist?list=PLblh5JKOoLUK0FLuzwntyYI10UQFUhsY9)
- [Zedstatistics: What is Statistics?](https://www.youtube.com/watch?v=hjZJIVWHnPE)
- [Types of Data: NOIR](https://www.youtube.com/watch?v=hZxnzfnt5v8) — Scribbr

**Books:**

- Bland M. *An Introduction to Medical Statistics*. Ch 1–2.
- Indrayan A. *Medical Biostatistics*. 4th ed. (Indian examples)
