---
title: "Probability and Clinical Reasoning"
subtitle: "Why Experienced Clinicians Order Fewer Tests — and Get More Answers"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    footer: "Biostatistics for Clinicians | Module 4"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
set.seed(42)
```

## Learning Objectives

::: {.incremental}
1. Recognise that **history → examination → investigation** IS Bayesian updating
2. Apply the **complement, addition, and multiplication** rules
3. Distinguish **independent** from **dependent** events
4. Understand why P(Disease|Test+) $\neq$ P(Test+|Disease)
5. Convert between **probability and odds**
6. Apply **Bayes' theorem** using natural frequencies
7. Trace a **diagnostic cascade** as sequential updates
8. Identify the **base rate**, **prosecutor's**, and **gambler's** fallacies
:::

## The CBC Mystery in Pathology {.smaller}

::: {.columns}
::: {.column width="55%"}
**Monthly lab audit finding:**

CBCs ordered by **senior faculty** → **40% abnormality rate**

CBCs ordered by **PG residents** → **12% abnormality rate**

Are residents ordering unnecessary tests?
Are faculty missing patients?

. . .

**Neither.** This is **Bayes' theorem** at work in the hospital corridors.

Faculty order CBCs when clinical assessment has *already raised suspicion* — **high pre-test probability**.

Residents cast a wider net — **low pre-test probability**.

**Same test. Same lab. Different yield.**
:::

::: {.column width="45%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5

tibble(
  orderer = c("Senior\nFaculty", "PG\nResidents"),
  abnormal = c(40, 12),
  normal = c(60, 88)
) %>%
  pivot_longer(-orderer, names_to = "result", values_to = "pct") %>%
  mutate(result = factor(result, levels = c("normal", "abnormal"))) %>%
  ggplot(aes(x = orderer, y = pct, fill = result)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(pct, "%")), position = position_stack(vjust = 0.5),
            colour = "white", fontface = "bold", size = 6) +
  scale_fill_manual(values = c("abnormal" = "#e74c3c", "normal" = "#bdc3c7"),
                    labels = c("Normal", "Abnormal"), name = "") +
  labs(title = "CBC Abnormality Rates", y = "Percentage", x = NULL) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```
:::
:::

## You Already Think Like Bayes

```{r}
#| fig-height: 5.5
#| fig-width: 11

cascade <- tibble(
  step = factor(c("OPD Base Rate\n(any chest pain)",
                  "History\n(60M, smoker,\ncrushing pain)",
                  "Examination\n(diaphoresis,\nS3, hypotension)",
                  "ECG\n(ST elevation\nin V1-V4)",
                  "Troponin\n(15 ng/mL)"),
                levels = c("OPD Base Rate\n(any chest pain)",
                           "History\n(60M, smoker,\ncrushing pain)",
                           "Examination\n(diaphoresis,\nS3, hypotension)",
                           "ECG\n(ST elevation\nin V1-V4)",
                           "Troponin\n(15 ng/mL)")),
  probability = c(5, 45, 75, 92, 99),
  stage = c("Prior", "History", "Examination", "Investigation", "Investigation")
)

ggplot(cascade, aes(x = step, y = probability)) +
  geom_col(aes(fill = stage), width = 0.6) +
  geom_segment(data = cascade[-nrow(cascade), ],
    aes(x = as.numeric(step) + 0.33, xend = as.numeric(step) + 0.67,
        y = probability, yend = lead(probability)),
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    colour = "#7f8c8d", linewidth = 0.8, inherit.aes = FALSE) +
  geom_text(aes(label = paste0(probability, "%")),
            vjust = -0.5, fontface = "bold", size = 5.5) +
  scale_fill_manual(values = c("Prior" = "#bdc3c7", "History" = "#e67e22",
               "Examination" = "#2980b9", "Investigation" = "#27ae60"), name = "Stage") +
  scale_y_continuous(limits = c(0, 115), labels = function(y) paste0(y, "%")) +
  labs(title = "The Diagnostic Cascade for Acute MI",
       subtitle = "Each clinical step is a Bayesian update", x = NULL, y = "P(MI)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "right",
        axis.text.x = element_text(size = 9, lineheight = 1.0))
```

> Faculty order tests at steps 3-4 (high prior). Residents order at step 1 (low prior). **That's the CBC mystery solved.**

## Three Basic Probability Rules {.smaller}

::: {.columns}
::: {.column width="50%"}
**Rule 1: Complement**

$$P(\text{not } A) = 1 - P(A)$$

Drug causes nausea in 15% → P(no nausea) = 85%

**Rule 2: Addition (OR)**

Mutually exclusive:
$$P(A \text{ or } B) = P(A) + P(B)$$

Non-mutually exclusive:
$$P(A \text{ or } B) = P(A) + P(B) - P(A \cap B)$$

Screening camp: DM=12%, HTN=25%, Both=6%
$$P(\text{DM or HTN}) = 12 + 25 - 6 = 31\%$$
:::

::: {.column width="50%"}
**Rule 3: Multiplication (AND)**

Independent:
$$P(A \text{ and } B) = P(A) \times P(B)$$

Dependent:
$$P(A \text{ and } B) = P(A) \times P(B|A)$$

Blood culture contamination = 3% per bottle.
$$P(\text{both contaminated}) = 0.03 \times 0.03 = 0.09\%$$

**This is why 2 positive cultures are far more convincing than 1!**
:::
:::

## Visualising the Addition Rule

```{r}
#| fig-height: 4
#| fig-width: 10

circle_pts <- function(cx, cy, r, n = 200) {
  theta <- seq(0, 2 * pi, length.out = n)
  tibble(x = cx + r * cos(theta), y = cy + r * sin(theta))
}

p1 <- ggplot() +
  geom_polygon(data = circle_pts(1.8, 2, 1.0), aes(x, y), fill = "#3498db", alpha = 0.45) +
  geom_polygon(data = circle_pts(3.8, 2, 1.0), aes(x, y), fill = "#e74c3c", alpha = 0.45) +
  annotate("text", x = 1.8, y = 2, label = "Group A\n28%", size = 4.5, fontface = "bold") +
  annotate("text", x = 3.8, y = 2, label = "Group B\n8%", size = 4.5, fontface = "bold") +
  annotate("text", x = 2.8, y = 0.5, label = "Mutually Exclusive\nP(A or B) = 28% + 8% = 36%",
           size = 3.5, colour = "grey30") +
  coord_equal(xlim = c(0.3, 5.3), ylim = c(0, 3.5)) +
  labs(title = "Blood Groups (No Overlap)") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p2 <- ggplot() +
  geom_polygon(data = circle_pts(2.0, 2, 1.3), aes(x, y), fill = "#3498db", alpha = 0.4) +
  geom_polygon(data = circle_pts(3.5, 2, 1.3), aes(x, y), fill = "#e74c3c", alpha = 0.4) +
  annotate("text", x = 1.3, y = 2, label = "DM\nonly 6%", size = 4, fontface = "bold") +
  annotate("text", x = 4.2, y = 2, label = "HTN\nonly 19%", size = 4, fontface = "bold") +
  annotate("text", x = 2.75, y = 2, label = "Both\n6%", size = 4, fontface = "bold", colour = "#8e44ad") +
  annotate("text", x = 2.75, y = 0.3, label = "Overlapping\nP(DM or HTN) = 12 + 25 - 6 = 31%",
           size = 3.5, colour = "grey30") +
  coord_equal(xlim = c(0.3, 5.3), ylim = c(0, 3.5)) +
  labs(title = "DM & HTN (Overlap)") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

p1 + p2
```

## Independence vs Dependence

| Scenario | Type | Why? |
|:---------|:-----|:-----|
| Two blood cultures from the **same bacteraemic patient** | **Dependent** | Both bottles grow the same organism |
| Blood cultures from **two unrelated patients** | **Independent** | One doesn't cause the other |
| Side effects in Patient A and Patient B | **Independent** | Different biology |
| Two joints in **same RA patient** | **Dependent** | Systemic disease |

. . .

::: {.callout-warning}
## Clinical Caution
Two bottles drawn from the **same arm** are NOT independent — contamination is correlated. Guidelines require **separate venepuncture sites**.
:::

## Conditional Probability: The Diagnostic Trap {.smaller}

$$P(A|B) = \frac{P(A \text{ and } B)}{P(B)}$$

The two most critical conditional probabilities in medicine:

| What we know | What we want | Name | Who uses it? |
|:-------------|:-------------|:-----|:-------------|
| Patient **has** disease | P(Test+ \| Disease) | **Sensitivity** | Lab scientists |
| Test **is** positive | P(Disease \| Test+) | **PPV** | Clinicians |

. . .

::: {.callout-important}
## The Most Dangerous Confusion in Medicine
$$P(\text{Disease} | \text{Test+}) \neq P(\text{Test+} | \text{Disease})$$

**Sensitivity is about the test. PPV is about the patient.**

A test can be 95% sensitive and have a PPV of only 16% if the disease is rare.
:::

## The CBC Mystery Explained

```{r}
#| fig-height: 5
#| fig-width: 11

faculty_data <- tibble(
  category = rep(c("Truly Significant\nAbnormality", "Clinically\nInsignificant", "Normal"), c(35, 5, 60)),
  source = "Senior Faculty (n = 100)"
) %>% mutate(id = row_number(), row = ((id-1) %% 10)+1, col = ((id-1) %/% 10)+1)

resident_data <- tibble(
  category = rep(c("Truly Significant\nAbnormality", "Clinically\nInsignificant", "Normal"), c(5, 7, 88)),
  source = "PG Residents (n = 100)"
) %>% mutate(id = row_number(), row = ((id-1) %% 10)+1, col = ((id-1) %/% 10)+1)

combined <- bind_rows(faculty_data, resident_data)

ggplot(combined, aes(x = col, y = row, fill = category)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  scale_fill_manual(values = c("Truly Significant\nAbnormality" = "#e74c3c",
                                "Clinically\nInsignificant" = "#f39c12",
                                "Normal" = "#bdc3c7"), name = "CBC Result") +
  facet_wrap(~ source) +
  coord_fixed() +
  labs(title = "100 CBCs: Faculty vs Residents",
       subtitle = "Red = truly significant. Faculty's higher pre-test probability → more red squares.") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5, colour = "grey40"),
        strip.text = element_text(face = "bold", size = 13), legend.position = "bottom")
```

> Faculty aren't better at reading CBCs. They're better at **selecting who needs one**.

## Probability vs Odds {.smaller}

::: {.columns}
::: {.column width="50%"}
**Conversion:**
$$\text{Odds} = \frac{P}{1-P} \qquad P = \frac{\text{Odds}}{1+\text{Odds}}$$
:::

::: {.column width="50%"}
| Probability | Odds | Clinical Example |
|:------------|:-----|:-----------------|
| 1% | 1:99 | Breast ca. screening |
| 10% | 1:9 | TB, symptomatic (MP) |
| 20% | 1:4 | PE, moderate Wells |
| 50% | 1:1 | No information |
| 80% | 4:1 | MI with ECG changes |
| 90% | 9:1 | 2x blood culture+ |
:::
:::

. . .

::: {.callout-tip}
## Why Odds Matter
$$\text{Post-test odds} = \text{Pre-test odds} \times \text{Likelihood Ratio}$$

One multiplication. That's the entire Bayesian update.
:::

## Bayes' Theorem: Two Forms

::: {.columns}
::: {.column width="50%"}
**Probability form (complex):**

$$P(D|T^+) = \frac{P(T^+|D) \cdot P(D)}{P(T^+|D) \cdot P(D) + P(T^+|\bar{D}) \cdot P(\bar{D})}$$

**Components:**

- $P(D)$ = pre-test probability
- $P(T^+|D)$ = sensitivity
- $P(T^+|\bar{D})$ = 1 - specificity
:::

::: {.column width="50%"}
**Odds form (simple!):**

$$\text{Post-test odds} = \text{Pre-test odds} \times LR$$

**Likelihood Ratios:**

$$LR^+ = \frac{\text{Sensitivity}}{1-\text{Specificity}}$$

$$LR^- = \frac{1-\text{Sensitivity}}{\text{Specificity}}$$
:::
:::

. . .

> **Natural frequencies** make this even easier — just think in 1000 patients.

## Dengue NS1 Test: Monsoon vs January

```{r}
#| fig-height: 5.5
#| fig-width: 11

nodes_m <- tibble(
  label = c("1000 febrile\npatients", "150 dengue\n(15%)", "850 no\ndengue",
            "128 TP", "22 FN", "43 FP", "807 TN"),
  x = c(4, 2, 6, 1, 3, 5, 7), y = c(5, 3, 3, 1, 1, 1, 1),
  fill = c("#495057","#e67e22","#3498db","#27ae60","#e74c3c","#e74c3c","#27ae60"))

edges_m <- tibble(
  x = c(4,4,2,2,6,6), xend = c(2,6,1,3,5,7),
  y = c(4.5,4.5,2.5,2.5,2.5,2.5), yend = c(3.5,3.5,1.5,1.5,1.5,1.5),
  label = c("15%","85%","Sens 85%","Miss 15%","FPR 5%","Spec 95%"))

ggplot() +
  geom_segment(data = edges_m, aes(x=x,y=y,xend=xend,yend=yend), colour="grey50", linewidth=0.7) +
  geom_text(data = edges_m, aes(x=(x+xend)/2+0.3, y=(y+yend)/2, label=label), size=3, colour="grey40") +
  geom_label(data = nodes_m, aes(x=x,y=y,label=label,fill=fill),
             colour="white", fontface="bold", size=3.5, label.padding=unit(0.4,"lines")) +
  scale_fill_identity() +
  annotate("rect", xmin=0.2, xmax=7.8, ymin=-0.5, ymax=0.2,
           fill="#fdf2e9", colour="#e67e22", linewidth=0.5) +
  annotate("text", x=4, y=-0.15,
           label="MONSOON: PPV = 128/171 = 75%    |    JANUARY (0.5%): PPV = 4/54 = 7%",
           size=4, fontface="bold", colour="#c0392b") +
  coord_cartesian(xlim=c(0,8), ylim=c(-0.6,5.8)) +
  labs(title="Dengue NS1: Natural Frequency Tree (Monsoon, Bhopal)") +
  theme_void(base_size=12) +
  theme(plot.title = element_text(face="bold", hjust=0.5, size=14))
```

> Same test, same lab, same technician. **PPV drops from 75% to 7%** when prevalence changes.

## PPV vs Prevalence: The Whole Picture

```{r}
#| fig-height: 5
#| fig-width: 10

prev <- seq(0.005, 0.50, by = 0.005)
ppv <- sapply(prev, function(p) { tp <- p*0.85; fp <- (1-p)*0.05; tp/(tp+fp) })
npv <- sapply(prev, function(p) { tn <- (1-p)*0.95; fn <- p*0.15; tn/(tn+fn) })

curves <- tibble(prevalence = rep(prev*100, 2), value = c(ppv*100, npv*100),
                 metric = rep(c("PPV","NPV"), each=length(prev)))

pts <- tibble(prevalence = c(0.5,5,15,40),
              ppv = sapply(c(0.005,0.05,0.15,0.40), function(p) {
                tp <- p*0.85; fp <- (1-p)*0.05; tp/(tp+fp)*100 }),
              label = c("January\n(0.5%)","Post-\nmonsoon\n(5%)","Monsoon\n(15%)","Outbreak\n(40%)"))

ggplot(curves, aes(x=prevalence, y=value, colour=metric)) +
  geom_line(linewidth=1.3) +
  geom_point(data=pts, aes(x=prevalence,y=ppv), colour="#e74c3c", size=4, inherit.aes=FALSE) +
  geom_text(data=pts, aes(x=prevalence, y=ppv+7, label=label),
            colour="#c0392b", size=3, fontface="bold", inherit.aes=FALSE) +
  scale_colour_manual(values = c("PPV"="#e74c3c","NPV"="#3498db")) +
  labs(title="Dengue NS1: PPV and NPV Across Seasons",
       subtitle="Sensitivity = 85%, Specificity = 95%",
       x="Prevalence (%)", y="Predictive Value (%)", colour=NULL) +
  theme_minimal(base_size=13) +
  theme(plot.title=element_text(face="bold"), legend.position="bottom")
```

## The Mammography Dilemma

```{r}
#| fig-height: 5.5
#| fig-width: 10

nodes_mam <- tibble(
  label = c("1000 women\nscreened","10 cancer\n(1%)","990 no\ncancer",
            "9 TP","1 FN","69 FP","921 TN"),
  x = c(4,2,6,1,3,5,7), y = c(5,3,3,1,1,1,1),
  fill = c("#495057","#e67e22","#3498db","#27ae60","#e74c3c","#e74c3c","#27ae60"))

edges_mam <- tibble(
  x = c(4,4,2,2,6,6), xend = c(2,6,1,3,5,7),
  y = c(4.5,4.5,2.5,2.5,2.5,2.5), yend = c(3.5,3.5,1.5,1.5,1.5,1.5),
  label = c("1%","99%","Sens 90%","Miss 10%","FPR 7%","Spec 93%"))

ggplot() +
  geom_segment(data=edges_mam, aes(x=x,y=y,xend=xend,yend=yend), colour="grey50", linewidth=0.7) +
  geom_text(data=edges_mam, aes(x=(x+xend)/2+0.3, y=(y+yend)/2, label=label), size=3, colour="grey40") +
  geom_label(data=nodes_mam, aes(x=x,y=y,label=label,fill=fill),
             colour="white", fontface="bold", size=3.5, label.padding=unit(0.4,"lines")) +
  scale_fill_identity() +
  annotate("rect", xmin=0.2, xmax=7.8, ymin=-0.5, ymax=0.2,
           fill="#fdedec", colour="#e74c3c", linewidth=0.5) +
  annotate("text", x=4, y=-0.15,
           label="PPV = 9/78 = 11.5%  —  Only 1 in ~9 positive mammograms is cancer!",
           size=4, fontface="bold", colour="#c0392b") +
  coord_cartesian(xlim=c(0,8), ylim=c(-0.6,5.8)) +
  labs(title="Mammography Screening (n = 1000, Prevalence = 1%)") +
  theme_void(base_size=12) +
  theme(plot.title = element_text(face="bold", hjust=0.5, size=14))
```

> Gigerenzer showed that presenting **natural frequencies** raised doctors' accuracy from ~15% to ~85%.

## The SLE Diagnostic Cascade {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 11

sle <- tibble(
  step = factor(c("Base Rate\n(young woman,\njoint pain + rash)",
                  "History\n(oral ulcers,\nphotosensitivity)",
                  "Examination\n(malar rash,\nsynovitis)",
                  "ANA+\n(titre 1:320)",
                  "Anti-dsDNA+",
                  "Low C3/C4\n+ proteinuria"),
               levels = c("Base Rate\n(young woman,\njoint pain + rash)",
                  "History\n(oral ulcers,\nphotosensitivity)",
                  "Examination\n(malar rash,\nsynovitis)",
                  "ANA+\n(titre 1:320)",
                  "Anti-dsDNA+",
                  "Low C3/C4\n+ proteinuria")),
  prob = c(8, 30, 55, 78, 93, 98),
  stage = c("Prior","History","Examination","Investigation","Investigation","Investigation"))

ggplot(sle, aes(x=step, y=prob)) +
  geom_col(aes(fill=stage), width=0.6, alpha=0.85) +
  geom_text(aes(label=paste0(prob,"%")), vjust=-0.5, fontface="bold", size=5.5) +
  geom_segment(data=sle[-nrow(sle),],
    aes(x=as.numeric(step)+0.33, xend=as.numeric(step)+0.67, y=prob, yend=lead(prob)),
    arrow=arrow(length=unit(0.2,"cm"), type="closed"), colour="#7f8c8d", linewidth=0.8, inherit.aes=FALSE) +
  scale_fill_manual(values=c("Prior"="#bdc3c7","History"="#e67e22",
    "Examination"="#2980b9","Investigation"="#27ae60"), name="Stage") +
  scale_y_continuous(limits=c(0,115), labels=function(y) paste0(y,"%")) +
  labs(title="SLE Diagnosis: Sequential Bayesian Updates",
       subtitle="No single test diagnoses SLE — the cascade IS the diagnosis",
       x=NULL, y="P(SLE)") +
  theme_minimal(base_size=13) +
  theme(plot.title=element_text(face="bold"), axis.text.x=element_text(size=8, lineheight=1.0),
        legend.position="right")
```

> ANA (Sens 95%, Spec 60%) is for **screening**. Anti-dsDNA (Sens 60%, Spec 95%) is for **confirmation**.

## Ruling OUT Pulmonary Embolism

```{r}
#| fig-height: 4.5
#| fig-width: 10

pe <- tibble(
  step = factor(c("ED Presentation\n(chest pain + SOB)",
                  "Wells Score Low\n(2 points)",
                  "D-dimer Negative\n(<500 ng/mL)",
                  "PE Ruled Out"),
               levels = c("ED Presentation\n(chest pain + SOB)",
                  "Wells Score Low\n(2 points)",
                  "D-dimer Negative\n(<500 ng/mL)",
                  "PE Ruled Out")),
  prob = c(15, 6, 0.7, 0.7),
  col = c("#e74c3c","#f39c12","#27ae60","#27ae60"))

ggplot(pe, aes(x=step, y=prob)) +
  geom_col(fill=pe$col, width=0.6) +
  geom_text(aes(label=paste0(prob,"%")), vjust=-0.5, fontface="bold", size=6) +
  annotate("text", x=1.5, y=12, label="Wells score\nLR ~ 0.3", size=3.5, colour="#e67e22", fontface="bold") +
  annotate("text", x=2.5, y=5, label="D-dimer neg\nLR- ~ 0.1", size=3.5, colour="#27ae60", fontface="bold") +
  scale_y_continuous(limits=c(0,20), labels=function(y) paste0(y,"%")) +
  labs(title="Ruling Out PE: The Downward Bayesian Cascade",
       subtitle="15% → 6% → 0.7%. Safe to avoid CT angiography.",
       x=NULL, y="P(PE)") +
  theme_minimal(base_size=13) +
  theme(plot.title=element_text(face="bold"), axis.text.x=element_text(size=9, lineheight=1.0))
```

> Bayesian reasoning saves the patient from **unnecessary radiation, contrast dye, and cost**.

## Three Probability Fallacies {.smaller}

```{r}
#| fig-height: 4.5
#| fig-width: 11

fallacy_df <- tibble(
  fallacy = factor(c("Base Rate\nFallacy","Prosecutor's\nFallacy","Gambler's\nFallacy"),
                   levels = c("Base Rate\nFallacy","Prosecutor's\nFallacy","Gambler's\nFallacy")),
  error = c("Ignores prevalence\nwhen interpreting\ntest results",
            "Confuses\nP(E|H) with P(H|E)\ni.e. sensitivity with PPV",
            "Past independent\nevents seem to\naffect future ones"),
  consequence = c("Unnecessary isolation,\novertreatment",
                   "False certainty\nin diagnosis",
                   "Irrational\nexpectations"),
  danger = c(9, 8, 5),
  colour = c("#e74c3c","#e67e22","#f1c40f"))

ggplot(fallacy_df, aes(x=fallacy, y=danger, fill=colour)) +
  geom_col(width=0.55) +
  geom_text(aes(label=error, y=danger/2), size=3.5, colour="white", fontface="bold", lineheight=0.9) +
  geom_text(aes(label=consequence, y=danger+0.4), size=3, colour="#2c3e50", vjust=0, lineheight=0.9) +
  scale_fill_identity() +
  scale_y_continuous(limits=c(0,13), breaks=NULL) +
  labs(title="Three Probability Fallacies That Harm Patients", x=NULL, y=NULL) +
  theme_minimal(base_size=13) +
  theme(plot.title=element_text(face="bold"), axis.text.y=element_blank())
```

> **Base rate fallacy** is #1: A 95% sensitive COVID test at 2% prevalence gives PPV of only 16%.

## Bayesian Update: Worked Example

**Patient:** 20% pre-test probability of PE. D-dimer negative (LR- = 0.1).

. . .

| Step | Calculation | Result |
|:-----|:-----------|:-------|
| Pre-test odds | 0.20 / 0.80 | 0.25 |
| Post-test odds | 0.25 $\times$ 0.1 | 0.025 |
| Post-test probability | 0.025 / 1.025 | **2.4%** |

. . .

::: {.callout-tip}
## Clinical Implication
Negative D-dimer drops PE probability from 20% to ~2.4% — **safe to rule out** without CT angiography. This is Bayes at the bedside.
:::

## Key Takeaways

1. **Clinical diagnosis IS Bayesian reasoning** — history → examination → investigation is sequential probability updating

2. **Pre-test probability is king** — it's why faculty CBCs yield more abnormals than residents' CBCs

3. **P(Disease|Test+) $\neq$ P(Test+|Disease)** — sensitivity is NOT PPV

4. **Natural frequencies** ("think in 1000 patients") make Bayes intuitive

5. **Post-test odds = Pre-test odds $\times$ LR** — one formula for bedside reasoning

6. **The base rate fallacy** is the most dangerous probability error in clinical medicine

## Further Learning

**Videos:**

- [3Blue1Brown: Bayes' Theorem — Geometry of Changing Beliefs](https://www.youtube.com/watch?v=HZGCoVF3YvM)
- [StatQuest: Bayes' Theorem](https://www.youtube.com/watch?v=9wCnvr7Xw4E)
- [Veritasium: The Bayesian Trap](https://www.youtube.com/watch?v=R13BD8qKeTg)

**Books:**

- Gigerenzer G. *Reckoning with Risk.* Penguin; 2002.
- Sackett DL et al. *Clinical Epidemiology.* 3rd ed. Ch 4.
- Bland M. *An Introduction to Medical Statistics.* 4th ed. Ch 20.
