---
title: "Diagnostic Test Evaluation"
subtitle: "The Clinician's Toolkit for Measuring How Well Tests Work"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    footer: "Biostatistics for Clinicians | Module 5"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
set.seed(42)

# Shared parameters
total_n <- 1000
prev <- 0.05
sens_via <- 0.80; spec_via <- 0.85
sens_pap <- 0.60; spec_pap <- 0.95

diseased <- total_n * prev
healthy  <- total_n * (1 - prev)

tp_via <- round(diseased * sens_via); fn_via <- diseased - tp_via
fp_via <- round(healthy * (1 - spec_via)); tn_via <- healthy - fp_via

tp_pap <- round(diseased * sens_pap); fn_pap <- diseased - tp_pap
fp_pap <- round(healthy * (1 - spec_pap)); tn_pap <- healthy - fp_pap

ppv_via <- tp_via / (tp_via + fp_via)
npv_via <- tn_via / (fn_via + tn_via)
ppv_pap <- tp_pap / (tp_pap + fp_pap)
npv_pap <- tn_pap / (fn_pap + tn_pap)

lr_pos_via <- sens_via / (1 - spec_via)
lr_neg_via <- (1 - sens_via) / spec_via
lr_pos_pap <- sens_pap / (1 - spec_pap)
lr_neg_pap <- (1 - sens_pap) / spec_pap

calc_post <- function(pre, lr) {
  odds_pre <- pre / (1 - pre)
  odds_post <- odds_pre * lr
  odds_post / (1 + odds_post)
}
```

## Learning Objectives

::: {.incremental}
1. Construct a **2×2 table** with index test in rows, gold standard in columns
2. Calculate **sensitivity, specificity, PPV, NPV** from the 2×2 table
3. Explain how **prevalence transforms** predictive values
4. Compute and interpret **likelihood ratios** (LR+, LR−)
5. Use the **Fagan nomogram** for pre-test → post-test probability
6. Construct and interpret **ROC curves** and AUC
7. Choose wisely: when to prioritise **sensitivity vs specificity**
:::

## Clinical Hook: Cervical Cancer Screening in Rural India {.smaller}

::: {.columns}
::: {.column width="55%"}
India bears **¼ of the global cervical cancer burden**.

Rural districts lack Pap smear infrastructure — so the government promotes **VIA (Visual Inspection with Acetic Acid)**.

A community health worker screens **1000 women** with VIA. Results are compared against colposcopy-directed biopsy.

| Test | Sensitivity | Specificity |
|:-----|:------------|:------------|
| **VIA** | 80% | 85% |
| **Pap Smear** | 60% | 95% |

*"VIA is more sensitive — isn't it the better test?"*

**It depends on what you need the test to do.**
:::

::: {.column width="45%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5

tibble(
  test = c("VIA", "Pap Smear"),
  sensitivity = c(80, 60),
  specificity = c(85, 95)
) %>%
  pivot_longer(-test, names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = metric, y = value, fill = test)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = paste0(value, "%")),
            position = position_dodge(width = 0.6),
            vjust = -0.5, fontface = "bold", size = 5) +
  scale_fill_manual(values = c("VIA" = "#e74c3c", "Pap Smear" = "#2980b9")) +
  scale_y_continuous(limits = c(0, 110)) +
  labs(title = "VIA vs Pap Smear: The Trade-Off",
       y = "%", x = NULL, fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```
:::
:::

## The 2×2 Contingency Table {.smaller}

**Standard layout:** Index test in rows, gold standard in columns.

```{r}
#| fig-height: 4.5
#| fig-width: 9

cells <- tibble(
  x = c(1, 2, 1, 2),
  y = c(2, 2, 1, 1),
  label = c("True Positive\n(TP)", "False Positive\n(FP)",
            "False Negative\n(FN)", "True Negative\n(TN)"),
  fill = c("#27ae60", "#e74c3c", "#f39c12", "#3498db"),
  count = c(tp_via, fp_via, fn_via, tn_via)
)

ggplot(cells) +
  geom_tile(aes(x = x, y = y, fill = fill), width = 0.95, height = 0.95, alpha = 0.85) +
  geom_text(aes(x = x, y = y + 0.15, label = label), colour = "white",
            fontface = "bold", size = 5) +
  geom_text(aes(x = x, y = y - 0.2, label = paste0("n = ", count)),
            colour = "white", size = 5) +
  scale_fill_identity() +
  annotate("text", x = 1, y = 2.6, label = "Disease\nPresent", fontface = "bold", size = 4.5) +
  annotate("text", x = 2, y = 2.6, label = "Disease\nAbsent", fontface = "bold", size = 4.5) +
  annotate("text", x = 0.35, y = 2, label = "Test +", fontface = "bold", size = 5) +
  annotate("text", x = 0.35, y = 1, label = "Test −", fontface = "bold", size = 5) +
  coord_cartesian(xlim = c(0, 2.7), ylim = c(0.4, 2.8)) +
  labs(title = "2×2 Table: VIA Screening (n = 1000, Prevalence = 5%)",
       subtitle = "Index test in ROWS, Gold standard in COLUMNS — FP is top-right") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
        plot.subtitle = element_text(hjust = 0.5, colour = "#e74c3c", size = 11))
```

::: {.callout-warning}
## Convention Alert
**FP is in the top-right corner** (test +, disease −). Most textbooks, NEET PG, and USMLE follow this layout.
:::

## VIA Screening: 1000 Women Icon Array

```{r}
#| fig-height: 5.5
#| fig-width: 11

icon_data <- tibble(
  id = 1:total_n,
  category = c(
    rep("True Positive", tp_via),
    rep("False Negative", fn_via),
    rep("False Positive", fp_via),
    rep("True Negative", tn_via)
  )
) %>%
  mutate(
    row = ((id - 1) %% 20) + 1,
    col = ((id - 1) %/% 20) + 1,
    category = factor(category,
                      levels = c("True Positive", "False Negative",
                                 "False Positive", "True Negative"))
  )

icon_colours <- c("True Positive" = "#27ae60", "False Negative" = "#e74c3c",
                   "False Positive" = "#f39c12", "True Negative" = "#bdc3c7")

ggplot(icon_data, aes(x = col, y = row, fill = category)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = icon_colours, name = "Result") +
  coord_fixed() +
  labs(
    title = "VIA Screening: 1000 Women in Rural MP (Prevalence = 5%)",
    subtitle = paste0("TP = ", tp_via, " | FN = ", fn_via,
                      " | FP = ", fp_via, " | TN = ", tn_via,
                      "  —  Notice the large orange FP cluster")
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, colour = "grey40"),
    legend.position = "bottom"
  )
```

## Sensitivity and Specificity {.smaller}

::: {.columns}
::: {.column width="50%"}
**Sensitivity (True Positive Rate)**

$$\text{Sens} = \frac{TP}{TP + FN}$$

*"Of patients WITH disease, how many test positive?"*

. . .

**VIA:** $\frac{40}{50} = 80\%$

**Pap:** $\frac{30}{50} = 60\%$

. . .

**Use:** Screening — **don't miss cases**

Mnemonic: **SnNOut** — Sensitive test, Negative result, rules **Out**
:::

::: {.column width="50%"}
**Specificity (True Negative Rate)**

$$\text{Spec} = \frac{TN}{FP + TN}$$

*"Of patients WITHOUT disease, how many test negative?"*

. . .

**VIA:** $\frac{808}{950} = 85\%$

**Pap:** $\frac{902}{950} = 95\%$

. . .

**Use:** Confirmation — **don't overdiagnose**

Mnemonic: **SpPIn** — Specific test, Positive result, rules **In**
:::
:::

## VIA vs Pap: The Sensitivity-Specificity Trade-Off

```{r}
#| fig-height: 5
#| fig-width: 10

comparison_key <- tibble(
  test = rep(c("VIA\n(Sens 80%, Spec 85%)", "Pap Smear\n(Sens 60%, Spec 95%)"), each = 3),
  metric = rep(c("True Positive\n(Detected)", "False Negative\n(Missed)",
                 "False Positive\n(False alarm)"), 2),
  count = c(tp_via, fn_via, fp_via, tp_pap, fn_pap, fp_pap),
  fill = rep(c("#27ae60", "#e74c3c", "#f39c12"), 2)
)

ggplot(comparison_key, aes(x = metric, y = count, fill = fill)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = count), vjust = -0.5, fontface = "bold", size = 5) +
  scale_fill_identity() +
  facet_wrap(~ test) +
  scale_y_continuous(limits = c(0, 180)) +
  labs(
    title = "VIA vs Pap: Fewer Missed Cases or Fewer False Alarms?",
    subtitle = "VIA: 10 missed, 142 false alarms. Pap: 20 missed, 48 false alarms.",
    x = NULL, y = "Number of Women (out of 1000)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold", size = 11),
        axis.text.x = element_text(size = 9, lineheight = 1.0))
```

## Predictive Values: What Clinicians Actually Need

::: {.columns}
::: {.column width="50%"}
**Positive Predictive Value**

$$PPV = \frac{TP}{TP + FP}$$

*"If test positive, what's the chance of disease?"*

. . .

**VIA PPV** = $\frac{40}{40 + 142}$ = **22%**

**Pap PPV** = $\frac{30}{30 + 48}$ = **38%**

At 5% prevalence, **78% of VIA positives are false alarms!**
:::

::: {.column width="50%"}
**Negative Predictive Value**

$$NPV = \frac{TN}{FN + TN}$$

*"If test negative, what's the chance of no disease?"*

. . .

**VIA NPV** = $\frac{808}{10 + 808}$ = **98.8%**

**Pap NPV** = $\frac{902}{20 + 902}$ = **97.8%**

Both excellent for ruling out.
:::
:::

. . .

::: {.callout-important}
## The Key Insight
**Sensitivity and specificity are test properties.** PPV and NPV depend on **prevalence** — the same test behaves differently in different populations!
:::

## The Prevalence Effect: VIA vs Pap

```{r}
#| fig-height: 5
#| fig-width: 10

prev_range <- seq(0.005, 0.40, by = 0.005)

ppv_via_curve <- sapply(prev_range, function(p) {
  tp <- p * sens_via; fp <- (1-p) * (1-spec_via); tp / (tp + fp)
})
ppv_pap_curve <- sapply(prev_range, function(p) {
  tp <- p * sens_pap; fp <- (1-p) * (1-spec_pap); tp / (tp + fp)
})

ppv_curves <- tibble(
  prevalence = rep(prev_range * 100, 2),
  ppv = c(ppv_via_curve * 100, ppv_pap_curve * 100),
  test = rep(c("VIA (Sens=80%, Spec=85%)", "Pap (Sens=60%, Spec=95%)"),
             each = length(prev_range))
)

key_pts <- tibble(
  prevalence = c(2, 5, 15, 30),
  ppv_via = sapply(c(0.02, 0.05, 0.15, 0.30), function(p) {
    tp <- p*sens_via; fp <- (1-p)*(1-spec_via); tp/(tp+fp)*100 }),
  label = c("Population\nscreening", "District\nhospital",
            "Colposcopy\nclinic", "Symptomatic\nreferral")
)

ggplot(ppv_curves, aes(x = prevalence, y = ppv, colour = test)) +
  geom_line(linewidth = 1.3) +
  geom_point(data = key_pts, aes(x = prevalence, y = ppv_via),
             colour = "#2c3e50", size = 3, inherit.aes = FALSE) +
  geom_text(data = key_pts, aes(x = prevalence, y = ppv_via + 6, label = label),
            colour = "#2c3e50", size = 3, fontface = "bold",
            inherit.aes = FALSE, lineheight = 0.9) +
  scale_colour_manual(values = c("VIA (Sens=80%, Spec=85%)" = "#e74c3c",
                                  "Pap (Sens=60%, Spec=95%)" = "#2980b9")) +
  scale_y_continuous(limits = c(0, 100), labels = function(y) paste0(y, "%")) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  labs(title = "PPV vs Prevalence: VIA vs Pap Smear",
       subtitle = "Pap's higher specificity gives it a PPV advantage at low prevalence. VIA catches up at higher prevalence.",
       x = "CIN2+ Prevalence (%)", y = "PPV (%)", colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

## Likelihood Ratios: Prevalence-Independent {.smaller}

::: {.columns}
::: {.column width="50%"}
$$LR^+ = \frac{\text{Sensitivity}}{1 - \text{Specificity}}$$

$$LR^- = \frac{1 - \text{Sensitivity}}{\text{Specificity}}$$

| Test | LR+ | LR− |
|:-----|:----|:----|
| **VIA** | 5.3 | 0.24 |
| **Pap** | 12.0 | 0.42 |
:::

::: {.column width="50%"}
| LR+ Value | Interpretation |
|:----------|:---------------|
| > 10 | Strong rule-in |
| 5 – 10 | Moderate rule-in |
| 2 – 5 | Weak evidence |

| LR− Value | Interpretation |
|:----------|:---------------|
| < 0.1 | Strong rule-out |
| 0.1 – 0.2 | Moderate rule-out |
| 0.2 – 0.5 | Weak evidence |
:::
:::

. . .

**Key insight:** Pap+ is more convincing (LR+ = 12 → strong rule-in), but VIA− is more reassuring (LR− = 0.24 < 0.42).

This **quantifies SnNOut / SpPIn** precisely.

## Fagan Nomogram: VIA vs Pap

```{r}
#| fig-height: 5.5
#| fig-width: 10

pre_test <- seq(0.01, 0.99, by = 0.01)

fagan_df <- tibble(
  pre = rep(pre_test * 100, 4),
  post = c(
    calc_post(pre_test, lr_pos_via) * 100,
    calc_post(pre_test, lr_neg_via) * 100,
    calc_post(pre_test, lr_pos_pap) * 100,
    calc_post(pre_test, lr_neg_pap) * 100
  ),
  result = rep(c("VIA Positive (LR+ = 5.3)", "VIA Negative (LR− = 0.24)",
                  "Pap Positive (LR+ = 12)", "Pap Negative (LR− = 0.42)"),
               each = length(pre_test))
)

scenarios <- tibble(
  test = c("VIA+", "VIA−", "Pap+", "Pap−"),
  pre = 5,
  post = c(
    calc_post(0.05, lr_pos_via) * 100,
    calc_post(0.05, lr_neg_via) * 100,
    calc_post(0.05, lr_pos_pap) * 100,
    calc_post(0.05, lr_neg_pap) * 100
  )
)

ggplot(fagan_df, aes(x = pre, y = post, colour = result)) +
  geom_line(linewidth = 1.1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              colour = "grey60", linewidth = 0.7) +
  geom_point(data = scenarios, aes(x = pre, y = post),
             size = 4, inherit.aes = FALSE, colour = "#2c3e50") +
  geom_text(data = scenarios,
            aes(x = pre + 4, y = post,
                label = paste0(test, " → ", round(post, 1), "%")),
            size = 3.5, fontface = "bold", inherit.aes = FALSE, colour = "#2c3e50") +
  scale_colour_manual(values = c(
    "VIA Positive (LR+ = 5.3)" = "#e74c3c",
    "VIA Negative (LR− = 0.24)" = "#3498db",
    "Pap Positive (LR+ = 12)" = "#c0392b",
    "Pap Negative (LR− = 0.42)" = "#2471a3"
  )) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  scale_y_continuous(labels = function(y) paste0(y, "%")) +
  labs(
    title = "Fagan Nomogram: Pre-Test → Post-Test Probability",
    subtitle = "At 5% pre-test: Pap+ shifts to 39% (strong), VIA+ only to 22% (moderate). VIA− drops to 1.2% (good rule-out).",
    x = "Pre-test Probability", y = "Post-test Probability", colour = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

## ROC Curves: Fasting Glucose for Diabetes {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 11

set.seed(42)
n_diabetic <- 300; n_healthy_dm <- 2700
glucose_diabetic <- rnorm(n_diabetic, mean = 145, sd = 35)
glucose_healthy <- rnorm(n_healthy_dm, mean = 90, sd = 15)

dist_df <- tibble(
  glucose = c(glucose_diabetic, glucose_healthy),
  status = c(rep("Diabetic", n_diabetic), rep("Non-Diabetic", n_healthy_dm))
)

p_dist <- ggplot(dist_df, aes(x = glucose, fill = status)) +
  geom_density(alpha = 0.5, colour = NA) +
  geom_vline(xintercept = c(100, 126), linetype = c("dashed", "solid"),
             colour = c("#f39c12", "#e74c3c"), linewidth = 0.8) +
  annotate("text", x = 100, y = 0.025, label = "IFG\n(100)",
           size = 3, fontface = "bold", colour = "#f39c12") +
  annotate("text", x = 126, y = 0.025, label = "DM\n(126)",
           size = 3, fontface = "bold", colour = "#e74c3c") +
  scale_fill_manual(values = c("Diabetic" = "#e74c3c", "Non-Diabetic" = "#3498db")) +
  labs(title = "Glucose Distributions", x = "Fasting Glucose (mg/dL)",
       y = "Density", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

cutoffs_dm <- seq(60, 220, by = 2)
roc_data <- lapply(cutoffs_dm, function(c) {
  tpr <- sum(glucose_diabetic >= c) / n_diabetic
  fpr <- sum(glucose_healthy >= c) / n_healthy_dm
  tibble(cutoff = c, tpr = tpr, fpr = fpr)
}) %>% bind_rows() %>% arrange(fpr)

auc_val <- sum(diff(roc_data$fpr) *
               (roc_data$tpr[-1] + roc_data$tpr[-nrow(roc_data)]) / 2)

cut_100 <- roc_data[which.min(abs(roc_data$cutoff - 100)), ]
cut_126 <- roc_data[which.min(abs(roc_data$cutoff - 126)), ]

p_roc <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
  geom_path(colour = "#2c3e50", linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              colour = "grey60", linewidth = 0.7) +
  geom_point(data = cut_100, colour = "#f39c12", size = 5) +
  geom_point(data = cut_126, colour = "#e74c3c", size = 5) +
  annotate("text", x = cut_100$fpr + 0.06, y = cut_100$tpr,
           label = paste0("Cutoff 100\nSens=", round(cut_100$tpr*100),
                          "% Spec=", round((1-cut_100$fpr)*100), "%"),
           size = 3, fontface = "bold", colour = "#f39c12") +
  annotate("text", x = cut_126$fpr + 0.06, y = cut_126$tpr - 0.05,
           label = paste0("Cutoff 126\nSens=", round(cut_126$tpr*100),
                          "% Spec=", round((1-cut_126$fpr)*100), "%"),
           size = 3, fontface = "bold", colour = "#e74c3c") +
  annotate("text", x = 0.55, y = 0.25,
           label = paste0("AUC = ", round(auc_val, 3)),
           size = 5, fontface = "bold", colour = "#2c3e50") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "ROC Curve: Fasting Glucose",
       x = "FPR (1 − Specificity)", y = "TPR (Sensitivity)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"), aspect.ratio = 1)

p_dist + p_roc + plot_layout(widths = c(1, 1.2))
```

## AUC: Summarising Test Discrimination

| AUC Range | Discrimination | Clinical Analogy |
|:----------|:---------------|:-----------------|
| 0.90 – 1.00 | **Excellent** | Troponin for MI |
| 0.80 – 0.90 | Good | GeneXpert for TB |
| 0.70 – 0.80 | Fair | CRP for infection |
| 0.60 – 0.70 | Poor | ESR for inflammation |
| 0.50 | None | Coin flip |

. . .

::: {.callout-tip}
## Practical Meaning
AUC = 0.87 means: pick one random diabetic and one random non-diabetic — there's an **87% chance** the diabetic has a higher fasting glucose.
:::

## Comparing Tests: HbA1c vs Random Blood Glucose

```{r}
#| fig-height: 5
#| fig-width: 9

set.seed(42)
hba1c_diabetic <- rnorm(300, mean = 8.5, sd = 1.5)
hba1c_healthy <- rnorm(2700, mean = 5.5, sd = 0.6)
rbg_diabetic <- rnorm(300, mean = 200, sd = 60)
rbg_healthy <- rnorm(2700, mean = 120, sd = 40)

calc_roc <- function(diseased_vals, healthy_vals) {
  all_vals <- c(diseased_vals, healthy_vals)
  cutoffs_seq <- seq(min(all_vals) - 5, max(all_vals) + 5,
                     by = (max(all_vals) - min(all_vals)) / 100)
  roc <- lapply(cutoffs_seq, function(c) {
    tpr <- sum(diseased_vals >= c) / length(diseased_vals)
    fpr <- sum(healthy_vals >= c) / length(healthy_vals)
    tibble(tpr = tpr, fpr = fpr)
  }) %>% bind_rows() %>% arrange(fpr)
  roc_sorted <- roc %>% distinct()
  auc <- sum(diff(roc_sorted$fpr) *
    (roc_sorted$tpr[-1] + roc_sorted$tpr[-nrow(roc_sorted)]) / 2)
  roc$auc <- auc
  return(roc)
}

roc_hba1c <- calc_roc(hba1c_diabetic, hba1c_healthy) %>%
  mutate(test = paste0("HbA1c (AUC = ", round(first(auc), 2), ")"))
roc_rbg <- calc_roc(rbg_diabetic, rbg_healthy) %>%
  mutate(test = paste0("Random Blood Glucose (AUC = ", round(first(auc), 2), ")"))

roc_both <- bind_rows(roc_hba1c, roc_rbg)

ggplot(roc_both, aes(x = fpr, y = tpr, colour = test)) +
  geom_path(linewidth = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              colour = "grey60", linewidth = 0.7) +
  scale_colour_manual(values = c("#27ae60", "#e67e22")) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Comparing Tests: HbA1c vs Random Blood Glucose",
       subtitle = "HbA1c hugs top-left corner → better discrimination",
       x = "FPR (1 − Specificity)", y = "TPR (Sensitivity)", colour = NULL) +
  theme_minimal(base_size = 13) +
  theme(aspect.ratio = 1, legend.position = "bottom",
        plot.title = element_text(face = "bold"))
```

## Test Selection: Sensitivity vs Specificity {.smaller}

| Clinical Scenario | Priority | Reason | Indian Example |
|:------------------|:---------|:-------|:---------------|
| **Community screening** | HIGH Sensitivity | Don't miss cases | VIA for cervical cancer |
| **Confirming diagnosis** | HIGH Specificity | Don't overdiagnose | Biopsy after VIA+ |
| **Ruling out danger** | HIGH Sensitivity | FN = patient sent home | D-dimer for PE |
| **Low prevalence** | HIGH Specificity | Reduce false positives | HIV screening |
| **Toxic treatment** | HIGH Specificity | FP → unnecessary harm | Chemo based on biopsy |

. . .

> **SnNOut / SpPIn:** Sensitive Negative rules Out. Specific Positive rules In.

## The Two-Step Strategy: Screen Then Confirm

```{r}
#| fig-height: 5
#| fig-width: 10

step_data <- tibble(
  step = factor(c("Population\n(1000 women)", "VIA Positive\n(Screen)",
                  "Biopsy Confirmed\n(Diagnose)"),
                levels = c("Population\n(1000 women)", "VIA Positive\n(Screen)",
                           "Biopsy Confirmed\n(Diagnose)")),
  total = c(1000, tp_via + fp_via, tp_via)
)

ggplot(step_data, aes(x = step, y = total)) +
  geom_col(fill = c("#3498db", "#f39c12", "#27ae60"), width = 0.6, alpha = 0.85) +
  geom_text(aes(label = total), vjust = -0.5, fontface = "bold", size = 6) +
  geom_text(data = step_data[2, ],
            aes(x = step, y = total/2,
                label = paste0("TP=", tp_via, " + FP=", fp_via)),
            colour = "white", fontface = "bold", size = 4.5) +
  geom_text(data = step_data[3, ],
            aes(x = step, y = total/2,
                label = paste0("TP=", tp_via, "\n(confirmed)")),
            colour = "white", fontface = "bold", size = 4) +
  labs(
    title = "Cervical Cancer Screening: The Funnel Approach",
    subtitle = paste0("Screen 1000 → VIA+ = ", tp_via + fp_via,
                      " → Biopsy confirms ", tp_via, " true CIN2+ cases"),
    x = NULL, y = "Number of Women"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(size = 11, lineheight = 1.0))
```

## Indian Scenario: TB GeneXpert {.smaller}

::: {.columns}
::: {.column width="50%"}
**Setting:** Rural PHC in Bihar, TB prevalence 3%

**Test:** GeneXpert MTB/RIF — Sens 98%, Spec 99%

```{r}
#| fig-height: 4
#| fig-width: 5

tibble(
  category = c("Sensitivity", "Specificity", "PPV\n(3% prev)", "NPV\n(3% prev)"),
  value = c(98, 99, 75, 99.9),
  fill = c("#3498db", "#3498db", "#e67e22", "#27ae60")
) %>%
  ggplot(aes(x = category, y = value, fill = fill)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(value, "%")), vjust = -0.5, fontface = "bold", size = 5) +
  scale_fill_identity() +
  scale_y_continuous(limits = c(0, 115)) +
  labs(title = "GeneXpert MTB/RIF", y = "%", x = NULL) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```
:::

::: {.column width="50%"}
**Why GeneXpert works at low prevalence:**

- **LR+ = 98** — dramatically shifts probability toward TB
- **PPV = 75%** — useful even at 3% prevalence
- **NPV = 99.9%** — negative GeneXpert powerfully excludes TB
- **LR− = 0.02** — near-perfect rule-out

. . .

::: {.callout-tip}
## The Specificity Secret
99% specificity means only **10 false positives per 1000** non-diseased patients. Compare to 90% specificity → 100 FPs. **High specificity protects PPV at low prevalence.**
:::
:::
:::

## Indian Scenario: Malaria RDT Across Seasons

```{r}
#| fig-height: 5
#| fig-width: 10

mal_sens <- 0.85; mal_spec <- 0.95

seasons <- tibble(
  season = c("Dry Season\n(2% prev)", "Post-Monsoon\n(10%)",
             "Monsoon Peak\n(25%)", "Outbreak\n(50%)"),
  prevalence = c(0.02, 0.10, 0.25, 0.50),
  ppv = sapply(c(0.02, 0.10, 0.25, 0.50), function(p) {
    tp <- p * mal_sens; fp <- (1-p) * (1-mal_spec)
    tp / (tp + fp) * 100
  }),
  npv = sapply(c(0.02, 0.10, 0.25, 0.50), function(p) {
    tn <- (1-p) * mal_spec; fn <- p * (1-mal_sens)
    tn / (tn + fn) * 100
  })
)

seasons_long <- seasons %>%
  pivot_longer(c(ppv, npv), names_to = "metric", values_to = "value") %>%
  mutate(metric = ifelse(metric == "ppv", "PPV", "NPV"))

ggplot(seasons_long, aes(x = season, y = value, fill = metric)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = paste0(round(value), "%")),
            position = position_dodge(width = 0.6),
            vjust = -0.5, fontface = "bold", size = 4) +
  scale_fill_manual(values = c("PPV" = "#e74c3c", "NPV" = "#3498db")) +
  scale_y_continuous(limits = c(0, 110)) +
  labs(
    title = "Malaria RDT: Same Test, Different Seasons (Sens=85%, Spec=95%)",
    subtitle = "PPV swings from 26% to 94%. Dry season positive → confirm with smear. Monsoon positive → treat immediately.",
    x = NULL, y = "%", fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

## Indian Scenario: VIA "See and Treat" Model

```{r}
#| fig-height: 5
#| fig-width: 10

see_treat <- tibble(
  step = factor(c("Women\nScreened\n(1000)", "VIA\nPositive\n(182)",
                  "Cryotherapy\nSame Day\n(182)", "Overtreatment\n(142 FP)",
                  "Cancers\nPrevented\n(~40 TP)"),
                levels = c("Women\nScreened\n(1000)", "VIA\nPositive\n(182)",
                  "Cryotherapy\nSame Day\n(182)", "Overtreatment\n(142 FP)",
                  "Cancers\nPrevented\n(~40 TP)")),
  count = c(1000, tp_via + fp_via, tp_via + fp_via, fp_via, tp_via),
  fill = c("#3498db", "#f39c12", "#e67e22", "#e74c3c", "#27ae60")
)

ggplot(see_treat, aes(x = step, y = count, fill = fill)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = count), vjust = -0.5, fontface = "bold", size = 5) +
  scale_fill_identity() +
  labs(
    title = "VIA 'See and Treat': Accept Overtreatment to Prevent Cancer",
    subtitle = "142 women overtreated (safe cryotherapy) to prevent ~40 cancers. Favourable trade-off in resource-limited settings.",
    x = NULL, y = "Number of Women"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(size = 9, lineheight = 1.0))
```

**The programme logic:** In rural India where follow-up is unreliable, VIA + same-day cryotherapy catches **80% of precancers in a single visit**. Test selection depends on **healthcare context**, not just statistics.

## Key Takeaways

1. **2×2 table** — index test in rows, gold standard in columns, **FP in top-right**

2. **Sensitivity & specificity** are test properties; **PPV & NPV** depend on prevalence

3. **Low prevalence destroys PPV** — even good tests produce mostly false positives

4. **Likelihood ratios** are prevalence-independent: LR+ > 10 = strong rule-in, LR− < 0.1 = strong rule-out

5. **ROC curves** show the full sensitivity-specificity trade-off; **AUC** summarises discrimination

6. **Two-step strategy** — screen (high sensitivity) then confirm (high specificity)

7. **Context matters** — VIA was chosen for India not for best PPV, but for the "see and treat" model

## Further Learning

**Videos:**

- [StatQuest: Sensitivity and Specificity](https://www.youtube.com/watch?v=vP06aMoz4v8)
- [StatQuest: ROC and AUC, Clearly Explained](https://www.youtube.com/watch?v=4jRBRDbJemM)
- [Zedstatistics: Diagnostic Test Evaluation](https://www.youtube.com/watch?v=Z5TtopYXNuc)

**Textbooks:**

- Sackett DL et al. *Clinical Epidemiology*. 3rd ed. Ch 4 (Diagnosis).
- Kirkwood BR. *Essential Medical Statistics*. 2nd ed. Ch 36.
