---
title: "Sampling Methods and Sample Size"
subtitle: "From Population to Sample — How Many and Who?"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    footer: "Biostatistics for Clinicians | Module 6"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
set.seed(42)
```

## Learning Objectives

::: {.incremental}
1. Distinguish **target, accessible, and study populations**
2. Choose among **SRS, systematic, stratified, cluster, and multi-stage** sampling
3. Recognize when **non-probability sampling** is acceptable
4. Explain **sampling error, standard error,** and the **√n law**
5. Calculate sample size for **estimating a proportion**
6. Calculate sample size for **comparing two means or proportions**
7. Adjust for **dropouts, clustering,** and **design effect**
:::

## Clinical Hook: Sampling 1.4 Billion People {.smaller}

::: {.columns}
::: {.column width="55%"}
**ICMR COVID-19 Serosurvey (2020):**

- Goal: Estimate seroprevalence in India
- Population: 1.4 billion people
- Sample: **29,000** individuals
- Ratio: 1 in 48,000

Two fundamental questions:

1. **Who** to sample? → Sampling methods
2. **How many** to sample? → Sample size determination

*How can 29,000 people represent 1.4 billion?*
:::

::: {.column width="45%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5

circle_pts <- function(x0, y0, r, n = 200) {
  theta <- seq(0, 2 * pi, length.out = n)
  data.frame(x = x0 + r * cos(theta), y = y0 + r * sin(theta))
}

ggplot() +
  geom_path(data = circle_pts(0, 0, 3), aes(x, y), color = "#1f77b4", linewidth = 1.2, linetype = "dashed") +
  geom_path(data = circle_pts(0, 0, 2), aes(x, y), color = "#ff7f0e", linewidth = 1.2, linetype = "dashed") +
  geom_polygon(data = circle_pts(0, 0, 0.8), aes(x, y), fill = "#2ca02c", alpha = 0.3, linewidth = 1.2) +
  annotate("text", x = 0, y = 2.6, label = "TARGET\n(All India)", size = 5, fontface = "bold", color = "#1f77b4") +
  annotate("text", x = 0, y = 1.5, label = "ACCESSIBLE\n(Selected districts)", size = 4, fontface = "bold", color = "#ff7f0e") +
  annotate("text", x = 0, y = 0, label = "SAMPLE\n(n = 29,000)", size = 4.5, fontface = "bold", color = "#2ca02c") +
  coord_equal() + xlim(-3.5, 3.5) + ylim(-3.5, 3.5) +
  theme_void() +
  labs(title = "Three Levels of Population")
```
:::
:::

# Part I: Sampling Methods

## Simple Random Sampling (SRS) {.smaller}

::: {.columns}
::: {.column width="50%"}
**Every individual has an equal probability of selection.**

- Use random number table or computer generator
- Gold standard for unbiased selection
- Requires a complete sampling frame (list of all individuals)

**When to use:**

- Small, well-defined populations
- Complete sampling frame available

**Limitation:**

- Impractical for large/geographically dispersed populations
- May miss small subgroups
:::

::: {.column width="50%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5

set.seed(42)
pop <- tibble(x = runif(200), y = runif(200))
srs <- pop %>% slice_sample(n = 20)

ggplot() +
  geom_point(data = pop, aes(x, y), color = "gray70", size = 2) +
  geom_point(data = srs, aes(x, y), color = "#e74c3c", size = 4) +
  labs(title = "Simple Random Sampling", subtitle = "20 selected from 200 (red = selected)") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 14))
```
:::
:::

## Systematic Sampling {.smaller}

::: {.columns}
::: {.column width="50%"}
**Select every k-th individual from the frame.**

1. Calculate interval: k = N/n
2. Pick random start (1 to k)
3. Select every k-th person thereafter

**Example (NFHS):**

- 500 households in a village, need 50
- k = 500/50 = 10
- Random start = 3
- Select: 3, 13, 23, 33, ...

**Advantage:** Simpler than SRS, spreads evenly

**Risk:** Periodicity bias (if list has hidden pattern)
:::

::: {.column width="50%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5

set.seed(42)
sys_pop <- tibble(id = 1:100, x = rep(1:10, 10), y = rep(10:1, each = 10))
sys_selected <- sys_pop %>% filter((id - 3) %% 10 == 0)

ggplot() +
  geom_point(data = sys_pop, aes(x, y), color = "gray70", size = 3) +
  geom_point(data = sys_selected, aes(x, y), color = "#2980b9", size = 5) +
  labs(title = "Systematic Sampling (k = 10)", subtitle = "Every 10th individual starting from position 3") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 14))
```
:::
:::

## Stratified Sampling {.smaller}

::: {.columns}
::: {.column width="50%"}
**Divide population into strata, sample from each.**

1. Define strata (age groups, districts, urban/rural)
2. Sample independently within each stratum
3. Can use proportional or disproportional allocation

**Why stratify?**

- Ensures representation of all subgroups
- Reduces sampling error (more homogeneous within strata)
- Enables stratum-specific estimates

**Example (NFHS-5):**

- Strata: Urban vs Rural within each state
- Sample proportionally from each
:::

::: {.column width="50%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5

set.seed(42)
strat_pop <- tibble(
  x = c(runif(80, 0, 0.45), runif(60, 0.55, 1)),
  y = runif(140),
  stratum = c(rep("Urban", 80), rep("Rural", 60))
)
strat_sel <- strat_pop %>% group_by(stratum) %>% slice_sample(n = 10) %>% ungroup()

ggplot() +
  geom_point(data = strat_pop, aes(x, y, color = stratum), size = 2, alpha = 0.4) +
  geom_point(data = strat_sel, aes(x, y, color = stratum), size = 4) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray40") +
  scale_color_manual(values = c("Urban" = "#e74c3c", "Rural" = "#27ae60")) +
  labs(title = "Stratified Sampling", subtitle = "10 selected from each stratum", color = NULL) +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 14), legend.position = "bottom")
```
:::
:::

## Cluster Sampling {.smaller}

::: {.columns}
::: {.column width="50%"}
**Randomly select whole clusters, sample everyone within.**

1. Divide population into clusters (villages, schools, PHCs)
2. Randomly select clusters
3. Include all individuals in selected clusters

**Why use clusters?**

- No need for complete individual-level frame
- Cheaper — travel to fewer locations
- Standard for WHO EPI surveys

**Disadvantage:**

- Higher sampling error (individuals within clusters are similar)
- Need **design effect** correction for sample size
:::

::: {.column width="50%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5

set.seed(42)
cluster_data <- tibble(
  cluster = rep(1:9, each = 12),
  x = rep(c(1, 2, 3, 1, 2, 3, 1, 2, 3), each = 12) + runif(108, -0.25, 0.25),
  y = rep(c(3, 3, 3, 2, 2, 2, 1, 1, 1), each = 12) + runif(108, -0.25, 0.25),
  selected = cluster %in% c(2, 5, 7)
)

ggplot(cluster_data, aes(x, y, color = selected)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("FALSE" = "gray70", "TRUE" = "#e74c3c"), labels = c("Not selected", "Selected cluster")) +
  labs(title = "Cluster Sampling", subtitle = "3 of 9 clusters randomly selected; all individuals within sampled", color = NULL) +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 14), legend.position = "bottom")
```
:::
:::

## Multi-Stage Sampling {.smaller}

**The real-world approach — used by NFHS and ICMR.**

```{r}
#| fig-height: 4
#| fig-width: 11

stages <- tibble(
  x = c(1, 3, 5, 7, 9),
  y = rep(1, 5),
  label = c("All States\n(28 + 8 UTs)", "Select\nDistricts", "Select\nVillages/Wards", "Select\nHouseholds", "Select\nIndividuals"),
  stage = paste("Stage", 1:5)
)

ggplot(stages) +
  geom_rect(aes(xmin = x - 0.7, xmax = x + 0.7, ymin = 0.5, ymax = 1.5),
            fill = c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd"), alpha = 0.8, color = "white", linewidth = 1) +
  geom_text(aes(x = x, y = 1, label = label), color = "white", fontface = "bold", size = 4.5) +
  geom_segment(data = tibble(x = c(1.7, 3.7, 5.7, 7.7), xend = c(2.3, 4.3, 6.3, 8.3)),
               aes(x = x, xend = xend, y = 1, yend = 1),
               arrow = arrow(length = unit(0.3, "cm")), linewidth = 1.5, color = "gray40") +
  xlim(0, 10) + ylim(0, 2) +
  labs(title = "Multi-Stage Sampling: NFHS-5 Design") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5))
```

**Each stage introduces sampling error** — multi-stage designs require larger samples than SRS to achieve the same precision.

## Sampling Methods: Summary {.smaller}

```{r}
methods <- tibble(
  Method = c("Simple Random", "Systematic", "Stratified", "Cluster", "Multi-Stage"),
  `How It Works` = c("Random selection from complete list",
                     "Every k-th person from list",
                     "Divide into strata, sample each",
                     "Select whole clusters, sample all within",
                     "Hierarchical selection at multiple levels"),
  `Best For` = c("Small well-defined populations",
                 "Large lists without subgroups",
                 "Ensuring subgroup representation",
                 "Large geographic spread, no frame",
                 "National surveys (NFHS, ICMR)"),
  Precision = c("High", "High", "Highest", "Lower", "Lower")
)

kable(methods, format = "html") %>%
  kable_styling(full_width = TRUE, font_size = 18) %>%
  row_spec(0, background = "#2c3e50", color = "white", bold = TRUE) %>%
  column_spec(1, bold = TRUE, width = "15%") %>%
  column_spec(4, width = "12%")
```

## Non-Probability Sampling {.smaller}

| Method | How | When Acceptable | Risk |
|--------|-----|-----------------|------|
| **Convenience** | Whoever is available | Pilot studies, feasibility | Selection bias |
| **Purposive** | Researcher selects specific cases | Qualitative, rare diseases | Not generalizable |
| **Snowball** | Participants recruit others | Hidden populations (MSM, PWID) | Network bias |
| **Quota** | Fill predefined categories | Market research, quick surveys | Subjective selection |

. . .

**Key rule:** Non-probability sampling **cannot support statistical inference** (no valid CIs or p-values). Use only when probability sampling is impossible.

# Part II: Sampling Error & Sample Size

## The Standard Error (SE) {.smaller}

::: {.columns}
::: {.column width="50%"}
**SE quantifies typical sampling error.**

$$SE_{\bar{x}} = \frac{\sigma}{\sqrt{n}}$$

$$SE_{\hat{p}} = \sqrt{\frac{p(1-p)}{n}}$$

**Key insight (CLT):**

- Sampling error is random and follows a normal distribution
- SE decreases with √n — the **√n law**
- Doubling precision requires **4× the sample size**
:::

::: {.column width="50%"}
```{r}
#| fig-height: 5.5
#| fig-width: 5.5

population <- rgamma(100000, shape = 2, rate = 0.5)
pop_sd <- sd(population)

n_vals <- seq(5, 500, by = 5)
se_vals <- pop_sd / sqrt(n_vals)

tibble(n = n_vals, SE = se_vals) %>%
  ggplot(aes(n, SE)) +
  geom_line(linewidth = 1.5, color = "#2980b9") +
  geom_point(data = tibble(n = c(10, 30, 100, 200), SE = pop_sd / sqrt(c(10, 30, 100, 200))),
             aes(n, SE), size = 4, color = "#c0392b") +
  geom_text(data = tibble(n = c(10, 30, 100, 200), SE = pop_sd / sqrt(c(10, 30, 100, 200)),
                          lab = paste0("n=", c(10,30,100,200))),
            aes(label = lab), vjust = -1, size = 4.5, fontface = "bold") +
  labs(title = "The √n Law", subtitle = "Diminishing returns with increasing n",
       x = "Sample Size (n)", y = "Standard Error") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```
:::
:::

## Margin of Error and Confidence Intervals {.smaller}

**Margin of Error (MOE)** = how far our estimate might be from the truth.

$$MOE = z_{\alpha/2} \times SE$$

**For 95% confidence:** MOE = 1.96 × SE

. . .

**Example:** HbA1c in 64 diabetic patients, mean = 11.5, SD = 1.6

- SE = 1.6 / √64 = 0.2
- MOE = 1.96 × 0.2 = **0.39 g/dL**
- 95% CI: 11.5 ± 0.39 = **(11.11, 11.89)**

. . .

> "We are 95% confident the true mean HbA1c is between 11.11 and 11.89 g/dL."

## Sample Size for Estimating a Proportion {.smaller}

::: {.columns}
::: {.column width="50%"}
$$n = \frac{z^2_{\alpha/2} \times p(1-p)}{E^2}$$

**Where:**

- $z_{\alpha/2}$ = 1.96 (for 95% CI)
- $p$ = expected proportion
- $E$ = desired margin of error

**If p is unknown:** Use p = 0.5 (maximizes sample size — conservative)
:::

::: {.column width="50%"}
**Worked Example:**

Estimate diabetes prevalence in Bhopal.

- Expected: p = 0.10 (10%)
- Precision: E = 0.02 (±2%)
- Confidence: 95% (z = 1.96)

$$n = \frac{1.96^2 \times 0.10 \times 0.90}{0.02^2}$$

$$n = \frac{3.8416 \times 0.09}{0.0004} = \frac{0.346}{0.0004}$$

$$\boxed{n = 864}$$
:::
:::

## Type I and Type II Errors {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 11

x <- seq(-4, 6, length.out = 1000)
h0 <- dnorm(x, 0, 1)
h1 <- dnorm(x, 2.5, 1)
crit <- qnorm(0.95)

df <- tibble(x = x, h0 = h0, h1 = h1)

ggplot(df, aes(x = x)) +
  geom_area(aes(y = h0), fill = "#3498db", alpha = 0.4) +
  geom_area(aes(y = h1), fill = "#e74c3c", alpha = 0.4) +
  geom_area(data = df %>% filter(x >= crit), aes(y = h0), fill = "#c0392b", alpha = 0.7) +
  geom_area(data = df %>% filter(x < crit), aes(y = h1), fill = "#f39c12", alpha = 0.7) +
  geom_vline(xintercept = crit, linetype = "dashed", linewidth = 1.2, color = "#2c3e50") +
  annotate("label", x = -2, y = 0.35, label = "H₀: No effect", fill = "#3498db", color = "white", fontface = "bold", size = 5) +
  annotate("label", x = 4, y = 0.35, label = "H₁: Real effect", fill = "#e74c3c", color = "white", fontface = "bold", size = 5) +
  annotate("label", x = 2.5, y = 0.15, label = "α (Type I)\nFalse Positive", fill = "#c0392b", color = "white", fontface = "bold", size = 4) +
  annotate("label", x = -0.5, y = 0.15, label = "β (Type II)\nFalse Negative", fill = "#f39c12", color = "white", fontface = "bold", size = 4) +
  labs(title = "Type I and Type II Errors", subtitle = "α = P(reject H₀ | H₀ true)  •  β = P(fail to reject H₀ | H₁ true)  •  Power = 1 − β",
       x = "Test Statistic", y = "Density") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"), axis.text.y = element_blank())
```

## The Four Outcomes {.smaller}

```{r}
outcomes <- tibble(
  ` ` = c("**H₀ TRUE** (No effect)", "**H₀ FALSE** (Effect exists)"),
  `Reject H₀` = c("❌ **Type I Error (α)**\nFalse Positive\nP = 0.05",
                   "✅ **Correct** (Power)\nTrue Positive\nP = 1 − β = 0.80"),
  `Fail to Reject H₀` = c("✅ **Correct**\nTrue Negative\nP = 1 − α = 0.95",
                            "❌ **Type II Error (β)**\nFalse Negative\nP = 0.20")
)

kable(outcomes, format = "html", escape = FALSE) %>%
  kable_styling(full_width = TRUE, font_size = 18) %>%
  row_spec(0, background = "#2c3e50", color = "white", bold = TRUE) %>%
  column_spec(1, width = "25%", bold = TRUE)
```

. . .

**Key trade-off:** Decreasing α increases β. **Solution:** increase sample size!

**Convention:** α = 0.05, β = 0.20 (Power = 80%)

## Sample Size for Comparing Two Means {.smaller}

::: {.columns}
::: {.column width="50%"}
$$n_{per\ group} = \frac{2(z_{\alpha/2} + z_{\beta})^2 \times \sigma^2}{\delta^2}$$

**Where:**

- $z_{\alpha/2}$ = 1.96 (α = 0.05, two-tailed)
- $z_{\beta}$ = 0.84 (β = 0.20, power = 80%)
- σ = common SD in both groups
- δ = difference to detect
:::

::: {.column width="50%"}
**Worked Example:**

New antihypertensive vs standard.

- Detect: δ = 10 mmHg difference
- SD: σ = 15 mmHg
- Power: 80%, α = 0.05

$$n = \frac{2(1.96 + 0.84)^2 \times 15^2}{10^2}$$

$$= \frac{2 \times 7.84 \times 225}{100}$$

$$\boxed{n = 36 \text{ per group}}$$
:::
:::

## Sample Size for Comparing Two Proportions {.smaller}

::: {.columns}
::: {.column width="50%"}
$$n = \frac{2(z_{\alpha/2} + z_{\beta})^2 \times [p_1(1-p_1) + p_2(1-p_2)]}{(p_2 - p_1)^2}$$

**Where:**

- $p_1$ = proportion in control group
- $p_2$ = proportion in treatment group
:::

::: {.column width="50%"}
**Worked Example:**

TB cure rate: 70% → 85%.

$$n = \frac{2 \times 7.84 \times [0.70 \times 0.30 + 0.85 \times 0.15]}{(0.85 - 0.70)^2}$$

$$= \frac{2 \times 7.84 \times [0.21 + 0.1275]}{0.0225}$$

$$= \frac{2 \times 7.84 \times 0.3375}{0.0225}$$

$$\boxed{n \approx 94 \text{ per group}}$$
:::
:::

## Power Curves {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 11

sample_sizes <- seq(10, 300, by = 5)

power_d3 <- sapply(sample_sizes, function(n) pnorm(0.3 * sqrt(n/2) - qnorm(0.975)))
power_d5 <- sapply(sample_sizes, function(n) pnorm(0.5 * sqrt(n/2) - qnorm(0.975)))
power_d8 <- sapply(sample_sizes, function(n) pnorm(0.8 * sqrt(n/2) - qnorm(0.975)))

tibble(n = rep(sample_sizes, 3),
       Power = pmin(c(power_d3, power_d5, power_d8), 1),
       Effect = rep(c("d = 0.3 (Small)", "d = 0.5 (Medium)", "d = 0.8 (Large)"), each = length(sample_sizes))) %>%
  ggplot(aes(n, Power, color = Effect)) +
  geom_line(linewidth = 1.3) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray40") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  scale_color_manual(values = c("#e74c3c", "#27ae60", "#2980b9")) +
  annotate("text", x = 270, y = 0.83, label = "80% Power", fontface = "bold", color = "gray40", size = 5) +
  labs(title = "Power Curves: Larger Effects Need Fewer Subjects",
       subtitle = "α = 0.05 (two-tailed) | Minimum n per group to achieve 80% power",
       x = "Sample Size (per group)", y = "Statistical Power", color = NULL) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom", legend.text = element_text(size = 13))
```

# Part III: Practical Adjustments

## Adjusting for Dropouts {.smaller}

Not all enrolled subjects complete the study. Adjust upward:

$$n_{adjusted} = \frac{n}{1 - \text{dropout rate}}$$

. . .

**Example:** TB drug trial needs 94 per group. Expected dropout = 15%.

$$n_{adjusted} = \frac{94}{1 - 0.15} = \frac{94}{0.85} \approx 111 \text{ per group}$$

. . .

**Always report both:** "We calculated n = 94 per group, inflated to 111 to account for 15% expected dropout."

## Design Effect for Cluster Sampling {.smaller}

Cluster sampling inflates variance because individuals within clusters are similar.

$$DE = 1 + (m - 1) \times \rho$$

::: {.columns}
::: {.column width="50%"}
**Where:**

- m = cluster size
- ρ = intracluster correlation (ICC)

$$n_{adjusted} = n_{SRS} \times DE$$
:::

::: {.column width="50%"}
**Example:**

- SRS sample size: 36 per group
- Cluster size: m = 12 patients per PHC
- ICC: ρ = 0.02

$$DE = 1 + (12 - 1) \times 0.02 = 1.22$$

$$n_{adjusted} = 36 \times 1.22 \approx 44 \text{ per group}$$
:::
:::

. . .

**Key insight:** Even small ICC (ρ = 0.02) with large clusters can substantially inflate required n.

## Sample Size Decision Flowchart {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 11

flow <- tibble(
  x = c(5.5, 3, 8, 1.5, 4.5, 6.5, 9.5),
  y = c(5, 3, 3, 1, 1, 1, 1),
  label = c("What is your\nstudy objective?",
            "ESTIMATION\n(single group)", "COMPARISON\n(two+ groups)",
            "Proportion?\nn = z²p(1-p)/E²", "Mean?\nn = z²σ²/E²",
            "Two means?\nn = 2(zα+zβ)²σ²/δ²", "Two proportions?\nSpecific formula"),
  fill = c("#2c3e50", "#3498db", "#e74c3c", "#3498db", "#3498db", "#e74c3c", "#e74c3c")
)

arrows <- tibble(
  x = c(5.5, 5.5, 3, 3, 8, 8),
  xend = c(3, 8, 1.5, 4.5, 6.5, 9.5),
  y = c(4.6, 4.6, 2.6, 2.6, 2.6, 2.6),
  yend = c(3.4, 3.4, 1.4, 1.4, 1.4, 1.4)
)

ggplot() +
  geom_rect(data = flow, aes(xmin = x-1.1, xmax = x+1.1, ymin = y-0.5, ymax = y+0.5, fill = fill),
            color = "white", linewidth = 1) +
  scale_fill_identity() +
  geom_text(data = flow, aes(x = x, y = y, label = label), color = "white", fontface = "bold", size = 3.8) +
  geom_segment(data = arrows, aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.3, "cm")), linewidth = 1, color = "gray40") +
  xlim(-0.5, 11.5) + ylim(0, 6) +
  labs(title = "Sample Size Decision Flowchart") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5))
```

Then adjust for: **dropouts** → **design effect** → **finite population correction**

## Key Takeaways {.smaller}

::: {.incremental}
1. **Probability sampling** enables valid inference; non-probability does not
2. **Stratified** sampling gives the highest precision; **cluster** trades precision for feasibility
3. **SE = σ/√n** — the √n law means diminishing returns with larger samples
4. **Sample size depends on:** desired precision, expected variability, effect size, α, and power
5. **For estimation:** n = z²p(1−p)/E² (proportions) or n = z²σ²/E² (means)
6. **For comparison:** also need effect size (δ) and power (1−β)
7. **Always adjust** for dropouts (inflate by 1/(1−dropout)) and clustering (multiply by DE)
8. **Calculate sample size BEFORE the study** — underpowered studies are unethical
:::

## Formula Quick Reference {.smaller}

| Purpose | Formula | Key Inputs |
|---------|---------|------------|
| **Estimate proportion** | $n = z^2 p(1-p) / E^2$ | p, margin of error |
| **Estimate mean** | $n = z^2 \sigma^2 / E^2$ | σ, margin of error |
| **Compare two means** | $n = 2(z_\alpha + z_\beta)^2 \sigma^2 / \delta^2$ | σ, effect size, power |
| **Compare two proportions** | See formula above | p₁, p₂, power |
| **Dropout adjustment** | $n_{adj} = n / (1 - d)$ | dropout rate |
| **Design effect** | $DE = 1 + (m-1)\rho$ | cluster size, ICC |

. . .

**Remember:** z₀.₀₂₅ = 1.96 (95% CI) • z₀.₂₀ = 0.84 (80% power) • z₀.₁₀ = 1.28 (90% power)
