---
title: "Study Design, Bias, and Confounding"
subtitle: "Why Study Design Determines What You Can Conclude"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    footer: "Biostatistics for Clinicians | Module 7"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
library(DiagrammeR)
set.seed(42)
```

## Learning Objectives

::: {.incremental}
1. Describe the **purpose, logic, and key features** of cross-sectional, case-control, cohort, and RCT designs
2. Choose the **right design** for a given research question
3. Calculate **prevalence, OR, RR, and AR** from 2×2 tables
4. Distinguish **selection bias, information bias,** and **confounding**
5. Use **DAGs** to reason about confounding
6. Evaluate the **hierarchy of evidence**
:::

## Clinical Hook: Does Wine Prevent Heart Disease? {.smaller}

::: {.columns}
::: {.column width="55%"}
A 10-year cohort study (n = 20,000) found:

**Light wine drinkers had 30% lower MI risk** vs non-drinkers.

Headlines: *"Wine is Cardioprotective!"*

. . .

**But wait —** wine drinkers also:

- Had higher incomes (better healthcare access)
- Exercised more regularly
- Had lower smoking rates
- Ate more fruits and vegetables

**The wine–MI association may be entirely due to confounding by socioeconomic status and lifestyle.**

This module teaches you to think critically about *how* a study was designed before trusting *what* it found.
:::

::: {.column width="45%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5
grViz("
digraph wine_confound {
  graph[rankdir = TB, bgcolor = 'transparent', nodesep = 1.2, ranksep = 0.8]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 12, margin = '0.25,0.12']

  C [label = 'SES & Lifestyle\\n(Confounder)', fillcolor = '#E8C07A', fontcolor = '#333']
  E [label = 'Wine\\nConsumption', fillcolor = '#B8D4E3', fontcolor = '#333']
  O [label = 'Heart\\nDisease', fillcolor = '#E89595', fontcolor = '#333']

  C -> E [label = '  causes  ', fontsize = 10, fontcolor = '#555']
  C -> O [label = '  causes  ', fontsize = 10, fontcolor = '#555']
  E -> O [label = '  ??  ', fontsize = 10, fontcolor = '#c0392b', style = dashed]
}
")
```
:::
:::

## Cross-Sectional Study {.smaller}

**Purpose:** Snapshot — measure exposure and outcome **at the same time**

::: {.columns}
::: {.column width="55%"}
**How it works:**

1. Select a sample from the population
2. Measure exposure and outcome simultaneously
3. Compare prevalence of outcome in exposed vs non-exposed

**Key features:**

- Quick, inexpensive
- Cannot establish temporal sequence (chicken or egg?)
- Best for: prevalence estimation, hypothesis generation

**Measure:** Prevalence, Prevalence Odds Ratio (POR)
:::

::: {.column width="45%"}
```{r}
#| fig-height: 4.5
#| fig-width: 5.5
grViz("
digraph cross_sectional {
  graph[rankdir = LR, bgcolor = 'transparent', nodesep = 0.5, ranksep = 0.7]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 10, margin = '0.15,0.08']

  A [label = 'Population\\n(one time point)', fillcolor = '#B8D4E3']
  B [label = 'Exposed', fillcolor = '#E8C07A']
  C [label = 'Non-exposed', fillcolor = '#A8D5A2']
  D [label = 'Outcome +', fillcolor = '#E89595']
  E [label = 'Outcome −', fillcolor = '#D4EAF7']
  F [label = 'Outcome +', fillcolor = '#E89595']
  G [label = 'Outcome −', fillcolor = '#D4EAF7']

  edge [fontsize = 8, fontcolor = '#555']
  A -> B [label = '  classify  ']
  A -> C [label = '  classify  ']
  B -> D [label = '  assess  ']
  B -> E [label = '  assess  ']
  C -> F [label = '  assess  ']
  C -> G [label = '  assess  ']
}
")
```

**Example:** Measuring BP and salt intake in 500 adults at a health camp
:::
:::

## Case-Control Study {.smaller}

**Purpose:** Start with the **disease** and look **backward** for exposure

::: {.columns}
::: {.column width="55%"}
**How it works:**

1. Identify people WITH the disease (cases)
2. Select similar people WITHOUT the disease (controls)
3. Compare past exposure in cases vs controls

**Key features:**

- Efficient for **rare diseases** (don't need to wait for events)
- Relatively quick and cheap
- Prone to **recall bias** (cases remember exposures better)
- Cannot calculate incidence or RR directly

**Measure:** Odds Ratio (OR = ad/bc)
:::

::: {.column width="45%"}
```{r}
#| fig-height: 4.5
#| fig-width: 5.5
grViz("
digraph case_control {
  graph[rankdir = LR, bgcolor = 'transparent', nodesep = 0.5, ranksep = 0.7]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 10, margin = '0.15,0.08']

  A [label = 'Study\\nBase', fillcolor = '#B8D4E3']
  B [label = 'Cases\\n(Disease +)', fillcolor = '#E89595']
  C [label = 'Controls\\n(Disease −)', fillcolor = '#D4EAF7']
  D [label = 'Exposed', fillcolor = '#E8C07A']
  E [label = 'Not exposed', fillcolor = '#A8D5A2']
  F [label = 'Exposed', fillcolor = '#E8C07A']
  G [label = 'Not exposed', fillcolor = '#A8D5A2']

  edge [fontsize = 8, fontcolor = '#555']
  A -> B [label = '  select  ']
  A -> C [label = '  select  ']
  B -> D [label = '  ascertain  ']
  B -> E [label = '  ascertain  ']
  C -> F [label = '  ascertain  ']
  C -> G [label = '  ascertain  ']
}
")
```

**Example:** 100 oral cancer patients + 200 controls — asked about paan chewing history
:::
:::

## Cohort Study {.smaller}

**Purpose:** Start with **exposure** and follow **forward** to see who develops disease

::: {.columns}
::: {.column width="55%"}
**How it works:**

1. Identify exposed and non-exposed groups (disease-free at start)
2. Follow both groups over time
3. Compare incidence of outcome in exposed vs non-exposed

**Key features:**

- Establishes **temporal sequence** (exposure precedes outcome)
- Can calculate **incidence** and **RR** directly
- Efficient for **rare exposures**
- Expensive and time-consuming (especially prospective)
- Loss to follow-up is the main threat

**Measure:** Relative Risk (RR), Absolute Risk, Incidence
:::

::: {.column width="45%"}
```{r}
#| fig-height: 4.5
#| fig-width: 5.5
grViz("
digraph cohort {
  graph[rankdir = LR, bgcolor = 'transparent', nodesep = 0.5, ranksep = 0.7]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 10, margin = '0.15,0.08']

  A [label = 'Disease-free\\nPopulation', fillcolor = '#B8D4E3']
  B [label = 'Exposed', fillcolor = '#E8C07A']
  C [label = 'Non-exposed', fillcolor = '#A8D5A2']
  D [label = 'Disease +', fillcolor = '#E89595']
  E [label = 'Disease −', fillcolor = '#D4EAF7']
  F [label = 'Disease +', fillcolor = '#E89595']
  G [label = 'Disease −', fillcolor = '#D4EAF7']

  edge [fontsize = 8, fontcolor = '#555']
  A -> B [label = '  ascertain  ']
  A -> C [label = '  ascertain  ']
  B -> D [label = '  follow-up  ']
  B -> E [label = '  follow-up  ']
  C -> F [label = '  follow-up  ']
  C -> G [label = '  follow-up  ']
}
")
```

**Example:** Follow 5,000 factory workers (smokers vs non-smokers) for 10 years for lung disease
:::
:::

## Randomized Controlled Trial (RCT) {.smaller}

**Purpose:** The **gold standard** — test whether an intervention *causes* the outcome

::: {.columns}
::: {.column width="55%"}
**How it works:**

1. Randomly assign participants to treatment or control
2. Apply intervention
3. Follow up and compare outcomes

**Key features:**

- **Randomization** distributes ALL confounders (known and unknown) equally
- **Only design that can establish causation**
- Highest internal validity
- Expensive, time-consuming, not always ethical or feasible
- Requires equipoise (genuine uncertainty about which is better)

**Measure:** RR, ARR, NNT
:::

::: {.column width="45%"}
```{r}
#| fig-height: 4.5
#| fig-width: 5.5
grViz("
digraph rct {
  graph[rankdir = LR, bgcolor = 'transparent', nodesep = 0.5, ranksep = 0.7]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 10, margin = '0.15,0.08']

  A [label = 'Eligible\\nParticipants', fillcolor = '#B8D4E3']
  B [label = 'Treatment', fillcolor = '#E8C07A']
  C [label = 'Control', fillcolor = '#A8D5A2']
  D [label = 'Outcome +', fillcolor = '#E89595']
  E [label = 'Outcome −', fillcolor = '#D4EAF7']
  F [label = 'Outcome +', fillcolor = '#E89595']
  G [label = 'Outcome −', fillcolor = '#D4EAF7']

  edge [fontsize = 8, fontcolor = '#c0392b', fontname = 'Helvetica-Bold']
  A -> B [label = '  randomize  ']
  A -> C [label = '  randomize  ']
  edge [fontsize = 8, fontcolor = '#555', fontname = 'Helvetica']
  B -> D [label = '  follow-up  ']
  B -> E [label = '  follow-up  ']
  C -> F [label = '  follow-up  ']
  C -> G [label = '  follow-up  ']
}
")
```

**Example:** Randomly assign 200 diabetics to yoga vs standard care for 6 months
:::
:::

## Design Comparison at a Glance {.smaller}

```{r}
comparison <- tibble(
  Feature = c("Starting point", "Time direction", "Measure", "Can show causation?",
              "Best for rare...", "Time & cost", "Main threat"),
  `Cross-sectional` = c("Population", "Snapshot", "Prevalence, POR", "No", "Neither", "Low", "Temporal ambiguity"),
  `Case-control` = c("Outcome", "Backward", "OR", "Suggests", "**Disease**", "Low", "Recall bias"),
  Cohort = c("Exposure", "Forward", "RR, AR", "Suggests", "**Exposure**", "High", "Loss to follow-up"),
  RCT = c("Random assignment", "Forward", "RR, ARR, NNT", "**Yes**", "N/A", "Very high", "Ethical constraints")
)

kable(comparison, format = "html", escape = FALSE) %>%
  kable_styling(full_width = TRUE, font_size = 15) %>%
  row_spec(0, background = "#2c3e50", color = "white", bold = TRUE) %>%
  column_spec(1, bold = TRUE, width = "16%")
```

. . .

**Key identifier:** What is the starting point? Disease → case-control. Exposure → cohort. Random allocation → RCT.

## How to Choose a Study Design {.smaller}

| Research Question | Best Design | Why |
|---|---|---|
| What is the prevalence of diabetes in rural MP? | Cross-sectional | Snapshot of a population |
| Is gutka associated with oral cancer? | Case-control | Oral cancer is rare; start from cases |
| Does air pollution increase asthma in children? | Cohort | Follow exposed vs non-exposed over time |
| Does a new drug lower BP more than standard? | RCT | Need causal evidence for treatment |
| Is a screening test effective at reducing mortality? | RCT (or large cohort) | Lead-time bias threatens observational designs |

. . .

**The choice depends on:** research question, outcome frequency, time, ethics, and resources.

## Selection Bias {.smaller}

**Systematic error in how participants are selected or retained.**

::: {.columns}
::: {.column width="50%"}
**Common types:**

- **Berkson's bias:** Hospital controls overrepresent comorbidities
- **Volunteer bias:** Participants differ from non-participants
- **Loss to follow-up:** Differential dropout distorts results
- **Healthy worker effect:** Workers are healthier than general population
:::

::: {.column width="50%"}
**Example:**

Study of phone radiation and brain cancer using hospital controls.

- Hospital controls may have headache disorders (more phone use)
- Spurious protective association!
- **Fix:** Use population-based controls
:::
:::

. . .

**Key principle:** Selection bias cannot be corrected after data collection. **It must be prevented by design.**

## Information (Measurement) Bias {.smaller}

**Systematic error in how exposure or outcome is measured.**

| Type | What Happens | Example |
|------|-------------|---------|
| **Recall bias** | Cases remember exposure better than controls | Mothers of malformed babies recall drug use more |
| **Observer bias** | Assessor knows exposure status | Unblinded radiologist reads scans differently |
| **Misclassification** | Wrong category assignment | BP cuff too small → systematic overestimation |
| **Hawthorne effect** | Behavior changes when observed | Hand hygiene improves during audit |

. . .

**Solutions:** Blinding, standardized instruments, objective measures, validated questionnaires

## Lead-Time Bias {.smaller}

```{r}
#| fig-height: 5
#| fig-width: 11

ggplot() +
  geom_segment(aes(x = 2010, xend = 2020, y = 2, yend = 2), linewidth = 2, color = "#3498db") +
  geom_segment(aes(x = 2010, xend = 2020, y = 1, yend = 1), linewidth = 2, color = "#e74c3c") +
  geom_point(aes(x = 2015, y = 2), size = 5, color = "#2c3e50") +
  geom_point(aes(x = 2012, y = 1), size = 5, color = "#2c3e50") +
  geom_point(aes(x = 2020, y = c(1, 2)), size = 5, shape = 4, stroke = 3, color = "#2c3e50") +
  annotate("text", x = 2015, y = 2.3, label = "Symptom diagnosis\n(2015)", fontface = "bold", size = 4) +
  annotate("text", x = 2012, y = 0.7, label = "Screen diagnosis\n(2012)", fontface = "bold", size = 4) +
  annotate("text", x = 2020.5, y = 1.5, label = "Both die\n2020", fontface = "bold", size = 4, color = "#c0392b") +
  annotate("segment", x = 2012, xend = 2015, y = 1.5, yend = 1.5, linewidth = 1.5, color = "#f39c12",
           arrow = arrow(ends = "both", length = unit(0.2, "cm"))) +
  annotate("text", x = 2013.5, y = 1.7, label = "Lead time\n(3 years)", fontface = "bold", size = 4, color = "#f39c12") +
  annotate("text", x = 2017.5, y = 2.3, label = "Survival = 5 years", size = 4, color = "#3498db") +
  annotate("text", x = 2016, y = 0.7, label = "Survival = 8 years", size = 4, color = "#e74c3c") +
  scale_x_continuous(breaks = 2010:2020) +
  ylim(0.3, 2.8) +
  labs(title = "Lead-Time Bias: Screening Looks Better Than It Is",
       subtitle = "Both patients die at the same time — screening just detects earlier",
       x = "Year", y = NULL) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"), axis.text.y = element_blank(), panel.grid.major.y = element_blank())
```

## What Is Confounding? {.smaller}

A **confounder** is a variable that:

1. Is associated with the **exposure**
2. Is an independent cause of the **outcome**
3. Is NOT on the causal pathway between exposure and outcome

```{r}
#| fig-height: 3.5
#| fig-width: 8
grViz("
digraph confound_generic {
  graph[rankdir = TB, bgcolor = 'transparent', nodesep = 1.0, ranksep = 0.8]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 13, margin = '0.3,0.15']

  C [label = 'CONFOUNDER\\n(C)', fillcolor = '#E8C07A', fontcolor = '#333']
  E [label = 'EXPOSURE\\n(E)', fillcolor = '#B8D4E3', fontcolor = '#333']
  O [label = 'OUTCOME\\n(O)', fillcolor = '#E89595', fontcolor = '#333']

  C -> E [label = '  associated  ', fontsize = 10, fontcolor = '#555']
  C -> O [label = '  causes  ', fontsize = 10, fontcolor = '#555']
  E -> O [label = '  ??  ', fontsize = 10, fontcolor = '#c0392b', style = dashed, penwidth = 2]
}
")
```

**Example:** SES confounds the wine → heart disease association (see clinical hook)

## The Wine Paradox DAG {.smaller}

```{r}
#| fig-height: 4.5
#| fig-width: 10
grViz("
digraph wine {
  graph[rankdir = TB, bgcolor = 'transparent', nodesep = 1.2, ranksep = 0.8]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 12, margin = '0.3,0.15']

  SES [label = 'Socioeconomic\\nStatus & Lifestyle', fillcolor = '#E8C07A']
  Wine [label = 'Wine\\nConsumption', fillcolor = '#B8D4E3']
  CVD [label = 'Cardiovascular\\nDisease', fillcolor = '#E89595']
  Ex [label = 'Exercise', fillcolor = '#D4EAF7']
  Diet [label = 'Diet Quality', fillcolor = '#D4EAF7']
  Access [label = 'Healthcare\\nAccess', fillcolor = '#D4EAF7']

  SES -> Wine [label = '  ↑income → ↑wine  ', fontsize = 9, fontcolor = '#555']
  SES -> Ex [fontsize = 9]
  SES -> Diet [fontsize = 9]
  SES -> Access [fontsize = 9]
  Ex -> CVD [label = '  protects  ', fontsize = 9, fontcolor = '#27ae60']
  Diet -> CVD [label = '  protects  ', fontsize = 9, fontcolor = '#27ae60']
  Access -> CVD [label = '  protects  ', fontsize = 9, fontcolor = '#27ae60']
  Wine -> CVD [label = '  ??  ', fontsize = 10, fontcolor = '#c0392b', style = dashed, penwidth = 2]
}
")
```

**After adjusting for SES and lifestyle, the wine–CVD association largely disappears.**

## Methods to Control Confounding {.smaller}

| Stage | Method | How It Works |
|-------|--------|-----|
| **Design** | Randomization | Distributes ALL confounders equally (known + unknown) |
| **Design** | Restriction | Limit to one stratum (e.g., only males) |
| **Design** | Matching | Select controls with same confounder values |
| **Analysis** | Stratification | Analyze within confounder strata separately |
| **Analysis** | Multivariable regression | Adjust for confounders in model |

. . .

**Only randomization controls for UNKNOWN confounders.** All other methods require you to identify and measure confounders in advance.

## Hierarchy of Evidence {.smaller}

```{r}
grViz("
digraph evidence_pyramid {
  graph[rankdir = TB, bgcolor = 'transparent', nodesep = 0.3, ranksep = 0.4]
  node [fontname = 'Helvetica', shape = box, style = filled, fontsize = 11, margin = '0.2,0.1']

  SR  [label = 'Systematic Reviews\\n& Meta-analyses', fillcolor = '#27ae60', fontcolor = 'white', width = 3]
  RCT [label = 'Randomized Controlled Trials', fillcolor = '#2ecc71', fontcolor = 'white', width = 4]
  COH [label = 'Cohort Studies', fillcolor = '#f1c40f', fontcolor = '#333', width = 5]
  CC  [label = 'Case-Control Studies', fillcolor = '#e67e22', fontcolor = 'white', width = 6]
  CS  [label = 'Cross-Sectional Surveys', fillcolor = '#e74c3c', fontcolor = 'white', width = 7]
  CR  [label = 'Case Reports / Expert Opinion', fillcolor = '#c0392b', fontcolor = 'white', width = 8]

  SR -> RCT [arrowhead = none]
  RCT -> COH [arrowhead = none]
  COH -> CC [arrowhead = none]
  CC -> CS [arrowhead = none]
  CS -> CR [arrowhead = none]
}
")
```

**Higher = stronger evidence for causation** — but not always feasible or ethical!

## Exercise: Identify the Design {.smaller}

For each scenario, identify the study design:

::: {.incremental}
1. *"Measured BP and salt intake in 500 adults at a health camp"* → **Cross-sectional** (snapshot of exposure + outcome)

2. *"Identified 100 oral cancer patients and 200 controls, asked about paan chewing"* → **Case-control** (start from disease, look back at exposure)

3. *"Followed 5,000 factory workers for 10 years, comparing smokers vs non-smokers for lung disease"* → **Cohort** (classify by exposure, follow forward for outcome)

4. *"Randomly assigned 200 diabetics to yoga vs standard care for 6 months"* → **RCT** (random allocation of intervention)
:::

## The Sweden–India Paradox {.smaller}

::: {.columns}
::: {.column width="50%"}
| Country | Crude Death Rate | Life Expectancy |
|---------|:---:|:---:|
| India | **6.4** | 70.8 years |
| Sweden | **9.0** | 83.2 years |

Sweden's crude death rate is **41% higher** — yet Swedes live 12 years longer!

. . .

**Why?** Sweden has 29% of its population aged 60+, vs 18% in India. More elderly = more deaths = higher crude rate.

**Age is confounding the comparison.**
:::

::: {.column width="50%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5
age_df <- tibble(
  age_group = factor(rep(c("0–14", "15–44", "45–59", "60–74", "75+"), 2),
                     levels = c("0–14", "15–44", "45–59", "60–74", "75+")),
  country = rep(c("India", "Sweden"), each = 5),
  pct = c(24.2, 41.4, 16.0, 12.0, 6.4,
          17.2, 35.3, 18.5, 16.0, 13.0)
)
ggplot(age_df, aes(x = age_group, y = pct, fill = country)) +
  geom_col(position = "dodge", color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c("India" = "#FF9933", "Sweden" = "#006AA7")) +
  labs(title = "Population Age Structure", x = NULL, y = "% of Population", fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```
:::
:::

## Direct Standardization {.smaller}

**Question:** What would each country's death rate be if both had the **same** age structure?

**Method:** Apply each country's age-specific rates to a common standard population.

| Age Group | Standard Weight (Wᵢ) | India ASDR (Rᵢ) | Wᵢ × Rᵢ | Sweden ASDR (Rᵢ) | Wᵢ × Rᵢ |
|---|:---:|:---:|:---:|:---:|:---:|
| 0–14 | 0.264 | 2.8 | 0.739 | 0.3 | 0.079 |
| 15–44 | 0.378 | 1.5 | 0.567 | 0.4 | 0.151 |
| 45–59 | 0.172 | 5.2 | 0.894 | 2.8 | 0.482 |
| 60–74 | 0.118 | 22.0 | 2.596 | 12.0 | 1.416 |
| 75+ | 0.068 | 80.0 | 5.440 | 72.0 | 4.896 |
| **Total (ASR)** | | | **10.24** | | **7.02** |

. . .

**After standardization:** India (10.2) > Sweden (7.0) — the truth emerges!

$$\text{Age-Standardized Rate} = \sum_{i} W_i \times R_i$$

## Kerala vs Bihar: Same Paradox Within India {.smaller}

::: {.columns}
::: {.column width="50%"}
| | Kerala | Bihar |
|---|:---:|:---:|
| % Population 60+ | **26%** | **13%** |
| Crude Death Rate | 7.4 | 5.8 |
| **Age-Standardized Rate** | **5.6** | **8.1** |
| Life Expectancy | 76.3 yrs | 69.2 yrs |

Kerala's crude rate appears **higher** — but after standardization, Bihar's rate is **45% higher**.

Kerala is healthier at every age. It just has more elderly people (demographic transition).
:::

::: {.column width="50%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5
kb_df <- tibble(
  age_group = factor(rep(c("0–14", "15–44", "45–59", "60–74", "75+"), 2),
                     levels = c("0–14", "15–44", "45–59", "60–74", "75+")),
  state = rep(c("Kerala", "Bihar"), each = 5),
  asdr = c(0.8, 1.0, 4.5, 20.0, 85.0,
           4.5, 2.2, 7.0, 28.0, 95.0)
)
ggplot(kb_df, aes(x = age_group, y = asdr, fill = state)) +
  geom_col(position = "dodge", color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c("Kerala" = "#2E86C1", "Bihar" = "#E74C3C")) +
  labs(title = "Bihar Has Higher ASDR in Every Age Group",
       x = NULL, y = "Deaths per 1,000", fill = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```
:::
:::

## Indirect Standardization & SMR {.smaller}

**When to use:** Small populations where age-specific rates are unstable (few deaths per group).

**Method:** Apply the **standard population's** rates to your population's age structure → get **expected** deaths.

$$\text{SMR} = \frac{\text{Observed deaths}}{\text{Expected deaths}} \times 100$$

. . .

**Example:** A thermal power plant town in Chhattisgarh

| | Observed | Expected (based on India rates) |
|---|:---:|:---:|
| Total deaths | 103 | 75 |
| **SMR** | | **137** |

**Interpretation:** 37% excess mortality compared to national average, after accounting for age structure.

. . .

**Critical rule:** SMRs from different populations are **NOT comparable** to each other. Only compare each SMR to 100 (the reference).

## Direct vs Indirect: Quick Guide {.smaller}

| | Direct | Indirect |
|---|---|---|
| **Apply** | Study rates → standard weights | Standard rates → study weights |
| **Need** | Reliable age-specific rates | Only age structure + total deaths |
| **Get** | Age-standardized rate | SMR (ratio) |
| **Compare?** | Yes — rates are directly comparable | No — SMR only vs reference (100) |
| **Best for** | Large populations | Small populations |

. . .

**Remember:** Standardized rates are **fictional** — they tell you what *would* happen under a hypothetical age structure. Use crude rates for healthcare planning; use standardized rates for fair comparisons.

## Key Takeaways

::: {.incremental}
1. **Cross-sectional** = snapshot; quick but cannot establish causation
2. **Case-control** = start from disease, look backward; efficient for rare diseases
3. **Cohort** = start from exposure, follow forward; can calculate RR and incidence
4. **RCT** = random allocation; the only design that establishes causation
5. **Selection bias** must be prevented by design — cannot be fixed in analysis
6. **Confounders** must satisfy three criteria; use **DAGs** to reason about them
7. **Never compare crude rates** across populations with different age structures — use **standardized rates**
8. Choose your design based on: **research question, outcome frequency, time, ethics, and resources**
:::
