---
title: "Statistical Inference"
subtitle: "CLT, Confidence Intervals, Hypothesis Testing, and Effect Sizes"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    footer: "Biostatistics for Clinicians | Module 8"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
set.seed(42)

theme_slide <- function() {
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12, face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )
}
```

## Learning Objectives

::: {.incremental}
1. Explain the **Central Limit Theorem** and why it enables all inference
2. Calculate and interpret **confidence intervals** for means and proportions
3. Describe the logic of **hypothesis testing** (proof by contradiction)
4. Interpret **p-values** correctly and avoid common misconceptions
5. Distinguish **Type I error (α)**, **Type II error (β)**, and **power (1 − β)**
6. Separate **statistical significance** from **clinical significance**
7. Compute and interpret **effect sizes** (Cohen's d)
8. Recognize the **multiple testing problem** and apply corrections
:::

## Clinical Hook: When "Significant" Doesn't Mean Important {.smaller}

::: {.columns}
::: {.column width="55%"}
**Phase III Antihypertensive Trial (n = 10,000):**

New drug reduced SBP by **2 mmHg** more than standard therapy.

**p = 0.03** → Headlines: *"Significant BP reduction!"*

. . .

**But the clinical reality:**

- MCID for BP = **5–10 mmHg**
- 2 mmHg is **clinically irrelevant**
- With n = 10,000, even trivial differences reach p < 0.05

**This module builds inference from the ground up — and teaches you when to trust a "significant" result.**
:::

::: {.column width="45%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5

tibble(
  Category = c("Effect\nFound", "MCID\n(Minimum)"),
  Value = c(2, 5),
  fill = c("#e74c3c", "#27ae60")
) %>%
  ggplot(aes(x = Category, y = Value, fill = fill)) +
  geom_col(width = 0.6) +
  scale_fill_identity() +
  geom_hline(yintercept = 5, linetype = "dashed", color = "#27ae60", linewidth = 1) +
  annotate("text", x = 1, y = 2.5, label = "2 mmHg\np = 0.03", fontface = "bold", size = 5, color = "white") +
  annotate("text", x = 2, y = 5.5, label = "5 mmHg\nMinimum to matter", fontface = "bold", size = 4, color = "#27ae60") +
  annotate("label", x = 1.5, y = 9, label = '"Significant" ≠ Important', size = 5, fontface = "bold",
           fill = "#fadbd8", color = "#c0392b", label.size = 0.5) +
  labs(title = "Statistical vs Clinical Significance",
       y = "SBP Reduction (mmHg)", x = NULL) +
  ylim(0, 10) +
  theme_slide() +
  theme(legend.position = "none")
```
:::
:::

## The Fundamental Problem

You measure haemoglobin in **50 patients** → sample mean = 11.8 g/dL

. . .

But a **different** set of 50 patients → mean = 12.1 g/dL

. . .

And **another** 50 → mean = 11.5 g/dL

. . .

**Key question:** How can ONE sample tell us about the POPULATION?

. . .

**Answer: The Central Limit Theorem (CLT)** — the single most important result in statistics.

## The Central Limit Theorem: Statement {.smaller}

::: {.callout-tip}
Draw repeated random samples of size **n** from *any* population with mean **μ** and SD **σ**.

As **n** increases, the distribution of sample means ($\bar{x}$) approaches a **normal distribution** with:

$$\text{Mean of } \bar{x} = \mu \qquad \text{SD of } \bar{x} = \frac{\sigma}{\sqrt{n}} \text{ (Standard Error)}$$
:::

. . .

**Three key insights:**

| Insight | Implication |
|---------|-------------|
| Works for **any** population shape | Even skewed data → normal sampling distribution |
| SE shrinks with **√n** | More data → more precision (but diminishing returns) |
| "Large enough" ≈ **n ≥ 30** | Depends on skewness; very skewed → need n ≥ 100 |

## CLT in Action: Skewed Data Becomes Normal {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 11

population <- rgamma(100000, shape = 2, rate = 0.5)
pop_mean <- mean(population)
pop_sd <- sd(population)

draw_means <- function(n, reps = 5000) replicate(reps, mean(sample(population, n)))

means_n5 <- draw_means(5)
means_n30 <- draw_means(30)
means_n100 <- draw_means(100)

p1 <- ggplot(tibble(x = population[1:5000]), aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "#e74c3c", alpha = 0.7, color = "white") +
  geom_vline(xintercept = pop_mean, linetype = "dashed", linewidth = 1) +
  labs(title = "Population (Skewed)", subtitle = paste0("μ = ", round(pop_mean,1)), x = "Value", y = "Density") +
  theme_slide() + theme(axis.text.y = element_blank())

p2 <- ggplot(tibble(x = means_n5), aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "#3498db", alpha = 0.7, color = "white") +
  stat_function(fun = dnorm, args = list(mean = pop_mean, sd = pop_sd/sqrt(5)), linewidth = 1, linetype = "dashed") +
  labs(title = "n = 5", subtitle = paste0("SE = ", round(pop_sd/sqrt(5),2)), x = "Sample Mean", y = NULL) +
  theme_slide() + theme(axis.text.y = element_blank())

p3 <- ggplot(tibble(x = means_n30), aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "#27ae60", alpha = 0.7, color = "white") +
  stat_function(fun = dnorm, args = list(mean = pop_mean, sd = pop_sd/sqrt(30)), linewidth = 1, linetype = "dashed") +
  labs(title = "n = 30", subtitle = paste0("SE = ", round(pop_sd/sqrt(30),2)), x = "Sample Mean", y = NULL) +
  theme_slide() + theme(axis.text.y = element_blank())

p4 <- ggplot(tibble(x = means_n100), aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "#8e44ad", alpha = 0.7, color = "white") +
  stat_function(fun = dnorm, args = list(mean = pop_mean, sd = pop_sd/sqrt(100)), linewidth = 1, linetype = "dashed") +
  labs(title = "n = 100", subtitle = paste0("SE = ", round(pop_sd/sqrt(100),2)), x = "Sample Mean", y = NULL) +
  theme_slide() + theme(axis.text.y = element_blank())

p1 + p2 + p3 + p4 + plot_layout(nrow = 1)
```

**Even heavily skewed hospital-stay data → normal sampling distribution as n increases**

## The Standard Error: Bridge to Inference

$$SE = \frac{\sigma}{\sqrt{n}}$$

```{r}
#| fig-height: 4
#| fig-width: 10

n_vals <- seq(5, 500, by = 5)
se_vals <- pop_sd / sqrt(n_vals)

key_points <- tibble(
  n = c(10, 30, 100, 200, 400),
  SE = pop_sd / sqrt(c(10, 30, 100, 200, 400))
)

tibble(n = n_vals, SE = se_vals) %>%
  ggplot(aes(x = n, y = SE)) +
  geom_line(linewidth = 1.2, color = "#2980b9") +
  geom_point(data = key_points, size = 3, color = "#c0392b") +
  geom_text(data = key_points, aes(label = paste0("n=", n, "\nSE=", round(SE,2))),
            vjust = -1, size = 4, color = "#2c3e50") +
  labs(title = "The √n Law: SE Decreases with Sample Size",
       subtitle = "To halve SE → need 4× the sample size (diminishing returns)",
       x = "Sample Size (n)", y = "Standard Error") +
  theme_slide()
```

## CLT Connects to Everything

The SE connects the CLT to all inference tools:

| Tool | Formula Pattern | Uses SE? |
|------|----------------|----------|
| **Confidence Interval** | Estimate ± critical value × SE | ✅ |
| **Test Statistic** | (Observed − Expected) / SE | ✅ |
| **Sample Size** (Module 6) | Manipulate SE for desired precision | ✅ |

. . .

**The CLT is the engine; SE is the transmission; CIs and tests are the wheels.**

## From Point Estimate to Confidence Interval {.smaller}

**Point estimate:** $\bar{x}$ = 9.2 g/dL (how precise is this?)

**Confidence interval:** provides a *range* of plausible values

. . .

$$\text{CI} = \text{Point Estimate} \pm (\text{Critical Value}) \times \text{Standard Error}$$

::: {.columns}
::: {.column width="50%"}
**For a mean (σ unknown):**

$$\bar{x} \pm t_{\alpha/2, \, df} \times \frac{s}{\sqrt{n}}$$

- df = n − 1
- For 95% CI: $t_{0.025}$ ≈ 1.96 when n is large
:::
::: {.column width="50%"}
**For a proportion:**

$$\hat{p} \pm z_{\alpha/2} \times \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

- For 95% CI: $z_{0.025}$ = 1.96
- Check: np ≥ 5 and n(1−p) ≥ 5
:::
:::

## Worked Example: CI for a Mean {.smaller}

**64 newly diagnosed Type 2 DM patients — Mean FBS:**

| Given | Value |
|-------|-------|
| $\bar{x}$ | 156 mg/dL |
| s | 32 mg/dL |
| n | 64 |

. . .

**Step 1:** SE = 32 / √64 = 32 / 8 = **4 mg/dL**

. . .

**Step 2:** Critical value: $t_{0.025, 63}$ ≈ 2.00

. . .

**Step 3:** CI = 156 ± 2.00 × 4 = 156 ± 8 = **(148, 164) mg/dL**

. . .

**Interpretation:** We are 95% confident that the true mean FBS lies between 148 and 164 mg/dL.

## Worked Example: CI for a Proportion {.smaller}

**400 adults in rural Bhopal; 72 had hypertension (BP ≥ 140/90)**

| Given | Value |
|-------|-------|
| $\hat{p}$ | 72/400 = 0.18 (18%) |
| n | 400 |

. . .

**Step 1:** SE = $\sqrt{0.18 \times 0.82 / 400}$ = $\sqrt{0.000369}$ = **0.0192**

. . .

**Step 2:** CI = 0.18 ± 1.96 × 0.0192 = 0.18 ± 0.038 = **(14.2%, 21.8%)**

. . .

**Check:** np = 72 ≥ 5, n(1−p) = 328 ≥ 5 → normal approximation valid ✅

## What Does "95% Confidence" Actually Mean? {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 10

true_mean <- 120; true_sd <- 15; n_sim <- 40

set.seed(123)
ci_data <- tibble(
  study = 1:100,
  xbar = replicate(100, mean(rnorm(n_sim, true_mean, true_sd))),
  se = replicate(100, sd(rnorm(n_sim, true_mean, true_sd)) / sqrt(n_sim))
) %>%
  mutate(
    lo = xbar - qt(0.975, n_sim - 1) * se,
    hi = xbar + qt(0.975, n_sim - 1) * se,
    captures = lo <= true_mean & hi >= true_mean
  )

n_cap <- sum(ci_data$captures)

ggplot(ci_data, aes(x = study, y = xbar, color = captures)) +
  geom_hline(yintercept = true_mean, linewidth = 1.2, color = "#2c3e50") +
  geom_point(size = 1.2) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0, linewidth = 0.4) +
  scale_color_manual(values = c("TRUE" = "#3498db", "FALSE" = "#e74c3c"),
                     labels = c("TRUE" = "Captures μ", "FALSE" = "Misses μ"), name = NULL) +
  annotate("text", x = 50, y = true_mean + 8, label = paste0("True μ = ", true_mean, " mmHg"),
           fontface = "bold", size = 5, color = "#2c3e50") +
  annotate("label", x = 85, y = true_mean - 8,
           label = paste0(n_cap, "/100 CIs capture μ"), fill = "#d6eaf8", size = 4.5, fontface = "bold") +
  coord_flip() +
  labs(title = "100 Simulated 95% CIs from the Same Population",
       x = "Study Number", y = "Systolic BP (mmHg)") +
  theme_slide() + theme(legend.position = "top")
```

## CI Interpretation: Right vs Wrong

| | Wrong Interpretation | Why It's Wrong |
|---|---|---|
| ❌ | "95% probability the true mean is in this interval" | True mean is fixed — no probability about it |
| ❌ | "95% of patients fall in this range" | CI is about the mean, not individual values |
| ❌ | "95% chance a new sample mean falls here" | New sample generates its own CI |

. . .

**✅ Correct:** "If we repeated this study many times, **95% of calculated CIs** would contain the true population mean."

## What Makes a CI Narrower? {.smaller}

```{r}
#| fig-height: 4.5
#| fig-width: 11

p_n <- tibble(n = c(10,25,50,100,200,500),
              width = 2 * qt(0.975, c(10,25,50,100,200,500)-1) * (32/sqrt(c(10,25,50,100,200,500)))) %>%
  ggplot(aes(x = factor(n), y = width)) +
  geom_col(fill = "#3498db", alpha = 0.7, width = 0.6) +
  geom_text(aes(label = round(width,1)), vjust = -0.3, fontface = "bold", size = 4) +
  labs(title = "↑ Sample Size → Narrower CI", x = "n", y = "CI Width") +
  theme_slide()

p_sd <- tibble(sd = c(10,20,30,40,50),
               width = 2 * qt(0.975, 63) * (c(10,20,30,40,50)/sqrt(64))) %>%
  ggplot(aes(x = factor(sd), y = width)) +
  geom_col(fill = "#e74c3c", alpha = 0.7, width = 0.6) +
  geom_text(aes(label = round(width,1)), vjust = -0.3, fontface = "bold", size = 4) +
  labs(title = "↑ Variability → Wider CI", x = "SD", y = "CI Width") +
  theme_slide()

p_conf <- tibble(conf = c("80%","90%","95%","99%"),
                 width = 2 * qnorm(1 - (1 - c(0.80,0.90,0.95,0.99))/2) * (32/sqrt(64))) %>%
  ggplot(aes(x = conf, y = width)) +
  geom_col(fill = "#27ae60", alpha = 0.7, width = 0.6) +
  geom_text(aes(label = round(width,1)), vjust = -0.3, fontface = "bold", size = 4) +
  labs(title = "↑ Confidence Level → Wider CI", x = "Confidence", y = "CI Width") +
  theme_slide()

p_n + p_sd + p_conf
```

## The Logic of Hypothesis Testing

**Mathematical analogy:**

> To prove √2 is irrational → assume it IS rational → derive a contradiction → reject assumption

. . .

**Statistical analogy:**

> To test if a drug works → assume it DOESN'T (H₀) → observe data → if data are very unlikely under H₀ → reject H₀

. . .

**Key:** We never *prove* the drug works; we only ask whether "no effect" is plausible.

## H₀ and H₁: The Two-Hypothesis Framework {.smaller}

```{r}
#| fig-height: 4.5
#| fig-width: 10

x <- seq(-4, 5, length.out = 1000)
h0 <- dnorm(x, 0, 1)
h1 <- dnorm(x, 2, 1)
crit <- qnorm(0.95)

df <- tibble(x, h0, h1)

ggplot(df, aes(x = x)) +
  geom_area(aes(y = h0, fill = "H₀ (No effect)"), alpha = 0.5) +
  geom_area(aes(y = h1, fill = "H₁ (Real effect)"), alpha = 0.5) +
  geom_vline(xintercept = crit, linetype = "dashed", color = "red", linewidth = 1) +
  geom_area(data = df %>% filter(x >= crit), aes(x, y = h0), fill = "#c0392b", alpha = 0.8) +
  geom_area(data = df %>% filter(x < crit), aes(x, y = h1), fill = "#f39c12", alpha = 0.8) +
  scale_fill_manual(values = c("H₀ (No effect)" = "#3498db", "H₁ (Real effect)" = "#e74c3c")) +
  annotate("text", x = crit + 0.5, y = 0.35, label = "Reject H₀", size = 5, fontface = "bold", color = "darkred") +
  annotate("text", x = -2, y = 0.35, label = "Fail to\nReject H₀", size = 5, fontface = "bold", color = "darkgreen") +
  labs(title = "Two Overlapping Distributions Under H₀ and H₁",
       x = "Test Statistic (z)", y = "Density", fill = NULL) +
  theme_slide() + theme(legend.text = element_text(size = 12))
```

| | Definition | Example |
|---|---|---|
| **H₀** (Null) | No effect / no difference | New drug = standard drug |
| **H₁** (Alternative) | Effect exists | New drug ≠ (or >) standard drug |

## One-Tailed vs Two-Tailed Tests {.smaller}

```{r}
#| fig-height: 4
#| fig-width: 11

x <- seq(-4, 4, length.out = 1000)
y <- dnorm(x)
df_t <- tibble(x, y)

p_two <- ggplot(df_t, aes(x, y)) +
  geom_area(fill = "lightblue", alpha = 0.7) +
  geom_area(data = df_t %>% filter(x > 1.96), fill = "red", alpha = 0.8) +
  geom_area(data = df_t %>% filter(x < -1.96), fill = "red", alpha = 0.8) +
  geom_vline(xintercept = c(-1.96, 1.96), linetype = "dashed", color = "darkred") +
  labs(title = "Two-Tailed (α = 0.05)", subtitle = "0.025 each tail | z = ±1.96",
       x = "Test Statistic", y = "Density") +
  theme_slide() + theme(axis.text.y = element_blank())

p_one <- ggplot(df_t, aes(x, y)) +
  geom_area(fill = "lightblue", alpha = 0.7) +
  geom_area(data = df_t %>% filter(x > 1.645), fill = "red", alpha = 0.8) +
  geom_vline(xintercept = 1.645, linetype = "dashed", color = "darkred") +
  labs(title = "One-Tailed (α = 0.05)", subtitle = "0.05 in right tail | z = 1.645",
       x = "Test Statistic", y = "Density") +
  theme_slide() + theme(axis.text.y = element_blank())

p_two + p_one
```

**Default: Two-tailed** | Use one-tailed only if direction pre-specified and opposite is irrelevant

## Test Statistics: Common Pattern {.smaller}

$$\text{Test Statistic} = \frac{\text{Observed} - \text{Expected under } H_0}{\text{Standard Error}}$$

| Scenario | Formula | Distribution |
|----------|---------|-------------|
| One sample mean | $t = (\bar{x} - \mu_0) / (s/\sqrt{n})$ | t (n−1 df) |
| Two independent means | $t = (\bar{x}_1 - \bar{x}_2) / SE_{diff}$ | t (n₁+n₂−2 df) |
| Paired means | $t = \bar{x}_{diff} / (s_{diff}/\sqrt{n})$ | t (n−1 df) |
| One proportion | $z = (\hat{p} - p_0) / \sqrt{p_0(1-p_0)/n}$ | Normal (z) |
| Two proportions | $z = (\hat{p}_1 - \hat{p}_2) / SE_{diff}$ | Normal (z) |

**Always: (signal − expected) / uncertainty**

## What Is a P-Value? {.smaller}

::: {.callout-tip}
The **p-value** = probability of observing data **as extreme or more extreme** than what we got, **assuming H₀ is true**.
:::

```{r}
#| fig-height: 4
#| fig-width: 10

x <- seq(-4, 4, length.out = 1000)
y <- dnorm(x)
obs <- 2.5
df_p <- tibble(x, y)
df_shade <- df_p %>% filter(abs(x) >= abs(obs))

ggplot(df_p, aes(x, y)) +
  geom_area(fill = "lightblue", alpha = 0.7) +
  geom_area(data = df_shade, fill = "#e74c3c", alpha = 0.8) +
  geom_vline(xintercept = c(obs, -obs), color = "darkred", linewidth = 1.2) +
  annotate("label", x = 3.5, y = 0.25,
           label = "Red area = p-value\n≈ 0.012 (two-tailed)\nIf p < 0.05 → Reject H₀",
           size = 4, fontface = "bold", fill = "white", color = "darkred") +
  labs(title = "P-Value Visualized", subtitle = "Area in tails beyond the observed test statistic",
       x = "Test Statistic", y = "Density") +
  theme_slide() + theme(axis.text.y = element_blank())
```

## P-Value Misconceptions {.smaller}

| ❌ Misconception | ✅ Reality |
|---|---|
| "p = 0.03 → 3% chance H₀ is true" | P(data this extreme \| H₀ true), not P(H₀ true \| data) |
| "p = 0.03 → 97% sure result is real" | Replication depends on effect size, design, not p alone |
| "p = 0.06 → 'trending toward significance'" | Arbitrary! Report exact p-value and effect size |
| "p = 0.15 → treatment has no effect" | Absence of evidence ≠ evidence of absence |
| "p = 0.001 → very confident" | Bias doesn't care about your p-value |
| "Multiple p < 0.05 → strong evidence" | 20 tests → expect ~1 false positive by chance |

. . .

**ASA (2016):** Scientific conclusions require more than p < 0.05. Always report effect sizes and CIs.

## The Four Outcomes of a Hypothesis Test {.smaller}

|  | **Reject H₀** | **Fail to Reject H₀** |
|---|---|---|
| **H₀ TRUE** (no effect) | **TYPE I ERROR (α)** ❌ False Positive | ✅ Correct (1 − α) |
| **H₀ FALSE** (effect exists) | ✅ Correct — **Power (1 − β)** | **TYPE II ERROR (β)** ❌ False Negative |

. . .

**Convention:** α = 0.05, β = 0.20, Power = 0.80

**The 4:1 trade-off:** False positive (adopting useless drug) is considered 4× worse than false negative (missing an effective one).

## Type I Error (α): The False Positive

**Definition:** Concluding an effect exists when it DOESN'T

**Probability:** α (set by researcher, usually 0.05)

. . .

**Dangerous clinical scenarios:**

::: {.incremental}
- **Approving an ineffective drug** → patients receive it instead of proven therapy
- **Unnecessary surgery** → thousands undergo biopsies for a spurious biomarker association
- **Policy changes** → government bans a food additive based on a statistical fluke
:::

. . .

**What increases Type I risk?** Multiple testing, p-hacking, flexible analysis plans, not pre-registering

## Type II Error (β): The False Negative

**Definition:** Missing a real effect — declaring "no difference" when one exists

**Probability:** β (depends on sample size, effect size, variability)

. . .

**Dangerous clinical scenarios:**

::: {.incremental}
- **Shelving an effective drug** → immunotherapy that reduces mortality by 15% is abandoned because trial was too small (n = 80)
- **False equivalence** → concluding two drugs are "the same" with n = 40 per group
- **Premature stopping** → interim p = 0.08 stops a trial that would have reached p = 0.01
:::

. . .

**The #1 cause:** Small sample size

## Diagnostic Test Parallel

The concepts map directly to Module 5:

| Hypothesis Testing | Diagnostic Testing |
|---|---|
| Type I Error (α) | **False Positive** — healthy person tests positive |
| Type II Error (β) | **False Negative** — sick person tests negative |
| Power (1 − β) | **Sensitivity** — correctly detecting disease |
| Significance level (α) | 1 − Specificity |

. . .

**Power IS sensitivity for hypothesis tests** — the ability to detect a true effect.

## Power: What Determines It? {.smaller}

```{r}
#| fig-height: 4.5
#| fig-width: 11

effect_sizes <- seq(0.1, 1.5, by = 0.05)
power_eff <- sapply(effect_sizes, function(d) pnorm(d * sqrt(64/2) - qnorm(0.975)))

p_eff <- tibble(d = effect_sizes, power = power_eff) %>%
  ggplot(aes(d, power)) +
  geom_line(linewidth = 1.2, color = "#3498db") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "#c0392b") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  annotate("text", x = 1.3, y = 0.83, label = "80% Power", color = "#c0392b", fontface = "bold") +
  labs(title = "Power vs Effect Size", subtitle = "n = 64/group, α = 0.05",
       x = "Cohen's d", y = "Power") +
  theme_slide()

sample_sizes <- seq(10, 300, by = 5)
power_n <- sapply(sample_sizes, function(n) pnorm(0.5 * sqrt(n/2) - qnorm(0.975)))

p_ss <- tibble(n = sample_sizes, power = pmin(power_n, 1)) %>%
  ggplot(aes(n, power)) +
  geom_line(linewidth = 1.2, color = "#e74c3c") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "darkgreen") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  annotate("text", x = 220, y = 0.83, label = "80% Power", color = "darkgreen", fontface = "bold") +
  labs(title = "Power vs Sample Size", subtitle = "d = 0.5 (medium), α = 0.05",
       x = "n per group", y = "Power") +
  theme_slide()

p_eff + p_ss
```

**↑ Power with: ↑ sample size, ↑ effect size, ↑ α, ↓ variability**

## The α–β Trade-off {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 11

x <- seq(-4, 6, length.out = 1000)

plot_ab <- function(crit, label) {
  h0 <- dnorm(x, 0, 1)
  h1 <- dnorm(x, 2.5, 1)
  df <- tibble(x, h0, h1)

  alpha_val <- round(1 - pnorm(crit), 3)
  beta_val <- round(pnorm(crit, mean = 2.5), 3)

  ggplot(df, aes(x)) +
    geom_line(aes(y = h0), color = "#3498db", linewidth = 1) +
    geom_line(aes(y = h1), color = "#e74c3c", linewidth = 1) +
    geom_area(data = df %>% filter(x >= crit), aes(x, y = h0), fill = "#c0392b", alpha = 0.4) +
    geom_area(data = df %>% filter(x < crit), aes(x, y = h1), fill = "#f39c12", alpha = 0.4) +
    geom_vline(xintercept = crit, linetype = "dashed", linewidth = 1) +
    annotate("label", x = -2.5, y = 0.35, label = paste0("α = ", alpha_val),
             fill = "#fadbd8", color = "#c0392b", fontface = "bold", size = 4) +
    annotate("label", x = 4.5, y = 0.35, label = paste0("β = ", beta_val, "\nPower = ", round(1-beta_val,3)),
             fill = "#fdebd0", color = "#e67e22", fontface = "bold", size = 3.5) +
    labs(title = label, x = NULL, y = NULL) +
    theme_slide() + theme(axis.text = element_blank())
}

plot_ab(1.0, "Liberal (α ≈ 0.16)") / plot_ab(1.645, "Standard (α ≈ 0.05)") / plot_ab(2.33, "Stringent (α ≈ 0.01)")
```

## α–β Trade-off: Key Insight

- **Decrease α** (more stringent) → **β increases** (miss more real effects)
- You **cannot** minimize both errors simultaneously

. . .

**Solution:** Increase sample size — more data reduces BOTH errors

. . .

| Strategy | α | β | Power | Feasibility |
|----------|---|---|-------|-------------|
| Liberal | 0.10 | 0.10 | 90% | ⚠️ More false positives |
| Standard | 0.05 | 0.20 | 80% | ✅ Most common |
| Stringent | 0.01 | 0.30 | 70% | ⚠️ Miss real effects |
| Standard + large n | 0.05 | 0.10 | 90% | ✅ Best but expensive |

## Statistical vs Clinical Significance: The 2×2 Framework {.smaller}

```{r}
#| fig-height: 5
#| fig-width: 10

grid_data <- tibble(
  x = c(1, 2, 1, 2),
  y = c(2, 2, 1, 1),
  label = c(
    "MISSED FINDING\nClinically Important\nBUT Underpowered\n(Need larger trial)",
    "IDEAL FINDING\nStatistically +\nClinically Significant\n(Change practice!)",
    "CORRECT NULL\nNeither Significant\n(No effect)",
    "SPURIOUS FINDING\nStatistically Significant\nBUT Clinically Trivial\n(Don't change practice)"
  ),
  color = c("#f39c12", "#27ae60", "#3498db", "#e74c3c")
)

ggplot(grid_data, aes(x, y)) +
  geom_tile(aes(fill = color), width = 0.9, height = 0.9, color = "white", linewidth = 2) +
  scale_fill_identity() +
  geom_text(aes(label = label), size = 4.5, fontface = "bold", color = "white", lineheight = 1.1) +
  scale_x_continuous(breaks = c(1, 2), labels = c("Not\nSignificant", "Statistically\nSignificant"), position = "top") +
  scale_y_continuous(breaks = c(1, 2), labels = c("Not Clinically\nImportant", "Clinically\nImportant")) +
  labs(title = "Statistical vs Clinical Significance", x = "p-value", y = "Effect Size") +
  theme_slide() + theme(panel.grid = element_blank(), legend.position = "none",
                         axis.text = element_text(size = 12, face = "bold"))
```

## Real-World Examples from Indian Medicine {.smaller}

| Study | Effect | n | p-value | MCID | Verdict |
|-------|--------|---|---------|------|---------|
| **Hypertension** | SBP ↓ 2 mmHg | 10,000 | 0.03 | 5–10 mmHg | Stat YES, Clinical **NO** ❌ |
| **Sepsis** | Mortality ↓ 12% | 200 | 0.001 | ≥ 5% | Stat YES, Clinical **YES** ✅ |
| **Diabetes** | HbA1c ↓ 0.4% | 150 | 0.12 | 0.5% | Stat NO, Clinical borderline ❓ |
| **Anticoagulation** | Stroke ↓ 0.5% | 3,000 | 0.001 | ≥ 2% | Stat YES, Clinical **NO** ❌ |

. . .

**Key principle:** A statistically significant result below the MCID is clinically irrelevant — no matter how small the p-value.

## MCID: Minimal Clinically Important Difference

Always define the MCID **before** the trial:

| Outcome | MCID |
|---------|------|
| Systolic BP reduction | 5–10 mmHg |
| HbA1c reduction | 0.5% |
| Pain (VAS 0–100) | 10–15 points |
| Mortality reduction | 5% absolute |
| Quality of life (EQ-5D) | 0.05–0.08 units |

. . .

**Decision rule:** If 95% CI lies entirely above MCID → clinically important.
If CI straddles MCID → inconclusive. If CI below MCID → not important.

## Effect Size: Why It Matters More Than P-Values {.smaller}

```{r}
#| fig-height: 5
#| fig-width: 11

plot_eff <- function(d, title_text) {
  x <- seq(-4, 4 + d, length.out = 1000)
  tibble(x, Control = dnorm(x, 0, 1), Treatment = dnorm(x, d, 1)) %>%
    ggplot(aes(x)) +
    geom_area(aes(y = Control), fill = "#3498db", alpha = 0.5) +
    geom_area(aes(y = Treatment), fill = "#e74c3c", alpha = 0.5) +
    annotate("text", x = -1.5, y = 0.35, label = "Control", size = 4, fontface = "bold", color = "#2471a3") +
    annotate("text", x = d + 1, y = 0.35, label = "Treatment", size = 4, fontface = "bold", color = "#c0392b") +
    labs(title = title_text, x = NULL, y = NULL) +
    theme_slide() + theme(axis.text = element_blank())
}

plot_eff(0.2, "Small (d = 0.2)") + plot_eff(0.5, "Medium (d = 0.5)") + plot_eff(0.8, "Large (d = 0.8)")
```

**All three could have p < 0.05 with adequate n — but clinical relevance differs hugely!**

## Cohen's d: Formula and Benchmarks

$$d = \frac{\bar{x}_1 - \bar{x}_2}{s_p} \qquad s_p = \sqrt{\frac{(n_1-1)s_1^2 + (n_2-1)s_2^2}{n_1 + n_2 - 2}}$$

. . .

| Cohen's d | Size | Clinical Example | n needed/group |
|-----------|------|------------------|----------------|
| 0.2 | Small | New antacid vs old antacid | ~400 |
| 0.5 | Medium | Lisinopril for hypertension | ~64 |
| 0.8 | Large | Penicillin for strep throat | ~25 |
| > 1.2 | Very large | Insulin for new-onset diabetes | ~12 |

. . .

**Key advantage:** Effect size is **independent of sample size** — p-value is not!

## Worked Example: ACE Inhibitor Trial {.smaller}

**ACE inhibitor vs placebo (n = 50 per group, 8 weeks):**

| Group | SBP Reduction | SD |
|-------|--------------|-----|
| Placebo | 5 mmHg | 10 |
| ACE-I | 12 mmHg | 10 |

. . .

$$s_p = \sqrt{\frac{49 \times 100 + 49 \times 100}{98}} = 10 \qquad d = \frac{12 - 5}{10} = 0.7$$

. . .

| Question | Answer |
|----------|--------|
| Statistically significant? | ✅ p < 0.001 |
| Effect size meaningful? | ✅ d = 0.7 (medium-large) |
| Clinically important? | ✅ 7 mmHg > MCID of 5 mmHg |
| **Should we implement?** | **✅ YES** |

## The Multiple Testing Problem {.smaller}

```{r}
#| fig-height: 4.5
#| fig-width: 10

k <- seq(1, 50, by = 1)
prob <- 1 - (1 - 0.05)^k

tibble(k, prob) %>%
  ggplot(aes(k, prob)) +
  geom_line(linewidth = 1.2, color = "#e74c3c") +
  geom_area(alpha = 0.2, fill = "#e74c3c") +
  geom_vline(xintercept = 20, linetype = "dashed", color = "darkred") +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "blue") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  annotate("label", x = 35, y = 0.65,
           label = "At 20 tests:\nP(≥1 false positive) ≈ 64%",
           fontface = "bold", size = 4.5, fill = "lightyellow") +
  labs(title = "P(≥1 False Positive) = 1 − (0.95)^k",
       x = "Number of Tests (k)", y = "P(≥1 False Positive)") +
  theme_slide()
```

**Bonferroni correction:** α~corrected~ = α / k (e.g., 10 tests → use α = 0.005)

## P-Hacking: The Dark Side

**What is p-hacking?** Manipulating analysis to achieve p < 0.05:

::: {.incremental}
1. Test many variables → report only significant ones
2. Try multiple statistical tests on the same data
3. Selectively exclude outliers until p < 0.05
4. Stop data collection when p first crosses 0.05
5. Report only "interesting" subgroups
:::

. . .

**Solution:** Pre-register analysis plans (ClinicalTrials.gov, OSF Registries)

## CIs Tell You More Than P-Values {.smaller}

```{r}
#| fig-height: 4.5
#| fig-width: 10

studies <- tibble(
  Study = c("A: Large trial, clear effect", "B: Moderate trial, clear effect",
            "C: Small trial, significant", "D: Small trial, NOT significant",
            "E: Large trial, NOT significant"),
  Effect = c(5, 8, 12, 8, 1),
  lo = c(2, 5, 1, -3, -1),
  hi = c(8, 11, 23, 19, 3),
  Sig = c("Yes","Yes","Yes","No","No")
)

studies %>%
  mutate(Study = fct_inorder(Study) %>% fct_rev()) %>%
  ggplot(aes(Study, Effect, color = Sig)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.2, linewidth = 1.2) +
  scale_color_manual(values = c("Yes" = "#27ae60", "No" = "#e74c3c"),
                     labels = c("Yes" = "p < 0.05", "No" = "p ≥ 0.05"), name = NULL) +
  coord_flip() +
  labs(title = "CIs Show Magnitude, Precision, and Significance", x = NULL, y = "Effect Size (mmHg)") +
  theme_slide() + theme(legend.position = "top")
```

Study D: wide CI → underpowered (need more data) | Study E: narrow CI → truly no effect

## CI vs P-Value: Summary

| Aspect | P-Value | Confidence Interval |
|--------|---------|-------------------|
| Shows effect size? | ❌ No | ✅ Yes |
| Shows precision? | ❌ Implicit | ✅ Yes |
| Avoids dichotomy? | ❌ No | ✅ Yes |
| Assess clinical importance? | ❌ No | ✅ Compare to MCID |
| Subject to p-hacking? | ⚠️ Very | ✅ Less so |

. . .

**Best practice:** Report **both**. But if forced to choose one → choose the CI.

## Using CIs for Clinical Decisions

**Trial result:** SBP reduced by 6 mmHg (95% CI: 3 to 9, p = 0.001)

**MCID = 5 mmHg**

. . .

- Point estimate (6) > MCID ✅
- But lower bound (3) < MCID

. . .

**Interpretation:** Effect *likely* clinically meaningful, but we cannot rule out the true effect is below the MCID. Suggestive but not definitive.

**P-value alone** would just say "p = 0.001, significant!" — missing this nuance entirely.

## The Logical Flow of Inference {.smaller}

```{r}
#| fig-height: 5
#| fig-width: 11

flow <- tibble(
  step = c("Central\nLimit\nTheorem", "Standard\nError", "Confidence\nIntervals",
           "Hypothesis\nTesting", "P-Value", "Type I/II\nErrors",
           "Effect\nSize", "Clinical vs\nStatistical\nSignificance"),
  x = c(1, 2, 3, 3, 4, 4, 5, 6),
  y = c(2, 2, 3, 1, 3, 1, 2, 2),
  color = c("#2c3e50","#2980b9","#27ae60","#e74c3c","#27ae60","#e74c3c","#8e44ad","#f39c12")
)

arrows <- tibble(
  x = c(1, 2, 2, 3, 3, 4, 4, 5),
  xend = c(2, 3, 3, 4, 4, 5, 5, 6),
  y = c(2, 2, 2, 3, 1, 3, 1, 2),
  yend = c(2, 3, 1, 3, 1, 2, 2, 2)
)

ggplot() +
  geom_segment(data = arrows, aes(x = x, xend = xend, y = y, yend = yend),
               arrow = arrow(length = unit(0.2, "cm"), type = "closed"), color = "gray60", linewidth = 0.8) +
  geom_label(data = flow, aes(x, y, label = step, fill = color),
             color = "white", fontface = "bold", size = 4, label.padding = unit(0.4, "lines")) +
  scale_fill_identity() +
  xlim(0.5, 6.5) + ylim(0, 4) +
  labs(title = "Each Concept Builds on the Previous") +
  theme_void() + theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5))
```

## Key Takeaways

::: {.incremental}
1. **CLT** → sample means are normal regardless of population shape (foundation)
2. **SE = σ/√n** → precision improves with √n (diminishing returns)
3. **CI** = estimate ± critical value × SE → range of plausible values
4. **P-value** = P(data this extreme | H₀ true) — NOT P(H₀ true)
5. **Type I** = false positive (α) | **Type II** = false negative (β)
6. **Power** = 1 − β ≥ 0.80 before starting any study
7. **Statistical significance ≠ clinical significance** — always check MCID
8. **Effect size** (Cohen's d) is independent of n — report it always
9. **Multiple testing** inflates false positives — correct with Bonferroni
10. **CIs > p-values** — report both, but CIs tell you more
:::

## Do's and Don'ts

::: {.columns}
::: {.column width="50%"}
**✅ DO:**

- Pre-specify hypotheses
- Report CIs AND p-values
- Report effect sizes
- Power analysis BEFORE study
- Correct for multiple tests
- Report ALL tests conducted
- Consider clinical significance
:::
::: {.column width="50%"}
**❌ DON'T:**

- Interpret p as P(H₀ true)
- Claim non-significant = no effect
- Use one-tailed tests to dodge α
- Stop when p first hits 0.05
- Test many hypotheses uncorrected
- Say "trending toward significance"
- Confuse statistical and clinical significance
:::
:::

## Formula Reference Card {.smaller}

| Concept | Formula |
|---------|---------|
| **Standard Error (mean)** | $SE = s / \sqrt{n}$ |
| **95% CI for mean** | $\bar{x} \pm t_{0.025} \times SE$ |
| **95% CI for proportion** | $\hat{p} \pm 1.96 \sqrt{\hat{p}(1-\hat{p})/n}$ |
| **Test statistic (general)** | $(Observed - Expected_{H_0}) / SE$ |
| **Cohen's d** | $(\bar{x}_1 - \bar{x}_2) / s_p$ |
| **Bonferroni** | $\alpha_{adj} = \alpha / k$ |
| **Multiple testing P** | $P(\geq 1 \text{ FP}) = 1 - (1-\alpha)^k$ |

. . .

**Remember:** The CLT makes all of this possible.

## Thank You {.center}

**Module 8: Statistical Inference**

*From CLT to CIs to Hypothesis Testing — and knowing when "significant" truly matters*

::: {.callout-tip}
Next: Module 9 — Comparing Two Groups (t-tests, Mann-Whitney, Chi-Square)
:::
