---
title: "Comparing Groups"
subtitle: "t-tests, ANOVA, and Non-parametric Alternatives"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    footer: "Biostatistics for Clinicians | Module 9"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
set.seed(42)

theme_clean <- function() {
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    legend.position = "bottom"
  )
}
```

## The Clinical Problem

An ICMR trial in Jharkhand randomises 90 women with iron-deficiency anaemia to three arms (n = 30 each):

| Arm | Mean Hb ± SD (g/dL) |
|-----|---------------------|
| Oral iron | 10.8 ± 1.4 |
| IV iron | 11.5 ± 1.6 |
| Placebo | 9.6 ± 1.5 |

. . .

**Questions that cascade:**

1. Is oral iron better than placebo? → **t-test**
2. Are all three groups different? → **ANOVA**
3. *Which* pairs differ? → **Post-hoc tests**
4. What if data aren't normal? → **Non-parametric alternatives**

## Roadmap

```{r}
#| fig-width: 10
#| fig-height: 5

roadmap <- tibble(
  x = 1:7,
  label = c("One-Sample t", "Paired t", "Independent t",
            "Assumptions", "ANOVA", "Post-Hoc", "Non-Parametric"),
  group = c(rep("t-tests", 3), "Check", "ANOVA", "ANOVA", "Alternatives")
)

ggplot(roadmap, aes(x = x, y = 0, label = label, fill = group)) +
  geom_tile(width = 0.9, height = 0.5, color = "white", linewidth = 1.5) +
  geom_text(size = 4, fontface = "bold", color = "white") +
  scale_fill_manual(values = c("t-tests" = "#2c3e50", "Check" = "#e67e22",
                                "ANOVA" = "#2980b9", "Alternatives" = "#27ae60")) +
  coord_fixed(ratio = 1.2) +
  theme_void() +
  theme(legend.position = "none")
```

## One-Sample t-Test: When to Use

**Scenario:** One group vs. a known reference value

. . .

**Example:** 120 MBBS students at KGMU Lucknow — is their mean BMI different from the national average of 22.5?

. . .

**Hypotheses:**

- H₀: μ = 22.5 (mean BMI equals national average)
- H₁: μ ≠ 22.5 (two-tailed)

. . .

**Formula:**

$$t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}, \quad df = n - 1$$

## One-Sample t: Worked Example

```{r}
#| fig-width: 10
#| fig-height: 5

bmi <- rnorm(120, mean = 23.4, sd = 3.8)
t_result <- t.test(bmi, mu = 22.5)

p1 <- ggplot(data.frame(bmi), aes(x = bmi)) +
  geom_histogram(fill = "#3498db", color = "white", bins = 20, alpha = 0.8) +
  geom_vline(xintercept = 22.5, linetype = "dashed", color = "#e74c3c", linewidth = 1) +
  geom_vline(xintercept = mean(bmi), color = "#2c3e50", linewidth = 1) +
  annotate("text", x = 22.5, y = Inf, label = "National avg = 22.5",
           vjust = 2, hjust = 1.1, color = "#e74c3c", fontface = "bold") +
  annotate("text", x = mean(bmi), y = Inf,
           label = paste0("Sample mean = ", round(mean(bmi), 1)),
           vjust = 2, hjust = -0.1, color = "#2c3e50", fontface = "bold") +
  labs(title = "BMI of 120 MBBS Students, KGMU Lucknow",
       x = "BMI (kg/m²)", y = "Count") +
  theme_clean()

print(p1)
```

t = `r round(t_result$statistic, 2)`, df = `r t_result$parameter`, p = `r round(t_result$p.value, 4)`

## Paired t-Test: Same Subjects, Two Timepoints

**Scenario:** 40 hypertensive patients at a PHC in Kolar, Karnataka — systolic BP before and after 6 weeks of amlodipine

. . .

$$t = \frac{\bar{d}}{s_d / \sqrt{n}}, \quad df = n - 1$$

where $\bar{d}$ = mean of within-subject differences

. . .

**Key:** Each patient is their own control — reduces between-subject variability

## Paired t: Worked Example

```{r}
#| fig-width: 10
#| fig-height: 5

bp_before <- rnorm(40, mean = 152, sd = 12)
bp_after  <- bp_before - rnorm(40, mean = 14, sd = 8)
diff_bp   <- bp_before - bp_after
t_paired  <- t.test(bp_before, bp_after, paired = TRUE)

df_bp <- data.frame(
  Patient = rep(1:40, 2),
  Time = rep(c("Before", "After"), each = 40),
  SBP = c(bp_before, bp_after)
)

ggplot(df_bp, aes(x = Time, y = SBP)) +
  geom_line(aes(group = Patient), alpha = 0.3, color = "gray60") +
  geom_point(aes(color = Time), size = 2) +
  stat_summary(fun = mean, geom = "point", size = 5, shape = 18, color = "#e74c3c") +
  stat_summary(fun = mean, geom = "line", aes(group = 1), linewidth = 1.5, color = "#e74c3c") +
  scale_color_manual(values = c("Before" = "#e74c3c", "After" = "#2ecc71")) +
  labs(title = "Systolic BP Before and After Amlodipine",
       subtitle = paste0("Mean drop = ", round(mean(diff_bp), 1),
                        " mmHg | p = ", format(t_paired$p.value, digits = 3)),
       x = "", y = "Systolic BP (mmHg)") +
  theme_clean() +
  theme(legend.position = "none")
```

## Independent Two-Sample t-Test

**Scenario:** HbA1c at 6 months — metformin (n = 50) vs. glimepiride (n = 50) at KEM Hospital, Pune

. . .

$$t = \frac{\bar{x}_1 - \bar{x}_2}{SE}, \quad SE = s_p \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}$$

. . .

**Two versions:**

| Condition | Test |
|-----------|------|
| Equal variances (Levene's p > 0.05) | Student's t-test |
| Unequal variances | Welch's t-test (default in R) |

## Independent t: Worked Example

```{r}
#| fig-width: 10
#| fig-height: 5

met   <- rnorm(50, mean = 7.1, sd = 0.9)
glim  <- rnorm(50, mean = 7.5, sd = 1.1)
t_ind <- t.test(met, glim)

df_hba1c <- data.frame(
  Drug = rep(c("Metformin", "Glimepiride"), each = 50),
  HbA1c = c(met, glim)
)

ggplot(df_hba1c, aes(x = Drug, y = HbA1c, fill = Drug)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "#e74c3c") +
  scale_fill_manual(values = c("Metformin" = "#3498db", "Glimepiride" = "#e67e22")) +
  labs(title = "HbA1c: Metformin vs. Glimepiride",
       subtitle = paste0("Mean diff = ", round(mean(met) - mean(glim), 2),
                        "% | p = ", round(t_ind$p.value, 4)),
       y = "HbA1c (%)") +
  theme_clean() +
  theme(legend.position = "none")
```

## Effect Size: Cohen's d

Statistical significance alone is not enough — **how large is the difference?**

. . .

$$d = \frac{\bar{x}_1 - \bar{x}_2}{s_p}$$

. . .

| Cohen's d | Interpretation |
|-----------|---------------|
| 0.2 | Small |
| 0.5 | Medium |
| 0.8 | Large |

. . .

**Clinical significance** adds context: a difference of 0.5% HbA1c may not meet the MCID of 0.5% — statistically significant but clinically marginal.

## Checking Assumptions

:::: {.columns}
::: {.column width="50%"}
**Normality**

- Shapiro-Wilk test (p > 0.05 → normal)
- Q-Q plot (points on diagonal → normal)
- Histograms
- CLT: n > 30 often suffices
:::

::: {.column width="50%"}
**Equal Variances**

- Levene's test (p > 0.05 → equal)
- Rule of thumb: ratio of larger to smaller SD < 2
- If unequal → use Welch's t-test
:::
::::

. . .

**When assumptions fail badly:**

→ Use **non-parametric alternatives** (Section 7)

## One-Way ANOVA: Three or More Groups

**Why not multiple t-tests?**

. . .

With 3 groups → 3 pairwise comparisons → inflated Type I error

$$\alpha_{\text{family}} = 1 - (1 - 0.05)^3 = 0.143$$

. . .

**ANOVA** tests all groups simultaneously:

- H₀: μ₁ = μ₂ = μ₃ (all means equal)
- H₁: At least one mean differs

. . .

$$F = \frac{MS_{\text{between}}}{MS_{\text{within}}} = \frac{\text{Signal (between-group variance)}}{\text{Noise (within-group variance)}}$$

## ANOVA: Anaemia Trial Example

```{r}
#| fig-width: 10
#| fig-height: 5

oral <- rnorm(30, mean = 10.8, sd = 1.4)
iv   <- rnorm(30, mean = 11.5, sd = 1.6)
plac <- rnorm(30, mean = 9.6, sd = 1.5)

df_anova <- data.frame(
  Treatment = rep(c("Oral Iron", "IV Iron", "Placebo"), each = 30),
  Hb = c(oral, iv, plac)
)
df_anova$Treatment <- factor(df_anova$Treatment,
                              levels = c("Placebo", "Oral Iron", "IV Iron"))

anova_result <- aov(Hb ~ Treatment, data = df_anova)
f_val <- summary(anova_result)[[1]]["Treatment", "F value"]
p_val <- summary(anova_result)[[1]]["Treatment", "Pr(>F)"]

ggplot(df_anova, aes(x = Treatment, y = Hb, fill = Treatment)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "#e74c3c") +
  scale_fill_manual(values = c("Placebo" = "#95a5a6", "Oral Iron" = "#3498db",
                                "IV Iron" = "#2ecc71")) +
  labs(title = "Haemoglobin at 12 Weeks — ICMR Trial, Jharkhand",
       subtitle = paste0("F = ", round(f_val, 2), ", p = ",
                        format(p_val, digits = 3, scientific = TRUE)),
       y = "Haemoglobin (g/dL)") +
  theme_clean() +
  theme(legend.position = "none")
```

## The ANOVA Table

| Source | df | SS | MS | F |
|--------|----|----|----|----|
| Between groups | k − 1 | SS_B | SS_B / (k−1) | MS_B / MS_W |
| Within groups | N − k | SS_W | SS_W / (N−k) | |
| Total | N − 1 | SS_T | | |

. . .

**Significant F → at least one group differs, but WHICH ones?**

→ Post-hoc tests

## Post-Hoc Comparisons

ANOVA says "something differs" — post-hoc says **what**.

. . .

| Method | When to Use | Correction |
|--------|------------|------------|
| **Tukey HSD** | All pairwise comparisons | Controls familywise error |
| **Bonferroni** | Conservative; few planned comparisons | α / number of tests |
| **Dunnett** | Compare each treatment to one control | Only vs. control |

. . .

```{r}
tukey <- TukeyHSD(anova_result)
tukey_df <- as.data.frame(tukey$Treatment)
tukey_df$Comparison <- rownames(tukey_df)

kable(tukey_df[, c("Comparison", "diff", "lwr", "upr", "p adj")],
      col.names = c("Comparison", "Mean Diff", "Lower CI", "Upper CI", "Adj. p"),
      digits = 3, format = "html") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("bordered"))
```

## Non-Parametric Alternatives: When to Use

Use when:

- Data are **ordinal** (e.g., pain VAS, Likert scales)
- Distribution is **heavily skewed** or has outliers
- **Small samples** (n < 15–20 per group) and normality uncertain

. . .

| Parametric | Non-Parametric |
|-----------|---------------|
| Independent t-test | Mann-Whitney U |
| Paired t-test | Wilcoxon signed-rank |
| One-way ANOVA | Kruskal-Wallis |

## Mann-Whitney U Test

**Example:** Length of stay — laparoscopic vs. open cholecystectomy at Safdarjung Hospital, Delhi

```{r}
#| fig-width: 10
#| fig-height: 5

los_lap  <- c(2, 3, 2, 3, 4, 2, 3, 3, 2, 4, 3, 2, 5, 3, 2, 3, 4, 2, 3, 3,
               2, 3, 3, 4, 2, 3, 2, 3, 4, 3)
los_open <- c(5, 6, 7, 5, 8, 6, 7, 5, 6, 9, 7, 6, 8, 5, 6, 7, 5, 8, 6, 7,
               10, 6, 7, 8, 5, 6, 7, 5, 6, 8)

mw <- wilcox.test(los_lap, los_open)

df_los <- data.frame(
  Method = rep(c("Laparoscopic", "Open"), each = 30),
  LOS = c(los_lap, los_open)
)

ggplot(df_los, aes(x = Method, y = LOS, fill = Method)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 2) +
  scale_fill_manual(values = c("Laparoscopic" = "#2ecc71", "Open" = "#e74c3c")) +
  labs(title = "Hospital Stay: Laparoscopic vs. Open Cholecystectomy",
       subtitle = paste0("Mann-Whitney p = ", format(mw$p.value, digits = 3)),
       y = "Length of Stay (days)") +
  theme_clean() +
  theme(legend.position = "none")
```

## Wilcoxon Signed-Rank Test

**Example:** Pain VAS (0–10) before and after physiotherapy at CMC Vellore — 25 knee OA patients

```{r}
#| fig-width: 10
#| fig-height: 5

pain_before <- c(7, 8, 6, 9, 7, 8, 6, 7, 8, 9, 7, 6, 8, 7, 9, 6, 8, 7, 7, 8,
                  6, 9, 7, 8, 7)
pain_after  <- c(4, 5, 3, 6, 4, 5, 4, 3, 5, 6, 4, 3, 5, 4, 5, 3, 5, 4, 4, 5,
                  3, 6, 4, 5, 4)

wt <- wilcox.test(pain_before, pain_after, paired = TRUE)

df_pain <- data.frame(
  Patient = rep(1:25, 2),
  Time = rep(c("Before", "After"), each = 25),
  VAS = c(pain_before, pain_after)
)

ggplot(df_pain, aes(x = Time, y = VAS)) +
  geom_line(aes(group = Patient), alpha = 0.3, color = "gray60") +
  geom_point(aes(color = Time), size = 2.5) +
  stat_summary(fun = median, geom = "point", size = 5, shape = 18, color = "#2c3e50") +
  scale_color_manual(values = c("Before" = "#e74c3c", "After" = "#2ecc71")) +
  labs(title = "Pain VAS Before and After Physiotherapy, CMC Vellore",
       subtitle = paste0("Wilcoxon signed-rank p = ", format(wt$p.value, digits = 3)),
       y = "Pain VAS (0–10)") +
  theme_clean() +
  theme(legend.position = "none")
```

## Kruskal-Wallis Test

**Example:** Patient satisfaction (Likert 1–5) across 3 CHCs in Chhattisgarh

```{r}
#| fig-width: 10
#| fig-height: 5

chc_a <- sample(3:5, 25, replace = TRUE, prob = c(0.2, 0.4, 0.4))
chc_b <- sample(2:4, 25, replace = TRUE, prob = c(0.3, 0.5, 0.2))
chc_c <- sample(1:4, 25, replace = TRUE, prob = c(0.1, 0.3, 0.4, 0.2))

kw <- kruskal.test(list(chc_a, chc_b, chc_c))

df_sat <- data.frame(
  CHC = rep(c("CHC Raipur", "CHC Bilaspur", "CHC Korba"), each = 25),
  Satisfaction = c(chc_a, chc_b, chc_c)
)

ggplot(df_sat, aes(x = CHC, y = Satisfaction, fill = CHC)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 2) +
  scale_fill_manual(values = c("#3498db", "#e67e22", "#2ecc71")) +
  labs(title = "Patient Satisfaction Across 3 CHCs, Chhattisgarh",
       subtitle = paste0("Kruskal-Wallis p = ", round(kw$p.value, 4)),
       y = "Satisfaction (1–5 Likert)") +
  theme_clean() +
  theme(legend.position = "none")
```

## Decision Flowchart

```{r}
#| fig-width: 11
#| fig-height: 6

flow <- tibble(
  x = c(1, 1, 1,   3, 3, 3,   5, 5, 5),
  y = c(3, 2, 1,   3, 2, 1,   3, 2, 1),
  label = c("One-sample t", "Paired t", "Independent t",
            "—", "Wilcoxon signed-rank", "Mann-Whitney U",
            "—", "—", "Kruskal-Wallis"),
  actual = c("One-sample t", "Paired t", "Independent t",
             "", "Wilcoxon\nsigned-rank", "Mann-Whitney U",
             "", "", "Kruskal-Wallis"),
  header = c(rep("Parametric", 3), rep("Non-Parametric\n(2 groups)", 3),
             rep("Non-Parametric\n(3+ groups)", 3))
)

decision_table <- tibble(
  Scenario = c("One group vs. reference value",
               "Two groups, same subjects (paired)",
               "Two independent groups",
               "Three+ independent groups"),
  Parametric = c("One-sample t-test", "Paired t-test",
                 "Independent t-test", "One-way ANOVA"),
  `Non-Parametric` = c("Wilcoxon signed-rank*", "Wilcoxon signed-rank",
                        "Mann-Whitney U", "Kruskal-Wallis")
)

kable(decision_table, format = "html") %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("bordered", "hover")) %>%
  row_spec(0, bold = TRUE, background = "#1e3a5f", color = "white") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "* One-sample Wilcoxon signed-rank tests median vs. hypothesised value")
```

## Reporting Template

**t-test:**

> Mean HbA1c was lower in the metformin group (7.1 ± 0.9%) than in the glimepiride group (7.5 ± 1.1%); t(98) = −2.14, p = 0.035, 95% CI for difference [−0.77, −0.03], Cohen's d = 0.40.

. . .

**ANOVA:**

> One-way ANOVA showed a significant difference in Hb across treatment groups, F(2, 87) = 12.4, p < 0.001. Post-hoc Tukey HSD revealed IV iron was significantly higher than placebo (mean diff = 1.9 g/dL, p < 0.001).

. . .

**Non-parametric:**

> Median LOS was significantly shorter in the laparoscopic group (3 days, IQR 2–4) than the open group (6 days, IQR 5–7); Mann-Whitney U, p < 0.001.

## Do's and Don'ts

:::: {.columns}
::: {.column width="50%"}
### Do

- Check normality and variance before choosing a test
- Report effect sizes (Cohen's d, η²)
- Use Welch's t-test by default for independent samples
- Apply post-hoc corrections after ANOVA
- Report medians and IQR for non-parametric tests
:::

::: {.column width="50%"}
### Don't

- Run multiple t-tests instead of ANOVA
- Ignore assumption violations
- Report only p-values without effect sizes
- Use parametric tests on ordinal data
- Forget to state whether the test was one-tailed or two-tailed
:::
::::

## Key Takeaways

1. **One-sample t** → one group vs. reference value
2. **Paired t** → same subjects, two timepoints
3. **Independent t** → two separate groups
4. **ANOVA** → three or more groups; follow with post-hoc if significant
5. **Non-parametric alternatives** for skewed data or ordinal scales
6. **Always report:** test statistic, df, p-value, CI, and effect size
7. **Clinical significance** ≠ statistical significance — interpret both
