---
title: "Categorical Data Analysis"
subtitle: "Chi-square, Fisher's, and Risk Measures"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    footer: "Biostatistics for Clinicians | Module 10"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
library(vcd)
set.seed(42)

theme_clean <- function() {
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    legend.position = "bottom"
  )
}
```

## The Clinical Problem

Monsoon season — a village near Sagar, MP. 200 people attend a wedding feast. Within 48 hours, **80 develop acute gastroenteritis**.

. . .

The district surveillance officer constructs a 2×2 table for biryani exposure:

|  | Ate Biryani | Did Not | Total |
|--|---|---|---|
| **Ill** | 65 | 15 | 80 |
| **Well** | 30 | 50 | 80 |
| **Total** | 95 | 65 | 160 |

. . .

**Questions:** Is biryani associated with illness? How strong is the association? What's the clinical impact?

## The 2×2 Contingency Table

The fundamental building block of categorical data analysis:

|  | **Disease** | **No Disease** | **Total** |
|--|---|---|---|
| **Exposed** | a | b | a + b |
| **Not Exposed** | c | d | c + d |
| **Total** | a + c | b + d | N |

. . .

From this single table we can derive:

- **Tests:** Chi-square, Fisher's exact
- **Association:** RR, OR
- **Impact:** ARR, NNT, AR, PAR

## Chi-square Test of Independence

**H₀:** Exposure and disease are independent (no association)

**H₁:** Exposure and disease are associated

. . .

**In words:** For each cell, take (Observed − Expected)², divide by Expected, sum across all cells.

$$\chi^2 = \sum \frac{(\text{Observed} - \text{Expected})^2}{\text{Expected}}$$

. . .

For a 2×2 table, the **cross-product shortcut:**

$$\chi^2 = \frac{N(ad - bc)^2}{(a+b)(c+d)(a+c)(b+d)}, \quad df = 1$$

. . .

**Expected frequency:** $E_{ij} = \frac{\text{Row Total} \times \text{Column Total}}{\text{Grand Total}}$

## Worked Example: Biryani Outbreak

```{r}
#| fig-width: 10
#| fig-height: 5

a <- 65; b <- 30; c <- 15; d <- 50; N <- 160

E_ab <- (80 * 95) / 160  # expected: ill + biryani
chi_sq <- (N * (a*d - b*c)^2) / ((a+b) * (c+d) * (a+c) * (b+d))
p_val <- 1 - pchisq(chi_sq, df = 1)

freq_data <- data.frame(
  Category = c("Ate, Ill", "Ate, Well", "No, Ill", "No, Well"),
  Observed = c(65, 30, 15, 50),
  Expected = c((80*95)/160, (80*95)/160, (80*65)/160, (80*65)/160)
)

freq_long <- freq_data %>%
  pivot_longer(cols = c("Observed", "Expected"),
               names_to = "Type", values_to = "Freq")

ggplot(freq_long, aes(x = Category, y = Freq, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = round(Freq, 1)),
            position = position_dodge(0.9), vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("Observed" = "#2c3e50", "Expected" = "#bdc3c7")) +
  labs(title = "Observed vs. Expected Frequencies",
       subtitle = paste0("χ² = ", round(chi_sq, 2), ", p = ",
                        format(p_val, digits = 3, scientific = TRUE)),
       x = "", y = "Frequency") +
  theme_clean() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

**Reject H₀** — biryani exposure and illness are significantly associated.

## When Chi-square Fails: Fisher's Exact Test

**Use Fisher's exact test when:**

- Any expected cell frequency < 5
- Total sample size N < 20–30
- Rare disease or rare exposure

. . .

**Example:** Leptospirosis cluster after flooding in Alappuzha, Kerala (N = 25):

```{r}
small <- matrix(c(8, 2, 3, 12), nrow = 2, byrow = TRUE)
dimnames(small) <- list(
  c("Waded in floodwater", "Did not wade"),
  c("Leptospirosis", "Healthy")
)
ft <- fisher.test(small)

kable(small, format = "html") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("bordered"))
```

Fisher's p = `r round(ft$p.value, 4)`, OR = `r round(ft$estimate, 2)`

## McNemar's Test: Paired Categorical Data

**Same subjects, before–after** → only **discordant pairs** matter

. . .

**Example:** TB knowledge in 100 contacts, Patna — before and after health education

|  | After: Correct | After: Incorrect |
|--|---|---|
| **Before: Correct** | 55 | **5** |
| **Before: Incorrect** | **25** | 15 |

. . .

$$\chi^2_{\text{McNemar}} = \frac{(b - c)^2}{b + c} = \frac{(5 - 25)^2}{30} = 13.33, \quad p < 0.001$$

. . .

Knowledge significantly improved after education. The 55 + 15 concordant pairs are **ignored**.

## Measures of Association: Overview

```{r}
measures <- tibble(
  Measure = c("Relative Risk (RR)", "Odds Ratio (OR)",
              "ARR", "RRR", "NNT", "NNH", "AR", "PAR"),
  `In Words` = c(
    "Risk in Exposed ÷ Risk in Unexposed",
    "Odds in Exposed ÷ Odds in Unexposed",
    "Risk in Control − Risk in Treatment",
    "ARR ÷ Risk in Control",
    "1 ÷ ARR (treat how many to save one?)",
    "1 ÷ Absolute Risk of Harm",
    "Risk in Exposed − Risk in Unexposed",
    "Fraction of population disease due to exposure"
  ),
  `Cell Notation` = c("a/(a+b) ÷ c/(c+d)", "ad / bc",
              "Rc − Rt", "ARR / Rc",
              "1 / ARR", "1 / ARH",
              "Re − Ru", "Pe(RR−1) / [Pe(RR−1)+1]"),
  Use = c("Cohort", "Case-control",
          "Trials", "Trials",
          "Clinical", "Safety",
          "Individual", "Public health")
)

kable(measures, format = "html") %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("bordered", "hover")) %>%
  row_spec(0, bold = TRUE, background = "#1e3a5f", color = "white") %>%
  column_spec(1, bold = TRUE)
```

## Relative Risk (RR)

**In words:** Risk of disease in Exposed ÷ Risk of disease in Unexposed

$$RR = \frac{\text{Incidence in Exposed}}{\text{Incidence in Unexposed}} = \frac{a/(a+b)}{c/(c+d)}$$

. . .

**Outbreak example:**

```{r}
risk_exp   <- a / (a + b)
risk_unexp <- c / (c + d)
RR <- risk_exp / risk_unexp
```

- Risk in exposed (biryani): `r round(risk_exp, 3)` (`r round(risk_exp*100, 1)`%)
- Risk in unexposed: `r round(risk_unexp, 3)` (`r round(risk_unexp*100, 1)`%)
- **RR = `r round(RR, 2)`** — biryani eaters were `r round(RR, 1)`× more likely to fall ill

. . .

**Key:** RR can only be calculated from **cohort studies** where you follow exposed and unexposed groups forward.

## Odds Ratio (OR)

**In words:** Odds of disease in Exposed ÷ Odds of disease in Unexposed

$$OR = \frac{\text{Odds in Exposed}}{\text{Odds in Unexposed}} = \frac{ad}{bc} \quad \text{(cross-product ratio)}$$

```{r}
OR <- (a * d) / (b * c)
```

- **OR = `r round(OR, 2)`**

. . .

**Why not just use RR?**

In **case-control studies**, you cannot calculate incidence → cannot get RR → must use OR.

. . .

**The critical question:** When does OR ≈ RR?

## OR ≈ RR: The Rare Disease Assumption

$$OR \approx RR \text{ when disease prevalence < 5–10%}$$

```{r}
#| fig-width: 10
#| fig-height: 5

prev_range <- seq(0.01, 0.60, by = 0.02)
rr_val <- 2.5
or_vals <- rr_val * ((1 - prev_range) / (1 - prev_range * rr_val))

df_or_rr <- data.frame(
  Prevalence = prev_range * 100,
  RR = rr_val,
  OR = or_vals
) %>%
  pivot_longer(cols = c("RR", "OR"), names_to = "Measure", values_to = "Value")

ggplot(df_or_rr, aes(x = Prevalence, y = Value, color = Measure,
                      linetype = Measure)) +
  geom_line(linewidth = 1.3) +
  scale_color_manual(values = c("RR" = "#2980b9", "OR" = "#e74c3c")) +
  annotate("rect", xmin = 0, xmax = 10, ymin = 0, ymax = Inf,
           alpha = 0.1, fill = "#2ecc71") +
  annotate("text", x = 5, y = 6, label = "Rare disease\nOR ≈ RR",
           fontface = "bold", color = "#27ae60", size = 4) +
  annotate("text", x = 45, y = 6, label = "Common disease\nOR >> RR",
           fontface = "bold", color = "#e74c3c", size = 4) +
  labs(title = "OR Diverges from RR as Prevalence Increases",
       subtitle = "True RR = 2.5 held constant",
       x = "Disease Prevalence (%)", y = "Measure Value") +
  theme_clean()
```

## 95% Confidence Interval for OR

$$OR_{95\%CI} = OR \times e^{\pm\, 1.96 \times SE(\ln OR)}$$

where $SE(\ln OR) = \sqrt{1/a + 1/b + 1/c + 1/d}$

. . .

```{r}
ln_OR <- log(OR)
se_ln_OR <- sqrt(1/a + 1/b + 1/c + 1/d)
or_ci <- exp(ln_OR + c(-1, 1) * 1.96 * se_ln_OR)
```

**Outbreak result:** OR = `r round(OR, 2)` (95% CI: `r round(or_ci[1], 2)` – `r round(or_ci[2], 2)`)

. . .

**Interpretation rule:**

- CI includes 1.0 → **not significant** (p > 0.05)
- CI excludes 1.0 → **significant** (p < 0.05)

## ARR, RRR, and NNT

**TB vaccine trial in Varanasi:** Vaccinated 5/500 (1%); Unvaccinated 20/500 (4%)

. . .

**ARR** (how much does treatment lower absolute risk?):

$$ARR = \text{Risk in Control} - \text{Risk in Treatment} = 4\% - 1\% = 3\%$$

. . .

**RRR** (what fraction of baseline risk is removed?):

$$RRR = \frac{ARR}{\text{Risk in Control}} = \frac{3\%}{4\%} = 75\%$$

. . .

**NNT** (how many to treat to prevent one case?):

$$NNT = \frac{1}{ARR} = \frac{1}{0.03} \approx 33$$

. . .

**Clinical meaning:** Vaccinate 33 people to prevent one case of TB.

:::{.callout-warning}
RRR (75%) sounds impressive; ARR (3%) gives the true clinical picture.
:::

## NNT and NNH: The Full Picture

```{r}
#| fig-width: 10
#| fig-height: 5

nnt_nnh <- tibble(
  Measure = c("NNT\n(benefit)", "NNH\n(harm)"),
  Value = c(33, 83),
  Color = c("#2ecc71", "#e74c3c")
)

ggplot(nnt_nnh, aes(x = Measure, y = Value, fill = Color)) +
  geom_col(width = 0.5, alpha = 0.85) +
  geom_text(aes(label = Value), vjust = -0.5, size = 7, fontface = "bold") +
  scale_fill_identity() +
  labs(title = "Benefit vs. Harm: TB Vaccine Trial",
       subtitle = "NNH/NNT = 2.5 → benefit outweighs harm",
       y = "Number Needed") +
  theme_clean() +
  theme(axis.title.x = element_blank())
```

**Therapeutic index** = NNH / NNT = 83 / 33 = 2.5 → Benefit outweighs harm → Recommend vaccination

## Attributable Risk (AR) and Population Attributable Risk (PAR)

:::: {.columns}
::: {.column width="50%"}
**AR** — excess risk from exposure (individual level)

$$AR = \text{Risk in Exposed} - \text{Risk in Unexposed}$$

*"How much extra disease does the exposure cause?"*

. . .

**Smoking example:**

AR = 8% − 1% = 7%

AR% = 7/8 = **87.5%** of lung cancer in smokers is attributable to smoking
:::

::: {.column width="50%"}
**PAR** — population-level impact (depends on both RR *and* exposure prevalence)

$$PAR\% = \frac{P_e(RR - 1)}{P_e(RR - 1) + 1} \times 100$$

*"If we eliminated this exposure, what fraction of ALL disease would disappear?"*

. . .

If 40% smoke, RR = 8:

PAR% = **73.7%** of all lung cancer could be prevented by eliminating smoking
:::
::::

## Decision Guide: Choosing the Right Test

```{r}
guide <- tibble(
  Scenario = c("2 independent groups, expected freq ≥ 5",
               "2 independent groups, expected freq < 5",
               "Paired / before–after",
               "Cohort study → risk measure",
               "Case-control study → association measure",
               "Clinical impact of treatment"),
  Test = c("Chi-square test",
           "Fisher's exact test",
           "McNemar's test",
           "Relative Risk (RR)",
           "Odds Ratio (OR)",
           "ARR, RRR, NNT, NNH")
)

kable(guide, col.names = c("Scenario", "Test / Measure"),
      format = "html") %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("bordered", "hover")) %>%
  row_spec(0, bold = TRUE, background = "#1e3a5f", color = "white") %>%
  column_spec(1, bold = TRUE, width = "55%") %>%
  column_spec(2, bold = TRUE, color = "#2c3e50")
```

## Common Pitfalls

:::: {.columns}
::: {.column width="50%"}
### Avoid

- Chi-square with expected freq < 5
- Interpreting OR as RR at high prevalence
- Claiming causation from chi-square alone
- Forgetting to report 95% CI for OR/RR
- Confusing ARR (absolute) with RRR (relative)
:::

::: {.column width="50%"}
### Remember

- Fisher's exact test for small samples
- OR ≈ RR only when disease is rare (< 5–10%)
- McNemar's for paired categorical data
- NNT = 1/ARR, not 1/RRR
- CI excluding 1.0 → statistically significant
:::
::::

## Key Takeaways

1. **Chi-square** tests association between two categorical variables (check expected freq ≥ 5)
2. **Fisher's exact** for small samples; **McNemar's** for paired before–after data
3. **RR** from cohort studies; **OR** from case-control studies
4. **OR ≈ RR** only when disease is rare (< 5–10% prevalence)
5. **ARR** gives the true clinical impact; **RRR** can be misleading
6. **NNT = 1/ARR** — lower is better; compare NNT to NNH for net benefit
7. **AR** for individual counselling; **PAR** for public health policy
8. Always report **95% CI** — if it excludes 1.0, the result is significant
