---
title: "Correlation and Regression"
subtitle: "From Measuring Association to Predicting Outcomes"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    footer: "Biostatistics for Clinicians | Module 11"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
library(broom)
set.seed(42)

theme_clean <- function() {
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.3),
    legend.position = "bottom"
  )
}

# Generate data
n <- 200
age <- round(rnorm(n, mean = 45, sd = 12))
bmi <- round(rnorm(n, mean = 26, sd = 4.5), 1)
fbs <- round(80 + 1.8 * bmi + 0.5 * age + rnorm(n, sd = 15))
health <- data.frame(ID = 1:n, Age = age, BMI = bmi, FBS = fbs,
                      Sex = sample(c("Male", "Female"), n, replace = TRUE))
health$diabetes_prob <- plogis(-8 + 0.15 * health$BMI + 0.05 * health$Age)
health$Diabetes <- rbinom(n, 1, health$diabetes_prob)
```

## The Clinical Problem

A research team at JIPMER, Puducherry studies 200 adults at a health screening camp.

. . .

**Questions that cascade:**

1. Is higher BMI **associated** with higher FBS? → Correlation
2. Can we **predict** FBS from BMI? → Linear regression
3. Does the association **persist** after adjusting for age and sex? → Multiple regression
4. What if the outcome is diabetes (**yes/no**)? → Logistic regression

## Roadmap

```{r}
#| fig-width: 11
#| fig-height: 3

road <- tibble(
  x = 1:6,
  label = c("Correlation", "Simple\nRegression", "Assumptions", "Multiple\nRegression",
            "Logistic\nRegression", "Framework"),
  fill = c("#2c3e50", "#2c3e50", "#e67e22", "#2980b9", "#8e44ad", "#27ae60")
)

ggplot(road, aes(x = x, y = 0, label = label, fill = fill)) +
  geom_tile(width = 0.9, height = 0.5, color = "white", linewidth = 1.5) +
  geom_text(size = 3.5, fontface = "bold", color = "white") +
  scale_fill_identity() +
  coord_fixed(ratio = 1.2) +
  theme_void()
```

## What Correlation Measures

**In words:** As one variable goes up, does the other tend to go up, go down, or show no pattern?

. . .

$$r = \frac{\text{Covariance}(x, y)}{SD_x \times SD_y}$$

. . .

| Range | Meaning |
|-------|---------|
| r = +1 | Perfect positive linear |
| r = 0 | No linear association |
| r = −1 | Perfect negative linear |

. . .

**Two versions:** Pearson (for continuous, linear) and Spearman (for ordinal, or when outliers/skew present)

## BMI vs. FBS: Scatter Plot

```{r}
#| fig-width: 10
#| fig-height: 6

r_val <- cor(health$BMI, health$FBS)

ggplot(health, aes(x = BMI, y = FBS)) +
  geom_point(alpha = 0.5, color = "#2c3e50") +
  geom_smooth(method = "lm", se = TRUE, color = "#e74c3c", linewidth = 1) +
  annotate("text", x = 36, y = 85,
           label = paste0("r = ", round(r_val, 3), "\np < 0.001"),
           fontface = "bold", color = "#e74c3c", size = 5) +
  labs(title = "BMI vs. Fasting Blood Sugar",
       subtitle = "JIPMER Health Camp, n = 200",
       x = "BMI (kg/m²)", y = "FBS (mg/dL)") +
  theme_clean()
```

r = `r round(r_val, 3)` — moderate positive correlation

## Pearson vs. Spearman

| Feature | Pearson (r) | Spearman (rₛ) |
|---------|------------|---------------|
| Measures | Linear association | Monotonic association |
| Data | Continuous, ~normal | Ordinal or skewed |
| Outlier-sensitive? | Yes | No (uses ranks) |
| Use when | Scatter looks linear, no extreme outliers | Ranks, ordinal data, non-linear monotonic |

. . .

**Rule of thumb:** If in doubt, compute both. If they agree, report Pearson. If they disagree substantially, investigate the scatter plot for non-linearity or outliers.

## Correlation ≠ Causation

:::: {.columns}
::: {.column width="50%"}
**Ice cream sales ↑ and drowning deaths ↑**

r ≈ 0.8 — very strong!

Should we ban ice cream?

. . .

**No.** Lurking variable = **temperature**
:::

::: {.column width="50%"}
**Sources of spurious association:**

- Confounding (third variable)
- Reverse causality (y → x)
- Selection bias
- Measurement error
:::
::::

. . .

A strong r means variables *move together* — not that one *causes* the other.

## Simple Linear Regression: The Model

**In words:** Predicted outcome = Baseline + (Rate of change × Predictor) + Error

$$y = \beta_0 + \beta_1 x + \varepsilon$$

. . .

- **β₀ (intercept):** Predicted y when x = 0
- **β₁ (slope):** Change in y for each 1-unit increase in x
- **ε:** What the model can't explain

. . .

**Least squares** finds β₀ and β₁ that minimise $\sum(y_i - \hat{y}_i)^2$

## Regression: BMI → FBS

```{r}
#| fig-width: 10
#| fig-height: 6

model <- lm(FBS ~ BMI, data = health)
b0 <- coef(model)[1]; b1 <- coef(model)[2]
r_sq <- summary(model)$r.squared

ggplot(health, aes(x = BMI, y = FBS)) +
  geom_point(alpha = 0.4, color = "#2c3e50") +
  geom_smooth(method = "lm", se = TRUE, color = "#e74c3c", fill = "#e74c3c",
              alpha = 0.15, linewidth = 1) +
  annotate("text", x = 36, y = 90,
           label = paste0("FBS = ", round(b0, 1), " + ", round(b1, 1),
                         " × BMI\nR² = ", round(r_sq, 3)),
           fontface = "bold", color = "#e74c3c", size = 4.5) +
  labs(title = "Simple Linear Regression: BMI → FBS",
       x = "BMI (kg/m²)", y = "FBS (mg/dL)") +
  theme_clean()
```

**Slope:** Each 1 kg/m² BMI increase → `r round(b1, 1)` mg/dL higher FBS

## R²: Variance Explained

**In words:** What fraction of the total scatter in y is captured by the regression line?

$$R^2 = 1 - \frac{\sum(y_i - \hat{y}_i)^2}{\sum(y_i - \bar{y})^2}$$

. . .

For simple regression: $R^2 = r^2$

. . .

**Our model:** R² = `r round(r_sq, 3)` → BMI explains `r round(r_sq * 100, 1)`% of FBS variation

. . .

The remaining `r round((1 - r_sq) * 100, 1)`% is due to age, diet, genetics, measurement error, etc.

## Confidence vs. Prediction Interval

```{r}
#| fig-width: 10
#| fig-height: 6

new_data <- data.frame(BMI = seq(16, 40, by = 0.5))
ci <- predict(model, newdata = new_data, interval = "confidence")
pi <- predict(model, newdata = new_data, interval = "prediction")
pd <- data.frame(BMI = new_data$BMI, Fit = ci[,1],
                  CI_lo = ci[,2], CI_hi = ci[,3],
                  PI_lo = pi[,2], PI_hi = pi[,3])

ggplot() +
  geom_ribbon(data = pd, aes(x = BMI, ymin = PI_lo, ymax = PI_hi),
              fill = "#3498db", alpha = 0.15) +
  geom_ribbon(data = pd, aes(x = BMI, ymin = CI_lo, ymax = CI_hi),
              fill = "#e74c3c", alpha = 0.25) +
  geom_line(data = pd, aes(x = BMI, y = Fit), color = "#2c3e50", linewidth = 1) +
  geom_point(data = health, aes(x = BMI, y = FBS), alpha = 0.3) +
  annotate("text", x = 37, y = 170, label = "Prediction interval\n(single patient)",
           color = "#3498db", fontface = "bold", size = 3.5) +
  annotate("text", x = 37, y = 135, label = "Confidence interval\n(population mean)",
           color = "#e74c3c", fontface = "bold", size = 3.5) +
  labs(title = "CI vs. PI: Two Types of Uncertainty",
       x = "BMI (kg/m²)", y = "FBS (mg/dL)") +
  theme_clean()
```

## Regression Assumptions: LINE

1. **L**inearity — scatter plot shows a linear trend
2. **I**ndependence — observations are independent
3. **N**ormality — *residuals* (not x!) are approximately normal
4. **E**qual variance — residual spread is constant (homoscedasticity)

. . .

**Check with diagnostic plots:**

- Residuals vs. Fitted → random cloud (not curved, not funnel)
- Q-Q plot → points on diagonal
- Scale-Location → flat trend
- Cook's distance → no single observation dominates

## What to Do When Assumptions Fail

| Problem | Sign | Fix |
|---------|------|-----|
| Non-linear | Curved residuals | Transform x or y; polynomial term |
| Non-normal residuals | Q-Q deviates at tails | Transform y; often OK if n > 30 |
| Heteroscedasticity | Fan-shaped residuals | Log-transform y; robust SE |
| Influential outlier | Cook's distance > 0.5 | Investigate; report with and without |

## Multiple Regression: Adjusting for Confounders

**Question:** Is the BMI–FBS association genuine, or confounded by age?

. . .

$$\text{FBS} = \beta_0 + \beta_1 \times BMI + \beta_2 \times Age + \beta_3 \times Sex + \varepsilon$$

. . .

**In words:** Each coefficient tells you the effect of that variable *while all others are held fixed*.

. . .

```{r}
model_multi <- lm(FBS ~ BMI + Age + Sex, data = health)
cs <- tidy(model, conf.int = TRUE)
cm <- tidy(model_multi, conf.int = TRUE)

comp <- tibble(
  Model = c("Simple (BMI only)", "Adjusted (+ Age, Sex)"),
  `BMI slope` = c(round(cs$estimate[2], 2), round(cm$estimate[2], 2)),
  `95% CI` = c(paste0(round(cs$conf.low[2], 2), " – ", round(cs$conf.high[2], 2)),
               paste0(round(cm$conf.low[2], 2), " – ", round(cm$conf.high[2], 2))),
  `Adj R²` = c(round(summary(model)$adj.r.squared, 3),
               round(summary(model_multi)$adj.r.squared, 3))
)

kable(comp, format = "html") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("bordered"))
```

If the slope changes after adjustment → **confounding was present**.

## Logistic Regression: Binary Outcomes

When the outcome is **yes/no** (diabetes, death, readmission), linear regression breaks down. Logistic regression models the **log-odds**:

. . .

$$\ln\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x$$

. . .

The **odds ratio** for a 1-unit increase in x:

$$OR = e^{\beta_1}$$

. . .

**Connection to Module 10:** The 2×2 table OR is a special case. Logistic regression is the **multivariable extension**.

## Logistic Regression: Predicted Probabilities

```{r}
#| fig-width: 10
#| fig-height: 6

logit <- glm(Diabetes ~ BMI + Age + Sex, data = health, family = binomial)
bmi_seq <- data.frame(BMI = seq(16, 40, by = 0.5), Age = 45, Sex = "Male")
bmi_seq$prob <- predict(logit, newdata = bmi_seq, type = "response")

ggplot(bmi_seq, aes(x = BMI, y = prob)) +
  geom_line(color = "#8e44ad", linewidth = 1.3) +
  geom_rug(data = health %>% filter(Diabetes == 1), aes(x = BMI, y = NULL),
           sides = "t", alpha = 0.3, color = "#e74c3c") +
  geom_rug(data = health %>% filter(Diabetes == 0), aes(x = BMI, y = NULL),
           sides = "b", alpha = 0.3, color = "#2ecc71") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(title = "Predicted Probability of Diabetes by BMI",
       subtitle = "Logistic regression (Age = 45, Male) | Rug = actual cases",
       x = "BMI (kg/m²)", y = "P(Diabetes)") +
  theme_clean()
```

The **sigmoid curve** — probability bounded between 0 and 1

## Decision Framework

```{r}
decision <- tibble(
  Scenario = c("Strength of association (two continuous)?",
               "Ordinal data or outliers?",
               "Predict continuous outcome (one predictor)?",
               "Adjust for confounders?",
               "Predict binary outcome (yes/no)?",
               "Predict time-to-event?"),
  Method = c("Pearson r", "Spearman rₛ",
             "Simple linear regression", "Multiple linear regression",
             "Logistic regression", "Cox regression (Module 12)")
)

kable(decision, format = "html") %>%
  kable_styling(full_width = TRUE, bootstrap_options = c("bordered", "hover")) %>%
  row_spec(0, bold = TRUE, background = "#1e3a5f", color = "white") %>%
  column_spec(1, bold = TRUE, width = "55%")
```

## Common Pitfalls

:::: {.columns}
::: {.column width="50%"}
### Avoid

- Assuming correlation proves causation
- Using linear regression for binary outcomes
- Ignoring residual plots
- Extrapolating beyond the data range
- Reporting only R² without checking assumptions
:::

::: {.column width="50%"}
### Remember

- r measures *linear* association only
- R² = r² for simple regression
- Adjusted R² for comparing models
- Logistic regression → OR, not risk ratio
- Regression adjusts for *measured* confounders only
:::
::::

## Key Takeaways

1. **Pearson r** = linear association; **Spearman rₛ** = rank-based alternative
2. **Correlation ≠ causation** — always consider confounding
3. **Simple regression:** y = β₀ + β₁x; slope = change per unit; R² = variance explained
4. **Check LINE assumptions** with diagnostic plots
5. **CI** for the mean; **PI** for a new individual — always report uncertainty
6. **Multiple regression** adjusts for confounders; use **adjusted R²**
7. **Logistic regression** for binary outcomes → coefficients exponentiate to **ORs**
8. **No regression proves causation** in observational data — only RCTs can
