---
title: "Meta-Analysis and Evidence Synthesis"
subtitle: "Combining Evidence from Multiple Studies"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    scrollable: true
    footer: "Biostatistics for Clinicians | Module 13"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
set.seed(42)
```

## Learning Objectives

::: {.incremental}
1. **Distinguish** systematic review from meta-analysis
2. **Read and interpret** forest plots — estimates, CIs, weights, and pooled effects
3. **Choose** between fixed-effect and random-effects models
4. **Calculate and interpret** I² for heterogeneity
5. **Assess** publication bias using funnel plots
6. **Evaluate** evidence quality using GRADE
7. **Recognise** when meta-analysis is inappropriate
:::

## Clinical Hook: TB in Healthcare Workers {.smaller}

::: {.columns}
::: {.column width="55%"}
**Rajnish Joshi** (now AIIMS Bhopal) led a systematic review in *PLoS Medicine* (2006):

**"TB among Health-Care Workers in Low- and Middle-Income Countries"**

Individual studies told different stories:

- Some: LTBI prevalence **33%**
- Others: LTBI prevalence **79%**
- Risk estimates ranged from **2× to 5×** the general population

. . .

The team synthesised **42 articles** (51 studies):

- Pooled LTBI prevalence: **54%** (range 33–79%)
- Revealed *why* estimates varied — TB burden, infection control, diagnostic method

**Meta-analysis turns conflicting signals into a coherent picture.**

*Joshi R, Reingold AL, Menzies D, Pai M. PLoS Med 2006;3(12):e494.*
:::

::: {.column width="45%"}
```{r}
#| fig-height: 5.5
#| fig-width: 5.5

# Conceptual diagram: many studies → one answer
concept_df <- tibble(
  study = paste("Study", 1:8),
  estimate = c(33, 55, 42, 79, 63, 45, 71, 54),
  y = 8:1
)

ggplot(concept_df, aes(x = estimate, y = y)) +
  geom_point(size = 4, color = "#2c3e50") +
  geom_segment(aes(x = estimate - runif(8, 5, 15), xend = estimate + runif(8, 5, 15),
                    yend = y), linewidth = 0.8, color = "#2c3e50") +
  geom_vline(xintercept = 54, linetype = "dashed", color = "#e74c3c", linewidth = 1.2) +
  annotate("text", x = 54, y = 0.3, label = "Pooled: 54%", fontface = "bold",
           color = "#e74c3c", size = 4.5) +
  scale_y_continuous(breaks = 1:8, labels = paste("Study", 8:1)) +
  labs(title = "Many Studies, One Synthesis",
       x = "LTBI Prevalence (%)", y = NULL) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
```
:::
:::

## Systematic Review vs Meta-Analysis {.smaller}

| | **Systematic Review** | **Meta-Analysis** |
|---|---|---|
| **What** | Structured method for finding, appraising, synthesising *all* evidence | Statistical technique combining results into a pooled estimate |
| **Nature** | Qualitative (can exist without statistics) | Quantitative (requires numbers) |
| **Stand-alone?** | Yes | No — needs a systematic review |
| **Output** | Evidence summary, quality assessment | Pooled effect, forest plot, I² |

. . .

::: {.callout-important title="The Hierarchy"}
A meta-analysis is only as good as the systematic review that feeds it. **Garbage in, garbage out.**
:::

## The PICO Framework {.smaller}

Every systematic review starts with a focused question:

| Component | Meaning | Example (Joshi et al.) |
|---|---|---|
| **P**opulation | Who? | Healthcare workers in LMICs |
| **I**ntervention/Exposure | What? | Working in healthcare settings |
| **C**omparator | Compared to? | General population |
| **O**utcome | What measured? | LTBI and active TB prevalence |

. . .

**Without a clear PICO, you don't know what to search for, what to include, or what to pool.**

## The Forest Plot — Anatomy {.smaller}

```{r}
#| fig-height: 6
#| fig-width: 10

# Simulated teaching data inspired by TB in healthcare workers
set.seed(42)
studies <- tibble(
  study = c("Pai 2005 (India)", "Yanai 2003 (Thailand)", "Joshi 2006 (Ethiopia)",
            "Kassim 2000 (Malawi)", "Balt 1998 (Russia)", "Rao 2009 (India)",
            "Naidoo 2006 (S. Africa)", "Dimitrova 2006 (Russia)"),
  or = c(2.4, 3.1, 1.8, 5.2, 2.9, 1.5, 4.1, 2.2),
  or_lo = c(1.5, 1.9, 1.1, 2.4, 1.6, 0.9, 2.0, 1.3),
  or_hi = c(3.8, 5.0, 3.0, 11.3, 5.3, 2.5, 8.4, 3.7),
  weight = c(16, 14, 15, 6, 10, 18, 8, 13)
)

pooled_or <- 2.6; pooled_lo <- 1.9; pooled_hi <- 3.5

studies <- studies %>% mutate(study = fct_rev(fct_inorder(study)))

ggplot(studies, aes(y = study)) +
  geom_point(aes(x = or, size = weight), color = "#2c3e50", shape = 15) +
  geom_errorbarh(aes(xmin = or_lo, xmax = or_hi), height = 0.2, color = "#2c3e50") +
  annotate("polygon",
           x = c(pooled_lo, pooled_or, pooled_hi, pooled_or),
           y = c(0.3, 0.55, 0.3, 0.05),
           fill = "#e74c3c", alpha = 0.7) +
  annotate("text", x = pooled_or, y = -0.3,
           label = paste0("Pooled OR = ", pooled_or, " (", pooled_lo, "–", pooled_hi, ")"),
           fontface = "bold", size = 3.5, color = "#e74c3c") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  scale_x_log10(breaks = c(0.5, 1, 2, 4, 8, 12)) +
  scale_size_continuous(range = c(2, 8), guide = "none") +
  labs(title = "Forest Plot: TB Risk in Healthcare Workers vs General Population",
       subtitle = "Simulated teaching data inspired by Joshi R et al. (PLoS Med, 2006)",
       x = "Odds Ratio (log scale)", y = NULL) +
  annotate("text", x = 0.6, y = 8.5, label = "← Lower risk\n   in HCWs", size = 3, color = "gray50") +
  annotate("text", x = 6, y = 8.5, label = "Higher risk →\n   in HCWs", size = 3, color = "gray50") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"))
```

## Reading the Forest Plot {.smaller}

**Each row** = one study:

- **Square** = point estimate (OR, RR, or mean difference)
- **Horizontal line** = 95% confidence interval
- **Square size** = study weight (larger studies → larger squares)

. . .

**Bottom diamond** = pooled estimate:

- **Centre** = pooled effect size
- **Width** = 95% CI of pooled estimate
- If the diamond **does not cross the null line** (OR = 1) → statistically significant

. . .

**X-axis** is on a **log scale** for ratios — so OR = 0.5 and OR = 2.0 are equidistant from 1.0

## Fixed-Effect vs Random-Effects {.smaller}

| | **Fixed-Effect** | **Random-Effects** |
|---|---|---|
| **Assumption** | One true effect across all studies | Each study has a *different* true effect |
| **Between-study variation** | Sampling error only | Sampling error + real differences |
| **CI width** | Narrower | Wider (more conservative) |
| **Weighting** | Proportional to study size | More balanced — small studies get more weight |
| **Use when** | Studies very similar; I² < 25% | Studies diverse; I² ≥ 50% |

. . .

::: {.callout-tip title="Practical Rule"}
When in doubt, **use random-effects**. If heterogeneity is truly zero, both models give the same result anyway.
:::

## Fixed vs Random — Visualised {.smaller}

```{r}
#| fig-height: 4.5
#| fig-width: 10

fe_or <- 2.3; fe_lo <- 1.8; fe_hi <- 2.9
re_or <- 2.6; re_lo <- 1.9; re_hi <- 3.5

comp_df <- tibble(
  model = c("Fixed-effect", "Random-effects"),
  or = c(fe_or, re_or),
  lo = c(fe_lo, re_lo),
  hi = c(fe_hi, re_hi)
) %>% mutate(model = fct_rev(fct_inorder(model)))

ggplot(comp_df, aes(y = model)) +
  geom_point(aes(x = or), size = 6, shape = 18, color = c("#e74c3c", "#2c3e50")) +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.15,
                 color = c("#e74c3c", "#2c3e50"), linewidth = 1.2) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  scale_x_continuous(limits = c(0.5, 5), breaks = 1:5) +
  labs(title = "Same Data, Different Models",
       subtitle = "Random-effects gives wider CI — accounts for between-study variation",
       x = "Pooled Odds Ratio", y = NULL) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 13, face = "bold"),
        plot.title = element_text(face = "bold"))
```

## Heterogeneity — The I² Statistic {.smaller}

$$I^2 = \frac{Q - df}{Q} \times 100\%$$

| **I² value** | **Interpretation** |
|:---:|---|
| 0–25% | Low — studies agree well |
| 25–50% | Moderate — some differences |
| 50–75% | Substantial — results vary considerably |
| > 75% | Considerable — may be too different to pool |

. . .

**I² tells you what percentage of variation is real (not chance).**

High I² → investigate *why* before pooling.

## Heterogeneity Example: OSA in India {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 10

# Simulated data inspired by OSA meta-analysis
osa_studies <- tibble(
  study = c("Sharma 2006", "Udwadia 2004", "Reddy 2009", "Puri 2015",
            "Kale 2018", "Suri 2008", "Bhatia 2017", "BLESS 2024*"),
  prev = c(4.9, 7.5, 3.6, 13.7, 9.3, 19.5, 8.2, 30.5),
  lo = c(2.1, 4.2, 1.8, 9.1, 5.8, 12.3, 5.0, 27.6),
  hi = c(7.7, 10.8, 5.4, 18.3, 12.8, 26.7, 11.4, 33.4),
  method = c("Questionnaire", "PSG (clinic)", "Questionnaire", "PSG (clinic)",
             "Questionnaire", "PSG (clinic)", "Questionnaire", "PSG (population)")
) %>% mutate(study = fct_rev(fct_inorder(study)))

ggplot(osa_studies, aes(y = study)) +
  geom_point(aes(x = prev, color = method), size = 4, shape = 15) +
  geom_errorbarh(aes(xmin = lo, xmax = hi, color = method), height = 0.2) +
  scale_color_manual(values = c("Questionnaire" = "#7f8c8d", "PSG (clinic)" = "#3498db",
                                "PSG (population)" = "#e74c3c")) +
  labs(title = "OSA Prevalence in India — Why I² = 98%",
       subtitle = "BLESS cohort (Pakhare, Joshi A, 2024): 30.5% vs pooled meta-analysis: 11%",
       x = "Prevalence (%)", y = NULL, color = "Diagnostic method") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.y = element_blank(), legend.position = "right",
        plot.title = element_text(face = "bold"))
```

. . .

**Sources:** Diagnostic method, population type, OSA definition, age/BMI, scoring rules

## Publication Bias — The Funnel Plot {.smaller}

```{r}
#| fig-height: 5
#| fig-width: 10

set.seed(42)
n_studies <- 20
true_effect <- log(2.5)

se_sym <- runif(n_studies, 0.15, 0.8)
effect_sym <- rnorm(n_studies, true_effect, se_sym)
funnel_sym <- tibble(effect = effect_sym, se = se_sym, type = "Symmetric (No bias)")

se_asym <- runif(n_studies, 0.15, 0.8)
effect_asym <- rnorm(n_studies, true_effect, se_asym)
keep <- !(se_asym > 0.5 & effect_asym < log(1.5))
funnel_asym <- tibble(effect = effect_asym[keep], se = se_asym[keep], type = "Asymmetric (Publication bias)")

funnel_data <- bind_rows(funnel_sym, funnel_asym)

ggplot(funnel_data, aes(x = exp(effect), y = se)) +
  geom_point(size = 3, color = "#2c3e50", alpha = 0.7) +
  geom_vline(xintercept = exp(true_effect), linetype = "dashed", color = "#e74c3c") +
  scale_y_reverse() + scale_x_log10() +
  facet_wrap(~type) +
  labs(title = "Funnel Plots: Detecting Publication Bias",
       subtitle = "Left: symmetric (no bias) | Right: small null studies missing (bias)",
       x = "Odds Ratio (log scale)", y = "Standard Error (inverted)") +
  theme_minimal(base_size = 14) +
  theme(strip.text = element_text(face = "bold", size = 12),
        plot.title = element_text(face = "bold"))
```

. . .

**Asymmetric funnel** → small studies with null results are missing → **publication bias**

**Egger's test** provides a formal test for asymmetry (needs ≥ 10 studies)

## Risk of Bias — Cochrane RoB 2 {.smaller}

For RCTs, assess **five domains**:

```{r}
#| results: asis

rob <- tibble(
  Domain = c("Randomisation", "Deviations from intervention",
             "Missing outcome data", "Outcome measurement",
             "Selection of reported result"),
  `What can go wrong` = c(
    "Inadequate sequence generation / allocation concealment",
    "Unblinded participants change behaviour",
    "Differential dropout between groups",
    "Assessor knows group; subjective outcome",
    "Authors pick the 'best' outcome or time point"
  )
)

kable(rob, format = "html", align = c("l", "l")) %>%
  kable_styling(full_width = TRUE, font_size = 15) %>%
  row_spec(0, background = "#2c3e50", color = "white", bold = TRUE) %>%
  column_spec(1, bold = TRUE, width = "30%")
```

. . .

**Traffic light:** Each domain rated **Low** (green), **Some concerns** (yellow), **High** (red)

A study with **high risk in even one critical domain** can bias the whole meta-analysis.

## GRADE — Rating Evidence Quality {.smaller}

| | **Factor** | **Rating** |
|---|---|:---:|
| **Start** | RCTs | High (⊕⊕⊕⊕) |
| | Observational | Low (⊕⊕○○) |
| **Downgrade** | Risk of bias | −1 or −2 |
| | Inconsistency (I²) | −1 or −2 |
| | Indirectness | −1 or −2 |
| | Imprecision | −1 or −2 |
| | Publication bias | −1 or −2 |
| **Upgrade** | Large effect | +1 or +2 |
| | Dose-response | +1 |
| | Confounders reduce effect | +1 |

. . .

**Final ratings:** High (⊕⊕⊕⊕) → Very low (⊕○○○)

A statistically significant pooled effect from very-low-quality evidence is **not compelling**.

## PRISMA Flow Diagram {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 10

prisma_df <- tibble(
  x = c(5, 5, 5, 5, 5, 9, 9, 9),
  y = c(10, 8, 6, 4, 2, 8, 6, 4),
  label = c(
    "Records identified\n(n = 3,842)",
    "After duplicates removed\n(n = 2,156)",
    "Full-text assessed\n(n = 187)",
    "Qualitative synthesis\n(n = 51)",
    "Meta-analysis\n(n = 42)",
    "Excluded by title/abstract\n(n = 1,969)",
    "Full-text excluded (n = 136)\nNo denominator (58)\nNot LMIC (42), Review (36)",
    "Not suitable for pooling\n(n = 9)"
  ),
  box = c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, FALSE)
)

ggplot(prisma_df) +
  geom_label(data = prisma_df %>% filter(box),
             aes(x = x, y = y, label = label),
             fill = "#d4e4f7", color = "#1e3a5f", size = 3.2, fontface = "bold",
             label.padding = unit(0.4, "cm")) +
  geom_label(data = prisma_df %>% filter(!box),
             aes(x = x, y = y, label = label),
             fill = "#fde8e8", color = "#c0392b", size = 2.8,
             label.padding = unit(0.3, "cm")) +
  geom_segment(aes(x = 5, xend = 5, y = 9.5, yend = 8.5), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 5, xend = 5, y = 7.5, yend = 6.5), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 5, xend = 5, y = 5.5, yend = 4.7), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 5, xend = 5, y = 3.5, yend = 2.5), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 6.2, xend = 8, y = 8, yend = 8), arrow = arrow(length = unit(0.2, "cm")), color = "#c0392b") +
  geom_segment(aes(x = 6.2, xend = 8, y = 6, yend = 6), arrow = arrow(length = unit(0.2, "cm")), color = "#c0392b") +
  geom_segment(aes(x = 6.2, xend = 8, y = 4, yend = 4), arrow = arrow(length = unit(0.2, "cm")), color = "#c0392b") +
  xlim(2, 12) + ylim(1, 11) +
  labs(title = "PRISMA Flow Diagram",
       subtitle = "Simulated example based on Joshi R et al. TB systematic review") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11, color = "gray40"))
```

## When NOT to Meta-Analyse {.smaller}

::: {.callout-warning}

1. **Extreme clinical heterogeneity** — apples and oranges
2. **Very high I²** with no identifiable source
3. **Only 2–3 studies** — narrative synthesis is better
4. **High risk of bias** in most studies — precise but wrong
5. **Contradictory results** with plausible mechanistic reasons
:::

. . .

**Teaching example:** Pralidoxime in organophosphorus poisoning

- Earlier observational studies suggested benefit
- Eddleston, **Joshi R** et al. (*PLoS Med* 2009): double-blind RCT (n = 235)
- Result: pralidoxime group had **higher mortality** (24.8% vs 15.8%)
- Pooling the RCT with weaker observational studies would **dilute** the stronger signal

. . .

**A single strong RCT can be more informative than a meta-analysis of many weak studies.**

## Sensitivity Analysis — Testing Robustness {.smaller}

| **Analysis** | **What you change** | **What it tests** |
|---|---|---|
| Exclude high-RoB studies | Drop biased studies | Is the effect driven by bias? |
| Fixed vs random effects | Switch the model | Does between-study variation matter? |
| Leave-one-out | Remove each study in turn | Is one study driving the result? |
| Restrict to RCTs | Drop observational studies | Consistent in stronger designs? |
| Trim-and-fill | Impute missing studies | Publication bias impact? |

. . .

If the pooled estimate **changes substantially** → the conclusion is **not robust**.

Report this transparently.

## Identify the Analysis — Vignette 1 {.smaller}

A team finds **15 RCTs** comparing albendazole vs placebo for soil-transmitted helminths in Indian schoolchildren. Each trial reports cure rate. The trials use different doses (200 mg vs 400 mg) and follow-up periods (1 week vs 4 weeks). I² = 72%.

. . .

**Random-effects meta-analysis with subgroup analysis** — High heterogeneity (72%) warrants investigation. Subgroup by dose and follow-up duration. A fixed-effect model would be inappropriate given the substantial I².

## Identify the Analysis — Vignette 2 {.smaller}

A systematic review of **Mycobacterium w (Mw)** as COVID-19 adjunct therapy finds **2 small open-label trials** (n = 40 and n = 55) with opposite results. Trial 1 shows benefit; Trial 2 shows no effect. Both are high risk of bias (unblinded, small).

. . .

**Do NOT meta-analyse.** Only 2 studies, opposite results, both high RoB. A pooled estimate would be misleading — **narrative synthesis** is the appropriate approach. Wait for larger, blinded RCTs before pooling.

## Identify the Analysis — Vignette 3 {.smaller}

You are reviewing a Cochrane systematic review of statins for primary prevention of cardiovascular disease. The forest plot shows 12 RCTs with a pooled RR = 0.75 (95% CI: 0.70–0.81), I² = 15%. The funnel plot is symmetric. GRADE rates the evidence as **High**.

. . .

**This is a well-conducted meta-analysis.** Low heterogeneity (I² = 15%), symmetric funnel plot (no publication bias), high GRADE quality. The pooled RR = 0.75 means a **25% reduction** in cardiovascular events — a robust, clinically meaningful result.

## Common Mistakes {.smaller}

:::: {.callout-warning}

1. **Pooling apples and oranges** — combining clinically incompatible studies
2. **Ignoring I²** — reporting a pooled estimate without assessing heterogeneity
3. **Choosing the model post-hoc** — switching to fixed-effect because it gives a "better" p-value
4. **Equating statistical significance with clinical importance** — a significant pooled OR = 1.05 may be trivial
5. **No sensitivity analysis** — only reporting one set of results
6. **Funnel plot with < 10 studies** — insufficient power to detect asymmetry
7. **Confusing random-effects with random sampling** — the "random" refers to study effects, not patient selection
8. **Ignoring GRADE** — a significant result from very-low-quality evidence is not compelling
::::

## Key Takeaways

::: {.incremental}
1. **Systematic review first, meta-analysis second** — never pool without structured search
2. **Forest plot** = squares (studies), diamond (pooled), vertical line at null
3. **Random-effects** is the safer default when heterogeneity is uncertain
4. **I² > 75%** → stop, investigate, consider subgroup analysis before pooling
5. **Funnel plot** asymmetry suggests publication bias (needs ≥ 10 studies)
6. **GRADE** rates overall evidence quality — significant ≠ high quality
7. **Sensitivity analysis is not optional** — vary assumptions, test robustness
8. **A single strong RCT** (like the pralidoxime trial) can be more informative than a meta-analysis of many weak studies
:::
