---
title: "Bayesian Statistics — Basics"
subtitle: "Updating Beliefs with Evidence"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    scrollable: true
    footer: "Biostatistics for Clinicians | Module 14"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
set.seed(42)
```

## Learning Objectives

::: {.incremental}
1. Explain the **Bayesian framework**: prior, likelihood, and posterior
2. Distinguish **frequentist vs Bayesian** inference
3. Interpret **posterior distributions** and credible intervals
4. Understand how **prior choice** affects conclusions
5. Apply Bayesian thinking to **clinical diagnosis**
6. Interpret **Bayes factors** as evidence for or against hypotheses
7. Recognise when Bayesian methods offer advantages
:::

## Clinical Hook: Screening for OSA {.smaller}

::: {.columns}
::: {.column width="55%"}
A 48-year-old man: daytime sleepiness, loud snoring, morning headaches.

**BLESS cohort** (Pakhare, Joshi A, AIIMS Bhopal, 2024):

- 1,015 adults with gold-standard polysomnography
- **30.5%** had moderate-severe OSA
- This is your **prior probability**

. . .

You administer the **SONA questionnaire** (Joshi A, Goyal A, Pakhare A, 2025):

- Sensitivity: **73%**, Specificity: **78%**
- Patient scores positive

**What is the probability he has OSA now?**

This is **Bayesian reasoning** — updating a prior belief with new evidence.
:::

::: {.column width="45%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5

prior <- 0.305
sens <- 0.73
spec <- 0.78
post_pos <- (sens * prior) / (sens * prior + (1 - spec) * (1 - prior))
post_neg <- ((1 - sens) * prior) / ((1 - sens) * prior + spec * (1 - prior))

bar_df <- tibble(
  stage = c("Pre-test\n(prior)", "Post-test\n(SONA +)", "Post-test\n(SONA −)"),
  prob = c(prior, post_pos, post_neg) * 100,
  fill = c("#3498db", "#e74c3c", "#2c3e50")
) %>% mutate(stage = fct_inorder(stage))

ggplot(bar_df, aes(x = stage, y = prob, fill = stage)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(prob, 1), "%")), vjust = -0.5, fontface = "bold", size = 5) +
  scale_fill_manual(values = c("#3498db", "#e74c3c", "#2c3e50")) +
  ylim(0, 75) +
  labs(title = "Bayesian Updating with SONA",
       x = NULL, y = "Probability of OSA (%)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none", plot.title = element_text(face = "bold"))
```
:::
:::

## The Bayesian Formula {.smaller}

$$P(\text{Disease} \mid \text{Test}^+) = \frac{P(\text{Test}^+ \mid \text{Disease}) \times P(\text{Disease})}{P(\text{Test}^+)}$$

. . .

**In words:** Posterior = (Likelihood × Prior) / Evidence

. . .

| Component | Meaning | OSA Example |
|---|---|---|
| **Prior** | Belief before test | Prevalence = 30.5% |
| **Likelihood** | P(data \| hypothesis) | Sensitivity = 73% |
| **Evidence** | P(data overall) | P(SONA+) = 37.5% |
| **Posterior** | Updated belief | **`r round(post_pos*100, 1)`%** |

. . .

**The same positive test gives different answers depending on the prior.** In a low-prevalence population (5%), the post-test probability would be only `r round((sens * 0.05) / (sens * 0.05 + (1-spec) * 0.95) * 100, 1)`%.

## Prior × Likelihood → Posterior {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 10

x_bf <- seq(0, 1, length.out = 500L)
prior_d <- dbeta(x_bf, 6, 14)
like_d <- dbeta(x_bf, 45 + 1, 55 + 1)
like_d <- like_d / max(like_d) * max(prior_d) * 1.2
post_d <- dbeta(x_bf, 51, 69)

df_bayes <- bind_rows(
  tibble(theta = x_bf, density = prior_d, component = "Prior"),
  tibble(theta = x_bf, density = like_d, component = "Likelihood (scaled)"),
  tibble(theta = x_bf, density = post_d, component = "Posterior")
) %>% mutate(component = fct_inorder(component))

ggplot(df_bayes, aes(x = theta, y = density, color = component, linetype = component)) +
  geom_line(linewidth = 1.3) +
  scale_color_manual(values = c("#3498db", "#e67e22", "#e74c3c")) +
  scale_linetype_manual(values = c("dashed", "dotted", "solid")) +
  labs(title = "Bayesian Updating: Prior × Likelihood → Posterior",
       subtitle = "Prior centred ~30% | Data: 45/100 positive | Posterior: compromise",
       x = "Proportion (θ)", y = "Density", color = NULL, linetype = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = c(0.82, 0.8), plot.title = element_text(face = "bold"),
        legend.background = element_rect(fill = "white", color = "gray80"))
```

**The posterior is ALWAYS a compromise.** More data → posterior follows data. Less data → prior dominates.

## Frequentist vs Bayesian {.smaller}

| | **Frequentist** | **Bayesian** |
|---|---|---|
| **Probability** | Long-run frequency | Degree of belief |
| **Parameters** | Fixed but unknown | Random variables |
| **Prior info** | Not formally used | Explicitly incorporated |
| **Result** | p-value, CI | Posterior, credible interval |
| **Interpretation** | "If repeated infinitely..." | "Given the data..." |

. . .

**The p-value problem:** p = 0.04 does NOT mean "4% chance the null is true."

A **95% CI** does NOT mean "95% probability the true value is in here."

A **95% credible interval** DOES mean exactly that.

## Confidence Interval vs Credible Interval {.smaller}

```{r}
#| fig-height: 5
#| fig-width: 10

set.seed(42)
true_mean <- 120; n <- 25; n_sims <- 15

freq_sims <- tibble(
  sim = 1:n_sims,
  mean = rnorm(n_sims, true_mean, 15/sqrt(n)),
  se = 15/sqrt(n)
) %>% mutate(lo = mean - 1.96 * se, hi = mean + 1.96 * se,
             covers = lo <= true_mean & hi >= true_mean)

p1 <- ggplot(freq_sims, aes(y = sim)) +
  geom_point(aes(x = mean, color = covers), size = 2) +
  geom_errorbarh(aes(xmin = lo, xmax = hi, color = covers), height = 0.3) +
  geom_vline(xintercept = true_mean, linetype = "dashed") +
  scale_color_manual(values = c("TRUE" = "#2c3e50", "FALSE" = "#e74c3c"), guide = "none") +
  labs(title = "Frequentist: many experiments,\nmany different CIs", x = "SBP (mmHg)", y = "Experiment") +
  theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold", size = 11))

theta_b <- seq(110, 130, length.out = 500)
post_sbp <- dnorm(theta_b, 119.5, 15/sqrt(n))
cri_lo <- qnorm(0.025, 119.5, 15/sqrt(n))
cri_hi <- qnorm(0.975, 119.5, 15/sqrt(n))

p2 <- ggplot(tibble(x = theta_b, y = post_sbp), aes(x, y)) +
  geom_line(linewidth = 1.3, color = "#e74c3c") +
  geom_area(data = tibble(x = theta_b, y = post_sbp) %>% filter(x >= cri_lo & x <= cri_hi),
            fill = "#e74c3c", alpha = 0.3) +
  geom_vline(xintercept = true_mean, linetype = "dashed") +
  annotate("text", x = 126, y = max(post_sbp) * 0.6,
           label = paste0("95% CrI:\n[", round(cri_lo,1), ", ", round(cri_hi,1), "]"),
           color = "#e74c3c", fontface = "bold") +
  labs(title = "Bayesian: one posterior,\none credible interval", x = "SBP (mmHg)", y = "Density") +
  theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold", size = 11))

p1 + p2
```

## Types of Priors {.smaller}

```{r}
#| fig-height: 5
#| fig-width: 10

x_pt <- seq(0, 1, length.out = 500L)
priors_df <- bind_rows(
  tibble(theta = x_pt, density = dbeta(x_pt, 1, 1), type = "Uninformative\nBeta(1,1)"),
  tibble(theta = x_pt, density = dbeta(x_pt, 2, 2), type = "Weakly informative\nBeta(2,2)"),
  tibble(theta = x_pt, density = dbeta(x_pt, 6, 14), type = "Informative\nBeta(6,14)"),
  tibble(theta = x_pt, density = dbeta(x_pt, 60, 140), type = "Strongly informative\nBeta(60,140)")
) %>% mutate(type = fct_inorder(type))

ggplot(priors_df, aes(x = theta, y = density, color = type)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~type, scales = "free_y") +
  scale_color_manual(values = c("#95a5a6", "#3498db", "#e67e22", "#e74c3c")) +
  labs(title = "Types of Prior Distributions",
       x = "Proportion (θ)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold"))
```

. . .

**Uninformative** = know nothing | **Informative** = prior studies | **Strongly informative** = large prior data

## Who Wins: Prior or Data? {.smaller}

```{r}
#| fig-height: 5
#| fig-width: 10

x_pc <- seq(0, 1, length.out = 500L)
conflict_df <- bind_rows(
  tibble(theta = x_pc, density = dbeta(x_pc, 30, 70), label = "Prior (centred 30%)"),
  tibble(theta = x_pc, density = dbeta(x_pc, 42, 78), label = "n = 20 (data: 60%)"),
  tibble(theta = x_pc, density = dbeta(x_pc, 90, 110), label = "n = 100 (data: 60%)"),
  tibble(theta = x_pc, density = dbeta(x_pc, 330, 270), label = "n = 500 (data: 60%)")
) %>% mutate(label = fct_inorder(label))

ggplot(conflict_df, aes(x = theta, y = density, color = label)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("#3498db", "#27ae60", "#e67e22", "#e74c3c")) +
  geom_vline(xintercept = 0.30, linetype = "dashed", color = "#3498db", alpha = 0.5) +
  geom_vline(xintercept = 0.60, linetype = "dashed", color = "#e74c3c", alpha = 0.5) +
  labs(title = "With More Data, the Data Win",
       subtitle = "Prior centred at 30% vs data suggesting 60%",
       x = "Proportion (θ)", y = "Density", color = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right", plot.title = element_text(face = "bold"))
```

## Bayes Factor — Evidence for or Against {.smaller}

$$BF_{10} = \frac{P(\text{Data} \mid H_1)}{P(\text{Data} \mid H_0)}$$

| **BF₁₀** | **Interpretation** |
|:---:|---|
| < 1 | Evidence favours **null** (H₀) |
| 1–3 | Anecdotal evidence for H₁ |
| 3–10 | Moderate evidence for H₁ |
| 10–30 | Strong evidence |
| > 100 | Extreme evidence |

. . .

**Key advantage over p-values:** A p-value can never support the null. A Bayes factor **can**.

BF₁₀ = 0.15 means BF₀₁ = 6.7 → moderate evidence **for** no treatment effect.

## Diagnostic Reasoning Is Bayesian {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 10

prev_range <- seq(0.05, 0.80, by = 0.01)
sens <- 0.73; spec <- 0.78

post_pos <- (sens * prev_range) / (sens * prev_range + (1 - spec) * (1 - prev_range))
post_neg <- ((1 - sens) * prev_range) / ((1 - sens) * prev_range + spec * (1 - prev_range))

diag_df <- tibble(
  prior = rep(prev_range, 2),
  posterior = c(post_pos, post_neg),
  result = rep(c("SONA Positive", "SONA Negative"), each = length(prev_range))
)

ggplot(diag_df, aes(x = prior * 100, y = posterior * 100, color = result)) +
  geom_line(linewidth = 1.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("SONA Positive" = "#e74c3c", "SONA Negative" = "#2c3e50")) +
  labs(title = "Post-Test Probability Depends on Pre-Test Probability",
       subtitle = "SONA questionnaire for OSA screening (Pakhare, Joshi A, AIIMS Bhopal)",
       x = "Pre-test probability (%)", y = "Post-test probability (%)", color = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = c(0.2, 0.85), plot.title = element_text(face = "bold"),
        legend.background = element_rect(fill = "white", color = "gray80"))
```

## Bayesian Updating in Practice {.smaller}

**The Pralidoxime Story** (Eddleston, Joshi R et al., *PLoS Med* 2009):

```{r}
#| fig-height: 4.5
#| fig-width: 10

x_pral <- seq(-1.5, 1.5, length.out = 500L)
post_var <- 1 / (1/0.3^2 + 1/0.25^2)
post_mean <- post_var * (-0.36/0.3^2 + 0.52/0.25^2)

pral_df <- bind_rows(
  tibble(theta = x_pral, density = dnorm(x_pral, -0.36, 0.3), component = "Prior (beneficial, HR \u2248 0.7)"),
  tibble(theta = x_pral, density = dnorm(x_pral, 0.52, 0.25), component = "RCT data (harmful, HR \u2248 1.69)"),
  tibble(theta = x_pral, density = dnorm(x_pral, post_mean, sqrt(post_var)), component = "Posterior (updated)")
) %>% mutate(component = fct_inorder(component))

post_pral <- dnorm(x_pral, post_mean, sqrt(post_var))

ggplot(pral_df, aes(x = exp(theta), y = density, color = component, linetype = component)) +
  geom_line(linewidth = 1.3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("#3498db", "#e67e22", "#e74c3c")) +
  scale_linetype_manual(values = c("dashed", "dotted", "solid")) +
  scale_x_continuous(breaks = c(0.3, 0.5, 0.7, 1, 1.5, 2, 3), trans = "log", limits = c(0.2, 4)) +
  annotate("text", x = 0.4, y = max(post_pral) * 0.5, label = "← Benefit", color = "gray50") +
  annotate("text", x = 2.5, y = max(post_pral) * 0.5, label = "Harm →", color = "gray50") +
  labs(title = "Pralidoxime: Prior Belief Overturned by RCT Data",
       x = "Hazard Ratio (log scale)", y = "Density", color = NULL, linetype = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right", plot.title = element_text(face = "bold"))
```

. . .

**Being Bayesian means being willing to update** — even when the data contradict your beliefs.

## ROPE: Clinically Meaningful Decisions {.smaller}

**Region of Practical Equivalence** — is the effect big enough to matter?

```{r}
#| fig-height: 4.5
#| fig-width: 10

theta <- seq(-15, 15, length.out = 500)
post_effect <- dnorm(theta, 3.2, 2.5)
df_rope <- tibble(theta = theta, density = post_effect)

ggplot(df_rope, aes(x = theta, y = density)) +
  geom_line(linewidth = 1.3, color = "#2c3e50") +
  geom_area(data = df_rope %>% filter(theta >= -2 & theta <= 2),
            fill = "#f39c12", alpha = 0.3) +
  geom_vline(xintercept = c(-2, 2), linetype = "dashed", color = "#f39c12") +
  annotate("text", x = 0, y = max(post_effect) * 1.1, label = "ROPE\n(clinically negligible)",
           fontface = "bold", color = "#f39c12", size = 4) +
  labs(title = "ROPE Analysis: Is the SBP Reduction Clinically Meaningful?",
       subtitle = "ROPE = [−2, 2] mmHg | Posterior mean = 3.2 mmHg",
       x = "SBP Reduction (mmHg)", y = "Posterior Density") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))
```

. . .

- 95% outside ROPE → **clinically meaningful** | 95% inside ROPE → **practically null** | Straddles → **undecided**

## Identify the Approach — Vignette 1 {.smaller}

A new rapid diagnostic test for TB has sensitivity 92% and specificity 88%. In a tribal district of MP, TB prevalence is 500 per 100,000 (0.5%). A patient with chronic cough tests positive.

**What is the post-test probability of active TB?**

. . .

```{r}
#| results: asis
prev_tb <- 0.005; sens_tb <- 0.92; spec_tb <- 0.88
post_tb <- (sens_tb * prev_tb) / (sens_tb * prev_tb + (1 - spec_tb) * (1 - prev_tb))
```

**Bayesian calculation:** P(TB | Test+) = (0.92 × 0.005) / (0.92 × 0.005 + 0.12 × 0.995) = **`r round(post_tb * 100, 1)`%**

Even with a positive test, the probability is only ~3.7% because the **prior is extremely low**. The prior dominates when prevalence is very low. This is why mass screening for rare diseases generates many false positives.

## Identify the Approach — Vignette 2 {.smaller}

An RCT of a new antihypertensive (n = 40) finds mean SBP reduction of 6 mmHg (95% CI: −1 to 13, p = 0.09). A frequentist concludes "not significant." A Bayesian analysis with an informative prior (from a related drug class showing 5 mmHg benefit) gives a posterior mean of 5.8 mmHg, 95% CrI [1.2, 10.4].

. . .

**The Bayesian analysis borrows strength from prior knowledge** and concludes there is a >97.5% probability of *some* benefit. The frequentist analysis, ignoring prior information, cannot reject the null. Both are valid approaches — but the Bayesian one incorporates more information and gives a more clinically useful answer for a small trial.

## Identify the Approach — Vignette 3 {.smaller}

A pharmaceutical company tests whether a generic drug is equivalent to the branded version. The RCT (n = 500) yields p = 0.42 for the difference. The company claims "no difference — the drugs are equivalent."

. . .

**A p-value cannot prove equivalence** — it only fails to reject a difference. A Bayesian analysis with BF₁₀ = 0.08 (BF₀₁ = 12.5) would provide **strong evidence for equivalence**. The Bayes factor directly quantifies evidence for the null hypothesis, which is exactly what's needed here.

## Common Misconceptions {.smaller}

:::: {.callout-warning}

1. **"Bayesian = subjective = unreliable"** → The prior is explicit and transparent; frequentist assumptions are hidden
2. **"You can choose any prior you want"** → Sensitivity analysis tests robustness; reviewers check priors
3. **"Bayesian always disagrees with frequentist"** → For large samples with weak priors, they converge
4. **"A flat prior = no assumptions"** → Flat on [0,1] says all values equally likely — that IS an assumption
5. **"You need special software"** → R (brms), Python (PyMC), JASP (free GUI) all support Bayesian
6. **"Too complicated for clinical research"** → The concept (update beliefs with evidence) is simpler than p-value logic
::::

## Key Takeaways

::: {.incremental}
1. **You're already Bayesian** — clinical diagnosis uses pre-test probability + LR = post-test probability
2. **Prior × Likelihood → Posterior** — the posterior is a compromise
3. **With enough data, the prior barely matters** — Bayesian and frequentist converge
4. **Credible intervals** answer the question clinicians actually want
5. **Bayes factors** can support the null — p-values cannot
6. **ROPE** replaces "p < 0.05" with clinically meaningful decisions
7. **Always do sensitivity analysis** — show results with alternative priors
8. **Being Bayesian = being willing to update** — even when data contradict your beliefs
:::
