---
title: "Machine Learning in Medical Sciences"
subtitle: "When Algorithms Meet Clinical Data"
format:
  revealjs:
    theme: [default, ../styles-slides.css]
    embed-resources: true
    slide-number: true
    show-slide-number: all
    chalkboard: false
    scrollable: true
    footer: "Biostatistics for Clinicians | Module 15"
    transition: fade
    incremental: false
    code-fold: true
    fig-align: center
    width: 1280
    height: 720
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(kableExtra)
library(patchwork)
library(scales)
set.seed(42)
```

## Learning Objectives

::: {.incremental}
1. Distinguish **supervised** from **unsupervised** learning
2. Understand common algorithms: logistic regression, decision trees, random forests
3. Explain the **bias-variance trade-off** and overfitting
4. Interpret **AUC, sensitivity, specificity** as ML evaluation metrics
5. Understand **training, validation, and test sets**
6. Evaluate ML claims in medical literature critically
7. Recognise **ethical concerns** of ML in healthcare
:::

## Clinical Hook: The SONA Score {.smaller}

::: {.columns}
::: {.column width="55%"}
The AIIMS Bhopal team built the **SONA questionnaire** for OSA screening (Joshi A, Goyal A, Pakhare A, 2025):

1. **Data:** 1,015 adults (BLESS cohort), all with PSG
2. **Split:** Training (n = 732) / Testing (n = 244)
3. **Models:** Four predictive models compared
4. **Selection:** Best model by AUC = **0.83**
5. **Validation:** Confirmed on held-out test set
6. **Deploy:** Coefficients → simple point score

. . .

**This IS machine learning** — even though it used logistic regression. ML is not about fancy algorithms. It's about the **discipline of training on one dataset and testing on another.**
:::

::: {.column width="45%"}
```{r}
#| fig-height: 5
#| fig-width: 5.5

steps_df <- tibble(
  step = c("Collect\ndata", "Split\ntrain/test", "Build\nmodels", "Evaluate\non test", "Deploy\nscore"),
  y = 5:1,
  icon = c("📊", "✂️", "🔧", "✅", "🏥")
)

ggplot(steps_df, aes(x = 1, y = y)) +
  geom_tile(width = 1.8, height = 0.8, fill = c("#3498db", "#e67e22", "#9b59b6", "#27ae60", "#e74c3c"),
            alpha = 0.85) +
  geom_text(aes(label = step), color = "white", fontface = "bold", size = 4.5) +
  geom_segment(aes(x = 1, xend = 1, y = y - 0.45, yend = y - 0.55),
               data = steps_df %>% filter(y > 1),
               arrow = arrow(length = unit(0.15, "cm")), color = "gray40") +
  xlim(0, 2) + ylim(0.3, 5.7) +
  labs(title = "ML Pipeline: SONA Development") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
```
:::
:::

## What Is Machine Learning? {.smaller}

**Algorithms that learn patterns from data to make predictions — without explicit rules.**

| | **Traditional Statistics** | **Machine Learning** |
|---|---|---|
| **Goal** | Understand relationships (inference) | Make accurate predictions |
| **Model choice** | Theory-driven | Data-driven |
| **Interpretability** | Usually high | Variable |
| **Validation** | CI, p-values | Train/test split, AUC |
| **Overfitting** | Managed by theory | Managed by held-out test data |

. . .

::: {.callout-tip}
**Logistic regression is BOTH statistics and ML.** The difference isn't the math — it's how you use it. Inference → statistics. Prediction with train/test → ML.
:::

## Supervised vs Unsupervised Learning {.smaller}

::: {.columns}
::: {.column width="50%"}
### Supervised
Algorithm learns from **labelled** examples.

| Task | Output | Example |
|---|---|---|
| **Classification** | Category | OSA yes/no |
| **Regression** | Number | Predict AHI |

```{r}
#| fig-height: 4
#| fig-width: 5

set.seed(42)
osa_data <- tibble(
  waist = c(rnorm(80, 98, 10), rnorm(80, 85, 10)),
  neck = c(rnorm(80, 39, 3), rnorm(80, 34, 3)),
  osa = factor(rep(c("OSA", "No OSA"), each = 80))
)
ggplot(osa_data, aes(x = waist, y = neck, color = osa)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_color_manual(values = c("#e74c3c", "#2c3e50")) +
  geom_abline(intercept = -10, slope = 0.5, linetype = "dashed", color = "#f39c12", linewidth = 1) +
  labs(title = "Classification", x = "Waist (cm)", y = "Neck (cm)", color = NULL) +
  theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```
:::

::: {.column width="50%"}
### Unsupervised
Algorithm finds **hidden structure** — no labels.

| Task | What | Example |
|---|---|---|
| **Clustering** | Groups | HTN subtypes |
| **Dimensionality** | Compress | Gene expression |

```{r}
#| fig-height: 4
#| fig-width: 5

set.seed(42)
cl_data <- tibble(
  sbp = c(rnorm(50, 125, 8), rnorm(40, 155, 10), rnorm(35, 140, 12)),
  bmi = c(rnorm(50, 24, 3), rnorm(40, 32, 4), rnorm(35, 28, 3)),
  cluster = factor(c(rep("Low risk", 50), rep("High risk", 40), rep("Moderate", 35)))
)
ggplot(cl_data, aes(x = sbp, y = bmi, color = cluster)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_color_manual(values = c("#27ae60", "#e74c3c", "#f39c12")) +
  labs(title = "Clustering", x = "SBP (mmHg)", y = "BMI", color = NULL) +
  theme_minimal(base_size = 12) + theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```
:::
:::

## Common Algorithms {.smaller}

```{r}
#| results: asis

algo_comp <- tibble(
  Algorithm = c("Logistic regression", "Decision tree", "Random forest",
                "Gradient boosting", "Neural network"),
  Interpretability = c("★★★★★", "★★★★☆", "★★☆☆☆", "★★☆☆☆", "★☆☆☆☆"),
  `Data needed` = c("Small–medium", "Small–medium", "Medium–large", "Medium–large", "Very large"),
  `Best for` = c("Risk scores (SONA)", "Clinical decision rules",
                 "Tabular clinical data", "Prediction competitions",
                 "Images, text, ECG signals")
)

kable(algo_comp, format = "html") %>%
  kable_styling(full_width = TRUE, font_size = 15) %>%
  row_spec(0, background = "#2c3e50", color = "white", bold = TRUE) %>%
  row_spec(1, background = "#d4edda")
```

. . .

**For clinical screening tools** (like SONA), logistic regression is often the best choice — interpretable, works with smaller data, and coefficients convert to simple point scores.

## Decision Tree — Intuitive Splits {.smaller}

```{r}
#| fig-height: 5.5
#| fig-width: 10

tree_df <- tibble(
  x = c(5, 2.5, 7.5, 1.25, 3.75, 6.25, 8.75),
  y = c(10, 7, 7, 4, 4, 4, 4),
  label = c("Waist > 93 cm?", "Snoring?", "Neck > 37 cm?",
            "Low risk\n(8%)", "Moderate\n(35%)", "Moderate\n(28%)", "High risk\n(72%)"),
  fill = c("#d4e4f7", "#d4e4f7", "#d4e4f7", "#27ae60", "#f39c12", "#f39c12", "#e74c3c")
)

ggplot(tree_df) +
  geom_label(aes(x = x, y = y, label = label, fill = fill),
             size = 3.5, fontface = "bold", color = "white", label.padding = unit(0.4, "cm")) +
  scale_fill_identity() +
  geom_segment(aes(x = 5, xend = 2.5, y = 9.2, yend = 7.6), color = "gray50") +
  geom_segment(aes(x = 5, xend = 7.5, y = 9.2, yend = 7.6), color = "gray50") +
  geom_segment(aes(x = 2.5, xend = 1.25, y = 6.2, yend = 4.6), color = "gray50") +
  geom_segment(aes(x = 2.5, xend = 3.75, y = 6.2, yend = 4.6), color = "gray50") +
  geom_segment(aes(x = 7.5, xend = 6.25, y = 6.2, yend = 4.6), color = "gray50") +
  geom_segment(aes(x = 7.5, xend = 8.75, y = 6.2, yend = 4.6), color = "gray50") +
  annotate("text", x = 3.5, y = 8.3, label = "No", fontface = "bold", size = 4) +
  annotate("text", x = 6.5, y = 8.3, label = "Yes", fontface = "bold", size = 4) +
  annotate("text", x = c(1.7, 3.3, 6.7, 8.3), y = 5.3, label = c("No","Yes","No","Yes"),
           fontface = "bold", size = 3.5) +
  xlim(0, 10) + ylim(3, 11) +
  labs(title = "Decision Tree for OSA Screening",
       subtitle = "Simulated example — sequential yes/no splits") +
  theme_void() + theme(plot.title = element_text(face = "bold", size = 14))
```

## The Bias-Variance Trade-off {.smaller}

```{r}
#| fig-height: 5
#| fig-width: 10

set.seed(42)
complexity <- 1:15
train_acc <- 1 - 0.3 * exp(-0.5 * complexity) + rnorm(15, 0, 0.01)
test_acc <- 1 - 0.3 * exp(-0.3 * complexity) - 0.015 * (complexity - 5)^2 * (complexity > 5) / 25 + rnorm(15, 0, 0.01)
train_acc <- pmin(train_acc, 0.99)
test_acc <- pmin(test_acc, 0.95)

acc_df <- tibble(
  complexity = rep(complexity, 2),
  accuracy = c(train_acc, test_acc),
  set = rep(c("Training", "Test"), each = 15)
)

ggplot(acc_df, aes(x = complexity, y = accuracy * 100, color = set)) +
  geom_line(linewidth = 1.3) + geom_point(size = 2) +
  scale_color_manual(values = c("Training" = "#3498db", "Test" = "#e74c3c")) +
  geom_vline(xintercept = 6, linetype = "dashed", color = "#27ae60") +
  annotate("text", x = 7, y = 75, label = "Sweet\nspot", fontface = "bold", color = "#27ae60", size = 4) +
  annotate("text", x = 3, y = 92, label = "Underfitting", color = "gray50", size = 3.5) +
  annotate("text", x = 12, y = 80, label = "Overfitting", color = "gray50", size = 3.5) +
  labs(title = "Training Accuracy Keeps Rising — Test Accuracy Peaks Then Falls",
       x = "Model Complexity", y = "Accuracy (%)", color = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = c(0.15, 0.3), plot.title = element_text(face = "bold"),
        legend.background = element_rect(fill = "white", color = "gray80"))
```

. . .

**The gap between training and test accuracy = overfitting.** Training accuracy is advertising; test accuracy is truth.

## Overfitting — The Central Problem {.smaller}

```{r}
#| fig-height: 4.5
#| fig-width: 10

set.seed(42)
x <- seq(0, 10, length.out = 30)
y_obs <- 3 + 0.5 * x + rnorm(30, 0, 1.5)
df_fit <- tibble(x = x, y = y_obs)

p1 <- ggplot(df_fit, aes(x, y)) + geom_point(size = 2, color = "#2c3e50") +
  geom_smooth(method = "lm", formula = y ~ 1, se = FALSE, color = "#3498db", linewidth = 1.2) +
  labs(title = "Underfit", subtitle = "Too simple") + theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

p2 <- ggplot(df_fit, aes(x, y)) + geom_point(size = 2, color = "#2c3e50") +
  geom_smooth(method = "lm", se = FALSE, color = "#27ae60", linewidth = 1.2) +
  labs(title = "Good fit", subtitle = "Captures the trend") + theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

p3 <- ggplot(df_fit, aes(x, y)) + geom_point(size = 2, color = "#2c3e50") +
  geom_smooth(method = "loess", span = 0.15, se = FALSE, color = "#e74c3c", linewidth = 1.2) +
  labs(title = "Overfit", subtitle = "Memorises noise") + theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

p1 + p2 + p3
```

. . .

**Signs:** Training 98% → Test 65% | Too many features for sample size | Absurd predictions at extremes

## Cross-Validation {.smaller}

```{r}
#| fig-height: 4
#| fig-width: 10

cv_df <- expand.grid(fold = 1:5, part = 1:5) %>%
  mutate(role = ifelse(fold == part, "Test", "Train"),
         label = ifelse(fold == part, "Test", ""))

ggplot(cv_df, aes(x = part, y = fct_rev(factor(fold)), fill = role)) +
  geom_tile(color = "white", linewidth = 1.5) +
  geom_text(aes(label = label), fontface = "bold", color = "white", size = 5) +
  scale_fill_manual(values = c("Train" = "#3498db", "Test" = "#e74c3c")) +
  labs(title = "5-Fold Cross-Validation",
       subtitle = "Each fold takes a turn as test set → average = final performance estimate",
       x = "Data Partition", y = "Iteration", fill = NULL) +
  theme_minimal(base_size = 14) + theme(panel.grid = element_blank(), plot.title = element_text(face = "bold"))
```

. . .

**Use when:** Not enough data for a separate test set. More robust than a single split.

## ROC Curve and AUC {.smaller}

```{r}
#| fig-height: 5
#| fig-width: 10

fpr <- seq(0, 1, length.out = 200)
tpr_sona <- pbeta(fpr, 0.5, 0.3)
tpr_stop <- pbeta(fpr, 0.65, 0.4)
tpr_random <- fpr

roc_df <- tibble(
  fpr = rep(fpr, 3),
  tpr = c(tpr_sona, tpr_stop, tpr_random),
  model = rep(c("SONA (AUC = 0.83)", "STOP-Bang (AUC = 0.72)", "Random (AUC = 0.50)"), each = 200)
) %>% mutate(model = fct_inorder(model))

ggplot(roc_df, aes(x = fpr, y = tpr, color = model)) +
  geom_line(linewidth = 1.3) +
  scale_color_manual(values = c("#e74c3c", "#3498db", "gray60")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray80") +
  labs(title = "ROC Curves: Comparing OSA Screening Tools",
       subtitle = "SONA outperforms STOP-Bang (Joshi A, Pakhare A, AIIMS Bhopal)",
       x = "False Positive Rate (1 − Specificity)", y = "True Positive Rate (Sensitivity)",
       color = NULL) +
  theme_minimal(base_size = 14) +
  theme(legend.position = c(0.65, 0.25), plot.title = element_text(face = "bold"),
        legend.background = element_rect(fill = "white", color = "gray80"))
```

. . .

**AUC:** 0.50 = coin flip | 0.70–0.79 = acceptable | **0.80–0.89 = excellent** | 0.90+ = outstanding

## Accuracy Is Misleading {.smaller}

Disease prevalence = 1%. Model predicts "no disease" for **everyone**.

. . .

| Metric | Value |
|---|:---:|
| **Accuracy** | **99%** ← looks great! |
| Sensitivity | **0%** ← catches NOTHING |
| Specificity | 100% |
| PPV | Undefined |

. . .

::: {.callout-warning}
**Never evaluate a model on accuracy alone.** With imbalanced classes, accuracy is meaningless. Always report sensitivity, specificity, PPV, NPV, and AUC.
:::

## ML in Indian Healthcare {.smaller}

| Application | Technology | Impact |
|---|---|---|
| **Diabetic retinopathy** | Deep learning on retinal images (Google + Aravind Eye Hospital) | Screening where ophthalmologists are scarce |
| **TB from chest X-ray** | CAD4TB, qXR (Qure.ai) — WHO recommended | Active case-finding in high-burden settings |
| **ECG interpretation** | AI on portable ECG (SanketLife) | PHCs without cardiologists |
| **OSA screening** | SONA score (AIIMS Bhopal) — logistic regression | Primary care tool for resource-limited settings |

. . .

**Common thread:** ML is most impactful where specialist access is limited — extending expert-level decisions to primary care.

## Critical Appraisal of ML Papers {.smaller}

```{r}
#| results: asis

tripod <- tibble(
  `#` = 1:6,
  Question = c(
    "Was data split into training and testing?",
    "Was external validation performed?",
    "Were multiple metrics reported (not just accuracy)?",
    "Was overfitting addressed?",
    "Were predictors available at time of decision?",
    "Is the model clinically useful (better than current practice)?"
  ),
  `Red flag` = c(
    "Only training accuracy reported",
    "Tested only where it was developed",
    "Only accuracy or AUC",
    "High accuracy + complex model + small sample",
    "Predictors include future information (data leakage)",
    "Marginal improvement over simple clinical rules"
  )
)

kable(tripod, format = "html") %>%
  kable_styling(full_width = TRUE, font_size = 14) %>%
  row_spec(0, background = "#2c3e50", color = "white", bold = TRUE)
```

. . .

**Cardinal rule:** If they only report training performance → the results are **meaningless**.

## Ethical Concerns {.smaller}

:::: {.callout-warning}

1. **Algorithmic bias** — model trained on private hospital data fails in government hospital (different demographics, severity)
2. **Black box problem** — if you can't explain *why* the model flagged a patient, should you trust it?
3. **Data quality** — "ML on dirty data = confidently wrong answers"
4. **Temporal drift** — disease patterns change; models need retraining
5. **Accountability** — who is responsible when AI makes a wrong prediction?
6. **Equity** — dermatology AI trained on light skin performs poorly on dark skin
::::

. . .

**Always validate across diverse populations** before deployment.

## Identify the Approach — Vignette 1 {.smaller}

A district hospital has 5 years of lab data on 10,000 patients. They want to identify **hidden patient subtypes** among diabetics based on HbA1c trajectories, comorbidity patterns, and treatment responses — without predefining the groups.

. . .

**Unsupervised learning (clustering)** — There are no predefined labels; the goal is to discover natural groupings. K-means or hierarchical clustering could identify phenotypic subtypes of diabetes, potentially guiding personalised treatment strategies.

## Identify the Approach — Vignette 2 {.smaller}

A research team develops an AI system to detect **pneumothorax on chest X-rays** using 50,000 labelled images from a tertiary hospital. They report AUC = 0.97 on an internal test set. A PHC wants to deploy it.

. . .

**Supervised learning (classification) — but needs external validation first.** The AUC = 0.97 is on internal data. The PHC has different X-ray machines, image quality, and patient population. The model MUST be tested on PHC data before deployment. Internal AUC ≠ real-world performance.

## Identify the Approach — Vignette 3 {.smaller}

A researcher with data from 80 patients and 200 clinical variables builds a random forest model for predicting ICU mortality. Training accuracy is 96%. She has no separate test set.

. . .

**This is almost certainly overfitted.** 200 features on 80 patients (features > patients!) is a recipe for overfitting. The 96% training accuracy is meaningless without a test set. She should use cross-validation, reduce features (variable selection), or use a simpler model (logistic regression with ≤5 predictors).

## Common Mistakes {.smaller}

:::: {.callout-warning}

1. **Testing on training data** — inflates accuracy
2. **Reporting only accuracy** — misleading with class imbalance
3. **Data leakage** — including future/outcome data in predictors
4. **Too many features** — 200 variables on 80 patients = overfitting
5. **No external validation** — model may only work at your hospital
6. **Ignoring clinical context** — AUC 0.85 with no clinical workflow is useless
7. **Claiming causation** — ML finds associations, not causes
8. **Ignoring bias** — training on one population, deploying on another
::::

## Key Takeaways

::: {.incremental}
1. **ML = disciplined prediction** using train/test separation — not magic
2. **Logistic regression IS ML** — SONA is a perfect example
3. **Always ask for test set performance** — training accuracy is advertising
4. **AUC alone is not enough** — report sensitivity, specificity, calibration
5. **External validation is essential** — internal performance ≠ real-world
6. **Interpretability matters** — if you can't explain it, clinicians won't trust it
7. **Check for bias** — was training data representative of deployment population?
8. **AI assists, humans decide** — ML is a diligent junior colleague, not a replacement
:::
